### 2021-08-19 README.Rmd
add Dockerfile and README

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# byocstarch

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2021-08-19 analysis/index.Rmd
add sections to manuscript

---
title: "Untitled"
author:
  - "Bjørn Peare Bartholdy"
  - Amanda G. Henry
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  bookdown::word_document2: default
  bookdown::html_document2: default
linestretch: 2.0 # double-spacing for review purposes
bibliography: paper/refs.bib
csl: paper/apa.csl
abstract: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures")
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
## Compile refs.bib using rbbt package ##
file_paths <- here("analysis/paper", c("intro.Rmd", "matmet.Rmd", "disc.Rmd"))
cite_keys <- sapply(file_paths, rbbt::bbt_detect_citations) 
cite_keys <- unlist(cite_keys, use.names = FALSE)
rbbt::bbt_write_bib("./paper/refs.bib", keys = cite_keys, overwrite = T)
## end refs.bib ##
```

```{r intro, child="paper/intro.Rmd"}

```

```{r matmet, child="paper/matmet.Rmd"}

```

```{r res, child="paper/results.Rmd"}

```

```{r disc, child="paper/disc.Rmd"}

```

# Acknowledgements {-}

<!-- If you're reading this, thank you for taking the time to read the source 
documents. You're awesome! -->

# References cited

### 2021-08-19 analysis/paper/abstract.Rmd
add sections to manuscript

# Abstract

<!-- One or two sentences providing a basic introduction to the field, comprehensible to a scientist in any discipline -->

<!-- Two to three sentences of more detailed background, comprehensible to scientists in related disciplines -->

<!-- One sentence clearly stating the general problem being addressed by this particular study -->

<!-- One sentence summarising the main result (with the words "here we show" or their equivalent) -->

<!-- Two or three sentences explaining what the main result reveals in direct comparison to what was thought to be the case previously, or how the main result adds to previous knowledge -->

<!-- One or two sentences to put the results into a more general context -->

<!-- Two or three sentences to provide a broader perspective, readily comprehensible to a scientist in any discipline, may be included in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced by their inclusion. Under these circumstances, the length of the paragraph can be yp to 300 words. (This example is 190 words without the final section, and 250 words with it).

### 2021-08-19 analysis/paper/disc.Rmd
add sections to manuscript

# Discussion

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likeliness of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples  were similar across all treatments
(`r paste(min(perc$total), max(perc$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(`r format(count_sol$total[4], big.mark = ",")`) 
and potato
(`r format(count_sol$total[3], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment; although, starch availability is unlikely
to be a main factor in starch incorporation. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc$total), max(perc$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010].

<!--
Given the low proportion of starch granule retention in our dental calculus model,
and the limited difference between treatments, we can conjecture that starch 
availability is not a main factor in starch incorporation. While there was a
proportional incorporation, i.e. more granules in the solution led to more granules
in the calculus, the difference between solutions [a number/percentage] was of a
higher magnitude than the difference between samples <!-- a number/percentage-->

<!-- although this study extrapolated counts to full sample content, which is
rarely done in archaeological studies-->

<!-- Scientific findings as extension of previous work -->
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of 
starch granules in dental calculus, and that starch granules extracted from dental 
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues [-@leonardDentalCalculus2015] found individual
calculus samples to be a poor predictor of diet in a population, as many of the 
consumed plants were missing from some individual samples, but were present in others.   
Power and colleagues [-@powerChimpCalculus2015] 
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules, and plants that 
produce predominantly larger starch (10-20 $\mu$m) granules were over-represented.
Both studies concluded that the dietary breadth of a population can
be captured in a sufficiently large sample, but interpretations on individual diets
from microremains in dental calculus is tenuous, at best 
[@leonardDentalCalculus2015; @powerChimpCalculus2015].

<!-- concentrations of solutions vs. starches in a mouthful of food 
starches in a potato, etc. 
Contact Steph -->

<!-- secondary findings -->

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
diameter were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10 -- 20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).
This was especially true for the potato starches, which can reach up to 100 $\mu$m
in diameter, whereas wheat generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004].
This is consistent with the findings by Power and collagues [-@powerChimpCalculus2015]
mentioned previously, which found that the larger (defined as around 10 -- 20 $\mu$m) 
granule-producing plants were overrepresented; although, the representation of granules 
larger than 20 $\mu$m in that study is unclear. Since those results were obtained
on a modern population of chimps that were only buried for a short period of time, 
degradation was likely limited [@powerChimpCalculus2015]. The taphonomy of
starch granules has been addressed previously, and depends on burial environment
(pH, temperature, microorganisms), as well as processing of the granules prior to
deposition [@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004]. 
The potential for detrimental diagenetic effects on starch granules increases as 
the size and amylose content decreases
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. 
Combined, the effect of intra-oral starch incorporation
and the bias against large granules, together with the increased effect of taphonomy
on  small granules, would eliminate a large portion of the consumed starch granules,
and would explain why the granules extracted from archaeological dental calculus 
are often sparse.

Additionally, the size of the calculus deposit seems to influence the quantity of 
starches extracted from the calculus, as we found a strong positive correlation
between biofilm size and retained starch granules
(Figure \@ref(fig:cor-plot)), which contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014]. 
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot2));
while the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
The lack of a correlation
in archaeological contexts could be due to diagenetic effects; or, the absence of 
amylase activity in our study could be impacting our interpretations.

<!-- How are starch granules trapped in dental calculus? Does anyone actually know? -->

The mechanism by which starch granules are incorporated into plaque and calculus 
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering 
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of significant age 
[@henryNeanderthalCalculus2014; @buckleyDentalCalculusCooking2014; @wuDietEarliest2021]. 
Studies have also shown that not all starch granules come from a dietary source. 
Other pathways include cross-contamination from plant interactions in soil, such 
as palm phytoliths adhering to the skin of sweet potatoes 
[@trompDietaryNondietary2015], 
or accidental ingestion not related to food consumption 
[@radiniFoodMultiplePathways2017; @radiniMedievalWomenEarly2019].

When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are 
only briefly retained in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition. These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by $\alpha$-amylase [@kashketFoodParticles1996].
Starch granules that are trapped in crevices and channels on the surface of the
mineralised plaque would be (at least to some extent) protected from salivary
clearance and mechanical cleaning actions of the tongue and lips. This may explain 
why a previous study found starches more commonly in clusters than dispersed over
the surface of the dental calculus [@powerSEMCalculus2014]. The authors hypothesised 
that the granules are either deposited in clusters, or group together in cracks and
crevices within the mineral matrix. Unmineralised Lacunae and channels within the 
calculus matrix have been shown to contain viable
bacteria, and may also be large enough to contain starch granules. These can 
range from the width of a single cell, to multiple-cell
width [@tanCalculusUltrastructure2004], of which the latter could feasibly contain 
starch granules. 
If this is indeed one of the incorporation mechanisms of starch granules, then the 
limit of incorporated starches is set by the number and size of channels. Since 
this is likely controlled by the size of the calculus deposit, then the number of 
starches will increase as the size of the biofilm increases, which is consistent 
with our results 
(Figures \@ref(fig:cor-plot1) and \@ref(fig:cor-plot2)).
The size bias against large granules (>20 $\mu$m) may give further credence to the 
incorporation pathway of starch granules primarily in cracks and crevices in the 
calculus, as the smaller starch granules have an advantage over larger granules, 
and can be stored in larger quantities.  
While potentially protected against clearance, granules trapped 
in plaque/calculus will still be susceptible to hydrolysis, as $\alpha$-amylase has
the ability to bind to both tooth enamel and bacteria within a biofilm and retain
a portion of its hydrolytic activity
[@nikitkovaStarchBiofilms2013; @tanStudyBacterialViability2004; @tanCalculusUltrastructure2004].
<!-- These channels and pockets of organic matter within the inorganic matrix are used by the 
bacteria for transportation of nutrients, and their persistence in the otherwise 
mineralised matrix may be the result of the differing 
rate of mineralisation displayed by the bacteria [@tanStudyBacterialViability2004].-->
The susceptibility of granules to hydrolysis depends on the crystallinity and size
of the starch granule, as well as the mode of processing. As mentioned previously,
smaller starch granules are more susceptible to enzymatic degradation, as are 
pre-processed starches (e.g. by cooking), while dehydrated starches will have a
reduced susceptibility
[@lingstromStarchyFood1994; @francoStarchDegradation1992; @haslamDecompositionStarch2004; 
@bjorckStarchProcessing1984; @henryCookingStarch2009]. 
<!--
Starches also have the ability to adhere to certain bacteria [@toppingResistantStarch2003],
which would allow starches to attach to bacterial communities within the biofilm.
Following deposition and/or mineralisation events, the granules that are able to 
resist hydrolysis and 
mechanical removal will then be trapped in pockets within the mineral matrix, and
will temporarily be protected from shearing action and salivary 
clearance. -->
This means, that the dietary picture we obtain from starch granules extracted from 
dental calculus reflects a number of individual, dietary, and environmental (both
pre- and post-mortem environments) factors, and that this will likely be a
somewhat random snapshot given the irregularity of plaque mineralisation within
and between individuals [@jinSupragingivalCalculus2002; @jepsenCalculusRemoval2011]. 
Although, certain factors may increase the likelihood of incorporation, such 
as abundance, size, and morphology of granules.

<!-- in summary: we know very little about the mechanism behind how starches 
become trapped in dental calculus, but it would be nice to find out more. -->

<!-- limitations of the platform show which questions would be unanswered -->
This article presents preliminary research exploring the potential of an oral 
biofilm model for dietary research in archaeology. A limitation of this study 
is the absence of amylase in the model. The presence of amylase will likely affect 
the total granule counts as well as the size ratios, as smaller starches may be more 
susceptible to hydrolysis 
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. 
In future experiments, we recommend the addition of $\alpha$-amylase to the model
in concentrations similar to those found in the oral cavity <!-- provide concentration and citation.--> 
<!-- foreshadowing... -->
While we are able to show the differing 
incorporation caused by absolute counts available, this
was merely a side-effect of the difference in the number of granules in potato and 
wheat solutions of the same concentration (w/v). Further research should test 
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown 
negligible morphological changes caused by exposure to EDTA 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that 
was feasible to count in a reasonable time period, so we calculated the total 
counts by extrapolating from three slide transects. 
Thus, we reasonably assume that the three transects are a good representation 
of the entire slide, and that the distribution of all granules on the slide is 
relatively homogenous.   
Finally, we only used native starches in the experimental procedure and the results 
will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
Based on counts obtained by Leonard
and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

<!-- How this engineering will facilitate future research -->
The oral biofilm model described in this study, and in 
Bartholdy and colleagues (in prep.),<!-- insert citation --> provides a method to directly
explore the incorporation and extraction of dietary compounds from dental calculus 
in a controlled laboratory setting. The addition of known species in separate
samples also removes the identification bias against the smaller undiagnostic starches.<!-- rephrase -->
It can also address the call for more baseline testing of biases associated with 
dietary research conducted on dental calculus 
[@lemoyneCalculusPretreatments2021].
<!-- How it will facilitate the specific topic of research (e.g. archaeology) -->
*In vivo* studies on humans and non-human primates will be limited by the variability 
occurring
between individuals, and difficulties documenting the entirety of starch granules
entering the oral cavity at any given moment, especially accidental intake. 
Our experimental setup allows us a
higher degree of control over the factors that influence starch incorporation, such as
dietary intake, processing of starches leading to differential survivability, 
and inter- and intra-individual variation in plaque accumulation and 
mineralisation.
The latter is especially difficult to control *in vivo* as it is influenced by 
numerous factors including genetics, diet, and tooth morphology [@jepsenCalculusRemoval2011]<!-- more citations -->.  
Further research is necessary to explore the mechanisms of the intra-oral starch 
incorporation and retention in dental calculus, and the dental calculus model
presented in this study is uniquely suited to explore these questions and 
may improve interpretations of dietary research in past populations.
<!-- still need to replicate results and do more science, but pretty cool, right? -->

# Conclusions

Our preliminary study shows that a very small proportion of the ingested starch
granules are retained in a dental calculus model.

Calculus has a low capacity for retention of starch granules, and this effect is
compounded by diagenetic effects in archaeological remains, resulting in low overall
counts of extracted granules. Additionally, the size of granules will influence the 
likelihood of incorporation, as large starches have a decreased incorporation, medium
starches increased, and small starches the only representative of intake.

Given the direct proportional incorporation of starches with quantity of exposure<!-- rephrase -->, 
we can say that the proportion of starches consumed will in many cases be reflected 
in the quantity
of starches extracted from the dental calculus, at least in modern calculus samples
unaffected by diagenesis.
Whether or not this also applies to archaeological samples remains to be tested.

The size of calculus deposit also influences the maximum capacity of granule incorporation; as the size of the deposit increases, so does the absolute count of
incorporated granules.

The data from this study are not sufficient to confirm or deny any proposed methods
of starch incorporation; however, while we are unable to address the mechanism(s) of
starch incorporation with the current experimental setup, this is something that can
potentially be addressed with the calculus model presented in this paper.

<!-- Forward-looking statements about the field as a whole -->
While we have shown multiple factors that influence the likelihood 
of incorporation, it seems the process is somewhat stochastic, and possibly related 
to the biofilm mineralisation events (as discussed above), which are irregular and
related to various factors<!-- citation needed -->. 
This means that the dietary components extracted from dental calculus provide a 
snapshot of the diet around the time of the mineralisation event, which may not 
be an accurate picture of the diet as a whole.

The dental calculus biofilm model presented in this study is a powerful tool to 
investigate various aspects
of dietary research.

### 2021-08-19 analysis/paper/fig_tab.Rmd
add sections to manuscript

---
title: "Tables and Figures"
output:
  bookdown::pdf_document2:
    toc: no
---

```{r setup-figs-and-tbls, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
#code to include R scripts in .Rmd: r knitr::spin_child("tables.R")
```

\listoftables
\listoffigures

<!-- ### Tables ### -->

<!-- ### Figures ### -->

```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring). l = large, m = medium, s = small."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

<!-- change colour of medium starches in D to match C -->
```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring). l = large, m = medium, s = small."}
#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

### 2021-08-19 analysis/paper/intro.Rmd
add sections to manuscript

# Introduction

<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
Dental calculus, mineralised dental plaque, has proven to contain a wealth of 
dietary information in the form of plant microfossils 
[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
proteins [@hendyProteomicCalculus2018; @warinnerDirectEvidenceMilk2014],
and other organic residues [@buckleyDentalCalculusCooking2014]. 
It has also proven to be able to retain and preserve dietary information for millenea,
providing detailed dietary information on past populations 
[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
and species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].

<!-- more detailed background, comprehensible to scientists in related disciplines -->

Until recently, few studies directly investigated the presence of plant microremains
from the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. The ability to detect coca
consumption by looking at human dental calculus was explored by Klepinger and Kuhn
[-@klepingerCalculusCoca1977], and starches and phytoliths were extracted from 
human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].

In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has been, 
to a larger extent, acknowledged. The applications of dental calculus span a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncientCalcified2013; @warinnerPathogensHostImmunity2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@mickleburghNewInsightsConsumption2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculusCooking2014; @hendyProteomicCalculus2018], 
and nicotine-use [@eerkensDentalCalculusSource2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

<!-- starches from dental calculus -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
knowledge of information-loss caused by these pathways and the methods we use to 
extract and analyse dental calculus and make inferences on past diets. 
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. Other studies have been conducted on modern populations 
[@leonardDentalCalculus2015] and primates [@powerChimpCalculus2015], both with
well-documented dietary intake, to reconcile our interpretations on the recovery 
of microremains (both phytoliths and starches) from dental calculus and how well
they represent actual dietary intake. These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved when studying living organisms. Dental calculus is a complex 
oral biofilm with a
multifactorial aetiology and variable formation rates both within and between
individuals [@jepsenCalculusRemoval2011].
Additionally, the effects of the many different methods
of starch processing, as well as post-depositional processes, are still being 
explored [@graneroStarchTaphonomy2020].

<!-- clearly stating the general problem being addressed by this particular study -->
<!-- Dental calculus research has its limitations, all of which I will solve with this
single method. -->
In this exploratory study, we use an oral biofilm model to investigate the 
retention of starch granules within dental calculus, in a controlled laboratory 
setting. Our main questions concern the representation of granules extracted
from the calculus compared to the actual intake. How much of the original diet is 
incorporated into the calculus, and how much is recovered? 
Is there differential loss of information from specific dietary markers that affects 
the obtained dietary information, and how does this affect the representation of diet from
extracted microremains? We find that a limited proportion of the starch input is actually 
retained in the calculus, as well as a shift in the size ratios of individual starch granules
that are incoporated into the calculus. We also show that the number of starch
granules that are incorporated increases as the size of the calculus deposit 
increases.

### 2021-08-19 analysis/paper/matmet.Rmd
add sections to manuscript

# Materials and Methods

```{r protocol-fig, fig.cap="Overview of experiment protocol"}
knitr::include_graphics(here("analysis/figures", "ExpL_protocol.png"))
```

<!-- remove sampling from protocol picture -->

## Biofilm formation

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues [-@sissonsMultistationDentalPlaque1991] and
Shellis [-@shellisSyntheticSalivaCultural1978]. The setup comprises a polypropylene 
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which 
is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins.

The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
described by Sissons and colleagues [-@sissonsMultistationDentalPlaque1991]. 
It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl~2~, 
5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationDentalPlaque1991], 6.4 mmol/l NaHCO~3~ 
[@shellisSyntheticSalivaCultural1978], 2.5 mg/l haemin. This is subsequently 
adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120&deg;C, 
1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
and 1 mmol/l arginine [@sissonsMultistationDentalPlaque1991].

<!-- add to protocols.io -->
Fresh whole saliva (WS) for inoculation was provided by a 31-year-old male donor 
with no history of caries, who abstained from oral hygiene for 24 hours, and no 
food was consumed two hours prior to donation. No antibiotics were taken up to six 
months prior to donation. 
The saliva was filtered through a sterilised (with bleach) nylon cloth to remove
particulates.
Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
20% glycerine for four hours at 36&deg;C, to allow attachment of the
salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
substrata were transferred to a new plate containing 1 ml/well AS and incubated 
at 36&deg;C, 30 rpm. The inoculation process was repeated on days 3 and 5. 
AS was partially refreshed once per day and fully refreshed every three days, 
throughout the experiment, by transferring the substrata to a new plate containing 
AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
(days 0, 3, and 5), where the samples only received one sucrose treatment after
inoculation.

Starch treatments were initiated on day 9 to avoid starch granule counts being 
affected by $\alpha$-amylase hydrolysis. An $\alpha$-amylase (EC 3.2.1.1) activity
assay was conducted to confirm that no amylase was present in the system (see 
Bartholdy et al. ???? <!-- ## [bartholdy] ## --> for more details). Starch treatments replaced sucrose treatments, occurring twice per day
for six minutes. The starch treatments involved transferring the substrata to a 
new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
concentrations (w/v) wheat and potato. All starch solutions were created in a 5%
(w/v) sucrose solution. Before transferring biofilm samples to the starch treatments, 
the starch plates were agitated to keep the starches in suspension in the solutions, 
and during treatments, the rpm was increased to 60.

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991], 
and (0.04 g/l MgCl). The substrata were submerged in 1 ml/well CPMU five times daily, 
every two hours, for six minutes. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This cycle was repeated 
for 10 days until the end of the experiment on day 24 
(see \@ref(fig:protocol-fig) for an overview of protocol). More detailed protocols
are available on protocols.io<!--insert DOI-->.

All laboratory work was conducted in sterile conditions under a laminar flow hood 
to prevent starch and bacterial contamination. Control samples were included to 
detect starch contamination.

## Starch counts

Starch granule counts were divided into three size categories: small (<10 $\mu$m),
medium (10 -- 20 $\mu$m), and large (>20 $\mu$m). Counts for small wheat and
potato starches were combined for the mixed treatment samples, as it is not 
possible to distinguish the small starch granules between the two species.

### Treatment solutions

A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
(Zeiss Axioscope A1). Counting starches on a full slide was not feasible, so three slide transects were 
counted (at 1/3, 1/2, and 2/3 of the slide), and the sample counts were extrapolated 
(see Supplementary material for more details).

### Extraction from samples

Extraction of starches from the calculus samples was performed by dissolving the
calculus in ethylenediaminetetraacetic acid (EDTA), as it has been shown to
have little to no effect on starch granules (especially native granules)
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When dispersing the liquid (EDTA) sample on the slide, the sample was homogenised using the pipette. 
It was not possible to count all wheat starches
on a slide, so wheat and mix counts were extrapolated as described for the treatment
solutions.

## Statistical analysis

Statistical analysis was conducted in R versions 4.0.5 ("Shake and Throw") 
and 4.0.3 ("Bunny-Wunnies Freak Out") [@R-base].
<!-- R packages used -->
A one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
as the grouping variable (GV) was conducted to explore the effect of the different 
starch treatments on biofilm growth.

To calculate size proportions, mean counts for each treatment were taken across 
both experimental plates for each treatment, resulting in a mean count for each
granule size category within each treatment.

Pearson's r was conducted on sample weight and total starch count, as well as sample
weight and starch count per mg calculus. The total count for each sample within a
treatment was standardised by z-score to account for the differences in magnitude
between the potato and wheat counts.

Scripts and data will be made available on osf.io upon publication.
Until then, they can be found on the first author's GitHub repository
(https://github.com/bbartholdy/byocStarch).

<!-- script and data availability: provide link (or doi?) to osf.io repository -->

### 2021-08-19 analysis/paper/results.Rmd
add sections to manuscript

# Results

```{r setup-res, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
#opts_knit$set(root.dir = '/Users/username/path_to_project')
#corr_counts <- readr::read_csv(here("analysis/data/derived_data", "corr_counts.csv"))
#corr_comb <- readr::read_csv(here("analysis/data/derived_data", "corr_comb.csv"))
#sol_corr <- readr::read_csv(here("analysis/data/derived_data", "sol_corr.csv"))
#source(here("analysis/scripts/analysis.R"))
#source(here("analysis/scripts/tables.R"))
#source(here("analysis/scripts/figures.R"))
#source(c("tables.R", "figures.R", "analysis.R"))
#code to include R scripts in .Rmd: r knitr::spin_child("tables.R")
```

All samples yielded sufficient biofilm growth and starch incorporation to be 
included in the analysis, resulting in a total of 48 biofilm samples (two plates of 24),
45 of which were used for analysis (three samples were set aside for separate analysis).
Most control samples contained no starch granules, while some contained negligible
quantities (supp. mat).

<!-- exclude problematic samples: st1C5 wheat and st2C6 potato, and  st2C5 
Use more formal outlier detection -->

## Treatment type had no effect on biofilm growth

```{r anova-tbl}
knitr::kable(anova_tbl, 
             col.names = c("Treatment", "Mean", "SD", "Min", "Max"),
             caption = "Summary statistics for biofilm dry-weights (in mg) by treatment.")
```

A one-way ANOVA suggests that the type of starch used during the biofilm 
growth period had a minimal effect on the growth of the biofilm 
(expressed as total dry weight of the sample), F(`r paste(treat_df[1], treat_df[2], sep = ", ")`)
= `r treat_f.stat`, p = `r treat_p`. 
A summary of sample weights is available in Table \@ref(tab:anova-tbl).

## Starch counts

<!-- starch counts in the solutions -->

```{r solution-count-out}
count_sol_out <- count_sol %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))

### adding the 'both' row to the data frame ###
count_sol_comb_out <- count_sol_comb %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))

count_sol_out <- count_sol %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")"))) %>%
  ungroup() %>%
  add_row(tibble_row(count_sol_comb_out), .before = 3)
```

```{r solution-count-tbl}
cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
knitr::kable(count_sol_out, col.names = cnames, 
             caption = "Mean starch counts from solutions, including the
             proportional makeup of the different sizes of granules.")
```

<!-- starch counts in the samples -->

```{r sample-count}
count_sample <- count_sample %>%
  ungroup() %>%
  add_row(count_sample_comb, .after = 2)

# convert columns to contain mean count and percentage of total count
count_samp_per <- count_sample %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), ")")))

# count_samp_comb_per <- count_sample_comb %>%
#   mutate(across(c(s, m, l, total), 
#                 function(x) 
#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))

# count_samp_comb_out <- tibble(count_samp_comb_per, count_sample_comb_sd, .name_repair = "unique") %>%
#   select(1,2,3,7,4,8,5,9,6,10) #%>%
  #add_row(tibble_row(count_sample_comb_out), .before = 3)

count_samp_out <- tibble(count_samp_per, count_sample_sd, .name_repair = "unique") %>%
  select(1,2,3,7,4,8,5,9,6,10) #%>%
  #add_row(tibble_row(count_sample_comb_out), .before = 3)
```

```{r sample-count-tbl}
cnames <- c("Treatment", "Starch", "Small (%)", "SD", "Medium (%)", "SD", 
            "Large (%)", "SD", "Total (%)", "SD")
knitr::kable(count_samp_out, col.names = cnames, 
             caption = "Mean starch counts extracted from samples with standard 
             deviation (SD), including the proportion of granule sizes of the total count.")
```

It was not possible to differentiate between potato and 
wheat starches smaller than ca. 10 $\mu$m,
small potato starches were counted as wheat starches in the mixed-treatment samples. This is reasonable given that the small potato granules make up an insignificant proportion of the 
total count of small granules within mixed-treatment solutions which are predominantly
wheat granules
(`r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`).
<!-- count of small granules from wheat solution divided by the sum of small granules 
from the potato and wheat solutions -->

<!-- absolute solution counts compared to absolute sample counts. does exposure
to higher quantities of starches mean more incorporated starches? -->
The separate wheat and potato solutions were made with a 0.25% (w/v) starch
concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
starch, with a total concentration of 0.50% (w/v).
The mixed-treatment had the highest absolute count of starch granules in solution,
while the biofilms exposed to the wheat solution preserved the greatest number of
granules. The potato treatment had the lowest absolute counts in both the solution 
and in the biofilm samples <!-- insert counts? -->
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).

<!-- look at proportions of incorporated starches in each sample before looking at overall means -->

### Proportion of available starches incorporated in samples

```{r sample-perc}
# convert proportions to percent labels
perc_comb <- prop_comb %>% 
  mutate(across(c(s, m, l, total), scales::percent, 0.001))

perc <- prop %>%
  mutate(across(c(s, m, l, total), scales::percent, 0.001)) %>% 
  add_row(tibble_row(perc_comb), .before = 3)
```

```{r sample-prop-tbl}
cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
knitr::kable(perc, col.names = cnames, 
             caption = "The mean percentage of starches from the solutions that 
              were incorported into the samples.")
```

The proportion of total starches from the solutions that were incorporated 
into the samples ranged from 
`r paste(min(perc$total), max(perc$total), sep = " to ")`,
with potato granules being more readily incorporated than wheat in both the
separated- and mixed-treatment samples
(Table \@ref(tab:sample-prop-tbl)). 
There is an inverse relationship between the absolute starch count in the solutions 
and the proportional incorporation of starches in the biofilm samples, i.e. potato 
had the lowest absolute count in solutions, but the highest proportional 
incorporation, and vice versa for the mixed treatment.

<!-- separated treatments -->

<!-- mixed treatments -->
Wheat incorporation was most affected in the mixed-treatment samples, with only
`r paste0(perc$total[2])` of the total available starches being incorporated into
the sample, compared to
`r paste0(perc$total[4])`
in the separated wheat treatment.

### Size ratios differ between solutions and samples

<!-- size proportions -->
<!-- pie charts -->

```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring). l = large, m = medium, s = small."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

<!-- change colour of medium starches in D to match C -->
```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring). l = large, m = medium, s = small."}
#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```
<!-- Combine to a single figure? -->

<!-- overall size trends -->

Overall, medium starch granules had a higher mean rate of incorporation 
(`r scales::percent(mean(prop$m, na.rm = T), 0.001)`) 
than small 
(`r scales::percent(mean(prop$s, na.rm = T), 0.001)`) 
and large 
(`r scales::percent(mean(prop$l, na.rm = T), 0.001)`) 
starch granules across all treatments, while large potato starches had the lowest 
rate of incorporation across all treatments.

<!-- separated and mixed treatments -->
The difference in incorporation between the size categories resulted in a change
in size ratios between the original starch solutions and the extracted samples.
Large potato granules (> 20 $\mu$m) were most affected, with a 
`r paste0(abs(diff_sep$potato[3]), "%")` 
decrease in relative abundance in the potato-only treatment, and a
`r paste0(abs(diff_mix$potato[3]), "%")` 
decrease in mixed treatments. Medium granules increased in relative abundance 
across all samples, while small granules decreased in wheat treatments and 
increased in potato treatments 
(Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).

<!-- Chi square test for size counts? -->

<!-- Proportion and size ratios are the most important points to be made -->

### Biofilm weight correlated positively with extracted starch counts

```{r cor, echo=FALSE}
# prepare cor output for reporting
cor_r <- signif(count_cor$estimate, 3)
cor_ci <- signif(count_cor$conf.int, 3)
cor_p <- if(count_cor$p.value > 0.001){
  signif(count_cor$p.value, 3)
} else {
  paste("< 0.001")
}

cor2_r <- signif(starch_cor$estimate, 3)
cor2_ci <- signif(starch_cor$conf.int, 3)
cor2_p <- if(starch_cor$p.value > 0.001){
  signif(starch_cor$p.value, 3)
} else {
  paste("< 0.001")
}
# Convert r correlation result to text
direct_cor <- ifelse(cor_r < 0, "negative", "positive")
strength_cor <- if(cor_r >= 0.8){
  "very strong"
} else if(cor_r < 0.8 & cor_r >= 0.6){
  "strong"
} else if(cor_r < 0.6 & cor_r >= 0.4){
  "moderate"
} else if(cor_r < 0.4 & cor_r >= 0.2){
  "weak"
} else{
  "very weak"
}
```

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments."}
pl_cor
```
<!-- with Z-score standardisation -->

Pearson's *r* suggests a 
`r strength_cor` `r direct_cor` 
correlation between the total weight of the biofilms and the total starch count 
(standardised by z-score) extracted from the samples across treatments, 
*r* = `r cor_r`, 
90%CI[`r paste(cor_ci[1], cor_ci[2], sep = ", ")`],
p = `r cor_p`
(Figure \@ref(fig:cor-plot)).

The same test was applied to total biofilm weight and starch count per mg 
calculus (also standardised by z-score), resulting in a weak positive correlation, 
*r* = `r cor2_r`, 
90%CI[`r paste(cor2_ci[1], cor2_ci[2], sep = ", ")`],
p `r cor2_p`
(Figure \@ref(fig:cor-plot2)).

<!-- size ratios in the larger deposits vs the smaller deposits?
  Are larger starches more proportionately represented in larger deposits?
  -->

```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
pl_cor2
```

### 2021-08-19 analysis/supplementary-materials/supp-mat.Rmd
add sections to manuscript

---
title: Supplementary Materials
---

## Metadata

**raw_counts**
<!-- insert data type for each column -->
Counts represent the absolute number of starches counted on a slide

sample: Sample number.
plate: Plate number that the sample came from.
row: Which row on the plate the sample came from.
s: Small starch count.
m: Medium starch count.
l: Large starch count.
total: Sum of s, m, and l.
treatment: Treatment solution to which the samples were exposed.
starch: Type of starch that was counted
weight: Weight of the biofilm sample
vol: Total volume of EDTA in which the sample was dissolved.
portion_slide: Proportion of the microscope slide that was counted.

## Count corrections

Sample counts were extrapolated using the following formula:

$$ \text{Corrected count} = \text{raw count} \times (\text{portion of slide})^{-1} \times \frac{V_{sample}}{V_{slide}} $$

Solution counts were extrapolated to the quantity in a 1 ml solution with
the following formula:

$$ \text{corrected count} = \text{raw count} \times \frac{29}{3} \times 100 \mu l $$

## Functions

### 2021-08-19 analysis/scripts/analysis.R

@@ -1,0 +1,49 @@
+# Analysis script
+
+library(here)
+library(tidyverse)
+library(broom)
+
+corr_counts <- readr::read_csv(here("analysis/data/derived_data/corr_counts.csv"))
+corr_comb <- readr::read_csv(here("analysis/data/derived_data/corr_comb.csv"))
+sol_corr <- readr::read_csv(here("analysis/data/derived_data/sol_corr.csv"))
+
+# ANOVA -------------------------------------------------------------------
+
+anova_treat <- corr_comb %>%
+  group_by(treatment) %>%
+  do(tidy(aov(weight ~ treatment, data = corr_comb)))
+treat_df <- c(anova_treat$df[1], anova_treat$df[2])
+treat_f.stat <- signif(anova_treat$statistic[1], 3)
+treat_p <- signif(anova_treat$p.value[1], 3)
+
+
+# Chi-squared --------------------------------------------------------------
+
+# Chi-squared test on size ratios
+# Expected: solution ratios
+# Observed: sample ratios
+
+# Pearson correlation -----------------------------------------------------
+
+# Standardise counts using Z-score
+z_count <- corr_comb %>%
+  filter(treatment != "control") %>%
+  group_by(treatment) %>%
+  mutate(std_total = (total - mean(total, na.rm = T)) / sd(total, na.rm = T)) %>%
+  mutate(log_total = log(total))
+
+# Pearson correlation
+count_cor <- cor.test(z_count$weight, z_count$std_total,
+                      use = "pairwise.complete", conf.level = 0.9)
+
+# Calculate number of starches per mg calculus
+
+starch_per_mg <- corr_comb %>%
+  mutate(starch_per = total / weight) %>%
+  group_by(treatment) %>%
+  mutate(std_starch_per = (starch_per - mean(starch_per, na.rm = T)) / sd(starch_per, na.rm = T)) %>%
+  ungroup()
+
+starch_cor <- cor.test(starch_per_mg$weight, starch_per_mg$std_starch_per,
+                       use = "pairwise.complete", conf.level = 0.9)

### 2021-08-19 analysis/scripts/figures.R

@@ -1,0 +1,201 @@
+library(tidyverse)
+
+corr_counts <- readr::read_csv(here("analysis/data/derived_data/", "corr_counts.csv"))
+corr_comb <- readr::read_csv(here("analysis/data/derived_data/", "corr_comb.csv"))
+sol_corr <- readr::read_csv(here("analysis/data/derived_data/", "sol_corr.csv"))
+
+# group by starch category
+# the mix treatment/solution is separated into starch types
+sol_sep <- sol_corr %>%
+  group_by(solution, starch) %>%
+  summarise(across(c(s, m, l, total), mean))
+
+samp_sep <- corr_counts %>%
+  group_by(treatment, starch) %>%
+  filter(treatment != "control") %>%
+  summarise(across(c(s, m, l, total), mean, na.rm = T))
+
+# transpose data frames to create pie charts
+# solution count data
+ratio_sol_sep <- sol_sep %>%
+  filter(solution != "mix") %>% # removed mixed treatment
+  select(!total) %>%
+  t %>%
+  as.data.frame %>%
+  rownames_to_column("size") # create size variable
+colnames(ratio_sol_sep)[2:3] <- c("potato", "wheat")
+ratio_sol_sep <- ratio_sol_sep[-c(1:2),]
+ratio_sol_sep <- ratio_sol_sep %>%
+  mutate(across(c(potato, wheat), as.character)) %>%
+  mutate(across(c(potato, wheat), as.numeric)) %>%
+  mutate(wheat = wheat / sum(wheat)) %>%
+  mutate(label_wheat = scales::percent(wheat)) %>%
+  mutate(potato = potato/ sum(potato)) %>%
+  mutate(label_potato = scales::percent(potato))
+
+# sample count data
+ratio_samp_sep <- samp_sep %>%
+  filter(treatment != "mix") %>% # removed mixed treatment
+  select(!total) %>%
+  t %>%
+  as.data.frame %>%
+  rownames_to_column("size") # create size variable
+colnames(ratio_samp_sep)[2:3] <- c("potato", "wheat")
+ratio_samp_sep <- ratio_samp_sep[-c(1:2),]
+ratio_samp_sep <- ratio_samp_sep %>%
+  mutate(across(c(potato, wheat), as.character)) %>%
+  mutate(across(c(potato, wheat), as.numeric)) %>%
[…]

### 2021-08-19 analysis/scripts/script1.R

@@ -1,0 +1,73 @@
+# R Script for correcting counts according to volume and proportion of slide counted
+
+# Dependencies ------------------------------------------------------------
+
+library(magrittr)
+library(dplyr)
+library(here)
+library(readr)
+
+# Upload data -------------------------------------------------------------
+
+ # starch counts in samples
+raw_counts <- readr::read_csv(here("analysis/data/derived_data/", "starch_counts.csv"))
+
+ # starch counts in solutions
+sol_counts <- readr::read_csv(here("analysis/data/derived_data/", "solution_counts.csv"))
+
+# Functions ---------------------------------------------------------------
+
+ # function to correct counts for volume and proportion of slide counted
+
+correct_count <- function(raw, prop, vol){
+  corrected <- raw * (1 / prop) * (vol / 20)
+  corrected
+}
+
+ # function to extrapolate solution counts to 1 ml exposure per day
+
+extrap_count <- function(raw, prop, vol, days){
+  extrap <- raw * (1 / prop) * 100 * days
+  extrap
+}
+
+# Outlier detection and removal -------------------------------------------
+
+raw_counts %>%
+  filter(treatment != "control") %>%
+  ggplot() +
+  geom_boxplot(aes(x = starch, y = total, fill = starch))
+
+# Only lower outliers were considered, as this would suggest that there may have
+  # been a problem with the sample and incorporation of starches. Given that all
+  # samples were treated equally, there is no reason to consider upper outliers
+  # problematic.
+
+
+# Solution counts ---------------------------------------------------------
+
+sol_corr <- sol_counts %>%
+  mutate(across(c(s, m, l, total), extrap_count, portion_slide, vol_slide, days = 16))
+readr::write_csv(sol_corr, here("data", "sol_corr.csv"))
+
+# Apply correction --------------------------------------------------------
+
+corr_counts <- raw_counts %>%
+  #filter(treatment != "control") %>%
+  mutate(across(c(s, m, l, total), correct_count, portion_slide, vol)) %>%
+  select(!c(vol, portion_slide)) %>%
+  filter(sample != "st2C6")
+readr::write_csv(corr_counts, here("data", "corr_counts.csv"))
+
+# Combining mixed treatment counts ----------------------------------------
+
+  # combining potato and wheat counts from mixed treatment samples
+
+# merge two adjacent rows and add variables s, m , l, total
+
+corr_comb <- corr_counts %>%
[…]

### 2021-08-19 analysis/scripts/tables.R

@@ -1,0 +1,98 @@
+# Script for tables
+
+library(here)
+library(tidyverse)
+
+corr_counts <- readr::read_csv(here("analysis/data/derived_data/", "corr_counts.csv"))
+corr_comb <- readr::read_csv(here("analysis/data/derived_data/", "corr_comb.csv"))
+sol_corr <- readr::read_csv(here("analysis/data/derived_data/", "sol_corr.csv"))
+
+# Sample weights ----------------------------------------------------------
+
+# create data frame with mean and sd by treatment
+anova_tbl <- corr_comb %>%
+  group_by(treatment) %>%
+  summarise(mean = mean(weight, na.rm = T),
+            sd = sd(weight, na.rm = T),
+            min = min(weight, na.rm = T),
+            max = max(weight, na.rm = T)) %>%
+  mutate(across(where(is.numeric), signif, 3))
+
+
+# Solution counts ---------------------------------------------------------
+
+count_sol <- sol_corr %>% # create a data frame with solution count means
+  group_by(solution, starch) %>%
+  summarise(across(c(s, m, l, total), mean, na.rm = T))
+
+# add a row containing the combined counts for the mixed treatment
+count_sol_comb <- sol_corr %>%
+  filter(solution == "mix") %>%
+  group_by(starch) %>%
+  summarise(across(c(s, m, l, total), mean, na.rm = T)) %>%
+  ungroup() %>%
+  summarise(across(c(s, m, l, total), sum, na.rm = T)) %>%
+  mutate(starch = "both",
+         solution = "mix") %>%
+  select(solution, starch, s, m, l, total)
+
+# Sample counts -----------------------------------------------------------
+
+count_sample <- corr_counts %>% # create a data frame with treatment count means
+  group_by(treatment, starch) %>%
+  filter(treatment != "control") %>%
+  summarise(across(c(s, m, l, total), mean, na.rm = T))
+
+corr_comb %>%
+  filter(treatment == "mix") %>%
+  summarise(across(c(s, m, l, total), mean, na.rm = T))# %>%
+
+count_sample_comb <- corr_comb %>% # create a data frame with treatment count means
+  filter(treatment == "mix") %>%
+  #group_by(starch) %>%
+  summarise(across(c(s, m, l, total), mean, na.rm = T)) %>%
+  ungroup() %>%
[…]

### 2021-08-20 analysis/index.Rmd
further development of manuscript

@@
   bookdown::pdf_document2:
+    df_print: kable
     toc: no
@@
   bookdown::word_document2: default
-  bookdown::html_document2: default
+  bookdown::html_document2:
+    df_print: kable
+    keep_md: true
 linestretch: 2.0 # double-spacing for review purposes
@@
   message = FALSE,
-  fig.path = "../figures")
+  fig.path = "./figures/")
 library(tidyverse)

### 2021-08-20 analysis/paper/disc.Rmd
further development of manuscript

@@
 [@lingstromStarchyFood1994; @francoStarchDegradation1992; @haslamDecompositionStarch2004; 
-@bjorckStarchProcessing1984; @henryCookingStarch2009]. 
+@bjorckStarchProcessing1984; @henryCookingStarch2009]. Given the lack of $\alpha$-amylase 
+activity detected in our model, hydrolysis is unlikely to be the main cause of our
+reported low incorporation. More details on the absence of $\alpha$-amylase 
+activity is discussed elsewhere<!--Bartholdy et al. in prep-->.
 <!--
@@
 clearance. -->
-This means, that the dietary picture we obtain from starch granules extracted from 
-dental calculus reflects a number of individual, dietary, and environmental (both
-pre- and post-mortem environments) factors, and that this will likely be a
+What we can say, is that the dietary picture we obtain from starch granules extracted 
+from dental calculus reflects a number of individual, dietary, and environmental 
+(both pre- and post-mortem environments) factors, and that this will likely be a
 somewhat random snapshot given the irregularity of plaque mineralisation within

### 2021-08-20 analysis/supplementary-materials/supp-mat.Rmd
further development of manuscript

@@
 title: Supplementary Materials
+output: 
+  html_document:
+    keep_md: true
+    df_print: paged
 ---
+
+```{r setup}
+library(here)
+library(tidyverse)
+raw_data <- read_csv(here("analysis/data/raw_data/starch_counts.csv"))
+```
 
@@
 
+## Control samples
+
+```{r}
+raw_data %>%
+  filter(treatment == "control") %>%
+  select(!c(vol, portion_slide, s, m, l))
+```
+
+Only the total starch count was considered for control samples, as size was
+deemed irrelevant.
+
 ## Count corrections

### 2021-08-20 analysis/paper/paper.Rmd
combine sections into single document

--- 2021-08-20 analysis/paper/disc.Rmd
@@
+---
+title: "Untitled"
+author:
+  - "Bjørn Peare Bartholdy"
+  - Amanda G. Henry
+date: "`r Sys.Date()`"
+output:
+  bookdown::pdf_document2:
+    df_print: kable
+    toc: no
+    latex_engine: xelatex
+    fig_caption: yes
+    number_sections: yes
+  bookdown::word_document2: default
+  bookdown::html_document2:
+    df_print: kable
+    keep_md: true
+linestretch: 2.0 # double-spacing for review purposes
+bibliography: paper/refs.bib
+csl: paper/apa.csl
+abstract: |
+  Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Curabitur eget porta erat.
+Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem.
+Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
+Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
+Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
+Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
+Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(
+  echo = FALSE,
+  warning = FALSE,
+  message = FALSE,
+  fig.path = "./figures/")
+library(tidyverse)
+library(broom)
+library(here)
+library(patchwork)
+# load R scripts
+source(here("analysis/scripts/analysis.R"))
+source(here("analysis/scripts/tables.R"))
+source(here("analysis/scripts/figures.R"))
+## Compile refs.bib using rbbt package ##
+file_paths <- here("analysis/paper", c("intro.Rmd", "matmet.Rmd", "disc.Rmd"))
+cite_keys <- sapply(file_paths, rbbt::bbt_detect_citations) 
+cite_keys <- unlist(cite_keys, use.names = FALSE)
+rbbt::bbt_write_bib("./paper/refs.bib", keys = cite_keys, overwrite = T)
+## end refs.bib ##
+```
+
+# Introduction
+
+<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
+<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
+Dental calculus, mineralised dental plaque, has proven to contain a wealth of 
+dietary information in the form of plant microfossils 
+[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
+proteins [@hendyProteomicCalculus2018; @warinnerDirectEvidenceMilk2014],
+and other organic residues [@buckleyDentalCalculusCooking2014]. 
+It has also proven to be able to retain and preserve dietary information for millenea,
+providing detailed dietary information on past populations 
+[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
+and species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].
+
+<!-- more detailed background, comprehensible to scientists in related disciplines -->
+
+Until recently, few studies directly investigated the presence of plant microremains
+in the dental calculus of archaeological remains. The ability to extract phytoliths
+from the dental calculus of archaeological fauna to investigate diet was first 
+noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
+and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
+and Fox and colleagues [-@foxPhytolithCalculus1996]. The ability to detect coca
+consumption by looking at human dental calculus was explored by Klepinger and Kuhn
+[-@klepingerCalculusCoca1977], and starches and phytoliths were extracted from 
+human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].
+
+In more recent years, the study of dental calculus has increased exponentially, 
+and the wealth of information contained within the mineralised matrix has been, 
+to a larger extent, acknowledged. The applications of dental calculus span a wide
+variety of archaeological research areas, such as oral microbiome 
+characterisation (including pathogens) through the analysis of DNA and proteins 
+[@adlerSequencingAncientCalcified2013; @warinnerPathogensHostImmunity2014],
+microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
+@mickleburghNewInsightsConsumption2012], 
+other organic residues and proteins from dietary compounds 
+[@buckleyDentalCalculusCooking2014; @hendyProteomicCalculus2018], 
+and nicotine-use [@eerkensDentalCalculusSource2018]. Especially the extraction 
+of starch granules has become a rich source of dietary
+information, as starch granules have proven to preserve well within dental calculus
+over a variety of geographical and temporal ranges 
+[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
+@jovanovicNeolithicCalculus2021].
+
+<!-- starches from dental calculus -->
+Despite this, our knowledge of dental calculus and the incorporation pathways of 
+the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
+knowledge of information-loss caused by these pathways and the methods we use to 
+extract and analyse dental calculus and make inferences on past diets. 
+Extraction methods were tested 
+by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
+decalcification using HCl or EDTA. 
+The authors found significantly more starches with the EDTA extraction method 
+than the HCl extraction method; however, as noted by the authors, comparisons 
+involving archaeological calculus are problematic due to variability between and 
+within individuals. Other studies have been conducted on modern populations 
+[@leonardDentalCalculus2015] and primates [@powerChimpCalculus2015], both with
+well-documented dietary intake, to reconcile our interpretations on the recovery 
+of microremains (both phytoliths and starches) from dental calculus and how well
+they represent actual dietary intake. These studies are justifiably limited, 
+despite meticulous documentation and observation, due to unknown variables and 
+uncertainty involved when studying living organisms. Dental calculus is a complex 
+oral biofilm with a
+multifactorial aetiology and variable formation rates both within and between
+individuals [@jepsenCalculusRemoval2011].
+Additionally, the effects of the many different methods
+of starch processing, as well as post-depositional processes, are still being 
+explored [@graneroStarchTaphonomy2020].
+
+<!-- clearly stating the general problem being addressed by this particular study -->
+<!-- Dental calculus research has its limitations, all of which I will solve with this
+single method. -->
+In this exploratory study, we use an oral biofilm model to investigate the 
+retention of starch granules within dental calculus, in a controlled laboratory 
+setting. Our main questions concern the representation of granules extracted
+from the calculus compared to the actual intake. How much of the original diet is 
+incorporated into the calculus, and how much is recovered? 
+Is there differential loss of information from specific dietary markers that affects 
+the obtained dietary information, and how does this affect the representation of diet from
+extracted microremains? We find that a limited proportion of the starch input is actually 
+retained in the calculus, as well as a shift in the size ratios of individual starch granules
+that are incoporated into the calculus. We also show that the number of starch
+granules that are incorporated increases as the size of the calculus deposit 
+increases.
+
+# Materials and Methods
+
+```{r protocol-fig, fig.cap="Overview of experiment protocol"}
+knitr::include_graphics(here("analysis/figures", "ExpL_protocol.png"))
+```
+
+<!-- remove sampling from protocol picture -->
+
+## Biofilm formation
+
+In this study we employ a multispecies oral biofilm model following a modified
+protocol from Sissons and colleagues [-@sissonsMultistationDentalPlaque1991] and
+Shellis [-@shellisSyntheticSalivaCultural1978]. The setup comprises a polypropylene 
+24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which 
+is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins.
+
+The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
+described by Sissons and colleagues [-@sissonsMultistationDentalPlaque1991]. 
+It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
+5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
+5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl~2~, 
+5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationDentalPlaque1991], 6.4 mmol/l NaHCO~3~ 
+[@shellisSyntheticSalivaCultural1978], 2.5 mg/l haemin. This is subsequently 
+adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120&deg;C, 
+1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
+and 1 mmol/l arginine [@sissonsMultistationDentalPlaque1991].
+
+<!-- add to protocols.io -->
+Fresh whole saliva (WS) for inoculation was provided by a 31-year-old male donor 
+with no history of caries, who abstained from oral hygiene for 24 hours, and no 
+food was consumed two hours prior to donation. No antibiotics were taken up to six 
+months prior to donation. 
+The saliva was filtered through a sterilised (with bleach) nylon cloth to remove
+particulates.
+Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
+20% glycerine for four hours at 36&deg;C, to allow attachment of the
+salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
+substrata were transferred to a new plate containing 1 ml/well AS and incubated 
+at 36&deg;C, 30 rpm. The inoculation process was repeated on days 3 and 5. 
+AS was partially refreshed once per day and fully refreshed every three days, 
+throughout the experiment, by transferring the substrata to a new plate containing 
+AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
+5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
+(days 0, 3, and 5), where the samples only received one sucrose treatment after
+inoculation.
+
+Starch treatments were initiated on day 9 to avoid starch granule counts being 
+affected by $\alpha$-amylase hydrolysis. An $\alpha$-amylase (EC 3.2.1.1) activity
+assay was conducted to confirm that no amylase was present in the system (see 
+Bartholdy et al. ???? <!-- ## [bartholdy] ## --> for more details). Starch treatments replaced sucrose treatments, occurring twice per day
+for six minutes. The starch treatments involved transferring the substrata to a 
+new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
+concentrations (w/v) wheat and potato. All starch solutions were created in a 5%
+(w/v) sucrose solution. Before transferring biofilm samples to the starch treatments, 
+the starch plates were agitated to keep the starches in suspension in the solutions, 
+and during treatments, the rpm was increased to 60.
+
+After 15 days, mineralisation was encouraged with a 
+calcium phosphate monofluorophosphate urea (CPMU) solution containing 
+20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea
+[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991], 
+and (0.04 g/l MgCl). The substrata were submerged in 1 ml/well CPMU five times daily, 
+every two hours, for six minutes. During the mineralisation period, starch treatments
+were reduced to once per day after the five CPMU treatments. This cycle was repeated 
+for 10 days until the end of the experiment on day 24 
+(see \@ref(fig:protocol-fig) for an overview of protocol). More detailed protocols
+are available on protocols.io<!--insert DOI-->.
+
+All laboratory work was conducted in sterile conditions under a laminar flow hood 
+to prevent starch and bacterial contamination. Control samples were included to 
+detect starch contamination.
+
+## Amylase activity detection
+
+<!-- amylase protocol description -->
+
+## Starch counts
+
+Starch granule counts were divided into three size categories: small (<10 $\mu$m),
+medium (10 -- 20 $\mu$m), and large (>20 $\mu$m). Counts for small wheat and
+potato starches were combined for the mixed treatment samples, as it is not 
+possible to distinguish the small starch granules between the two species.
+
+### Treatment solutions
+
+A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
+on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
+(Zeiss Axioscope A1). Counting starches on a full slide was not feasible, so three slide transects were 
+counted (at 1/3, 1/2, and 2/3 of the slide), and the sample counts were extrapolated 
+(see Supplementary material for more details).
+
+### Extraction from samples
+
+Extraction of starches from the calculus samples was performed by dissolving the
+calculus in ethylenediaminetetraacetic acid (EDTA), as it has been shown to
+have little to no effect on starch granules (especially native granules)
+[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
+and vortexing for 3 days until the sample was completely dissolved. 
+Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
+When dispersing the liquid (EDTA) sample on the slide, the sample was homogenised using the pipette. 
+It was not possible to count all wheat starches
+on a slide, so wheat and mix counts were extrapolated as described for the treatment
+solutions.
+
+## Statistical analysis
+
+Statistical analysis was conducted in R versions 4.0.5 ("Shake and Throw") 
+and 4.0.3 ("Bunny-Wunnies Freak Out") [@R-base].
+<!-- R packages used -->
+A one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
+as the grouping variable (GV) was conducted to explore the effect of the different 
+starch treatments on biofilm growth.
+
+To calculate size proportions, mean counts for each treatment were taken across 
+both experimental plates for each treatment, resulting in a mean count for each
+granule size category within each treatment.
+
+Pearson's r was conducted on sample weight and total starch count, as well as sample
+weight and starch count per mg calculus. The total count for each sample within a
+treatment was standardised by z-score to account for the differences in magnitude
+between the potato and wheat counts.
+
+Scripts and data will be made available on osf.io upon publication.
+Until then, they can be found on the first author's GitHub repository
+(https://github.com/bbartholdy/byocStarch).
+
+<!-- script and data availability: provide link (or doi?) to osf.io repository -->
+
+# Results
+
+All samples yielded sufficient biofilm growth and starch incorporation to be 
+included in the analysis, resulting in a total of 48 biofilm samples (two plates of 24),
+45 of which were used for analysis (three samples were set aside for separate analysis).
+Most control samples contained no starch granules, while some contained negligible
+quantities (supp. mat).
+
+## No amylase activity detected in the model
+
+<!-- insert amylase results -->
+
+## Treatment type had no effect on biofilm growth
+
+```{r anova-tbl}
+knitr::kable(anova_tbl, 
+             col.names = c("Treatment", "Mean", "SD", "Min", "Max"),
+             caption = "Summary statistics for biofilm dry-weights (in mg) by treatment.")
+```
+
+A one-way ANOVA suggests that the type of starch used during the biofilm 
+growth period had a minimal effect on the growth of the biofilm 
+(expressed as total dry weight of the sample), F(`r paste(treat_df[1], treat_df[2], sep = ", ")`)
+= `r treat_f.stat`, p = `r treat_p`. 
+A summary of sample weights is available in Table \@ref(tab:anova-tbl).
+
+## Starch counts
+
+<!-- starch counts in the solutions -->
+
+```{r solution-count-out}
+count_sol_out <- count_sol %>%
+  mutate(across(c(s, m, l, total), 
+                function(x) 
+                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
+
+### adding the 'both' row to the data frame ###
+count_sol_comb_out <- count_sol_comb %>%
+  mutate(across(c(s, m, l, total), 
+                function(x) 
+                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
+
+count_sol_out <- count_sol %>%
+  mutate(across(c(s, m, l, total), 
+                function(x) 
+                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")"))) %>%
+  ungroup() %>%
+  add_row(tibble_row(count_sol_comb_out), .before = 3)
+```
+
+```{r solution-count-tbl}
+cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
+knitr::kable(count_sol_out, col.names = cnames, 
+             caption = "Mean starch counts from solutions, including the
+             proportional makeup of the different sizes of granules.")
+```
+
+<!-- starch counts in the samples -->
+
+```{r sample-count}
+count_sample <- count_sample %>%
+  ungroup() %>%
+  add_row(count_sample_comb, .after = 2)
+
+# convert columns to contain mean count and percentage of total count
+count_samp_per <- count_sample %>%
+  mutate(across(c(s, m, l, total), 
+                function(x) 
+                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), ")")))
+
+# count_samp_comb_per <- count_sample_comb %>%
+#   mutate(across(c(s, m, l, total), 
+#                 function(x) 
+#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
+
+# count_samp_comb_out <- tibble(count_samp_comb_per, count_sample_comb_sd, .name_repair = "unique") %>%
+#   select(1,2,3,7,4,8,5,9,6,10) #%>%
+  #add_row(tibble_row(count_sample_comb_out), .before = 3)
+
+count_samp_out <- tibble(count_samp_per, count_sample_sd, .name_repair = "unique") %>%
+  select(1,2,3,7,4,8,5,9,6,10) #%>%
+  #add_row(tibble_row(count_sample_comb_out), .before = 3)
+```
+
+```{r sample-count-tbl}
+cnames <- c("Treatment", "Starch", "Small (%)", "SD", "Medium (%)", "SD", 
+            "Large (%)", "SD", "Total (%)", "SD")
+knitr::kable(count_samp_out, col.names = cnames, 
+             caption = "Mean starch counts extracted from samples with standard 
+             deviation (SD), including the proportion of granule sizes of the total count.")
+```
+
+It was not possible to differentiate between potato and 
+wheat starches smaller than ca. 10 $\mu$m,
+small potato starches were counted as wheat starches in the mixed-treatment samples. This is reasonable given that the small potato granules make up an insignificant proportion of the 
+total count of small granules within mixed-treatment solutions which are predominantly
+wheat granules
+(`r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`).
+<!-- count of small granules from wheat solution divided by the sum of small granules 
+from the potato and wheat solutions -->
+
+<!-- absolute solution counts compared to absolute sample counts. does exposure
+to higher quantities of starches mean more incorporated starches? -->
+The separate wheat and potato solutions were made with a 0.25% (w/v) starch
+concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
+starch, with a total concentration of 0.50% (w/v).
+The mixed-treatment had the highest absolute count of starch granules in solution,
+while the biofilms exposed to the wheat solution preserved the greatest number of
+granules. The potato treatment had the lowest absolute counts in both the solution 
+and in the biofilm samples <!-- insert counts? -->
+(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
+
+<!-- look at proportions of incorporated starches in each sample before looking at overall means -->
+
+### Proportion of available starches incorporated in samples
+
+```{r sample-perc}
+# convert proportions to percent labels
+perc_comb <- prop_comb %>% 
+  mutate(across(c(s, m, l, total), scales::percent, 0.001))
+
+perc <- prop %>%
+  mutate(across(c(s, m, l, total), scales::percent, 0.001)) %>% 
+  add_row(tibble_row(perc_comb), .before = 3)
+```
+
+```{r sample-prop-tbl}
+cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
+knitr::kable(perc, col.names = cnames, 
+             caption = "The mean percentage of starches from the solutions that 
+              were incorported into the samples.")
+```
+
+The proportion of total starches from the solutions that were incorporated 
+into the samples ranged from 
+`r paste(min(perc$total), max(perc$total), sep = " to ")`,
+with potato granules being more readily incorporated than wheat in both the
+separated- and mixed-treatment samples
+(Table \@ref(tab:sample-prop-tbl)). 
+There is an inverse relationship between the absolute starch count in the solutions 
+and the proportional incorporation of starches in the biofilm samples, i.e. potato 
+had the lowest absolute count in solutions, but the highest proportional 
+incorporation, and vice versa for the mixed treatment.
+
+<!-- separated treatments -->
+
+<!-- mixed treatments -->
+Wheat incorporation was most affected in the mixed-treatment samples, with only
+`r paste0(perc$total[2])` of the total available starches being incorporated into
+the sample, compared to
+`r paste0(perc$total[4])`
+in the separated wheat treatment.
+
+### Size ratios differ between solutions and samples
+
+<!-- size proportions -->
+<!-- pie charts -->
+
+```{r sep-plot, fig.cap="Proportion (%) of sizes of (A) starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring).
+l = large, m = medium, s = small."}
+#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
+pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+```
+
+<!-- change colour of medium starches in D to match C -->
+```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring).
+l = large, m = medium, s = small."}
+#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
+pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+```
+<!-- Combine to a single figure? -->
+
+<!-- overall size trends -->
+
+Overall, medium starch granules had a higher mean rate of incorporation 
+(`r scales::percent(mean(prop$m, na.rm = T), 0.001)`) 
+than small 
+(`r scales::percent(mean(prop$s, na.rm = T), 0.001)`) 
+and large 
+(`r scales::percent(mean(prop$l, na.rm = T), 0.001)`) 
+starch granules across all treatments, while large potato starches had the lowest 
+rate of incorporation across all treatments.
+
+<!-- separated and mixed treatments -->
+The difference in incorporation between the size categories resulted in a change
+in size ratios between the original starch solutions and the extracted samples.
+Large potato granules (> 20 $\mu$m) were most affected, with a 
+`r paste0(abs(diff_sep$potato[3]), "%")` 
+decrease in relative abundance in the potato-only treatment, and a
+`r paste0(abs(diff_mix$potato[3]), "%")` 
+decrease in mixed treatments. Medium granules increased in relative abundance 
+across all samples, while small granules decreased in wheat treatments and 
+increased in potato treatments 
+(Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).
+
+<!-- Chi square test for size counts? -->
+
+<!-- Proportion and size ratios are the most important points to be made -->
+
+### Biofilm weight correlated positively with extracted starch counts
+
+```{r cor, echo=FALSE}
+# prepare cor output for reporting
+cor_r <- signif(count_cor$estimate, 3)
+cor_ci <- signif(count_cor$conf.int, 3)
+cor_p <- if(count_cor$p.value > 0.001){
+  signif(count_cor$p.value, 3)
+} else {
+  paste("< 0.001")
+}
+
+cor2_r <- signif(starch_cor$estimate, 3)
+cor2_ci <- signif(starch_cor$conf.int, 3)
+cor2_p <- if(starch_cor$p.value > 0.001){
+  signif(starch_cor$p.value, 3)
+} else {
+  paste("< 0.001")
+}
+# Convert r correlation result to text
+direct_cor <- ifelse(cor_r < 0, "negative", "positive")
+strength_cor <- if(cor_r >= 0.8){
+  "very strong"
+} else if(cor_r < 0.8 & cor_r >= 0.6){
+  "strong"
+} else if(cor_r < 0.6 & cor_r >= 0.4){
+  "moderate"
+} else if(cor_r < 0.4 & cor_r >= 0.2){
+  "weak"
+} else{
+  "very weak"
+}
+```
+
+```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments."}
+pl_cor
+```
+<!-- with Z-score standardisation -->
+
+Pearson's *r* suggests a 
+`r strength_cor` `r direct_cor` 
+correlation between the total weight of the biofilms and the total starch count 
+(standardised by z-score) extracted from the samples across treatments, 
+*r* = `r cor_r`, 
+90%CI[`r paste(cor_ci[1], cor_ci[2], sep = ", ")`],
+p = `r cor_p`
+(Figure \@ref(fig:cor-plot)).
+
+The same test was applied to total biofilm weight and starch count per mg 
+calculus (also standardised by z-score), resulting in a weak positive correlation, 
+*r* = `r cor2_r`, 
+90%CI[`r paste(cor2_ci[1], cor2_ci[2], sep = ", ")`],
+p `r cor2_p`
+(Figure \@ref(fig:cor-plot2)).
+
+<!-- size ratios in the larger deposits vs the smaller deposits?
+  Are larger starches more proportionately represented in larger deposits?
+  -->
+
+```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
+pl_cor2
+```
+
 # Discussion
@@
 of dietary research.
+
+# Acknowledgements {-}
+
+<!-- If you're reading this, thank you for taking the time to read the source 
+documents. You're awesome! -->
+
+# References cited

### 2021-08-20 analysis/paper/results.Rmd
further development of manuscript
combine sections into single document

@@
 # Results
-
-```{r setup-res, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message = FALSE)
-#opts_knit$set(root.dir = '/Users/username/path_to_project')
-#corr_counts <- readr::read_csv(here("analysis/data/derived_data", "corr_counts.csv"))
-#corr_comb <- readr::read_csv(here("analysis/data/derived_data", "corr_comb.csv"))
-#sol_corr <- readr::read_csv(here("analysis/data/derived_data", "sol_corr.csv"))
-#source(here("analysis/scripts/analysis.R"))
-#source(here("analysis/scripts/tables.R"))
-#source(here("analysis/scripts/figures.R"))
-#source(c("tables.R", "figures.R", "analysis.R"))
-#code to include R scripts in .Rmd: r knitr::spin_child("tables.R")
-```
 
@@
 
-<!-- exclude problematic samples: st1C5 wheat and st2C6 potato, and  st2C5 
-Use more formal outlier detection -->
+## No amylase activity detected in the model
+
+<!-- insert amylase results -->
 
@@
 
-```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring).
+```{r sep-plot, fig.cap="Proportion (%) of sizes of (A) starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring).
 l = large, m = medium, s = small."}

### 2021-08-27 analysis/paper/paper.Rmd
combine sections to paper.Rmd

@@
   bookdown::pdf_document2:
+    #template: "../templates/sn-article-template/sn-article.tex"
     df_print: kable
@@
     number_sections: yes
-  bookdown::word_document2: default
   bookdown::html_document2:
@@
     keep_md: true
+  bookdown::word_document2:
+    reference_docx: "../templates/template.docx"
 linestretch: 2.0 # double-spacing for review purposes
-bibliography: paper/refs.bib
-csl: paper/apa.csl
+bibliography: refs.bib
+csl: apa.csl
 abstract: |
@@
 Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+keywords: |
+  keyword 1; keyword 2
 ---
@@
   message = FALSE,
-  fig.path = "./figures/")
+  fig.path = "../figures/")
 library(tidyverse)
@@
 source(here("analysis/scripts/figures.R"))
+source(here("analysis/scripts/amylase-assay.R"))
 ## Compile refs.bib using rbbt package ##
-file_paths <- here("analysis/paper", c("intro.Rmd", "matmet.Rmd", "disc.Rmd"))
-cite_keys <- sapply(file_paths, rbbt::bbt_detect_citations) 
-cite_keys <- unlist(cite_keys, use.names = FALSE)
-rbbt::bbt_write_bib("./paper/refs.bib", keys = cite_keys, overwrite = T)
+cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
+#cite_keys <- unlist(cite_keys, use.names = FALSE)
+rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T)
 ## end refs.bib ##
@@
 <!-- more detailed background, comprehensible to scientists in related disciplines -->
-
 Until recently, few studies directly investigated the presence of plant microremains
@@
 
-<!-- starches from dental calculus -->
+<!-- clearly stating the general problem being addressed by this particular study -->
+<!-- Dental calculus research has its limitations, all of which I will solve with this
+single method. -->
 Despite this, our knowledge of dental calculus and the incorporation pathways of 
@@
 uncertainty involved when studying living organisms. Dental calculus is a complex 
-oral biofilm with a
-multifactorial aetiology and variable formation rates both within and between
-individuals [@jepsenCalculusRemoval2011].
-Additionally, the effects of the many different methods
-of starch processing, as well as post-depositional processes, are still being 
-explored [@graneroStarchTaphonomy2020].
+oral biofilm with a multifactorial aetiology and variable formation rates both 
+within and between individuals [@jepsenCalculusRemoval2011], contributing to
+the stochasticity of starch representation being observed in numerous studies.
+Additionally, the rate of $\alpha$-amylase differs both between and within
+individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
+causing different rates of hydrolysis of the starch granules present in the oral
+cavity. Add to this the effects of the many different methods
+of starch processing, as well as post-depositional processes that are still being 
+explored [@graneroStarchTaphonomy2020], and you have a highly unpredictable process.
 
-<!-- clearly stating the general problem being addressed by this particular study -->
-<!-- Dental calculus research has its limitations, all of which I will solve with this
-single method. -->
 In this exploratory study, we use an oral biofilm model to investigate the 
-retention of starch granules within dental calculus, in a controlled laboratory 
-setting. Our main questions concern the representation of granules extracted
+retention of starch granules within dental calculus in a controlled laboratory 
+setting, allowing us full control over dietary input. We use a multiwell setup for consistent, high-throughout analysis.
+Our main questions concern the representation of granules extracted
 from the calculus compared to the actual intake. How much of the original diet is 
@@
 Is there differential loss of information from specific dietary markers that affects 
-the obtained dietary information, and how does this affect the representation of diet from
-extracted microremains? We find that a limited proportion of the starch input is actually 
-retained in the calculus, as well as a shift in the size ratios of individual starch granules
+the obtained dietary information, and how does this affect the representation of 
+diet from extracted microremains?  
+We find that, despite an absence of $\alpha$-amylase in
+the model, a limited proportion of the starch input is actually 
+retained in the calculus, as well as a shift in the size ratios of individual 
+starch granules
 that are incoporated into the calculus. We also show that the number of starch
@@
 ```
-
-<!-- remove sampling from protocol picture -->
+<!-- modify sampling days from protocol picture to match amylase sampling -->
 
@@
 affected by $\alpha$-amylase hydrolysis. An $\alpha$-amylase (EC 3.2.1.1) activity
-assay was conducted to confirm that no amylase was present in the system (see 
-Bartholdy et al. ???? <!-- ## [bartholdy] ## --> for more details). Starch treatments replaced sucrose treatments, occurring twice per day
+assay was conducted to confirm that no amylase was present in the system following
+starch treatments. Starch treatments replaced sucrose treatments, occurring twice per day
 for six minutes. The starch treatments involved transferring the substrata to a 
@@
 calcium phosphate monofluorophosphate urea (CPMU) solution containing 
-20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea
-[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991], 
-and (0.04 g/l MgCl). The substrata were submerged in 1 ml/well CPMU five times daily, 
+20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea, 
+and (0.04 g/l MgCl)
+[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991].
+The substrata were submerged in 1 ml/well CPMU five times daily, 
 every two hours, for six minutes. During the mineralisation period, starch treatments
@@
 for 10 days until the end of the experiment on day 24 
-(see \@ref(fig:protocol-fig) for an overview of protocol). More detailed protocols
+(see \@ref(fig:protocol-fig) for an overview of the protocol). More detailed protocols
 are available on protocols.io<!--insert DOI-->.
@@
 All laboratory work was conducted in sterile conditions under a laminar flow hood 
-to prevent starch and bacterial contamination. Control samples were included to 
-detect starch contamination.
+to prevent starch and bacterial contamination. Control samples, that only received
+sucrose as a treatment, were included to detect starch contamination.
 
@@
 
-<!-- amylase protocol description -->
+An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
+saliva samples collected from the plate wells on days 3, 5, 6, 9, and 12. 
+Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
+Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
+All samples and standard curves were added in triplicates to two separate plates.
+Positive control saliva samples were compared against a standard curve containing 
+H~2~O, while artificial saliva samples were compared against a standard curve
+containing sterile artificial saliva (due to the colour of artificial saliva).
+Two photometric readings were conducted for each plate with a 540 nm filter on a... 
+<!-- photometric reader specs -->
+The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
+(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
+maltose released from starch by $\alpha$-amylase activity.
+For a more detailed protocol, see <!-- protocols.io DOI -->
 
@@
 (Zeiss Axioscope A1). Counting starches on a full slide was not feasible, so three slide transects were 
-counted (at 1/3, 1/2, and 2/3 of the slide), and the sample counts were extrapolated 
+counted (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts were extrapolated 
 (see Supplementary material for more details).
@@
 Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
-When dispersing the liquid (EDTA) sample on the slide, the sample was homogenised using the pipette. 
-It was not possible to count all wheat starches
+When dispersing the liquid (EDTA) sample on the slide, the sample was homogenised using 
+the pipette. It was not possible to count all wheat starches
 on a slide, so wheat and mix counts were extrapolated as described for the treatment
@@
 
-Statistical analysis was conducted in R versions 4.0.5 ("Shake and Throw") 
-and 4.0.3 ("Bunny-Wunnies Freak Out") [@R-base].
+Statistical analysis was conducted in `r version$version.string`  [@R-base].
 <!-- R packages used -->
@@
 
-Scripts and data will be made available on osf.io upon publication.
-Until then, they can be found on the first author's GitHub repository
-(https://github.com/bbartholdy/byocStarch).
-
-<!-- script and data availability: provide link (or doi?) to osf.io repository -->
+All scripts and data are available on OSF (https://osf.io/uc5qy/)
+and Github (https://github.com/bbartholdy/byoc-starch).
 
@@
 
-<!-- insert amylase results -->
+```{r amylase-activity}
+# table of results reported in units amylase per mL enzyme (but let's be honest,
+  # ...it doesn't really matter what the unit is. No activity is no activity)
+cols <- c("1", "2", "3")  
+rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
+plt1_ph1_result <- rbind(sal1_ph1, bmm1_ph1)
+  plt1_ph1_result[plt1_ph1_result < 0] <- 0
+```
+
+```{r amylase-activity-tbl}
+rownames(plt1_ph1_result) <- rows
+colnames(plt1_ph1_result) <- cols
+knitr::kable(signif(plt1_ph1_result, 3), row.names = TRUE, caption = "Amylase activity in U/mL enzyme, where a U is mg maltose released from starch in six minutes at 36 degrees Celsius.
+Plate 1, photometric reading 1.
+Negative values converted to 0.")
+```
+
+No appreciable amount of $\alpha$-amylase activity was detected in any of the artificial
+saliva samples from any of the days that were sampled. Only positive controls
+contained amylase activity that could be detected in the assay (Table
+\@ref(tab:amylase-activity-tbl)). The results from the additional plates and
+photometric readings can be found in the 
+[Supplementary Materials](../supplementary-materials/supp-mat.Rmd). The results
+are not comparable to other studies presenting $\alpha$-amylase activity levels
+in humans; however, they are sufficient to show that there is no activity in the
+system.
 
@@
 <!-- starch counts in the solutions -->
+
+<!-- INSERT PICTURES OF STARCHES -->
 
@@
 a portion of its hydrolytic activity
-[@nikitkovaStarchBiofilms2013; @tanStudyBacterialViability2004; @tanCalculusUltrastructure2004].
+[@scannapiecoSalivaryAmylase1993; @nikitkovaStarchBiofilms2013; @tanStudyBacterialViability2004; @tanCalculusUltrastructure2004].
 <!-- These channels and pockets of organic matter within the inorganic matrix are used by the 
@@
 @bjorckStarchProcessing1984; @henryCookingStarch2009]. Given the lack of $\alpha$-amylase 
-activity detected in our model, hydrolysis is unlikely to be the main cause of our
-reported low incorporation. More details on the absence of $\alpha$-amylase 
+activity detected in our model, hydrolysis is not the cause of the low
+incorporation rate. More details on the absence of $\alpha$-amylase 
 activity is discussed elsewhere<!--Bartholdy et al. in prep-->.
@@
 In future experiments, we recommend the addition of $\alpha$-amylase to the model
-in concentrations similar to those found in the oral cavity <!-- provide concentration and citation.--> 
-<!-- foreshadowing... -->
+in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. 
+<!-- foreshadowing...? -->
 While we are able to show the differing 
@@
 The oral biofilm model described in this study, and in 
-Bartholdy and colleagues (in prep.),<!-- insert citation --> provides a method to directly
+Bartholdy and colleagues (in prep.),<!-- insert citation --> provides a method
+to directly
 explore the incorporation and extraction of dietary compounds from dental calculus 
-in a controlled laboratory setting. The addition of known species in separate
-samples also removes the identification bias against the smaller undiagnostic starches.<!-- rephrase -->
+in a controlled laboratory setting. The addition of known starch species in
+separate samples also removes the identification bias against the smaller
+undiagnostic starches.<!-- rephrase -->
 It can also address the call for more baseline testing of biases associated with 
@@
 *In vivo* studies on humans and non-human primates will be limited by the variability 
-occurring
-between individuals, and difficulties documenting the entirety of starch granules
-entering the oral cavity at any given moment, especially accidental intake. 
+occurring between individuals, and difficulties in accurately documenting 
+the entirety of starch granules entering the oral cavity at any given moment,
+especially accidental intake. 
 Our experimental setup allows us a
@@
 
-Our preliminary study shows that a very small proportion of the ingested starch
-granules are retained in a dental calculus model.
-
-Calculus has a low capacity for retention of starch granules, and this effect is
-compounded by diagenetic effects in archaeological remains, resulting in low overall
-counts of extracted granules. Additionally, the size of granules will influence the 
-likelihood of incorporation, as large starches have a decreased incorporation, medium
-starches increased, and small starches the only representative of intake.
-
-Given the direct proportional incorporation of starches with quantity of exposure<!-- rephrase -->, 
-we can say that the proportion of starches consumed will in many cases be reflected 
+This preliminary study shows that a very small proportion of the input starch
+granules are retained in a dental calculus model. This and previous studies 
+have shown that calculus has a low capacity for retention of starch granules, 
+an effect that is compounded by diagenetic effects in archaeological remains, 
+resulting in low overall counts of extracted granules. 
+The proportion of starches consumed will in many cases be reflected 
 in the quantity
-of starches extracted from the dental calculus, at least in modern calculus samples
-unaffected by diagenesis.
+of starches extracted from the dental calculus---i.e. the more starch granules
+entering the oral cavity, the more will be recovered from extraction---at 
+least in modern calculus samples unaffected by diagenesis and hydrolysis.
 Whether or not this also applies to archaeological samples remains to be tested.
-
-The size of calculus deposit also influences the maximum capacity of granule incorporation; as the size of the deposit increases, so does the absolute count of
-incorporated granules.
-
-The data from this study are not sufficient to confirm or deny any proposed methods
-of starch incorporation; however, while we are unable to address the mechanism(s) of
-starch incorporation with the current experimental setup, this is something that can
-potentially be addressed with the calculus model presented in this paper.
+Additionally, we have
+shown that the size of granules will influence the likelihood of incorporation, 
+as large starches have a decreased incorporation rate, medium
+starches an increased rate, and small starches remained somewhat constant.
+The size of calculus deposit also seems to influence the capacity of 
+granule incorporation; as the size of the deposit increases, so does the
+absolute count of incorporated granules.
+<!--The data from this study are not sufficient to confirm or deny any proposed
+methods of starch incorporation; however, while we are unable to address the 
+mechanism(s) of starch incorporation with the data obtained in this study, 
+this is something that can potentially be addressed with the calculus model 
+presented in this paper.-->
 
@@
 to the biofilm mineralisation events (as discussed above), which are irregular and
-related to various factors<!-- citation needed -->. 
-This means that the dietary components extracted from dental calculus provide a 
-snapshot of the diet around the time of the mineralisation event, which may not 
-be an accurate picture of the diet as a whole.
-
+related to various factors. 
 The dental calculus biofilm model presented in this study is a powerful tool to 
-investigate various aspects
-of dietary research.
+investigate various aspects of dietary research and starch incorporation and
+retention in dental calculus.
 

### 2021-08-27 analysis/supplementary-materials/supp-mat.Rmd
combine sections to paper.Rmd

@@
 
+## Amylase activity
+
+Tables containing the amylase activity results for both plates and photometric
+readings.
+
+```{r amylase-activity-results}
+# table of results reported in units amylase per mL enzyme (but let's be honest,
+  # ...it doesn't really matter what the unit is. No activity is no activity)
+cols <- c("1", "2", "3")  
+rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
+(plt1_ph1_result <- rbind(sal1_ph1, bmm1_ph1))
+#  plt1_ph1_result[plt1_ph1_result < 0] <- 0
+(plt1_ph2_result <- rbind(sal1_ph2, bmm1_ph2))
+#  plt1_ph2_result[plt1_ph2_result < 0] <- 0
+(plt2_ph1_result <- rbind(sal2_ph1, bmm2_ph1))
+  #plt2_ph1_result[plt2_ph1_result < 0] <- 0
+(plt2_ph2_result <- rbind(sal2_ph2, bmm2_ph2))
+  #plt2_ph2_result[plt2_ph2_result < 0] <- 0
+```
+
 ## Control samples

### 2021-08-27 analysis/templates/template.Rmd
combine sections to paper.Rmd

---
title: "My title"
author:
- author 1
- other author
date: '`r Sys.Date()`'
output: word_document
abstract: "My abstract \n"
keywords: "keyword 1; keyword 2 \n"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Heading 1

## Heading 2

### Heading 3

#### Heading 4

##### Heading 5 - this is for the page break with MS Word

In the MS Word document, the following modifications must be made in the 'Modify Style' menu:

+ Set the font color to ‘white’ (rather than ‘Automatic’).
+ Select the smallest font size (8 rather than 11).
+ Select ‘Page break before’ in the ‘Line and Page Breaks’ tab.
+ Set the line spacing to ‘Exactly’ and ‘1 pt’ in the ‘Indents and Spacing’ tab.

After these tweaks, the ‘Heading 5’ style will no longer format a heading of level 5. Instead it will insert a very small and white (and, thus, invisible) line followed by a page break.

Now knit this doc, and edit the styles in the resulting docx file. Then rename as 'template' and use this in the yml:

```
output:
  word_document:
    reference_docx: templates/template.docx
```

Don't forget line numbers for JAS! Headings are bold, underline, nothing. And Arial Narrow 10pt for tables.

See here for more information: http://rmarkdown.rstudio.com/articles_docx.html

### 2021-08-27 R/amylase.R

@@ -1,0 +1,99 @@
+# Functions ---------------------------------------------------------------
+
+#' Standard curve calculation
+#' @param std_data Data frame with the photometric readings of the standard curve.
+#' @param cols Vector containing the column numbers containing the standard curve photometric readings. If NULL, all columns assumed to be standard curve measurements.
+#' @param ml_maltose vector of length 7 containing the volume (ml) of 0.2% maltose solution added to the assay.
+#' @importFrom magrittr %>%
+#' @importFrom dplyr slice n
+stdcurve <- function(std_data, cols = NULL, ml_maltose = c(3.75, 15, 30, 45, 60, 75, 150)){
+  if(!is.null(cols)){
+    std_data <- std_data[,cols]
+  }
+  # delta 540 calculation
+  mg_maltose <- 0.002 * ml_maltose
+  replicates <- dim(std_data)[2]
+  pht_blank <- std_data[8,] %>%
+    slice(rep(1:n(), each = 7))
+  pht_meas <- std_data[1:7,] - pht_blank
+
+  pht_meas <- c(pht_meas[, seq(1:replicates)]) %>%
+    unlist()
+  # Linear regression
+  stdCurve_data <- data.frame("d540" = pht_meas, "mgMaltose" = rep(mg_maltose, replicates))
+  std_curve <- lm(d540 ~ mgMaltose, stdCurve_data)
+  # Plot of standard curve
+  p_stdcurve <- stdCurve_data %>%
+    ggplot(aes(x = mgMaltose, y = d540)) +
+    geom_point() +
+    geom_smooth(formula = "y ~ x", method = "lm", se = F) +
+    theme_bw()
+  out <- list("std_curve" = std_curve, "plot" = p_stdcurve, "blank" = pht_blank[1,], "replicates" = replicates, "model" = stdCurve_data)
+  return(out)
+}
+
+#' Amylase activity in sample calculation
+#' @param samp_data data frame with the photometric readings of the samples.
+#' @param cols vector containing the column numbers of the sample photometric readings. If NULL, all columns assumed to be standard curve measurements.
+#' @param std_curve an object from the `stdcurve` function output.
[…]

### 2021-08-27 analysis/scripts/amylase-assay.R

@@ -1,0 +1,130 @@
+#library(byocstarch)
+devtools::load_all()
+library(tidyverse)
+
+# Upload data -------------------------------------------------------------
+
+plt1_ph1 <- readr::read_table(
+  here::here("analysis/Amylase/ExpM_plate1_photometric1.txt"),
+  col_names = F)
+plt1_ph2 <- readr::read_table(
+  here::here("analysis/Amylase/ExpM_plate1_photometric2.txt"),
+  col_names = F)
+
+plt2_ph1 <- readr::read_table(
+  here::here("analysis/Amylase/ExpM_plate2_photometric1.txt"),
+  col_names = F)
+plt2_ph2 <- readr::read_table(
+  here::here("analysis/Amylase/ExpM_plate2_photometric2.txt"),
+  col_names = F)
+
+
+# Data prep ---------------------------------------------------------------
+
+
+
+# Standard curve ----------------------------------------------------------
+
+# convert mL 0.2% Maltose to mg maltose
+#mg_malt <- c(3.75, 15, 30, 45, 60, 75, 150) * 0.002
+# two standard curves prepared: one in h2o and one in artificial saliva (BMM)
+stdH2O1.ph1 <- stdcurve(plt1_ph1, cols = 7:9)
+#stdH2O1.ph1 <- stdcurve(plt1_ph1, cols = 7:9, mg_maltose = mg_malt)
+stdBMM1.ph1 <- stdcurve(plt1_ph1, cols = 10:12)
+#stdBMM1.ph1 <- stdcurve(plt1_ph1, cols = 10:12, mg_maltose = mg_malt)
+
+stdH2O1.ph2 <- stdcurve(plt1_ph2, cols = 7:9)
+stdBMM1.ph2 <- stdcurve(plt1_ph2, cols = 10:12)
+
+stdH2O2.ph1 <- stdcurve(plt2_ph1, cols = 7:9)
+stdBMM2.ph1 <- stdcurve(plt2_ph1, cols = 10:12)
+
+stdH2O2.ph2 <- stdcurve(plt2_ph2, cols = 7:9)
+stdBMM2.ph2 <- stdcurve(plt2_ph2, cols = 10:12)
+
+
+# Samples -----------------------------------------------------------------
+
+# Data prep
+
+plates <- list("plate1_phot1" = plt1_ph1, "plate1_phot2" = plt1_ph2,
+               "plate2_phot1" = plt2_ph1, "plate2_phot2" = plt2_ph2)
+phot_results <- lapply(plates, sep_samples)
[…]

### 2021-08-27 analysis/paper/paper.Rmd
first draft...?

@@
 <!-- add to protocols.io -->
-Fresh whole saliva (WS) for inoculation was provided by a 31-year-old male donor 
+Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
 with no history of caries, who abstained from oral hygiene for 24 hours, and no 
@@
 Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
-20% glycerine for four hours at 36&deg;C, to allow attachment of the
+20% (v/v) glycerine for four hours at 36&deg;C, to allow attachment of the
 salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
@@
 throughout the experiment, by transferring the substrata to a new plate containing 
-AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
+stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
 5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
@@
 affected by $\alpha$-amylase hydrolysis. An $\alpha$-amylase (EC 3.2.1.1) activity
-assay was conducted to confirm that no amylase was present in the system following
-starch treatments. Starch treatments replaced sucrose treatments, occurring twice per day
+assay was conducted to confirm that no amylase was present in the system before
+starch treatments started. Starch treatments replaced sucrose treatments, occurring twice per day
 for six minutes. The starch treatments involved transferring the substrata to a 
@@
 concentrations (w/v) wheat and potato. All starch solutions were created in a 5%
-(w/v) sucrose solution. Before transferring biofilm samples to the starch treatments, 
-the starch plates were agitated to keep the starches in suspension in the solutions, 
-and during treatments, the rpm was increased to 60.
+(w/v) sucrose solution. Before transferring biofilm samples to the starch treatment plate, 
+the plates were agitated to keep the starches in suspension in the solutions. 
+During treatments, the rpm was increased to 60 to facilitate contact between
+starch granules and biofilms.
 
@@
 [@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991].
-The substrata were submerged in 1 ml/well CPMU five times daily, 
-every two hours, for six minutes. During the mineralisation period, starch treatments
-were reduced to once per day after the five CPMU treatments. This cycle was repeated 
+The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
+daily, in a two-hour cycle. During the mineralisation period, starch treatments
+were reduced to once per day after the five CPMU treatments. This process was repeated 
 for 10 days until the end of the experiment on day 24 
@@
 All laboratory work was conducted in sterile conditions under a laminar flow hood 
-to prevent starch and bacterial contamination. Control samples, that only received
-sucrose as a treatment, were included to detect starch contamination.
+to prevent starch and bacterial contamination. Control samples that only received
+sucrose as a treatment, were included to detect starch contamination from the
+environment, or cross-contamination from other wells in the same plate.
 
@@
 Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
-All samples and standard curves were added in triplicates to two separate plates.
+All samples and standard curves were run in triplicates on two separate plates.
 Positive control saliva samples were compared against a standard curve containing 
@@
 (https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
-maltose released from starch by $\alpha$-amylase activity.
+maltose released from starch by $\alpha$-amylase activity. Results are reported
+in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.
 For a more detailed protocol, see <!-- protocols.io DOI -->
@@
 medium (10 -- 20 $\mu$m), and large (>20 $\mu$m). Counts for small wheat and
-potato starches were combined for the mixed treatment samples, as it is not 
-possible to distinguish the small starch granules between the two species.
+potato starches were combined for the mixed-treatment samples, as it was not 
+possible to distinguish betweem the small starch granules from the two species.
 
@@
 
-### Extraction from samples
+### Extraction method
 
 Extraction of starches from the calculus samples was performed by dissolving the
-calculus in ethylenediaminetetraacetic acid (EDTA), as it has been shown to
-have little to no effect on starch granules (especially native granules)
+calculus in ethylenediaminetetraacetic acid (EDTA) 
 [@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
@@
 Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
-When dispersing the liquid (EDTA) sample on the slide, the sample was homogenised using 
-the pipette. It was not possible to count all wheat starches
-on a slide, so wheat and mix counts were extrapolated as described for the treatment
-solutions.
+When transferring the sample to the slide, the sample was homogenised using 
+the pipette to ensure that the counted transects were representative of the 
+whole slide. Wheat and mix counts were extrapolated as described above.
 
@@
 between the potato and wheat counts.
+This was applied to total biofilm weight and starch count per mg 
+calculus (also z-score standardised) to account for differences in starch 
+concentration in the calculus [as per 
+@wesolowskiEvaluatingMicrofossilContent2010].
 
-All scripts and data are available on OSF (https://osf.io/uc5qy/)
-and Github (https://github.com/bbartholdy/byoc-starch).
+All scripts and data used in the analysis are available on OSF 
+(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch).
+All protocols are available on protocols.io.
 
@@
 included in the analysis, resulting in a total of 48 biofilm samples (two plates of 24),
-45 of which were used for analysis (three samples were set aside for separate analysis).
+45 of which were used for analysis (three samples were set aside for later
+ analysis).
 Most control samples contained no starch granules, while some contained negligible
-quantities (supp. mat).
+quantities (see Supplementary Material).
 
@@
 small potato starches were counted as wheat starches in the mixed-treatment samples. This is reasonable given that the small potato granules make up an insignificant proportion of the 
-total count of small granules within mixed-treatment solutions which are predominantly
+total count of small granules within mixed-treatment solutions, which are predominantly
 wheat granules
@@
 starch, with a total concentration of 0.50% (w/v).
-The mixed-treatment had the highest absolute count of starch granules in solution,
+The mixed treatment had the highest absolute count of starch granules in solution (mean = `r subset(count_sol, treatment == "mix")$total`),
 while the biofilms exposed to the wheat solution preserved the greatest number of
-granules. The potato treatment had the lowest absolute counts in both the solution 
-and in the biofilm samples <!-- insert counts? -->
+granules 
+(mean = `r subset(count_sample, treatment == "wheat")$total`). 
+The potato treatment had the lowest absolute counts in both the solution 
+(`r subset(count_sol, treatment == "potato")$total`)
+and in the biofilm samples 
+(`r subset(count_sample, treatment == "potato")$total`) 
 (Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
@@
 
-```{r sep-plot, fig.cap="Proportion (%) of sizes of (A) starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring).
+```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring).
 l = large, m = medium, s = small."}
@@
 
-<!-- concentrations of solutions vs. starches in a mouthful of food 
-starches in a potato, etc. 
-Contact Steph -->
-
 <!-- secondary findings -->
@@
 This is consistent with the findings by Power and collagues [-@powerChimpCalculus2015]
-mentioned previously, which found that the larger (defined as around 10 -- 20 $\mu$m) 
+mentioned previously, which found that the large (defined as medium in this study) 
 granule-producing plants were overrepresented; although, the representation of granules 
@@
 on  small granules, would eliminate a large portion of the consumed starch granules,
-and would explain why the granules extracted from archaeological dental calculus 
-are often sparse.
-
-Additionally, the size of the calculus deposit seems to influence the quantity of 
+and could explain the large descrepency between our counts (unaffected by hydrolysis and taphonomy) and the aforementioned studies.  
+Another potentially important factor from our results is the size of the calculus
+deposit, which seems to influence the quantity of 
 starches extracted from the calculus, as we found a strong positive correlation
 between biofilm size and retained starch granules
-(Figure \@ref(fig:cor-plot)), which contradicts findings from archaeological contexts
+(Figure \@ref(fig:cor-plot)); a result that contradicts findings from archaeological contexts
 [@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014]. 
@@
 [@radiniFoodMultiplePathways2017; @radiniMedievalWomenEarly2019].  
-
 When starch granules enter the mouth, whether through ingestion of food or accidental
@@
 that the bulk of starch granules are swallowed along with the food, and are 
-only briefly retained in the oral cavity. Other granules that are broken off
+only briefly present in the oral cavity. Other granules that are broken off
 during mastication may be retained in the dentition. These granules are then
@@
 Starch granules that are trapped in crevices and channels on the surface of the
-mineralised plaque would be (at least to some extent) protected from salivary
-clearance and mechanical cleaning actions of the tongue and lips. This may explain 
-why a previous study found starches more commonly in clusters than dispersed over
-the surface of the dental calculus [@powerSEMCalculus2014]. The authors hypothesised 
-that the granules are either deposited in clusters, or group together in cracks and
+mineralised plaque are (at least to some extent) protected from salivary
+clearance and mechanical cleaning actions of the tongue and lips, especially once
+a new layer of plaque has covered the surface of the calculus. This may explain 
+why a previous study found starches more commonly in clusters, rather than 
+dispersed over the surface of the dental calculus [@powerSEMCalculus2014]. 
+The authors hypothesised that the granules are either deposited in clusters, or 
+group together in cracks and
 crevices within the mineral matrix. Unmineralised Lacunae and channels within the 
@@
 While potentially protected against clearance, granules trapped 
-in plaque/calculus will still be susceptible to hydrolysis, as $\alpha$-amylase has
+in plaque/calculus may still be susceptible to hydrolysis, as $\alpha$-amylase has
 the ability to bind to both tooth enamel and bacteria within a biofilm and retain
@@
 [@scannapiecoSalivaryAmylase1993; @nikitkovaStarchBiofilms2013; @tanStudyBacterialViability2004; @tanCalculusUltrastructure2004].
-<!-- These channels and pockets of organic matter within the inorganic matrix are used by the 
-bacteria for transportation of nutrients, and their persistence in the otherwise 
-mineralised matrix may be the result of the differing 
-rate of mineralisation displayed by the bacteria [@tanStudyBacterialViability2004].-->
 The susceptibility of granules to hydrolysis depends on the crystallinity and size
@@
 incorporation rate. More details on the absence of $\alpha$-amylase 
-activity is discussed elsewhere<!--Bartholdy et al. in prep-->.
-<!--
-Starches also have the ability to adhere to certain bacteria [@toppingResistantStarch2003],
-which would allow starches to attach to bacterial communities within the biofilm.
-Following deposition and/or mineralisation events, the granules that are able to 
-resist hydrolysis and 
-mechanical removal will then be trapped in pockets within the mineral matrix, and
-will temporarily be protected from shearing action and salivary 
-clearance. -->
+activity is discussed elsewhere (Bartholdy et al. in prep.).
 What we can say, is that the dietary picture we obtain from starch granules extracted 
@@
 
+<!--Starches also have the ability to adhere to certain bacteria [@toppingResistantStarch2003],
+which would allow starches to attach to bacterial communities within the biofilm.
+Following deposition and/or mineralisation events, the granules that are able to 
+resist hydrolysis and 
+mechanical removal will then be trapped in pockets within the mineral matrix, and
+will temporarily be protected from shearing action and salivary 
+clearance. -->
+
 <!-- in summary: we know very little about the mechanism behind how starches 
@@
 biofilm model for dietary research in archaeology. A limitation of this study 
-is the absence of amylase in the model. The presence of amylase will likely affect 
+is the absence of $\alpha$-amylase in the model. The presence of $\alpha$-amylase will likely affect 
 the total granule counts as well as the size ratios, as smaller starches may be more 
 susceptible to hydrolysis 
-[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. 
-In future experiments, we recommend the addition of $\alpha$-amylase to the model
-in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. 
-<!-- foreshadowing...? -->
+[@francoStarchDegradation1992; @haslamDecompositionStarch2004]; however, 
+the lack of $\alpha$-amylase in the system is a beneficial side effect, 
+as it can allow us to directly explore
+the effect of $\alpha$-amylase on starch counts in future experiments,
+where $\alpha$-amylase can be added to the model
+in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. <!-- foreshadowing...? -->
 While we are able to show the differing 
@@
 Bartholdy and colleagues (in prep.),<!-- insert citation --> provides a method
-to directly
-explore the incorporation and extraction of dietary compounds from dental calculus 
-in a controlled laboratory setting. The addition of known starch species in
+to explore the incorporation and extraction of dietary compounds from dental calculus 
+in a controlled laboratory setting, as many of the variables can be adjusted. 
+The addition of known starch species in
 separate samples also removes the identification bias against the smaller
@@
 [@lemoyneCalculusPretreatments2021].
-<!-- How it will facilitate the specific topic of research (e.g. archaeology) -->
 *In vivo* studies on humans and non-human primates will be limited by the variability 
@@
 numerous factors including genetics, diet, and tooth morphology [@jepsenCalculusRemoval2011]<!-- more citations -->.
-Further research is necessary to explore the mechanisms of the intra-oral starch 
-incorporation and retention in dental calculus, and the dental calculus model
-presented in this study is uniquely suited to explore these questions and 
-may improve interpretations of dietary research in past populations.
-<!-- still need to replicate results and do more science, but pretty cool, right? -->
+It can also facilitate training of students and researchers on methods of 
+dental calculus analysis, such as starch and phytolith extraction and 
+identification, where it can replace the use of finite archaeological resources.  
+<!-- we still need to replicate results and do more science... but pretty cool, right? -->
 
@@
 While we have shown multiple factors that influence the likelihood 
-of incorporation, it seems the process is somewhat stochastic, and possibly related 
-to the biofilm mineralisation events (as discussed above), which are irregular and
-related to various factors. 
-The dental calculus biofilm model presented in this study is a powerful tool to 
-investigate various aspects of dietary research and starch incorporation and
-retention in dental calculus.
+of incorporation, the process still appears to be somewhat stochastic. Further
+research is needed to make sense of the contributing factors, and to explore the
+mechanisms of intra-oral starch incorporation and retention in dental calculus.
+The dental calculus model presented in this study is uniquely suited to explore 
+these questions and may improve interpretations of dietary practices in past populations.
 

### 2021-08-27 analysis/paper/paper.Rmd
knit first draft... although, still needs abstract

@@
 output:
+  bookdown::word_document2:
+    reference_docx: "../templates/template.docx"
   bookdown::pdf_document2:
@@
     keep_md: true
-  bookdown::word_document2:
-    reference_docx: "../templates/template.docx"
 linestretch: 2.0 # double-spacing for review purposes
@@
 starch, with a total concentration of 0.50% (w/v).
-The mixed treatment had the highest absolute count of starch granules in solution (mean = `r subset(count_sol, treatment == "mix")$total`),
+The mixed treatment had the highest absolute count of starch granules in solution (mean = `r subset(count_sol, solution == "mix")$total`),
 while the biofilms exposed to the wheat solution preserved the greatest number of
@@
 The potato treatment had the lowest absolute counts in both the solution 
-(`r subset(count_sol, treatment == "potato")$total`)
+(`r subset(count_sol, solution == "potato")$total`)
 and in the biofilm samples 

### 2021-08-30 README.Rmd
add sections to manuscript
fix figure captions and plot colours

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bbartholdy/byoc-starch/main?urlpath=rstudio) <!-- DOI badge: [![DOI](.svg)](repo link) -->
+
+<!-- Document structure provided by rrtools package -->
 

### 2021-08-30 analysis/paper/fig_tab.Rmd
further development of manuscript
fix figure captions and plot colours

@@
 
+## Tables
 <!-- ### Tables ### -->
 
+```{r sample-perc}
+# convert proportions to percent labels
+perc_comb <- prop_comb %>% 
+  mutate(across(c(s, m, l, total), scales::percent, 0.001))
+
+perc <- prop %>%
+  mutate(across(c(s, m, l, total), scales::percent, 0.001)) %>% 
+  add_row(tibble_row(perc_comb), .before = 3)
+```
+
+```{r sample-prop-tbl}
+cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
+knitr::kable(perc, col.names = cnames, 
+             caption = "The mean percentage of starches from the solutions that 
+              were incorported into the samples.")
+```
+
+## Figures
 <!-- ### Figures ### -->
@@
 
-<!-- change colour of medium starches in D to match C -->
 ```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring).
@@
 ```
+
+```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments."}
+pl_cor
+```
+
+```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
+pl_cor2
+```

### 2021-08-30 analysis/paper/paper.Rmd
fix figure captions and plot colours

@@
 output:
-  bookdown::word_document2:
-    reference_docx: "../templates/template.docx"
   bookdown::pdf_document2:
@@
     number_sections: yes
+  bookdown::word_document2:
+    reference_docx: "../templates/template.docx"
   bookdown::html_document2:
@@
 oral biofilm with a multifactorial aetiology and variable formation rates both 
-within and between individuals [@jepsenCalculusRemoval2011], contributing to
+within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], contributing to
 the stochasticity of starch representation being observed in numerous studies.
@@
 
-<!-- INSERT PICTURES OF STARCHES -->
+<!-- INSERT MICROSCOPE PICTURES OF STARCHES -->
 
@@
 
-```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring).
-l = large, m = medium, s = small."}
+```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring)."}
 #plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
@@
 
-<!-- change colour of medium starches in D to match C -->
-```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring).
-l = large, m = medium, s = small."}
+```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring)."}
 #plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
@@
 ```
+
 <!-- Combine to a single figure? -->
@@
 ```
+
 <!-- with Z-score standardisation -->
@@
 The latter is especially difficult to control *in vivo* as it is influenced by 
-numerous factors including genetics, diet, and tooth morphology [@jepsenCalculusRemoval2011]<!-- more citations -->.
+numerous factors including genetics, diet, salivary flow, tooth position and morphology 
+[@jepsenCalculusRemoval2011; @simon-soroOralGeography2013; @proctorSpatialGradient2018; @fagernasMicrobialBiogeography2021; @haffajeeBiofilmPosition2009], 
+as well as evolutionary differences through time [@yatesOralMicrobiome2021].
 It can also facilitate training of students and researchers on methods of 

### 2021-08-30 analysis/scripts/figures.R

@@ -138,19 +138,26 @@ pl_wheat2 <- ratio_mix %>%
   coord_polar(theta = "y") +
   theme_void() +
   #geom_text(aes(x = 2, label = label_wheat)) +
-  scale_fill_viridis_d(begin = 0.7)
+  scale_fill_viridis_d(begin = 0.7,
+                       name = "Size",
+                       labels = c("Large", "Medium", "Small"))
+  #scale_fill_manual(values = viridisLite::viridis(3, begin = 0.7),
+  #                  name = "Size", labels = c("Large", "Medium", "Small"))
 
 pl_potato2 <- ratio_mix %>%
+  mutate(potato = if_else(is.na(potato), 0, potato)) %>% # fix colour issue
   select(!contains("wheat")) %>%
-  filter(size != "s") %>%
+  #filter(size != "s") %>% # this caused m to have a different colour that pl_wheat2
   ggplot(aes(x = type, y = potato, fill = size)) +
-  #geom_rect(xmin = 2, xmax = 3, ymin = ratio_sep$ymin, ymax = ratio_sep$ymax_wheat) +
   geom_bar(stat = "identity", width = 1, colour = "white") +
   coord_polar(theta = "y") +
   theme_void() +
   #geom_text(aes(x = 2, label = label_wheat)) +
-  scale_fill_viridis_d(begin = 0.7)
-
+  scale_fill_viridis_d(begin = 0.7,
+                       name = "Size",
+                       labels = c("Large", "Medium", "Small"))
+  #scale_fill_manual(values = viridisLite::viridis(3, begin = 0.7)[c(1,2)],
+  #                  name = "Size", labels = c("Large", "Medium", "Small"))
 
 # Scatter plot ------------------------------------------------------------
 

### 2021-09-20 analysis/paper/paper.Rmd
revisions to first draft

@@
 ---
-title: "Untitled"
+title: "Investigating Biases Associated with Dietary Starch Incorporation into an Oral Biofilm Model"
 author:
-  - "Bjørn Peare Bartholdy"
+  - Bjørn Peare Bartholdy
   - Amanda G. Henry
@@
     #template: "../templates/sn-article-template/sn-article.tex"
+    includes:
+      in_header: "../templates/latex/pre-print/preamble.tex"
+      before_body: "../templates/latex/pre-print/before_body.tex"
     df_print: kable
@@
     number_sections: yes
-  bookdown::word_document2:
-    reference_docx: "../templates/template.docx"
   bookdown::html_document2:
@@
     keep_md: true
+  bookdown::word_document2:
+    reference_docx: "../templates/template.docx"
 linestretch: 2.0 # double-spacing for review purposes
@@
 Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
-keywords: |
-  keyword 1; keyword 2
+keywords: 
+  - oral biofilm
+  - starch
+  - alpha-amylase
+  - dental calculus
 ---
@@
 and other organic residues [@buckleyDentalCalculusCooking2014]. 
-It has also proven to be able to retain and preserve dietary information for millenea,
-providing detailed dietary information on past populations 
+This dietary information can be preserved within calculus over many millenia,
+providing a unique window into the food-related behaviours of past populations 
 [@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
-and species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].
+and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].
 
 <!-- more detailed background, comprehensible to scientists in related disciplines -->
-Until recently, few studies directly investigated the presence of plant microremains
+Until recently, only a few studies directly investigated the presence of plant
+microremains
 in the dental calculus of archaeological remains. The ability to extract phytoliths
@@
 and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
-and Fox and colleagues [-@foxPhytolithCalculus1996]. The ability to detect coca
-consumption by looking at human dental calculus was explored by Klepinger and Kuhn
-[-@klepingerCalculusCoca1977], and starches and phytoliths were extracted from 
-human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].
-
+and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
+extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
 In more recent years, the study of dental calculus has increased exponentially, 
@@
 <!-- clearly stating the general problem being addressed by this particular study -->
-<!-- Dental calculus research has its limitations, all of which I will solve with this
-single method. -->
 Despite this, our knowledge of dental calculus and the incorporation pathways of 
 the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
-knowledge of information-loss caused by these pathways and the methods we use to 
-extract and analyse dental calculus and make inferences on past diets. 
+knowledge of information-loss caused by these pathways. Additionally, the methods
+we use to extract and analyse dental calculus, and make inferences on past diets
+represent another potential source of bias. Studies on both archaeological and
+modern individuals have explored these biases in more detail.
 Extraction methods were tested 
@@
 involving archaeological calculus are problematic due to variability between and 
-within individuals. Other studies have been conducted on modern populations 
-[@leonardDentalCalculus2015] and primates [@powerChimpCalculus2015], both with
-well-documented dietary intake, to reconcile our interpretations on the recovery 
-of microremains (both phytoliths and starches) from dental calculus and how well
-they represent actual dietary intake. These studies are justifiably limited, 
+within individuals. 
+Studies conducted on modern humans [@leonardDentalCalculus2015] 
+and non-human primates [@powerChimpCalculus2015], 
+have explored the how well microremains (phytoliths and starches)
+extracted from dental calculus represent the actual dietary intake.
+These studies are justifiably limited, 
 despite meticulous documentation and observation, due to unknown variables and 
@@
 the stochasticity of starch representation being observed in numerous studies.
-Additionally, the rate of $\alpha$-amylase differs both between and within
-individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
+Additionally, the concentration of oral $\alpha$-amylase differs both between and
+within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
 causing different rates of hydrolysis of the starch granules present in the oral
@@
 retention of starch granules within dental calculus in a controlled laboratory 
-setting, allowing us full control over dietary input. We use a multiwell setup for consistent, high-throughout analysis.
-Our main questions concern the representation of granules extracted
+setting, allowing us full control over dietary input. Our main questions concern
+the representation of granules extracted
 from the calculus compared to the actual intake. How much of the original diet is 
@@
 diet from extracted microremains?  
-We find that, despite an absence of $\alpha$-amylase in
+We find that, despite the absence of $\alpha$-amylase in
 the model, a limited proportion of the starch input is actually 
@@
 starch granules
-that are incoporated into the calculus. We also show that the number of starch
+that are incorporated into the calculus. We also show that the number of starch
 granules that are incorporated increases as the size of the calculus deposit 
@@
 
-```{r protocol-fig, fig.cap="Overview of experiment protocol"}
-knitr::include_graphics(here("analysis/figures", "ExpL_protocol.png"))
-```
-<!-- modify sampling days from protocol picture to match amylase sampling -->
+## Biofilm formation
 
-## Biofilm formation
+<!-- insert figure of the setup -->
 
@@
 
-<!-- add to protocols.io -->
+```{r protocol-fig, fig.cap="Overview of experiment protocol"}
+knitr::include_graphics(here("analysis/figures", "ExpL_protocol.png"))
+```
+
+<!-- modify sampling days from protocol picture to match amylase sampling -->
+
 Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
@@
 months prior to donation. 
-The saliva was filtered through a sterilised (with bleach) nylon cloth to remove
-particulates.
+The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15% active chlorine) 
+nylon cloth to remove particulates.
 Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
@@
 substrata were transferred to a new plate containing 1 ml/well AS and incubated 
-at 36&deg;C, 30 rpm. The inoculation process was repeated on days 3 and 5. 
+in a shaking incubator (Infors HT Ecotron) at 36&deg;C, 30 rpm. 
+The inoculation process was repeated on days 3 and 5. 
 AS was partially refreshed once per day and fully refreshed every three days, 
@@
 Starch treatments were initiated on day 9 to avoid starch granule counts being 
-affected by $\alpha$-amylase hydrolysis. An $\alpha$-amylase (EC 3.2.1.1) activity
+affected by $\alpha$-amylase hydrolysis from saliva inoculation days. 
+An $\alpha$-amylase (EC 3.2.1.1) activity
 assay was conducted to confirm that no amylase was present in the system before
@@
 new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
-concentrations (w/v) wheat and potato. All starch solutions were created in a 5%
-(w/v) sucrose solution. Before transferring biofilm samples to the starch treatment plate, 
-the plates were agitated to keep the starches in suspension in the solutions. 
-During treatments, the rpm was increased to 60 to facilitate contact between
-starch granules and biofilms.
+concentrations (w/v) wheat and potato. All starch treatments were created in dH~2~O
+with 5% (w/v) sucrose. Before transferring biofilm samples to the starch treatment
+plate, the plates were agitated to keep the starches in suspension in the
+solutions. During treatments, the rpm was increased to 60 to facilitate contact
+between starch granules and biofilms.
 
@@
 containing sterile artificial saliva (due to the colour of artificial saliva).
-Two photometric readings were conducted for each plate with a 540 nm filter on a... 
-<!-- photometric reader specs -->
+Two photometric readings were conducted for each plate with a 540 nm filter on a 
+Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
 The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
@@
 
-## Starch counts
-
-Starch granule counts were divided into three size categories: small (<10 $\mu$m),
-medium (10 -- 20 $\mu$m), and large (>20 $\mu$m). Counts for small wheat and
-potato starches were combined for the mixed-treatment samples, as it was not 
-possible to distinguish betweem the small starch granules from the two species.
-
-### Treatment solutions
+## Treatment solutions
 
@@
 on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
-(Zeiss Axioscope A1). Counting starches on a full slide was not feasible, so three slide transects were 
-counted (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts were extrapolated 
-(see Supplementary material for more details).
+(Zeiss Axioscope A1). For wheat and mixed treatment samples, we counted three 
+slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts 
+were extrapolated (see Supplementary material for more details). For potato
+treatment samples, the whole slide was counted.
 
-### Extraction method
+## Extraction method
 
 Extraction of starches from the calculus samples was performed by dissolving the
-calculus in ethylenediaminetetraacetic acid (EDTA) 
+calculus in -0.5 M<!-- what percentage???--> ethylenediaminetetraacetic acid (EDTA) 
 [@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
@@
 whole slide. Wheat and mix counts were extrapolated as described above.
+Both wheat and potato granules were divided into three size categories: 
+small (<10 $\mu$m), medium (10 -- 20 $\mu$m), and large (>20 $\mu$m).
 
@@
 
-Statistical analysis was conducted in `r version$version.string`  [@R-base].
+Statistical analysis was conducted in `r version$version.string`  [@R-base] and
+the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
+here [@R-here], and patchwork [@R-patchwork].
 <!-- R packages used -->
@@
   add_row(tibble_row(count_sol_comb_out), .before = 3)
-```
 
-```{r solution-count-tbl}
-cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
-knitr::kable(count_sol_out, col.names = cnames, 
-             caption = "Mean starch counts from solutions, including the
-             proportional makeup of the different sizes of granules.")
+count_sol_out[1,3] <- NA
 ```
@@
   select(1,2,3,7,4,8,5,9,6,10) #%>%
+
+count_samp_out[1,3] <- NA # replace with NA
   #add_row(tibble_row(count_sample_comb_out), .before = 3)
+```
+
+It was not possible to differentiate between potato and 
+wheat starches smaller than ca. 10 $\mu$m. These were counted as wheat,
+as we assumed that the majority of the small granules were wheat. This is a
+reasonable assumption to make based on the small granule counts from the separate
+treatment solutions, where the wheat granules made up 
+`r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`
+of all small granule counts in the separated solutions.
+<!-- count of small granules from wheat solution divided by the sum of small granules 
+from the potato and wheat solutions -->
+
+```{r solution-count-tbl}
+cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
+options(knitr.kable.NA = "")
+knitr::kable(count_sol_out, col.names = cnames, 
+             caption = "Mean starch counts from solutions, including the
+             proportional makeup of the different sizes of granules.")
 ```
@@
 
-It was not possible to differentiate between potato and 
-wheat starches smaller than ca. 10 $\mu$m,
-small potato starches were counted as wheat starches in the mixed-treatment samples. This is reasonable given that the small potato granules make up an insignificant proportion of the 
-total count of small granules within mixed-treatment solutions, which are predominantly
-wheat granules
-(`r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`).
-<!-- count of small granules from wheat solution divided by the sum of small granules 
-from the potato and wheat solutions -->
-
 <!-- absolute solution counts compared to absolute sample counts. does exposure
@@
 starch, with a total concentration of 0.50% (w/v).
-The mixed treatment had the highest absolute count of starch granules in solution (mean = `r subset(count_sol, solution == "mix")$total`),
+The mixed treatment had the highest absolute count of starch granules in solution (mean = `r signif(subset(count_sol, solution == "mix")$total, 3)`),
 while the biofilms exposed to the wheat solution preserved the greatest number of
 granules 
-(mean = `r subset(count_sample, treatment == "wheat")$total`). 
+(mean = `r signif(subset(count_sample, treatment == "wheat")$total, 3)`). 
 The potato treatment had the lowest absolute counts in both the solution 
-(`r subset(count_sol, solution == "potato")$total`)
+(`r signif(subset(count_sol, solution == "potato")$total, 3)`)
 and in the biofilm samples 
-(`r subset(count_sample, treatment == "potato")$total`) 
+(`r signif(subset(count_sample, treatment == "potato")$total, 3)`) 
 (Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
-
-<!-- look at proportions of incorporated starches in each sample before looking at overall means -->
 
@@
 retained in the mineral matrix, and that the size of the starch granules
-may affect the likeliness of incorporation. The proportions of starch granules 
+may affect the likelihood of incorporation. The proportions of starch granules 
 (of all sizes) present in the extracted samples were similar across all treatments
@@
 despite large differences in absolute granule counts between wheat 
-(`r format(count_sol$total[4], big.mark = ",")`) 
+(mean = `r format(count_sol$total[4], big.mark = ",")`) 
 and potato
-(`r format(count_sol$total[3], big.mark = ",")`)
+(mean = `r format(count_sol$total[3], big.mark = ",")`)
 solutions.  
@@
 in higher quantities in the dental calculus, at least prior to inhumation and 
-degradation in the burial environment; although, starch availability is unlikely
-to be a main factor in starch incorporation. 
+degradation in the burial environment. 
 Despite the low proportion of granules recovered from the model calculus 
@@
 archaeological remains 
-[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010].
+[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010]. <!-- although this study extrapolated counts to full sample content, which is
+rarely done in archaeological studies -->This could, in part, be due to the 
+activity of oral amylase, which was absent in our model.
 
@@
 higher magnitude than the difference between samples <!-- a number/percentage-->
-
-<!-- although this study extrapolated counts to full sample content, which is
-rarely done in archaeological studies-->
 
@@
 presented similar findings in non-human primates, where phytoliths were more
-representative of individual diets than starch granules, and plants that 
-produce predominantly larger starch (10-20 $\mu$m) granules were over-represented.
-Both studies concluded that the dietary breadth of a population can
-be captured in a sufficiently large sample, but interpretations on individual diets
-from microremains in dental calculus is tenuous, at best 
-[@leonardDentalCalculus2015; @powerChimpCalculus2015].
+representative of individual diets than starch granules.
 
 <!-- secondary findings -->
-
 We have also shown that the size of the starch granules influences the likelihood
@@
 (Table \@ref(tab:sample-prop-tbl), 
-and Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).
+and 
+Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).
 This was especially true for the potato starches, which can reach up to 100 $\mu$m
@@
 [@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004].
-This is consistent with the findings by Power and collagues [-@powerChimpCalculus2015]
-mentioned previously, which found that the large (defined as medium in this study) 
-granule-producing plants were overrepresented; although, the representation of granules 
-larger than 20 $\mu$m in that study is unclear. Since those results were obtained
-on a modern population of chimps that were only buried for a short period of time, 
-degradation was likely limited [@powerChimpCalculus2015]. The taphonomy of
-starch granules has been addressed previously, and depends on burial environment
-(pH, temperature, microorganisms), as well as processing of the granules prior to
-deposition [@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004]. 
+The size bias is consistent with the findings by Power and colleagues
+[-@powerChimpCalculus2015],
+which found that plants producing starches 10--20 $\mu$m in
+size were over-represented; although, the representation of granules larger than
+20 $\mu$m is unclear.  
+Given the differences in proportional incorporation between wheat and potato,
+especially for small and medium granules, it is also possible that the morphology
+of granules plays a role. However, the proportional count differences between large 
+wheat and potato granules were less pronounced.
+
+<!-- Since those results were obtained -->
+<!-- on a modern population of chimps that were only buried for a short period of time,  -->
+<!-- degradation was likely limited [@powerChimpCalculus2015]. -->
+
+The taphonomy of starch granules has been addressed previously, and depends on 
+burial environment (pH, temperature, moisture content, microorganisms), as well 
+as processing of the granules prior to deposition 
+[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004]. 
 The potential for detrimental diagenetic effects on starch granules increases as 
@@
 [@francoStarchDegradation1992; @haslamDecompositionStarch2004].
-Combined, the effect of intra-oral starch incorporation
-and the bias against large granules, together with the increased effect of taphonomy
-on  small granules, would eliminate a large portion of the consumed starch granules,
-and could explain the large descrepency between our counts (unaffected by hydrolysis and taphonomy) and the aforementioned studies.  
-Another potentially important factor from our results is the size of the calculus
-deposit, which seems to influence the quantity of 
-starches extracted from the calculus, as we found a strong positive correlation
-between biofilm size and retained starch granules
-(Figure \@ref(fig:cor-plot)); a result that contradicts findings from archaeological contexts
+
+<!-- add potential for morphological differences affecting incorporation 
+ Medium potato starches had the highest rate of incorporation 
+ All large starches had a consistently low rate of incorporation 
+ Starch granule adhesion to bacteria and enzymes? Do they have a preference?
+ Binding to select species in a biofilm could explain clustering
+ Colonisation of bacteria onto starch @nagaraBacteriaNutrients2017-->
+
+<!-- Combined, the effect of intra-oral starch incorporation -->
+<!-- and the bias against large granules, together with the increased effect of taphonomy -->
+<!-- on  small granules, would eliminate a large portion of the consumed starch granules, -->
+<!-- and could explain the large descrepency between our counts (unaffected by hydrolysis  -->
+<!-- and taphonomy) and the aforementioned studies.  -->
+
+Another potentially important factor 
+from our results is the size of the calculus deposit, which seems to influence the 
+quantity of starches extracted from the calculus, as we found a strong positive 
+correlation between biofilm size and retained starch granules (Figure \@ref(fig:cor-plot));
+a result that contradicts findings from archaeological contexts
 [@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014]. 
@@
 The lack of a correlation
-in archaeological contexts could be due to diagenetic effects; or, the absence of 
+in archaeological contexts could be due to diagenetic effects. Alternately, the absence of 
 amylase activity in our study could be impacting our interpretations.
 
-<!-- How are starch granules trapped in dental calculus? Does anyone actually know? -->
+<!-- How are starch granules trapped in dental calculus? Does anyone actually know? - adhesion to bacteria
+- stuck in cracks and crevices
+- covered by new layer of plaque
 
+-->
 The mechanism by which starch granules are incorporated into plaque and calculus 
@@
 only briefly present in the oral cavity. Other granules that are broken off
-during mastication may be retained in the dentition. These granules are then
+during mastication may be retained in the dentition.
+
+<!-- initial incorporation of starch granules? 
+cracks and crevices? attachment to bacteria? -->
+
+These granules are then
 susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
@@
 clearance and mechanical cleaning actions of the tongue and lips, especially once
-a new layer of plaque has covered the surface of the calculus. This may explain 
-why a previous study found starches more commonly in clusters, rather than 
-dispersed over the surface of the dental calculus [@powerSEMCalculus2014]. 
-The authors hypothesised that the granules are either deposited in clusters, or 
-group together in cracks and
-crevices within the mineral matrix. Unmineralised Lacunae and channels within the 
+a new layer of plaque has covered the surface of the calculus. 
+The size bias against large granules (>20 $\mu$m) from both wheat and potato 
+(Table \@ref(tab:sample-prop-tbl)) may give further credence to
+this incorporation pathway, as the smaller starch granules have an advantage over
+larger granules, and can be stored in larger quantities.  
+This was also suggested by Power and colleagues [-@powerSEMCalculus2014], who
+observed clusters of starches within dental calculus, rather than an even 
+distribution across the surface of the dental calculus.
+
+<!--Bacteria also have the ability to adhere to starch [@toppingResistantStarch2003],
+which would allow starches to attach to bacterial communities within the biofilm.
+Following deposition and/or mineralisation events, the granules that are able to 
+resist hydrolysis and 
+mechanical removal will then be trapped in pockets within the mineral matrix, and
+will temporarily be protected from shearing action and salivary 
+clearance. -->
+<!-- Unmineralised lacunae and channels within the  
  calculus matrix have been shown to contain viable 
@@
  width [@tanCalculusUltrastructure2004], of which the latter could feasibly contain 
-starch granules. 
+   starch granules.  -->
 If this is indeed one of the incorporation mechanisms of starch granules, then the 
@@
 with our results 
-(Figures \@ref(fig:cor-plot1) and \@ref(fig:cor-plot2)).
-The size bias against large granules (>20 $\mu$m) may give further credence to the 
-incorporation pathway of starch granules primarily in cracks and crevices in the 
-calculus, as the smaller starch granules have an advantage over larger granules, 
-and can be stored in larger quantities.  
+(Figures \@ref(fig:cor-plot) and \@ref(fig:cor-plot2)).
+
 While potentially protected against clearance, granules trapped 
@@
 
-<!--Starches also have the ability to adhere to certain bacteria [@toppingResistantStarch2003],
-which would allow starches to attach to bacterial communities within the biofilm.
-Following deposition and/or mineralisation events, the granules that are able to 
-resist hydrolysis and 
-mechanical removal will then be trapped in pockets within the mineral matrix, and
-will temporarily be protected from shearing action and salivary 
-clearance. -->
-
 <!-- in summary: we know very little about the mechanism behind how starches 
@@
 <!-- limitations of the platform show which questions would be unanswered -->
-This article presents preliminary research exploring the potential of an oral 
-biofilm model for dietary research in archaeology. A limitation of this study 
+A limitation of this study 
 is the absence of $\alpha$-amylase in the model. The presence of $\alpha$-amylase will likely affect 
@@
 [@francoStarchDegradation1992; @haslamDecompositionStarch2004]; however, 
-the lack of $\alpha$-amylase in the system is a beneficial side effect, 
-as it can allow us to directly explore
+the absense can also allow us to directly explore
 the effect of $\alpha$-amylase on starch counts in future experiments,
@@
 <!-- How this engineering will facilitate future research -->
-The oral biofilm model described in this study, and in 
-Bartholdy and colleagues (in prep.),<!-- insert citation --> provides a method
+The oral biofilm model described in this study provides a method
 to explore the incorporation and extraction of dietary compounds from dental calculus 
@@
 shown that the size of granules will influence the likelihood of incorporation, 
-as large starches have a decreased incorporation rate, medium
-starches an increased rate, and small starches remained somewhat constant.
+as large (>20 $\mu$m) starches have a decreased incorporation rate, medium 
+(10--20 $\mu$m)
+starches an increased rate, and small (<10 $\mu$m) granules remained somewhat constant.
 The size of calculus deposit also seems to influence the capacity of 
@@
 
+This research has received funding from the European Research Council under the 
+European Union’s Horizon 2020 research and innovation program, grant agreement 
+number STG–677576 (“HARVEST”).
+
 # References cited

### 2021-09-20 README.Rmd
update README

@@
 
+This compendium can be accessed in a web browser using binder...
+
 This research compendium has been developed using the statistical programming
-language R. To work with the compendium, you will need
+language R. To work with the compendium locally, you will need
 installed on your computer the [R software](https://cloud.r-project.org/)

### 2021-09-20 R/amylase.R
modify amylase.R function

@@ -55,6 +55,8 @@ sample_calc <- function(samp_data, cols = NULL, std_curve, dilf, ml_enz){
   blank <- std_curve$blank
   n_rows <- nrow(samp_data)
   n_cols <- ncol(samp_data)
+  samp_data <- samp_data %>%
+    mutate(across(everything(), as.numeric))
   blank_mat <- blank %>%
      slice(rep(1:n(), each = n_rows))
   samp_data_corr <- samp_data - blank_mat

### 2021-09-22 analysis/scripts/amylase-assay.R
modify amylase assay script

@@ -4,18 +4,18 @@ library(tidyverse)
 
 # Upload data -------------------------------------------------------------
 
-plt1_ph1 <- readr::read_table(
-  here::here("analysis/Amylase/ExpM_plate1_photometric1.txt"),
+plt1_ph1 <- readr::read_tsv(
+  here::here("analysis/data/raw_data/ExpM_plate1_photometric1.csv"),
   col_names = F)
-plt1_ph2 <- readr::read_table(
-  here::here("analysis/Amylase/ExpM_plate1_photometric2.txt"),
+plt1_ph2 <- readr::read_tsv(
+  here::here("analysis/data/raw_data/ExpM_plate1_photometric2.csv"),
   col_names = F)
 
-plt2_ph1 <- readr::read_table(
-  here::here("analysis/Amylase/ExpM_plate2_photometric1.txt"),
+plt2_ph1 <- readr::read_tsv(
+  here::here("analysis/data/raw_data/ExpM_plate2_photometric1.csv"),
   col_names = F)
-plt2_ph2 <- readr::read_table(
-  here::here("analysis/Amylase/ExpM_plate2_photometric2.txt"),
+plt2_ph2 <- readr::read_tsv(
+  here::here("analysis/data/raw_data/ExpM_plate2_photometric2.csv"),
   col_names = F)
 
 

### 2021-09-30 analysis/paper/abstract.Rmd
revisions to first draft

@@
 # Abstract
+
+```{r include=FALSE}
+w_count <- wordcountaddin::word_count()
+```
 
@@
 
+Dental calculus, mineralised plaque, has proven to contain a wealth of information 
+on the dietary habits of past populations. The extraction and identification of
+starch granules contained within dental calculus has already provided valuable
+dietary insights from a wide range of regions and time periods, and seem to preserve 
+well in the archaeological record.
 <!-- One sentence clearly stating the general problem being addressed by this particular study -->
+While the majority of studies have focused on microfossil extraction and
+identification from archaeological remains, <!--
+as well as methods of extraction and decontamination,--> few studies have 
+addressed the biases of starch retention in dental calculus, and have been
+limited to *in vivo* studies on modern humans and non-human primates.
+<!--
+and have primarily been 
+conducted on modern humans, non-human primates, and archaeological remains; 
+all of which have 
+methodological challenges of dealing with living (or once living) organisms, 
+and the variables that are difficult to control. <!-- rephrase -->
 
 <!-- One sentence summarising the main result (with the words "here we show" or their equivalent) -->
+Here, we present a multispecies oral biofilm model, which allows experimental
+research to be conducted on *in vitro* dental calculus in a controlled laboratory
+setting. The biofilms were exposed treatment solutions with known quantities of
+dietary starches (wheat and potato) during the 25-day growth period. After this,
+the starch granules were extracted from the mature biofilm (dissolution in EDTA),
+and counted. We 
+show that the granule counts extracted from the model dental calculus represented 
+a low proportion of the total counts (ranging from 0.064% to 0.161%) exposed to 
+the biofilms throughout the experiment. Additionally, we found that the ratios
+of granule sizes from the extracted starch granules differed from the original
+treatment solutions, with large granules (>20 $\mu$m) consistently being
+under-represented.
+Additionally, the size ratios of granules within starch species shifted between the input starch solutions and the granules extracted from the model calculus, with large granules (>20 $\mu$m) consistently being underrepresented.
+We also found a correlation
+between the absolute granule counts and dry-weight of the biofilm (r = ), as well
+as between the concentration (count per mg) of granules and dry-weight (r = ).
 
@@
 <!-- Two or three sentences to provide a broader perspective, readily comprehensible to a scientist in any discipline, may be included in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced by their inclusion.
-Under these circumstances, the length of the paragraph can be yp to 300 words.
-(This example is 190 words without the final section, and 250 words with it).
+-->
+Our results reinforce previous *in vivo* studies suggesting that dental calculus 
+presents a very small, and partly biased picture of the original dietary intake 
+of starches, with an over-representation of plants producing granules smaller than
+20 $\mu$ in size. The experimental model presented here is well-suited to address
+the need for further validation of methods and biases associated with dietary
+research on dental calculus.
+
+Word count: `r w_count`

### 2021-09-30 analysis/supplementary-materials/supp-mat.Rmd
revisions to first draft

--- 2021-08-20 analysis/supplementary-materials/supp-mat.Rmd
@@
     df_print: paged
+    toc: true
+    toc_float: true
 ---
 
-```{r setup}
+```{r setup, include=FALSE}
 library(here)
@@
 
-## Metadata
+## R Session info
 
-**raw_counts**
-<!-- insert data type for each column -->
+```{r}
+print(sessionInfo(), locale = FALSE)
+```
+
+## Metadata for raw data files
+
 Counts represent the absolute number of starches counted on a slide
 
-sample: Sample number.
-plate: Plate number that the sample came from.
-row: Which row on the plate the sample came from.
-s: Small starch count.
-m: Medium starch count.
-l: Large starch count.
-total: Sum of s, m, and l.
-treatment: Treatment solution to which the samples were exposed.
-starch: Type of starch that was counted
-weight: Weight of the biofilm sample
-vol: Total volume of EDTA in which the sample was dissolved.
-portion_slide: Proportion of the microscope slide that was counted.
+**starch_counts.csv**
+
+| variable | description |
+|----|----|
+| sample | Sample number. |
+| plate | Plate number that the sample came from. |
+| row | Which row on the plate the sample came from. |
+| s | Small starch count. |
+| m | Medium starch count. |
+| l | Large starch count. |
+| total | Sum of s, m, and l. |
+| treatment | Treatment solution to which the samples were exposed. |
+| starch | Type of starch that was counted. |
+| weight | Weight of the biofilm sample. |
+| vol | Total volume of EDTA in which the sample was dissolved. |
+| portion_slide | Proportion of the microscope slide that was counted. Total transects on slide divided by counted transects. |
+
+**solution_counts**
+
+| variable | description |
+|----|----|
+| solution | Type of starch in solution. |
+| concentration | Concentration (%w/v) of starch in solution. | 
+| vol_slide | Volume of solution added to slide. |
+| vol_total | Total volume of solution in aliquot. |
+| portion_slide | Proportion of slide that was counted. Total transects on slide divided by counted transects. |
+| slide | Slide number. |
+| starch | Starch type counted. |
+| s | Small starch count. |
+| m | Medium starch count. |
+| l | Large starch count. |
+| total | Sum of s, m, and l. |
+
+## Amylase activity
+
+Amylase activity in U/mL enzyme, where a U is mg maltose released from starch in six minutes at 36 &deg;C.
+
+Tables containing the amylase activity results for both plates and photometric
+readings.
+
+```{r amylase-activity-results, message=FALSE}
+# table of results reported in units amylase per mL enzyme (but let's be honest,
+  # ...it doesn't really matter what the unit is. No activity is no activity)
+source(here("analysis/scripts/amylase-assay.R"))
+cols <- c("1", "2", "3")  
+rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
+plt1_ph1_result <- rbind(sal1_ph1, bmm1_ph1)
+rownames(plt1_ph1_result) <- rows
+plt1_ph2_result <- rbind(sal1_ph2, bmm1_ph2)
+rownames(plt1_ph2_result) <- rows
+plt2_ph1_result <- rbind(sal2_ph1, bmm2_ph1)
+rownames(plt2_ph1_result) <- rows
+plt2_ph2_result <- rbind(sal2_ph2, bmm2_ph2)
+rownames(plt2_ph2_result) <- rows
+```
+
+```{r amylase-activity-tables}
+plt1_ph1_result
+plt1_ph2_result
+plt2_ph1_result
+plt2_ph2_result
+```
 
@@
 
+Slide transects were calculated by counting the number of transects on the 
+cover slip under the microscope. This was done by starting in the bottom-left
+corner, and counting the total number of full fields-of-view across the cover 
+slip to the bottom-right corner. The total number of transects was 29 (verified
+multiple times).
+
+Solution counts were extrapolated to the quantity in a 1 ml solution with
+the following formula:
+
+$$ \text{corrected count} = \text{raw count} \times \frac{\text{total} \space \text{transects}}{\text{counted} \space \text{transects}} \times 100 \mu l $$
+
 Sample counts were extrapolated using the following formula:
@@
 
-Solution counts were extrapolated to the quantity in a 1 ml solution with
-the following formula:
+## Microscope images
 
-$$ \text{corrected count} = \text{raw count} \times \frac{29}{3} \times 100 \mu l $$
+```{r img-1, echo=FALSE, fig.cap="Image of starch granules extracted from a calculus sample"}
 
-## Functions
+```

### 2021-09-30 analysis/scripts/figures.R

@@ -139,7 +139,7 @@ pl_wheat2 <- ratio_mix %>%
   theme_void() +
   #geom_text(aes(x = 2, label = label_wheat)) +
   scale_fill_viridis_d(begin = 0.7,
-                       name = "Size",
+                       name = "",
                        labels = c("Large", "Medium", "Small"))
   #scale_fill_manual(values = viridisLite::viridis(3, begin = 0.7),
   #                  name = "Size", labels = c("Large", "Medium", "Small"))
@@ -154,7 +154,7 @@ pl_potato2 <- ratio_mix %>%
   theme_void() +
   #geom_text(aes(x = 2, label = label_wheat)) +
   scale_fill_viridis_d(begin = 0.7,
-                       name = "Size",
+                       name = "",
                        labels = c("Large", "Medium", "Small"))
   #scale_fill_manual(values = viridisLite::viridis(3, begin = 0.7)[c(1,2)],
   #                  name = "Size", labels = c("Large", "Medium", "Small"))

### 2021-10-08 analysis/paper/_archive/first_draft_rev-1/fig_tab.Rmd
incorporate AGH suggestions

--- 2021-08-30 analysis/paper/fig_tab.Rmd
@@
 
-```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring).
-l = large, m = medium, s = small."}
+```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."}
 #plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
-pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
-```
-
-```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring).
-l = large, m = medium, s = small."}
-#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
-pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
 ```

### 2021-10-08 analysis/paper/first_draft_rev-1/paper.Rmd
incorporate AGH suggestions

--- 2021-09-20 analysis/paper/paper.Rmd
@@
 ---
-title: "Investigating Biases Associated with Dietary Starch Incorporation into an Oral Biofilm Model"
+title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
 author:
@@
     number_sections: yes
+    keep_tex: true
   bookdown::html_document2:
@@
 abstract: |
-  Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+ Dental calculus, mineralised plaque, has proven to contain a wealth of information 
+ on the dietary habits of past populations. The extraction and identification of
+ starch granules contained within dental calculus has already provided valuable
+ dietary insights from a wide range of regions and time periods, and seem to preserve 
+ well in the archaeological record.
+ While the majority of studies have focused on microfossil extraction and
+ identification from archaeological remains, few studies have 
+ addressed the biases of starch retention in dental calculus, and have been
+ limited to *in vivo* studies on modern humans and non-human primates.
+ Here, we present a multispecies oral biofilm model, which allows experimental
+ research to be conducted on *in vitro* dental calculus in a controlled laboratory
+ setting. The biofilms were exposed treatment solutions with known quantities of
+ dietary starches (wheat and potato) during the 25-day growth period. After this,
+ the starch granules were extracted from the mature biofilm (dissolution in EDTA),
+ and counted. We 
+ show that the granule counts extracted from the model dental calculus represented 
+ a low proportion of the total counts (ranging from 0.064% to 0.161%) exposed to 
+ the biofilms throughout the experiment. Additionally, we found that the ratios
+ of granule sizes from the extracted starch granules differed from the original
+ treatment solutions, with large granules (>20 $\mu$m) consistently being
+ under-represented.
+ Additionally, the size ratios of granules within starch species shifted between 
+ the input  starch solutions and the granules extracted from the model calculus, 
+ with large granules  (>20 $\mu$m) consistently being underrepresented. 
+ We also found a correlation
+ between the absolute granule counts and dry-weight of the biofilm 
+ (*r* = 0.66, 90%CI[0.46,0.79]), as well
+ as between the concentration (count per mg) of granules and dry-weight 
+ (*r* = 0.3, 90%CI[0.06,0.51]).  
+ Our results reinforce previous *in vivo* studies suggesting that dental calculus 
+ presents a very small, and partly biased picture of the original dietary intake 
+ of starches, with an over-representation of plants producing granules smaller than
+ 20 $\mu$ in size. The experimental model presented here is well-suited to address
+ the need for further validation of methods and biases associated with dietary
+ research on dental calculus.
 keywords: 
@@
 ## end refs.bib ##
+
+## Word count ##
+wd_count <- wordcountaddin::word_count()
 ```
+
+Word count: `r wd_count`
 
@@
 characterisation (including pathogens) through the analysis of DNA and proteins 
-[@adlerSequencingAncientCalcified2013; @warinnerPathogensHostImmunity2014],
+[@adlerSequencingAncientCalcified2013; @warinnerPathogensHost2014],
 microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@@
 
-```{r protocol-fig, fig.cap="Overview of experiment protocol"}
-knitr::include_graphics(here("analysis/figures", "ExpL_protocol.png"))
+```{r protocol-fig, fig.cap="Overview of experiment protocol including the plate setup."}
+knitr::include_graphics(here("analysis/figures/protocol_overview.png"))
 ```
-
-<!-- modify sampling days from protocol picture to match amylase sampling -->
 
@@
 An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
-saliva samples collected from the plate wells on days 3, 5, 6, 9, and 12. 
+saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14. 
 Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
@@
 Extraction of starches from the calculus samples was performed by dissolving the
-calculus in -0.5 M<!-- what percentage???--> ethylenediaminetetraacetic acid (EDTA) 
+calculus in 0.5 $\tiny{M}$ ethylenediaminetetraacetic acid (EDTA) 
 [@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
@@
 
+```{r microscope-fig, fig.cap="Microscope image of potato starches on a microscope slide, as well as a bacterial community. Scale bar = 20 $\\mu$m."}
+knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
+```
+
 All samples yielded sufficient biofilm growth and starch incorporation to be 
-included in the analysis, resulting in a total of 48 biofilm samples (two plates of 24),
+included in the analysis (Figure \@ref(fig:microscope-fig)), resulting in a total of 48 biofilm samples (two plates of 24),
 45 of which were used for analysis (three samples were set aside for later
@@
   plt1_ph1_result[plt1_ph1_result < 0] <- 0
+
+amylase_results <- rbind(sal1_ph1, sal1_ph2, sal2_ph1, sal2_ph2, bmm1_ph1,
+                         bmm1_ph2, bmm2_ph1, bmm2_ph2)
+amylase_results[amylase_results < 0] <- NA
+positive_amylase <- na.omit(amylase_results)
 ```
+
+<!--
 
@@
 ```
+-->
 
-No appreciable amount of $\alpha$-amylase activity was detected in any of the artificial
+No $\alpha$-amylase activity was detected in any of the artificial
 saliva samples from any of the days that were sampled. Only positive controls
-contained amylase activity that could be detected in the assay (Table
-\@ref(tab:amylase-activity-tbl)). The results from the additional plates and
-photometric readings can be found in the 
-[Supplementary Materials](../supplementary-materials/supp-mat.Rmd). The results
-are not comparable to other studies presenting $\alpha$-amylase activity levels
-in humans; however, they are sufficient to show that there is no activity in the
-system.
+contained amylase activity that could be detected in the assay, ranging from
+`r paste(signif(range(positive_amylase),3), collapse = " to ")` U/mL enzyme 
+(Supplementary Material).
+The results are not comparable to other studies presenting $\alpha$-amylase activity
+levels in humans, as the unit definition may differ; 
+however, they are sufficient to show that there is no activity in the system.
 
@@
 
-<!-- separated treatments -->
-
 <!-- mixed treatments -->
@@
 
+```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."}
+#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
+(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+```
+
+<!--
 ```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring)."}
@@
 ```
-
-<!-- Combine to a single figure? -->
+ -->
 
@@
 increased in potato treatments 
-(Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).
-
-<!-- Chi square test for size counts? -->
+(Figure \@ref(fig:ratio-plots)).
 
@@
 
+We have also shown that the size of the starch granules influences the likelihood
+of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
+diameter were underrepresented in the calculus samples compared to the original 
+starch solutions, an effect that was consistent across all three treatments, 
+while medium granules (10--20 $\mu$m) were often over-represented
+(Table \@ref(tab:sample-prop-tbl), 
+and 
+Figure \@ref(fig:ratio-plots)).
+Large potato granules were most affected, and can reach up to 100 $\mu$m
+in diameter, whereas wheat generally only reach up to 35 $\mu$m 
+[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
+This may also suggest that granule morphology plays a role. Large wheat granules 
+are lenticular and have a larger surface area compared 
+to volume, whereas large potato granules are ovoid and have a larger volume
+compared to surface area 
+[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
+Another potentially important factor 
+from our results is the size of the calculus deposit. We found a 
+`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
+starch granules (Figure \@ref(fig:cor-plot));
+a result that contradicts findings from archaeological contexts
+[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
+When the concentration of starch granules
+per mg calculus is considered, the correlation is weaker, but still present
+(Figure \@ref(fig:cor-plot2));
+while the larger deposits contain a higher absolute count, our findings also suggest
+that they contain a slightly higher concentration of starches. 
+Further research is needed to determine why this differs from previous archaeological
+findings.
 <!--
@@
 representative of individual diets than starch granules.
-
-<!-- secondary findings -->
-We have also shown that the size of the starch granules influences the likelihood
-of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
-diameter were underrepresented in the calculus samples compared to the original 
-starch solutions, an effect that was consistent across all three treatments, 
-while medium granules (10 -- 20 $\mu$m) were often over-represented
-(Table \@ref(tab:sample-prop-tbl), 
-and 
-Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).
-This was especially true for the potato starches, which can reach up to 100 $\mu$m
-in diameter, whereas wheat generally only reach up to 35 $\mu$m 
-[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004].
-The size bias is consistent with the findings by Power and colleagues
+The size bias is also consistent with the findings by Power and colleagues
 [-@powerChimpCalculus2015],
-which found that plants producing starches 10--20 $\mu$m in
+who found that plants producing starches 10--20 $\mu$m in
 size were over-represented; although, the representation of granules larger than
 20 $\mu$m is unclear.
-Given the differences in proportional incorporation between wheat and potato,
-especially for small and medium granules, it is also possible that the morphology
-of granules plays a role. However, the proportional count differences between large 
-wheat and potato granules were less pronounced.
-
-<!-- Since those results were obtained -->
-<!-- on a modern population of chimps that were only buried for a short period of time,  -->
-<!-- degradation was likely limited [@powerChimpCalculus2015]. -->
-
-The taphonomy of starch granules has been addressed previously, and depends on 
-burial environment (pH, temperature, moisture content, microorganisms), as well 
-as processing of the granules prior to deposition 
-[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004]. 
-The potential for detrimental diagenetic effects on starch granules increases as 
-the size and amylose content decreases
-[@francoStarchDegradation1992; @haslamDecompositionStarch2004].
 
@@
  Binding to select species in a biofilm could explain clustering
- Colonisation of bacteria onto starch @nagaraBacteriaNutrients2017-->
+ Colonisation of bacteria onto starch @nagaraBacteriaNutrients2017 (although not
+ clear if the bacteria in our model are even 'consuming' starch-->
 
@@
 
-Another potentially important factor 
-from our results is the size of the calculus deposit, which seems to influence the 
-quantity of starches extracted from the calculus, as we found a strong positive 
-correlation between biofilm size and retained starch granules (Figure \@ref(fig:cor-plot));
-a result that contradicts findings from archaeological contexts
-[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014]. 
-When the concentration of starch granules
-per mg calculus is considered, the correlation is weaker, but still present
-(Figure \@ref(fig:cor-plot2));
-while the larger deposits contain a higher absolute count, our findings also suggest
-that they contain a slightly higher concentration of starches. 
-The lack of a correlation
-in archaeological contexts could be due to diagenetic effects. Alternately, the absence of 
-amylase activity in our study could be impacting our interpretations.
-
-<!-- How are starch granules trapped in dental calculus? Does anyone actually know? - adhesion to bacteria
-- stuck in cracks and crevices
-- covered by new layer of plaque
-
+<!-- How are starch granules trapped in dental calculus? Does anyone actually know? 
 -->
@@
 the mouth can become trapped in the plaque/calculus, and can be recovered from
-archaeological samples of significant age 
+archaeological samples of considerable age 
 [@henryNeanderthalCalculus2014; @buckleyDentalCalculusCooking2014; @wuDietEarliest2021]. 
@@
 only briefly present in the oral cavity. Other granules that are broken off
-during mastication may be retained in the dentition.
-
-<!-- initial incorporation of starch granules? 
-cracks and crevices? attachment to bacteria? -->
-
+during mastication may be retained in the dentition through attachment to 
+tooth surfaces (including plaque and dental calulus) and mucous membranes 
+[@kashketFoodRetention1991; @doddsCarbohydrateRetention1988].
+Bacteria also have the ability to adhere to starch granules 
+[@toppingResistantStarch2003],
+which would allow starches to attach to bacterial communities within the biofilm.
 These granules are then
@@
 by $\alpha$-amylase [@kashketFoodParticles1996].
-Starch granules that are trapped in crevices and channels on the surface of the
-mineralised plaque are (at least to some extent) protected from salivary
-clearance and mechanical cleaning actions of the tongue and lips, especially once
-a new layer of plaque has covered the surface of the calculus. 
+The susceptibility of granules to hydrolysis depends on the crystallinity and size
+of the starch granule, as well as the mode of processing. Smaller and pre-processed
+(e.g. cooking) starch granules are more susceptible to enzymatic degradation, 
+while dehydrated starches will have a reduced susceptibility
+[@lingstromStarchyFood1994; @francoStarchDegradation1992; @haslamDecompositionStarch2004; 
+@bjorckStarchProcessing1984; @henryCookingStarch2009].
+Cracks on the surface, as well as unmineralised islands and channels exist within 
+dental calculus and may also be able to contain starch granules 
+[@powerSEMCalculus2014; @tanCalculusUltrastructure2004; @charlierSEMCalculus2010].
+Starch granules that are trapped in these pockets are (at least to some extent) 
+protected from aforementioned clearance mechanisms, especially once
+a new layer of plaque has covered the surface of the plaque/calculus. 
 The size bias against large granules (>20 $\mu$m) from both wheat and potato 
@@
 distribution across the surface of the dental calculus.  
-
-<!--Bacteria also have the ability to adhere to starch [@toppingResistantStarch2003],
-which would allow starches to attach to bacterial communities within the biofilm.
-Following deposition and/or mineralisation events, the granules that are able to 
-resist hydrolysis and 
-mechanical removal will then be trapped in pockets within the mineral matrix, and
-will temporarily be protected from shearing action and salivary 
-clearance. -->
-<!-- Unmineralised lacunae and channels within the  
- calculus matrix have been shown to contain viable 
- bacteria, and may also be large enough to contain starch granules. These can 
- range from the width of a single cell, to multiple-cell 
- width [@tanCalculusUltrastructure2004], of which the latter could feasibly contain 
-   starch granules.  -->
-If this is indeed one of the incorporation mechanisms of starch granules, then the 
-limit of incorporated starches is set by the number and size of channels. Since 
-this is likely controlled by the size of the calculus deposit, then the number of 
-starches will increase as the size of the biofilm increases, which is consistent 
-with our results 
-(Figures \@ref(fig:cor-plot) and \@ref(fig:cor-plot2)).
-
-While potentially protected against clearance, granules trapped 
+Granules trapped 
 in plaque/calculus may still be susceptible to hydrolysis, as $\alpha$-amylase has
@@
 a portion of its hydrolytic activity
-[@scannapiecoSalivaryAmylase1993; @nikitkovaStarchBiofilms2013; @tanStudyBacterialViability2004; @tanCalculusUltrastructure2004].
-The susceptibility of granules to hydrolysis depends on the crystallinity and size
-of the starch granule, as well as the mode of processing. As mentioned previously,
-smaller starch granules are more susceptible to enzymatic degradation, as are 
-pre-processed starches (e.g. by cooking), while dehydrated starches will have a
-reduced susceptibility
-[@lingstromStarchyFood1994; @francoStarchDegradation1992; @haslamDecompositionStarch2004; 
-@bjorckStarchProcessing1984; @henryCookingStarch2009]. Given the lack of $\alpha$-amylase 
-activity detected in our model, hydrolysis is not the cause of the low
-incorporation rate. More details on the absence of $\alpha$-amylase 
-activity is discussed elsewhere (Bartholdy et al. in prep.).
-What we can say, is that the dietary picture we obtain from starch granules extracted 
-from dental calculus reflects a number of individual, dietary, and environmental 
-(both pre- and post-mortem environments) factors, and that this will likely be a
-somewhat random snapshot given the irregularity of plaque mineralisation within
-and between individuals [@jinSupragingivalCalculus2002; @jepsenCalculusRemoval2011]. 
-Although, certain factors may increase the likelihood of incorporation, such 
-as abundance, size, and morphology of granules.
+[@scannapiecoSalivaryAmylase1993; @nikitkovaStarchBiofilms2013; @tanBacterialViability2004; @tanCalculusUltrastructure2004].
+After the death of an individual, starches within dental calculus are susceptible 
+to further degradation by post-depositional processes, depending on burial environment 
+(pH, temperature, moisture content, microorganisms)  
+[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004; @francoStarchDegradation1992].
 
@@
 where $\alpha$-amylase can be added to the model
-in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. <!-- foreshadowing...? -->
-While we are able to show the differing 
-incorporation caused by absolute counts available, this
+in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. 
+We are able to show how absolute counts in the treatments cause a difference in
+incorporation; however, this
 was merely a side-effect of the difference in the number of granules in potato and 
@@
 will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
-Based on counts obtained by Leonard
+Based on the comparatively low counts obtained by Leonard
 and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
@@
 <!-- How this engineering will facilitate future research -->
-The oral biofilm model described in this study provides a method
-to explore the incorporation and extraction of dietary compounds from dental calculus 
-in a controlled laboratory setting, as many of the variables can be adjusted. 
-The addition of known starch species in
-separate samples also removes the identification bias against the smaller
-undiagnostic starches.<!-- rephrase -->
-It can also address the call for more baseline testing of biases associated with 
+
+The data from this study are not sufficient to confirm or deny any proposed
+methods of starch incorporation; however, while we are unable to address the 
+mechanism(s) of starch incorporation with the data obtained in this study, 
+the dental calculus model presented in this study is uniquely suited to explore 
+these questions and may improve interpretations of dietary practices in past populations.
+It can address the call for more baseline testing of biases associated with 
 dietary research conducted on dental calculus 
 [@lemoyneCalculusPretreatments2021].
-*In vivo* studies on humans and non-human primates will be limited by the variability 
-occurring between individuals, and difficulties in accurately documenting 
-the entirety of starch granules entering the oral cavity at any given moment,
-especially accidental intake. 
-Our experimental setup allows us a
-higher degree of control over the factors that influence starch incorporation, such as
-dietary intake, processing of starches leading to differential survivability, 
-and inter- and intra-individual variation in plaque accumulation and 
-mineralisation.
+Our high-throughput experimental setup allows us a
+higher degree of control over the factors that influence starch incorporation and
+retention, such as dietary intake, differential survivability of starches, 
+and inter- and intra-individual variation in plaque accumulation and mineralisation.
 The latter is especially difficult to control *in vivo* as it is influenced by 
-numerous factors including genetics, diet, salivary flow, tooth position and morphology 
+numerous factors including genetics, diet, salivary flow, and tooth position and
+morphology 
 [@jepsenCalculusRemoval2011; @simon-soroOralGeography2013; @proctorSpatialGradient2018; @fagernasMicrobialBiogeography2021; @haffajeeBiofilmPosition2009], 
-as well as evolutionary differences through time [@yatesOralMicrobiome2021].
+as well as evolutionary differences [@yatesOralMicrobiome2021].
 It can also facilitate training of students and researchers on methods of 
@@
 identification, where it can replace the use of finite archaeological resources.
+
 <!-- we still need to replicate results and do more science... but pretty cool, right? -->
@@
 (10--20 $\mu$m)
-starches an increased rate, and small (<10 $\mu$m) granules remained somewhat constant.
-The size of calculus deposit also seems to influence the capacity of 
+starches an increased rate, and small (<10 $\mu$m) granules remained somewhat
+constant. The size of calculus deposit also seems to influence the capacity of 
 granule incorporation; as the size of the deposit increases, so does the
 absolute count of incorporated granules.  
-<!--The data from this study are not sufficient to confirm or deny any proposed
-methods of starch incorporation; however, while we are unable to address the 
-mechanism(s) of starch incorporation with the data obtained in this study, 
-this is something that can potentially be addressed with the calculus model 
-presented in this paper.-->
-
-<!-- Forward-looking statements about the field as a whole -->
 While we have shown multiple factors that influence the likelihood 
@@
 mechanisms of intra-oral starch incorporation and retention in dental calculus.
-The dental calculus model presented in this study is uniquely suited to explore 
-these questions and may improve interpretations of dietary practices in past populations.
+The oral biofilm model described in this study provides a method
+to explore the incorporation and extraction of dietary compounds from dental calculus 
+in a controlled laboratory setting, and unearth the potential biases associated 
+with dietary research conducted on archaeological dental calculus.
 
@@
 documents. You're awesome! -->
-
 This research has received funding from the European Research Council under the 
 European Union’s Horizon 2020 research and innovation program, grant agreement 
-number STG–677576 (“HARVEST”).
+number STG–677576 (“HARVEST”). We also thank everyone in the general vicinity
+of the lab for enduring the smell of bacterial accumulation.
+We did NOT make use of Sci-Hub to access articles stuck behind a paywall, 
+and will, therefore, NOT acknowledge the use of Sci-Hub in this study.
 

### 2021-10-08 analysis/paper/first_draft_rev-1/paper.tex
incorporate AGH suggestions

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model},
  pdfauthor={Bjørn Peare Bartholdy; Amanda G. Henry},
  pdfkeywords={oral biofilm, starch, alpha-amylase, dental calculus},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{fancyhdr}
\fancyhf{}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0pt}
\fancyheadoffset{0pt}
\rhead{\scshape A preprint - \today}
\cfoot{\thepage}

%Handling Keywords
\def\keywordname{{\bfseries \emph Keywords}}%
\def\keywords#1{\par\addvspace\medskipamount{\rightskip=0pt plus1cm
\def\and{\ifhmode\unskip\nobreak\fi\ $\cdot$
}\noindent\keywordname\enspace\ignorespaces#1\par}}

% create title
% \providecommand{\maketitle}{}
% \renewcommand{\maketitle}{%
%   \par
%   \begingroup
%     \renewcommand{\thefootnote}{\fnsymbol{footnote}}
%     % for perfect author name centering
%     \renewcommand{\@makefnmark}{\hbox to \z@{$^{\@thefnmark}$\hss}}
%     % The footnote-mark was overlapping the footnote-text,
%     % added the following to fix this problem               (MK)
%     \long\def\@makefntext##1{%
%       \parindent 1em\noindent
%       \hbox to 1.8em{\hss $\m@th ^{\@thefnmark}$}##1
%     }
%     \thispagestyle{empty}
%     \@maketitle
%     \@thanks
%     %\@notice
%   \endgroup
%   \let\maketitle\relax
%   \let\thanks\relax
% }

% rules for title box at top of first page
% \newcommand{\@toptitlebar}{
%   \hrule height 2\p@
%   \vskip 0.25in
%   \vskip -\parskip%
% }
% \newcommand{\@bottomtitlebar}{
%   \vskip 0.29in
%   \vskip -\parskip
%   \hrule height 2\p@
%   \vskip 0.09in%
% }

% create title (includes both anonymized and non-anonymized versions)
% \providecommand{\@maketitle}{}
% \renewcommand{\@maketitle}{%
%   \vbox{%
%     \hsize\textwidth
%     \linewidth\hsize
%     \vskip 0.1in
%     \@toptitlebar
%     \centering
%     {\LARGE\sc \@title\par}
%     \@bottomtitlebar
%     \textsc{A Preprint}\\
%     \vskip 0.1in
%     \def\And{%
%       \end{tabular}\hfil\linebreak[0]\hfil%
%       \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\ignorespaces%
%     }
%     \def\AND{%
%       \end{tabular}\hfil\linebreak[4]\hfil%
%       \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\ignorespaces%
%     }
%     \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\@author\end{tabular}%
%   \vskip 0.4in \@minus 0.1in \center{\today}   \vskip 0.2in
%   }
% }

% add conference notice to bottom of first page
% \newcommand{\ftype@noticebox}{8}
% \newcommand{\@notice}{%
%   % give a bit of extra room back to authors on first page
%   \enlargethispage{2\baselineskip}%
%   \@float{noticebox}[b]%
%     \footnotesize\@noticestring%
%   \end@float%
% }

% abstract styling
\renewenvironment{abstract}
{
  \centerline
  {\large \bfseries \scshape Abstract}
  \begin{quote}
}
{
  \end{quote}
}

\usepackage{lineno}
\linenumbers
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model}
\author{Bjørn Peare Bartholdy \and Amanda G. Henry}
\date{2021-09-30}

\begin{document}
\maketitle
\begin{abstract}
Dental calculus, mineralised plaque, has proven to contain a wealth of information
on the dietary habits of past populations. The extraction and identification of
starch granules contained within dental calculus has already provided valuable
dietary insights from a wide range of regions and time periods, and seem to preserve
well in the archaeological record.
While the majority of studies have focused on microfossil extraction and
identification from archaeological remains, few studies have
addressed the biases of starch retention in dental calculus, and have been
limited to \emph{in vivo} studies on modern humans and non-human primates.
Here, we present a multispecies oral biofilm model, which allows experimental
research to be conducted on \emph{in vitro} dental calculus in a controlled laboratory
setting. The biofilms were exposed treatment solutions with known quantities of
dietary starches (wheat and potato) during the 25-day growth period. After this,
the starch granules were extracted from the mature biofilm (dissolution in EDTA),
and counted. We
show that the granule counts extracted from the model dental calculus represented
a low proportion of the total counts (ranging from 0.064\% to 0.161\%) exposed to
the biofilms throughout the experiment. Additionally, we found that the ratios
of granule sizes from the extracted starch granules differed from the original
treatment solutions, with large granules (\textgreater20 \(\mu\)m) consistently being
under-represented.
Additionally, the size ratios of granules within starch species shifted between
the input starch solutions and the granules extracted from the model calculus,
with large granules (\textgreater20 \(\mu\)m) consistently being underrepresented.
We also found a correlation
between the absolute granule counts and dry-weight of the biofilm
(\emph{r} = 0.66, 90\%CI{[}0.46,0.79{]}), as well
as between the concentration (count per mg) of granules and dry-weight
(\emph{r} = 0.3, 90\%CI{[}0.06,0.51{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that dental calculus
presents a very small, and partly biased picture of the original dietary intake
of starches, with an over-representation of plants producing granules smaller than
20 \(\mu\) in size. The experimental model presented here is well-suited to address
the need for further validation of methods and biases associated with dietary
research on dental calculus.
\end{abstract}

% \providecommand{\@maketitle}{}
% \renewcommand{\@maketitle}{%
%   \vbox{%
%     \hsize\textwidth
%     \linewidth\hsize
%     \vskip 0.1in
%     \@toptitlebar
%     \centering
%     {\LARGE\sc \@title\par}
%     \@bottomtitlebar
%     \textsc{A Preprint}\\
%     \vskip 0.1in
%     \def\And{%
%       \end{tabular}\hfil\linebreak[0]\hfil%
%       \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\ignorespaces%
%     }
%     \def\AND{%
%       \end{tabular}\hfil\linebreak[4]\hfil%
%       \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\ignorespaces%
%     }
%     \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\@author\end{tabular}%
%   \vskip 0.4in \@minus 0.1in \center{\today}   \vskip 0.2in
%   }
% }
%
% \providecommand{\maketitle}{}
% \renewcommand{\maketitle}{%
%   \par
%   \begingroup
%     \renewcommand{\thefootnote}{\fnsymbol{footnote}}
%     % for perfect author name centering
%     \renewcommand{\@makefnmark}{\hbox to \z@{$^{\@thefnmark}$\hss}}
%     % The footnote-mark was overlapping the footnote-text,
%     % added the following to fix this problem               (MK)
%     \long\def\@makefntext##1{%
%       \parindent 1em\noindent
%       \hbox to 1.8em{\hss $\m@th ^{\@thefnmark}$}##1
%     }
%     \thispagestyle{empty}
%     \@maketitle
%     \@thanks
%     %\@notice
%   \endgroup
%   \let\maketitle\relax
%   \let\thanks\relax
% }

\setstretch{2.0}
Word count: 4634

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Dental calculus, mineralised dental plaque, has proven to contain a wealth of
dietary information in the form of plant microfossils
(Hardy et al., 2009; Henry \& Piperno, 2008),
proteins (Hendy et al., 2018; Warinner, Hendy, et al., 2014),
and other organic residues (Buckley et al., 2014).
This dietary information can be preserved within calculus over many millenia,
providing a unique window into the food-related behaviours of past populations
(Henry \& Piperno, 2008; Jovanović et al., 2021; Tao et al., 2020)
and extinct species (Hardy et al., 2012; Henry et al., 2014).

Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first
noted by Armitage (1975),
and later by Middleton and Rovner (1994),
and Fox and colleagues (1996). Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis (1997).\\
In more recent years, the study of dental calculus has increased exponentially,
and the wealth of information contained within the mineralised matrix has been,
to a larger extent, acknowledged. The applications of dental calculus span a wide
variety of archaeological research areas, such as oral microbiome
characterisation (including pathogens) through the analysis of DNA and proteins
(Adler et al., 2013; Warinner, Rodrigues, et al., 2014),
microbotanical remains (Hardy et al., 2009; Henry \& Piperno, 2008; Mickleburgh \& Pagán-Jiménez, 2012),
other organic residues and proteins from dietary compounds
(Buckley et al., 2014; Hendy et al., 2018),
and nicotine-use (Eerkens et al., 2018). Especially the extraction
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges
(Henry et al., 2014; Jovanović et al., 2021; Piperno \& Dillehay, 2008; Tao et al., 2020).

Despite this, our knowledge of dental calculus and the incorporation pathways of
the various markers is limited (Radini et al., 2017), as is our
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested
by Tromp and colleagues (2017), specifically regarding
decalcification using HCl or EDTA.
The authors found significantly more starches with the EDTA extraction method
than the HCl extraction method; however, as noted by the authors, comparisons
involving archaeological calculus are problematic due to variability between and
within individuals.
Studies conducted on modern humans (Leonard et al., 2015)
and non-human primates (Power et al., 2015),
have explored the how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited,
despite meticulous documentation and observation, due to unknown variables and
uncertainty involved when studying living organisms. Dental calculus is a complex
oral biofilm with a multifactorial aetiology and variable formation rates both
within and between individuals (Haffajee et al., 2009; Jepsen et al., 2011), contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral \(\alpha\)-amylase differs both between and
within individuals (Froehlich et al., 1987; Nater et al., 2005),
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being
explored (García-Granero, 2020), and you have a highly unpredictable process.

In this exploratory study, we use an oral biofilm model to investigate the
retention of starch granules within dental calculus in a controlled laboratory
setting, allowing us full control over dietary input. Our main questions concern
the representation of granules extracted
from the calculus compared to the actual intake. How much of the original diet is
incorporated into the calculus, and how much is recovered?
Is there differential loss of information from specific dietary markers that affects
the obtained dietary information, and how does this affect the representation of
diet from extracted microremains?\\
We find that, despite the absence of \(\alpha\)-amylase in
the model, a limited proportion of the starch input is actually
retained in the calculus, as well as a shift in the size ratios of individual
starch granules
that are incorporated into the calculus. We also show that the number of starch
granules that are incorporated increases as the size of the calculus deposit
increases.

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

\hypertarget{biofilm-formation}{%
\subsection{Biofilm formation}\label{biofilm-formation}}

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues (1991) and
Shellis (1978). The setup comprises a polypropylene
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which
is autoclaved at 120°C, 1 bar overpressure, for 20 mins.

The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
described by Sissons and colleagues (1991).
It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778),
5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl\textsubscript{2},
5.2 mmol/l Na\textsubscript{2}HPO\textsubscript{4} (Sissons et al., 1991), 6.4 mmol/l NaHCO\textsubscript{3}
(Shellis, 1978), 2.5 mg/l haemin. This is subsequently
adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120°C,
1 bar overpressure), and supplemented with 5.8 \(\mu\)mol/l menadione, 5 mmol/l urea,
and 1 mmol/l arginine (Sissons et al., 1991).

\begin{figure}
\centering
\includegraphics{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview.png}
\caption{\label{fig:protocol-fig}Overview of experiment protocol including the plate setup.}
\end{figure}

Fresh whole saliva (WS) for inoculation was provided by a 31-year-old male donor
with no history of caries, who abstained from oral hygiene for 24 hours, and no
food was consumed two hours prior to donation. No antibiotics were taken up to six
months prior to donation.
The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15\% active chlorine)
nylon cloth to remove particulates.
Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised
20\% (v/v) glycerine for four hours at 36°C, to allow attachment of the
salivary pellicle and plaque-forming bacteria. After initial inoculation, the
substrata were transferred to a new plate containing 1 ml/well AS and incubated
in a shaking incubator (Infors HT Ecotron) at 36°C, 30 rpm.
The inoculation process was repeated on days 3 and 5.
AS was partially refreshed once per day and fully refreshed every three days,
throughout the experiment, by transferring the substrata to a new plate containing
stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing
5\% (w/v) sucrose, for six minutes twice daily, except on inoculation days
(days 0, 3, and 5), where the samples only received one sucrose treatment after
inoculation.

Starch treatments were initiated on day 9 to avoid starch granule counts being
affected by \(\alpha\)-amylase hydrolysis from saliva inoculation days.
An \(\alpha\)-amylase (EC 3.2.1.1) activity
assay was conducted to confirm that no amylase was present in the system before
starch treatments started. Starch treatments replaced sucrose treatments, occurring twice per day
for six minutes. The starch treatments involved transferring the substrata to a
new plate containing a 0.25\% (w/v) starch from potato (Roth 9441.1) solution, a 0.25\% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5\% (w/v) mixture of equal
concentrations (w/v) wheat and potato. All starch treatments were created in dH\textsubscript{2}O
with 5\% (w/v) sucrose. Before transferring biofilm samples to the starch treatment
plate, the plates were agitated to keep the starches in suspension in the
solutions. During treatments, the rpm was increased to 60 to facilitate contact
between starch granules and biofilms.

After 15 days, mineralisation was encouraged with a
calcium phosphate monofluorophosphate urea (CPMU) solution containing
20 mmol/l CaCl\textsubscript{2}, 12 mmol/l NaH\textsubscript{2}PO\textsubscript{4}, 5 mmol/l Na\textsubscript{2}PO\textsubscript{3}F, 500 mmol/l Urea,
and (0.04 g/l MgCl)
(Pearce \& Sissons, 1987; Sissons et al., 1991).
The substrata were submerged in 1 ml/well CPMU for six minutes, five times
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated
for 10 days until the end of the experiment on day 24
(see \ref{fig:protocol-fig} for an overview of the protocol). More detailed protocols
are available on protocols.io.

All laboratory work was conducted in sterile conditions under a laminar flow hood
to prevent starch and bacterial contamination. Control samples that only received
sucrose as a treatment, were included to detect starch contamination from the
environment, or cross-contamination from other wells in the same plate.

\hypertarget{amylase-activity-detection}{%
\subsection{Amylase activity detection}\label{amylase-activity-detection}}

An \(\alpha\)-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14.
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4°C until the assay was conducted on day 18.
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing
H\textsubscript{2}O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a
Multiskan FC Microplate Photometer (Thermo Scientific 51119000).
The protocol is a slightly modified version of an Enzymatic Assay of \(\alpha\)-Amylase
(\url{https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase}) (Bernfeld, 1955), which measures the amount of
maltose released from starch by \(\alpha\)-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.
For a more detailed protocol, see

\hypertarget{treatment-solutions}{%
\subsection{Treatment solutions}\label{treatment-solutions}}

A 1 ml aliquot of each starch solution was taken, from which 10 \(\mu\)l was mounted
on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
(Zeiss Axioscope A1). For wheat and mixed treatment samples, we counted three
slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts
were extrapolated (see Supplementary material for more details). For potato
treatment samples, the whole slide was counted.

\hypertarget{extraction-method}{%
\subsection{Extraction method}\label{extraction-method}}

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 \(\tiny{M}\) ethylenediaminetetraacetic acid (EDTA)
(Le Moyne \& Crowther, 2021; Modi et al., 2020; Tromp et al., 2017),
and vortexing for 3 days until the sample was completely dissolved.
Twenty \(\mu\)l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using
the pipette to ensure that the counted transects were representative of the
whole slide. Wheat and mix counts were extrapolated as described above.
Both wheat and potato granules were divided into three size categories:
small (\textless10 \(\mu\)m), medium (10 -- 20 \(\mu\)m), and large (\textgreater20 \(\mu\)m).

\hypertarget{statistical-analysis}{%
\subsection{Statistical analysis}\label{statistical-analysis}}

Statistical analysis was conducted in R version 4.1.1 (2021-08-10) (R Core Team, 2020) and
the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
here (Müller, 2020), and patchwork (Pedersen, 2020).
A one-way ANOVA with sample weight as the dependent variable (DV) and treatment
as the grouping variable (GV) was conducted to explore the effect of the different
starch treatments on biofilm growth.

To calculate size proportions, mean counts for each treatment were taken across
both experimental plates for each treatment, resulting in a mean count for each
granule size category within each treatment.

Pearson's r was conducted on sample weight and total starch count, as well as sample
weight and starch count per mg calculus. The total count for each sample within a
treatment was standardised by z-score to account for the differences in magnitude
between the potato and wheat counts.
This was applied to total biofilm weight and starch count per mg
calculus (also z-score standardised) to account for differences in starch
concentration in the calculus (as per Wesolowski et al., 2010).

All scripts and data used in the analysis are available on OSF
(\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}).
All protocols are available on protocols.io.

\hypertarget{results}{%
\section{Results}\label{results}}

\begin{figure}
\centering
\includegraphics{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar.jpg}
\caption{\label{fig:microscope-fig}Microscope image of potato starches on a microscope slide, as well as a bacterial community. Scale bar = 20 \(\mu\)m.}
\end{figure}

All samples yielded sufficient biofilm growth and starch incorporation to be
included in the analysis (Figure \ref{fig:microscope-fig}), resulting in a total of 48 biofilm samples (two plates of 24),
45 of which were used for analysis (three samples were set aside for later
analysis).
Most control samples contained no starch granules, while some contained negligible
quantities (see Supplementary Material).

\hypertarget{no-amylase-activity-detected-in-the-model}{%
\subsection{No amylase activity detected in the model}\label{no-amylase-activity-detected-in-the-model}}

No \(\alpha\)-amylase activity was detected in any of the artificial
saliva samples from any of the days that were sampled. Only positive controls
contained amylase activity that could be detected in the assay, ranging from
3.4 to 10.3 U/mL enzyme
(Supplementary Material).
The results are not comparable to other studies presenting \(\alpha\)-amylase activity
levels in humans, as the unit definition may differ;
however, they are sufficient to show that there is no activity in the system.

\hypertarget{treatment-type-had-no-effect-on-biofilm-growth}{%
\subsection{Treatment type had no effect on biofilm growth}\label{treatment-type-had-no-effect-on-biofilm-growth}}

\begin{table}

\caption{\label{tab:anova-tbl}Summary statistics for biofilm dry-weights (in mg) by treatment.}
\centering
\begin{tabular}[t]{l|r|r|r|r}
\hline
Treatment & Mean & SD & Min & Max\\
\hline
control & 5.44 & 2.45 & 1.67 & 11.20\\
\hline
mix & 4.28 & 1.95 & 1.50 & 8.44\\
\hline
potato & 6.25 & 2.07 & 2.54 & 8.92\\
\hline
wheat & 5.53 & 3.45 & 0.56 & 9.80\\
\hline
\end{tabular}
\end{table}

A one-way ANOVA suggests that the type of starch used during the biofilm
growth period had a minimal effect on the growth of the biofilm
(expressed as total dry weight of the sample), F(3, 43)
= 1.16, p = 0.335.
A summary of sample weights is available in Table \ref{tab:anova-tbl}.

\hypertarget{starch-counts}{%
\subsection{Starch counts}\label{starch-counts}}

It was not possible to differentiate between potato and
wheat starches smaller than ca. 10 \(\mu\)m. These were counted as wheat,
as we assumed that the majority of the small granules were wheat. This is a
reasonable assumption to make based on the small granule counts from the separate
treatment solutions, where the wheat granules made up
99.2\%
of all small granule counts in the separated solutions.

\begin{table}

\caption{\label{tab:solution-count-tbl}Mean starch counts from solutions, including the
             proportional makeup of the different sizes of granules.}
\centering
\begin{tabular}[t]{l|l|l|l|l|l}
\hline
Solution & Starch & Small (\%) & Medium (\%) & Large (\%) & Total (\%)\\
\hline
mix & potato &  & 1051733 (53.1\%) & 928000 (46.9\%) & 1979733 (100.0\%)\\
\hline
mix & wheat & 18838400 (69.7\%) & 6403200 (23.7\%) & 1794133 (6.6\%) & 27035733 (100.0\%)\\
\hline
mix & both & 18838400 (64.9\%) & 7454933 (25.7\%) & 2722133 (9.4\%) & 29015467 (100.0\%)\\
\hline
potato & potato & 123733 (4.1\%) & 1337867 (44.4\%) & 1554400 (51.5\%) & 3016000 (100.0\%)\\
\hline
wheat & wheat & 16139467 (63.5\%) & 6434133 (25.3\%) & 2830400 (11.1\%) & 25404000 (100.0\%)\\
\hline
\end{tabular}
\end{table}

\begin{table}

\caption{\label{tab:sample-count-tbl}Mean starch counts extracted from samples with standard 
             deviation (SD), including the proportion of granule sizes of the total count.}
\centering
\begin{tabular}[t]{l|l|l|r|l|r|l|r|l|r}
\hline
Treatment & Starch & Small (\%) & SD & Medium (\%) & SD & Large (\%) & SD & Total (\%) & SD\\
\hline
mix & potato &  &  & 1959 (79.6\%) & 1800 & 501 (20.40\%) & 446 & 2460 (100\%) & 2190\\
\hline
mix & wheat & 9515 (54.60\%) & 8860 & 6522 (37.4\%) & 6030 & 1381 (7.93\%) & 1200 & 17417 (100\%) & 15900\\
\hline
mix & both & 9515 (47.90\%) & 8860 & 8480 (42.7\%) & 7650 & 1882 (9.47\%) & 1600 & 19877 (100\%) & 17800\\
\hline
potato & potato & 351 (7.24\%) & 297 & 3565 (73.6\%) & 2400 & 930 (19.20\%) & 929 & 4846 (100\%) & 3320\\
\hline
wheat & wheat & 15235 (55.00\%) & 11900 & 12148 (43.9\%) & 11100 & 1953 (7.06\%) & 2020 & 27680 (100\%) & 23600\\
\hline
\end{tabular}
\end{table}

The separate wheat and potato solutions were made with a 0.25\% (w/v) starch
concentration, while the mixed-starch solution was made with 0.25\% (w/v) of each
starch, with a total concentration of 0.50\% (w/v).
The mixed treatment had the highest absolute count of starch granules in solution (mean = \ensuremath{1.98\times 10^{6}}, \ensuremath{2.7\times 10^{7}}),
while the biofilms exposed to the wheat solution preserved the greatest number of
granules
(mean = \ensuremath{2.77\times 10^{4}}).
The potato treatment had the lowest absolute counts in both the solution
(\ensuremath{3.02\times 10^{6}})
and in the biofilm samples
(4850)
(Tables \ref{tab:solution-count-tbl} and \ref{tab:sample-count-tbl}).

\hypertarget{proportion-of-available-starches-incorporated-in-samples}{%
\subsubsection{Proportion of available starches incorporated in samples}\label{proportion-of-available-starches-incorporated-in-samples}}

\begin{table}

\caption{\label{tab:sample-prop-tbl}The mean percentage of starches from the solutions that 
              were incorported into the samples.}
\centering
\begin{tabular}[t]{l|l|l|l|l|l}
\hline
Treatment & Starch & Small & Medium & Large & Total\\
\hline
mix & potato &  & 0.186\% & 0.054\% & 0.124\%\\
\hline
mix & wheat & 0.051\% & 0.102\% & 0.077\% & 0.064\%\\
\hline
mix & both & 0.051\% & 0.114\% & 0.069\% & 0.069\%\\
\hline
potato & potato & 0.284\% & 0.266\% & 0.060\% & 0.161\%\\
\hline
wheat & wheat & 0.094\% & 0.189\% & 0.069\% & 0.109\%\\
\hline
\end{tabular}
\end{table}

The proportion of total starches from the solutions that were incorporated
into the samples ranged from
0.064\% to 0.161\%,
with potato granules being more readily incorporated than wheat in both the
separated- and mixed-treatment samples
(Table \ref{tab:sample-prop-tbl}).
There is an inverse relationship between the absolute starch count in the solutions
and the proportional incorporation of starches in the biofilm samples, i.e.~potato
had the lowest absolute count in solutions, but the highest proportional
incorporation, and vice versa for the mixed treatment.

Wheat incorporation was most affected in the mixed-treatment samples, with only
0.064\% of the total available starches being incorporated into
the sample, compared to
0.161\%
in the separated wheat treatment.

\hypertarget{size-ratios-differ-between-solutions-and-samples}{%
\subsubsection{Size ratios differ between solutions and samples}\label{size-ratios-differ-between-solutions-and-samples}}

\begin{figure}
\centering
\includegraphics{../figures/ratio-plots-1.pdf}
\caption{\label{fig:ratio-plots}Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.}
\end{figure}

Overall, medium starch granules had a higher mean rate of incorporation
(0.186\%)
than small
(0.143\%)
and large
(0.065\%)
starch granules across all treatments, while large potato starches had the lowest
rate of incorporation across all treatments.

The difference in incorporation between the size categories resulted in a change
in size ratios between the original starch solutions and the extracted samples.
Large potato granules (\textgreater{} 20 \(\mu\)m) were most affected, with a
32.3\%
decrease in relative abundance in the potato-only treatment, and a
26.5\%
decrease in mixed treatments. Medium granules increased in relative abundance
across all samples, while small granules decreased in wheat treatments and
increased in potato treatments
(Figure \ref{fig:ratio-plots}).

\hypertarget{biofilm-weight-correlated-positively-with-extracted-starch-counts}{%
\subsubsection{Biofilm weight correlated positively with extracted starch counts}\label{biofilm-weight-correlated-positively-with-extracted-starch-counts}}

\begin{figure}
\centering
\includegraphics{../figures/cor-plot-1.pdf}
\caption{\label{fig:cor-plot}Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments.}
\end{figure}

Pearson's \emph{r} suggests a
strong positive
correlation between the total weight of the biofilms and the total starch count
(standardised by z-score) extracted from the samples across treatments,
\emph{r} = 0.659,
90\%CI{[}0.463, 0.794{]},
p = \textless{} 0.001
(Figure \ref{fig:cor-plot}).

The same test was applied to total biofilm weight and starch count per mg
calculus (also standardised by z-score), resulting in a weak positive correlation,
\emph{r} = 0.3,
90\%CI{[}0.0618, 0.506{]},
p 0.0403
(Figure \ref{fig:cor-plot2}).

\begin{figure}
\centering
\includegraphics{../figures/cor-plot2-1.pdf}
\caption{\label{fig:cor-plot2}Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus.}
\end{figure}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Here, we have provided a method for exploring the incorporation of dietary
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules
(of all sizes) present in the extracted samples were similar across all treatments
(0.064\% to 0.161\%),
despite large differences in absolute granule counts between wheat
(mean = 25,404,000)
and potato
(mean = 3,016,000)
solutions.\\
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat
and mixed solutions had the highest absolute mean count of starch granules, and
also had the highest absolute mean count of starch granules extracted from the
dental calculus
(Tables \ref{tab:solution-count-tbl} and \ref{tab:sample-count-tbl}).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and
degradation in the burial environment.
Despite the low proportion of granules recovered from the model calculus
(0.064\% to 0.161\%),
the absolute counts were still substantially greater than counts recovered from
archaeological remains
(Tromp et al., 2017; Tromp \& Dudgeon, 2015; Wesolowski et al., 2010). This could, in part, be due to the
activity of oral amylase, which was absent in our model.

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
diameter were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments,
while medium granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, and can reach up to 100 \(\mu\)m
in diameter, whereas wheat generally only reach up to 35 \(\mu\)m
(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, pp. 174--176).
This may also suggest that granule morphology plays a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
(Jane et al., 1994; Reichert, 1913, pp. 364--365; Seidemann, 1966, pp. 174--176; van de Velde et al., 2002).
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between biofilm size and retained
starch granules (Figure \ref{fig:cor-plot});
a result that contradicts findings from archaeological contexts
(Dudgeon \& Tromp, 2014; Wesolowski et al., 2010).
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot2});
while the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches.
Further research is needed to determine why this differs from previous archaeological
findings.

Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of
starch granules in dental calculus, and that starch granules extracted from dental
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues (2015) found individual
calculus samples to be a poor predictor of diet in a population, as many of the
consumed plants were missing from some individual samples, but were present in others.\\
Power and colleagues (2015)
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.
The size bias is also consistent with the findings by Power and colleagues
(2015),
who found that plants producing starches 10--20 \(\mu\)m in
size were over-represented; although, the representation of granules larger than
20 \(\mu\)m is unclear.

The mechanism by which starch granules are incorporated into plaque and calculus
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of considerable age
(Buckley et al., 2014; Henry et al., 2014; Wu et al., 2021).
Studies have also shown that not all starch granules come from a dietary source.
Other pathways include cross-contamination from plant interactions in soil, such
as palm phytoliths adhering to the skin of sweet potatoes
(Tromp \& Dudgeon, 2015),
or accidental ingestion not related to food consumption
(Radini et al., 2019; Radini et al., 2017).\\
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition through attachment to
tooth surfaces (including plaque and dental calulus) and mucous membranes
(Dodds \& Edgar, 1988; S. Kashket et al., 1991).
Bacteria also have the ability to adhere to starch granules
(Topping et al., 2003),
which would allow starches to attach to bacterial communities within the biofilm.
These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by \(\alpha\)-amylase (S. Kashket et al., 1996).
The susceptibility of granules to hydrolysis depends on the crystallinity and size
of the starch granule, as well as the mode of processing. Smaller and pre-processed
(e.g.~cooking) starch granules are more susceptible to enzymatic degradation,
while dehydrated starches will have a reduced susceptibility
(Björck et al., 1984; Franco et al., 1992; Haslam, 2004; Henry et al., 2009; Lingstrom et al., 1994).
Cracks on the surface, as well as unmineralised islands and channels exist within
dental calculus and may also be able to contain starch granules
(Charlier et al., 2010; Power et al., 2014; Tan, Gillam, et al., 2004).
Starch granules that are trapped in these pockets are (at least to some extent)
protected from aforementioned clearance mechanisms, especially once
a new layer of plaque has covered the surface of the plaque/calculus.
The size bias against large granules (\textgreater20 \(\mu\)m) from both wheat and potato
(Table \ref{tab:sample-prop-tbl}) may give further credence to
this incorporation pathway, as the smaller starch granules have an advantage over
larger granules, and can be stored in larger quantities.\\
This was also suggested by Power and colleagues (2014), who
observed clusters of starches within dental calculus, rather than an even
distribution across the surface of the dental calculus.\\
Granules trapped
in plaque/calculus may still be susceptible to hydrolysis, as \(\alpha\)-amylase has
the ability to bind to both tooth enamel and bacteria within a biofilm and retain
a portion of its hydrolytic activity
(Nikitkova et al., 2013; Scannapieco et al., 1993; Tan, Mordan, et al., 2004; Tan, Gillam, et al., 2004).
After the death of an individual, starches within dental calculus are susceptible
to further degradation by post-depositional processes, depending on burial environment
(pH, temperature, moisture content, microorganisms)\\
(Franco et al., 1992; García-Granero, 2020; Haslam, 2004; Henry et al., 2009).

A limitation of this study
is the absence of \(\alpha\)-amylase in the model. The presence of \(\alpha\)-amylase will likely affect
the total granule counts as well as the size ratios, as smaller starches may be more
susceptible to hydrolysis
(Franco et al., 1992; Haslam, 2004); however,
the absense can also allow us to directly explore
the effect of \(\alpha\)-amylase on starch counts in future experiments,
where \(\alpha\)-amylase can be added to the model
in concentrations similar to those found in the oral cavity (Scannapieco et al., 1993).
We are able to show how absolute counts in the treatments cause a difference in
incorporation; however, this
was merely a side-effect of the difference in the number of granules in potato and
wheat solutions of the same concentration (w/v). Further research should test
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown
negligible morphological changes caused by exposure to EDTA
(Le Moyne \& Crowther, 2021; Modi et al., 2020; Tromp et al., 2017),
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that
was feasible to count in a reasonable time period, so we calculated the total
counts by extrapolating from three slide transects.
Thus, we reasonably assume that the three transects are a good representation
of the entire slide, and that the distribution of all granules on the slide is
relatively homogenous.\\
Finally, we only used native starches in the experimental procedure and the results
will likely differ for processed starches (García-Granero, 2020).
Based on the comparatively low counts obtained by Leonard
and colleagues (2015, Supplement 2), processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

The data from this study are not sufficient to confirm or deny any proposed
methods of starch incorporation; however, while we are unable to address the
mechanism(s) of starch incorporation with the data obtained in this study,
the dental calculus model presented in this study is uniquely suited to explore
these questions and may improve interpretations of dietary practices in past populations.
It can address the call for more baseline testing of biases associated with
dietary research conducted on dental calculus
(Le Moyne \& Crowther, 2021).
Our high-throughput experimental setup allows us a
higher degree of control over the factors that influence starch incorporation and
retention, such as dietary intake, differential survivability of starches,
and inter- and intra-individual variation in plaque accumulation and mineralisation.
The latter is especially difficult to control \emph{in vivo} as it is influenced by
numerous factors including genetics, diet, salivary flow, and tooth position and
morphology
(Fagernäs et al., 2021; Haffajee et al., 2009; Jepsen et al., 2011; Proctor et al., 2018; Simón-Soro et al., 2013),
as well as evolutionary differences (Yates et al., 2021).
It can also facilitate training of students and researchers on methods of
dental calculus analysis, such as starch and phytolith extraction and
identification, where it can replace the use of finite archaeological resources.

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

This preliminary study shows that a very small proportion of the input starch
granules are retained in a dental calculus model. This and previous studies
have shown that calculus has a low capacity for retention of starch granules,
an effect that is compounded by diagenetic effects in archaeological remains,
resulting in low overall counts of extracted granules.
The proportion of starches consumed will in many cases be reflected
in the quantity
of starches extracted from the dental calculus---i.e.~the more starch granules
entering the oral cavity, the more will be recovered from extraction---at
least in modern calculus samples unaffected by diagenesis and hydrolysis.
Whether or not this also applies to archaeological samples remains to be tested.
Additionally, we have
shown that the size of granules will influence the likelihood of incorporation,
as large (\textgreater20 \(\mu\)m) starches have a decreased incorporation rate, medium
(10--20 \(\mu\)m)
starches an increased rate, and small (\textless10 \(\mu\)m) granules remained somewhat
constant. The size of calculus deposit also seems to influence the capacity of
granule incorporation; as the size of the deposit increases, so does the
absolute count of incorporated granules.\\
While we have shown multiple factors that influence the likelihood
of incorporation, the process still appears to be somewhat stochastic. Further
research is needed to make sense of the contributing factors, and to explore the
mechanisms of intra-oral starch incorporation and retention in dental calculus.
The oral biofilm model described in this study provides a method
to explore the incorporation and extraction of dietary compounds from dental calculus
in a controlled laboratory setting, and unearth the potential biases associated
with dietary research conducted on archaeological dental calculus.

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

This research has received funding from the European Research Council under the
European Union's Horizon 2020 research and innovation program, grant agreement
number STG--677576 (``HARVEST''). We also thank everyone in the general vicinity
of the lab for enduring the smell of bacterial accumulation.
We did NOT make use of Sci-Hub to access articles stuck behind a paywall,
and will, therefore, NOT acknowledge the use of Sci-Hub in this study.

\hypertarget{references-cited}{%
\section*{References cited}\label{references-cited}}
\addcontentsline{toc}{section}{References cited}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-adlerSequencingAncientCalcified2013}{}%
Adler, C. J., Dobney, K., Weyrich, L. S., Kaidonis, J., Walker, A. W., Haak, W., Bradshaw, C. J., Townsend, G., Soltysiak, A., Alt, K. W., Parkhill, J., \& Cooper, A. (2013). Sequencing ancient calcified dental plaque shows changes in oral microbiota with dietary shifts of the {Neolithic} and {Industrial} revolutions. \emph{Nature Genetics}, \emph{45}(4), 450--455, 455e1. \url{https://doi.org/10.1038/ng.2536}

\leavevmode\hypertarget{ref-armitageExtractionIdentificationOpal1975}{}%
Armitage, P. L. (1975). The {Extraction} and {Identification} of {Opal Phytoliths} from the {Teeth} of {Ungulates}. \emph{Journal of Archaeological Science}, \emph{2}, 187--197.

\leavevmode\hypertarget{ref-bernfeldAmylase1955}{}%
Bernfeld, P. (1955). Amylases, α and β. In \emph{Methods in {Enzymology}} (Vol. 1, pp. 149--158). {Academic Press}. \url{https://doi.org/10.1016/0076-6879(55)01021-5}

\leavevmode\hypertarget{ref-bjorckStarchProcessing1984}{}%
Björck, I., Asp, N.-G., Birkhed, D., Eliasson, A.-C., Sjöberg, L.-B., \& Lundquist, I. (1984). Effects of processing on starch availability {In} vitro and {In} vivo. {II}. {Drum}-drying of wheat flour. \emph{Journal of Cereal Science}, \emph{2}(3), 165--178. \url{https://doi.org/10.1016/S0733-5210(84)80030-2}

\leavevmode\hypertarget{ref-buckleyDentalCalculusCooking2014}{}%
Buckley, S., Usai, D., Jakob, T., Radini, A., \& Hardy, K. (2014). Dental {Calculus Reveals Unique Insights} into {Food Items}, {Cooking} and {Plant Processing} in {Prehistoric Central Sudan}. \emph{PLOS ONE}, \emph{9}(7), e100808. \url{https://doi.org/10.1371/journal.pone.0100808}

\leavevmode\hypertarget{ref-charlierSEMCalculus2010}{}%
Charlier, P., Huynh-Charlier, I., Munoz, O., Billard, M., Brun, L., \& Grandmaison, G. L. de la. (2010). The microscopic (optical and {SEM}) examination of dental calculus deposits ({DCD}). {Potential} interest in forensic anthropology of a bio-archaeological method. \emph{Legal Medicine}, \emph{12}(4), 163--171. \url{https://doi.org/10.1016/j.legalmed.2010.03.003}

\leavevmode\hypertarget{ref-cummingsMayanCalculus1997}{}%
Cummings, L. S., \& Magennis, A. (1997). A phytolith and starch record of food and grit in {Mayan} human tooth tartar. In A. Pinilla, J. Juan-Tresserras, \& M. J. Machado (Eds.), \emph{The {State}-of-the-{Art} of {Phytoliths} in {Soils} and {Plants}}. {CSIC Press}. \url{http://books.google.com?id=j66CDVfVhwEC}

\leavevmode\hypertarget{ref-doddsCarbohydrateRetention1988}{}%
Dodds, M. W. J., \& Edgar, W. M. (1988). The {Relationship Between Plaque pH}, {Plaque Acid Anion Profiles}, and {Oral Carbohydrate Retention After Ingestion} of {Several} '{Reference Foods}' by {Human Subjects}. \emph{Journal of Dental Research}, \emph{67}(5), 861--865. \url{https://doi.org/10.1177/00220345880670051301}

\leavevmode\hypertarget{ref-dudgeonDietGeographyDrinking2014}{}%
Dudgeon, J. V., \& Tromp, M. (2014). Diet, {Geography} and {Drinking Water} in {Polynesia}: {Microfossil Research} from {Archaeological Human Dental Calculus}, {Rapa Nui} ({Easter Island}). \emph{International Journal of Osteoarchaeology}, \emph{24}(5), 634--648. \url{https://doi.org/10.1002/oa.2249}

\leavevmode\hypertarget{ref-eerkensDentalCalculusSource2018}{}%
Eerkens, J. W., Tushingham, S., Brownstein, K. J., Garibay, R., Perez, K., Murga, E., Kaijankoski, P., Rosenthal, J. S., \& Gang, D. R. (2018). Dental calculus as a source of ancient alkaloids: {Detection} of nicotine by {LC}-{MS} in calculus samples from the {Americas}. \emph{Journal of Archaeological Science: Reports}, \emph{18}, 509--515. \url{https://doi.org/10.1016/j.jasrep.2018.02.004}

\leavevmode\hypertarget{ref-fagernasMicrobialBiogeography2021}{}%
Fagernäs, Z., Salazar-García, D. C., Avilés, A., Haber, M., Henry, A., Maurandi, J. L., Ozga, A., Velsko, I. M., \& Warinner, C. (2021). Understanding the microbial biogeography of ancient human dentitions to guide study design and interpretation. \emph{bioRxiv}, 2021.08.16.456492. \url{https://doi.org/10.1101/2021.08.16.456492}

\leavevmode\hypertarget{ref-foxPhytolithCalculus1996}{}%
Fox, C. L., Juan, J., \& Albert, R. M. (1996). Phytolith analysis on dental calculus, enamel surface, and burial soil: {Information} about diet and paleoenvironment. \emph{American Journal of Physical Anthropology}, \emph{101}(1), 101--113. \url{https://doi.org/10.1002/(SICI)1096-8644(199609)101:1\%3C101::AID-AJPA7\%3E3.0.CO;2-Y}

\leavevmode\hypertarget{ref-francoStarchDegradation1992}{}%
Franco, C. M. L., Preto, S. J. do R., \& Ciacco, C. F. (1992). Factors that {Affect} the {Enzymatic Degradation} of {Natural Starch Granules} -{Effect} of the {Size} of the {Granules}. \emph{Starch - Stärke}, \emph{44}(11), 422--426. \url{https://doi.org/10.1002/star.19920441106}

\leavevmode\hypertarget{ref-froehlichEffectOral1987}{}%
Froehlich, D. A., Pangborn, R. M., \& Whitaker, J. R. (1987). The effect of oral stimulation on human parotid salivary flow rate and alpha-amylase secretion. \emph{Physiology \& Behavior}, \emph{41}(3), 209--217. \url{https://doi.org/10.1016/0031-9384(87)90355-6}

\leavevmode\hypertarget{ref-graneroStarchTaphonomy2020}{}%
García-Granero, J. J. (2020). Starch taphonomy, equifinality and the importance of context: {Some} notes on the identification of food processing through starch grain analysis. \emph{Journal of Archaeological Science}, \emph{124}, 105267. \url{https://doi.org/10.1016/j.jas.2020.105267}

\leavevmode\hypertarget{ref-gismondiStarchGranulesData2019}{}%
Gismondi, A., D'Agostino, A., Canuti, L., Di Marco, G., Basoli, F., \& Canini, A. (2019). Starch granules: A data collection of 40 food species. \emph{Plant Biosystems - An International Journal Dealing with All Aspects of Plant Biology}, \emph{153}(2), 273--279. \url{https://doi.org/10.1080/11263504.2018.1473523}

\leavevmode\hypertarget{ref-haffajeeBiofilmPosition2009}{}%
Haffajee, A. D., Teles, R. P., Patel, M. R., Song, X., Yaskell, T., \& Socransky, S. S. (2009). Factors affecting human supragingival biofilm composition. {II}. {Tooth} position. \emph{Journal of Periodontal Research}, \emph{44}(4), 520--528. \url{https://doi.org/10.1111/j.1600-0765.2008.01155.x}

\leavevmode\hypertarget{ref-hardyStarchGranulesDental2009}{}%
Hardy, K., Blakeney, T., Copeland, L., Kirkham, J., Wrangham, R., \& Collins, M. (2009). Starch granules, dental calculus and new perspectives on ancient diet. \emph{Journal of Archaeological Science}, \emph{36}(2), 248--255. \url{https://doi.org/10.1016/j.jas.2008.09.015}

\leavevmode\hypertarget{ref-hardyNeanderthalMedics2012}{}%
Hardy, K., Buckley, S., Collins, M. J., Estalrrich, A., Brothwell, D., Copeland, L., García-Tabernero, A., García-Vargas, S., de la Rasilla, M., Lalueza-Fox, C., Huguet, R., Bastir, M., Santamaría, D., Madella, M., Wilson, J., Cortés, Á. F., \& Rosas, A. (2012). Neanderthal medics? {Evidence} for food, cooking, and medicinal plants entrapped in dental calculus. \emph{Naturwissenschaften}, \emph{99}(8), 617--626. \url{https://doi.org/10.1007/s00114-012-0942-0}

\leavevmode\hypertarget{ref-haslamDecompositionStarch2004}{}%
Haslam, M. (2004). The decomposition of starch grains in soils: Implications for archaeological residue analyses. \emph{Journal of Archaeological Science}, \emph{31}(12), 1715--1734. \url{https://doi.org/10.1016/j.jas.2004.05.006}

\leavevmode\hypertarget{ref-hendyProteomicCalculus2018}{}%
Hendy, J., Warinner, C., Bouwman, A., Collins, M. J., Fiddyment, S., Fischer, R., Hagan, R., Hofman, C. A., Holst, M., Chaves, E., Klaus, L., Larson, G., Mackie, M., McGrath, K., Mundorff, A. Z., Radini, A., Rao, H., Trachsel, C., Velsko, I. M., \& Speller, C. F. (2018). Proteomic evidence of dietary sources in ancient dental calculus. \emph{Proceedings. Biological Sciences}, \emph{285}(1883), 20180977. \url{https://doi.org/10.1098/rspb.2018.0977}

\leavevmode\hypertarget{ref-henryNeanderthalCalculus2014}{}%
Henry, A. G., Brooks, A. S., \& Piperno, D. R. (2014). Plant foods and the dietary ecology of {Neanderthals} and early modern humans. \emph{Journal of Human Evolution}, \emph{69}, 44--54. \url{https://doi.org/10.1016/j.jhevol.2013.12.014}

\leavevmode\hypertarget{ref-henryCookingStarch2009}{}%
Henry, A. G., Hudson, H. F., \& Piperno, D. R. (2009). Changes in starch grain morphologies from cooking. \emph{Journal of Archaeological Science}, \emph{36}(3), 915--922. \url{https://doi.org/10.1016/j.jas.2008.11.008}

\leavevmode\hypertarget{ref-henryCalculusSyria2008}{}%
Henry, A. G., \& Piperno, D. R. (2008). Using plant microfossils from dental calculus to recover human diet: A case study from {Tell} al-{Raqā}'i, {Syria}. \emph{Journal of Archaeological Science}, \emph{35}(7), 1943--1950. \url{https://doi.org/10.1016/j.jas.2007.12.005}

\leavevmode\hypertarget{ref-janeAnthologyStarch1994}{}%
Jane, J.-L., Kasemsuwan, T., Leas, S., Zobel, H., \& Robyt, J. F. (1994). Anthology of {Starch Granule Morphology} by {Scanning Electron Microscopy}. \emph{Starch - Stärke}, \emph{46}(4), 121--129. \url{https://doi.org/10.1002/star.19940460402}

\leavevmode\hypertarget{ref-jepsenCalculusRemoval2011}{}%
Jepsen, S., Deschner, J., Braun, A., Schwarz, F., \& Eberhard, J. (2011). Calculus removal and the prevention of its formation. \emph{Periodontology 2000}, \emph{55}(1), 167--188. \url{https://doi.org/10.1111/j.1600-0757.2010.00382.x}

\leavevmode\hypertarget{ref-jovanovicNeolithicCalculus2021}{}%
Jovanović, J., Power, R. C., de Becdelièvre, C., Goude, G., \& Stefanović, S. (2021). Microbotanical evidence for the spread of cereal use during the {Mesolithic}-{Neolithic} transition in the {Southeastern Europe} ({Danube Gorges}): {Data} from dental calculus analysis. \emph{Journal of Archaeological Science}, \emph{125}, 105288. \url{https://doi.org/10.1016/j.jas.2020.105288}

\leavevmode\hypertarget{ref-kashketFoodRetention1991}{}%
Kashket, S., Van Houte, J., Lopez, L. R., \& Stocks, S. (1991). Lack of {Correlation Between Food Retention} on the {Human Dentition} and {Consumer Perception} of {Food Stickiness}. \emph{Journal of Dental Research}, \emph{70}(10), 1314--1319. \url{https://doi.org/10.1177/00220345910700100101}

\leavevmode\hypertarget{ref-kashketFoodParticles1996}{}%
Kashket, S., Zhang, J., \& Houte, J. V. (1996). Accumulation of {Fermentable Sugars} and {Metabolic Acids} in {Food Particles} that {Become Entrapped} on the {Dentition}. \emph{Journal of Dental Research}, 8.

\leavevmode\hypertarget{ref-lemoyneCalculusPretreatments2021}{}%
Le Moyne, C., \& Crowther, A. (2021). Effects of chemical pre-treatments on modified starch granules: {Recommendations} for dental calculus decalcification for ancient starch research. \emph{Journal of Archaeological Science: Reports}, \emph{35}, 102762. \url{https://doi.org/10.1016/j.jasrep.2020.102762}

\leavevmode\hypertarget{ref-leonardDentalCalculus2015}{}%
Leonard, C., Vashro, L., O'Connell, J. F., \& Henry, A. G. (2015). Plant microremains in dental calculus as a record of plant consumption: {A} test with {Twe} forager-horticulturalists. \emph{Journal of Archaeological Science: Reports}, \emph{2}, 449--457. \url{https://doi.org/10.1016/j.jasrep.2015.03.009}

\leavevmode\hypertarget{ref-lingstromStarchyFood1994}{}%
Lingstrom, P., Birkhed, D., Ruben, J., \& Arends, J. (1994). Effect of {Frequent Consumption} of {Starchy Food Items} on {Enamel} and {Dentin Demineralization} and on {Plaque pH} in situ. \emph{Journal of Dental Research}, \emph{73}(3), 652--660. \url{https://doi.org/10.1177/00220345940730031101}

\leavevmode\hypertarget{ref-mickleburghNewInsightsConsumption2012}{}%
Mickleburgh, H. L., \& Pagán-Jiménez, J. R. (2012). New insights into the consumption of maize and other food plants in the pre-{Columbian Caribbean} from starch grains trapped in human dental calculus. \emph{Journal of Archaeological Science}, \emph{39}(7), 2468--2478. \url{https://doi.org/10.1016/j.jas.2012.02.020}

\leavevmode\hypertarget{ref-middletonExtractionOpalPhytoliths1994}{}%
Middleton, W. D., \& Rovner, I. (1994). Extraction of {Opal Phytoliths} from {Herbivore Dental Calculus}. \emph{Journal of Archaeological Science}, \emph{21}(4), 469--473. \url{https://doi.org/10.1006/jasc.1994.1046}

\leavevmode\hypertarget{ref-modiCalculusMethodologies2020}{}%
Modi, A., Pisaneschi, L., Zaro, V., Vai, S., Vergata, C., Casalone, E., Caramelli, D., Moggi-Cecchi, J., Mariotti Lippi, M., \& Lari, M. (2020). Combined methodologies for gaining much information from ancient dental calculus: Testing experimental strategies for simultaneously analysing {DNA} and food residues. \emph{Archaeological and Anthropological Sciences}, \emph{12}(1), 10. \url{https://doi.org/10.1007/s12520-019-00983-5}

\leavevmode\hypertarget{ref-R-here}{}%
Müller, K. (2020). \emph{Here: {A} simpler way to find your files} {[}Manual{]}. \url{https://CRAN.R-project.org/package=here}

\leavevmode\hypertarget{ref-naterHumanAmylase2005}{}%
Nater, U. M., Rohleder, N., Gaab, J., Berger, S., Jud, A., Kirschbaum, C., \& Ehlert, U. (2005). Human salivary alpha-amylase reactivity in a psychosocial stress paradigm. \emph{International Journal of Psychophysiology}, \emph{55}(3), 333--342. \url{https://doi.org/10.1016/j.ijpsycho.2004.09.009}

\leavevmode\hypertarget{ref-nikitkovaStarchBiofilms2013}{}%
Nikitkova, A. E., Haase, E. M., \& Scannapieco, F. A. (2013). Taking the {Starch} out of {Oral Biofilm Formation}: {Molecular Basis} and {Functional Significance} of {Salivary} α-{Amylase Binding} to {Oral Streptococci}. \emph{Applied and Environmental Microbiology}, \emph{79}(2), 416--423. \url{https://doi.org/10.1128/AEM.02581-12}

\leavevmode\hypertarget{ref-pearceConcomitantDepositionStrontium1987}{}%
Pearce, E. I. F., \& Sissons, C. H. (1987). The {Concomitant Deposition} of {Strontium} and {Fluoride} in {Dental Plaque}. \emph{Journal of Dental Research}, \emph{66}(10), 1518--1522. \url{https://doi.org/10.1177/00220345870660100101}

\leavevmode\hypertarget{ref-R-patchwork}{}%
Pedersen, T. L. (2020). \emph{Patchwork: {The} composer of plots} {[}Manual{]}. \url{https://CRAN.R-project.org/package=patchwork}

\leavevmode\hypertarget{ref-pipernoStarchGrains2008}{}%
Piperno, D. R., \& Dillehay, T. D. (2008). Starch grains on human teeth reveal early broad crop diet in northern {Peru}. \emph{Proceedings of the National Academy of Sciences}, \emph{105}(50), 19622--19627. \url{https://doi.org/10.1073/pnas.0808752105}

\leavevmode\hypertarget{ref-powerChimpCalculus2015}{}%
Power, R. C., Salazar-Garcia, D. C., Wittig, R. M., Freiberg, M., \& Henry, A. G. (2015). Dental calculus evidence of {Tai Forest Chimpanzee} plant consumption and life history transitions. \emph{Scientific Reports}, \emph{5}, 15161. \url{https://doi.org/10.1038/srep15161}

\leavevmode\hypertarget{ref-powerSEMCalculus2014}{}%
Power, R. C., Salazar-García, D. C., Wittig, R. M., \& Henry, A. G. (2014). Assessing use and suitability of scanning electron microscopy in the analysis of micro remains in dental calculus. \emph{Journal of Archaeological Science}, \emph{49}, 160--169. \url{https://doi.org/10.1016/j.jas.2014.04.016}

\leavevmode\hypertarget{ref-proctorSpatialGradient2018}{}%
Proctor, D. M., Fukuyama, J. A., Loomer, P. M., Armitage, G. C., Lee, S. A., Davis, N. M., Ryder, M. I., Holmes, S. P., \& Relman, D. A. (2018). A spatial gradient of bacterial diversity in the human oral cavity shaped by salivary flow. \emph{Nature Communications}, \emph{9}(1), 681. \url{https://doi.org/10.1038/s41467-018-02900-1}

\leavevmode\hypertarget{ref-R-base}{}%
R Core Team. (2020). \emph{R: {A} language and environment for statistical computing} {[}Manual{]}. {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing}. \url{https://www.R-project.org/}

\leavevmode\hypertarget{ref-radiniFoodMultiplePathways2017}{}%
Radini, A., Nikita, E., Buckley, S., Copeland, L., \& Hardy, K. (2017). Beyond food: {The} multiple pathways for inclusion of materials into ancient dental calculus. \emph{American Journal of Physical Anthropology}, \emph{162}, 71--83. \url{https://doi.org/10.1002/ajpa.23147}

\leavevmode\hypertarget{ref-radiniMedievalWomenEarly2019}{}%
Radini, A., Tromp, M., Beach, A., Tong, E., Speller, C., McCormick, M., Dudgeon, J. V., Collins, M. J., Rühli, F., Kröger, R., \& Warinner, C. (2019). Medieval women's early involvement in manuscript production suggested by lapis lazuli identification in dental calculus. \emph{Science Advances}, \emph{5}(1), eaau7126. \url{https://doi.org/10.1126/sciadv.aau7126}

\leavevmode\hypertarget{ref-reichertStarchBible1913b}{}%
Reichert, E. T. (1913). \emph{The differentiation and specificity of starches in relation to genera, species, etc: Stereochemistry applied to protoplasmic processes and products, and as a strictly scientific basis for the classification of plants and animals} (Vol. 2). {Carnegie institution of Washington}.

\leavevmode\hypertarget{ref-R-broom}{}%
Robinson, D., Hayes, A., \& Couch, S. (2021). \emph{Broom: {Convert} statistical objects into tidy tibbles} {[}Manual{]}. \url{https://CRAN.R-project.org/package=broom}

\leavevmode\hypertarget{ref-scannapiecoSalivaryAmylase1993}{}%
Scannapieco, F. A., Torres, G., \& Levine, M. J. (1993). Salivary α-amylase: Role in dental plaque and caries formation. \emph{Critical Reviews in Oral Biology \& Medicine}, \emph{4}(3), 301--307.

\leavevmode\hypertarget{ref-seidemannStarchAtlas1966}{}%
Seidemann, J. (1966). \emph{St\{\textbackslash"a\}rke-{Atlas}: {Grundlagen} der {St}\{\textbackslash"a\}rke-{Mikroskopie} und {Beschreibung} der wichtigsten {St}\{\textbackslash"a\}rkearten}. {Parey}.

\leavevmode\hypertarget{ref-shellisSyntheticSalivaCultural1978}{}%
Shellis, R. P. (1978). A synthetic saliva for cultural studies of dental plaque. \emph{Archives of Oral Biology}, \emph{23}(6), 485--489. \url{https://doi.org/10.1016/0003-9969(78)90081-X}

\leavevmode\hypertarget{ref-simon-soroOralGeography2013}{}%
Simón-Soro, A., Tomás, I., Cabrera-Rubio, R., Catalan, M. D., Nyvad, B., \& Mira, A. (2013). Microbial geography of the oral cavity. \emph{Journal of Dental Research}, \emph{92}(7), 616--621. \url{https://doi.org/10.1177/0022034513488119}

\leavevmode\hypertarget{ref-sissonsMultistationDentalPlaque1991}{}%
Sissons, C. H., Cutress, T. W., Hoffman, M. P., \& Wakefield, J. S. J. (1991). A {Multi}-station {Dental Plaque Microcosm} ({Artificial Mouth}) for the {Study} of {Plaque Growth}, {Metabolism}, {pH}, and {Mineralization}: \emph{Journal of Dental Research}. \url{https://doi.org/10.1177/00220345910700110301}

\leavevmode\hypertarget{ref-tanCalculusUltrastructure2004}{}%
Tan, B. T. K., Gillam, D. G., Mordan, N. J., \& Galgut, P. N. (2004). A preliminary investigation into the ultrastructure of dental calculus and associated bacteria. \emph{Journal of Clinical Periodontology}, \emph{31}(5), 364--369. \url{https://doi.org/10.1111/j.1600-051X.2004.00484.x}

\leavevmode\hypertarget{ref-tanBacterialViability2004}{}%
Tan, B. T. K., Mordan, N. J., Embleton, J., Pratten, J., \& Galgut, P. N. (2004). Study of {Bacterial Viability} within {Human Supragingival Dental Calculus}. \emph{Journal of Periodontology}, \emph{75}(1), 23--29. \url{https://doi.org/10.1902/jop.2004.75.1.23}

\leavevmode\hypertarget{ref-taoWheatCalculus2020}{}%
Tao, D., Zhang, G., Zhou, Y., \& Zhao, H. (2020). Investigating wheat consumption based on multiple evidences: {Stable} isotope analysis on human bone and starch grain analysis on dental calculus of humans from the {Laodaojing} cemetery, {Central Plains}, {China}. \emph{International Journal of Osteoarchaeology}, \emph{30}(5), 594--606. \url{https://doi.org/10.1002/oa.2884}

\leavevmode\hypertarget{ref-toppingResistantStarch2003}{}%
Topping, D. L., Fukushima, M., \& Bird, A. R. (2003). Resistant starch as a prebiotic and synbiotic: State of the art. \emph{Proceedings of the Nutrition Society}, \emph{62}(1), 171--176. \url{https://doi.org/10.1079/PNS2002224}

\leavevmode\hypertarget{ref-trompEDTACalculus2017}{}%
Tromp, M., Buckley, H., Geber, J., \& Matisoo-Smith, E. (2017). {EDTA} decalcification of dental calculus as an alternate means of microparticle extraction from archaeological samples. \emph{Journal of Archaeological Science: Reports}, \emph{14}, 461--466. \url{https://doi.org/10.1016/j.jasrep.2017.06.035}

\leavevmode\hypertarget{ref-trompDietaryNondietary2015}{}%
Tromp, M., \& Dudgeon, J. V. (2015). Differentiating dietary and non-dietary microfossils extracted from human dental calculus: The importance of sweet potato to ancient diet on {Rapa Nui}. \emph{Journal of Archaeological Science}, \emph{54}, 54--63. \url{https://doi.org/10.1016/j.jas.2014.11.024}

\leavevmode\hypertarget{ref-vandeveldeStarchMorphology2002}{}%
van de Velde, F., van Riel, J., \& Tromp, R. H. (2002). Visualisation of starch granule morphologies using confocal scanning laser microscopy ({CSLM}). \emph{Journal of the Science of Food and Agriculture}, \emph{82}(13), 1528--1536. \url{https://doi.org/10.1002/jsfa.1165}

\leavevmode\hypertarget{ref-warinnerDirectEvidenceMilk2014}{}%
Warinner, C., Hendy, J., Speller, C., Cappellini, E., Fischer, R., Trachsel, C., Arneborg, J., Lynnerup, N., Craig, O. E., Swallow, D. M., Fotakis, A., Christensen, R. J., Olsen, J. V., Liebert, A., Montalva, N., Fiddyment, S., Charlton, S., Mackie, M., Canci, A., \ldots{} Collins, M. J. (2014). Direct evidence of milk consumption from ancient human dental calculus. \emph{Scientific Reports}, \emph{4}, 7104. \url{https://doi.org/10.1038/srep07104}

\leavevmode\hypertarget{ref-warinnerPathogensHost2014}{}%
Warinner, C., Rodrigues, J. F., Vyas, R., Trachsel, C., Shved, N., Grossmann, J., Radini, A., Hancock, Y., Tito, R. Y., Fiddyment, S., Speller, C., Hendy, J., Charlton, S., Luder, H. U., Salazar-Garcia, D. C., Eppler, E., Seiler, R., Hansen, L. H., Castruita, J. A., \ldots{} Cappellini, E. (2014). Pathogens and host immunity in the ancient human oral cavity. \emph{Nature Genetics}, \emph{46}(4), 336--344. \url{https://doi.org/10.1038/ng.2906}

\leavevmode\hypertarget{ref-wesolowskiEvaluatingMicrofossilContent2010}{}%
Wesolowski, V., Ferraz Mendonça de Souza, S. M., Reinhard, K. J., \& Ceccantini, G. (2010). Evaluating microfossil content of dental calculus from {Brazilian} sambaquis. \emph{Journal of Archaeological Science}, \emph{37}(6), 1326--1338. \url{https://doi.org/10.1016/j.jas.2009.12.037}

\leavevmode\hypertarget{ref-tidyverse2019}{}%
Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L. D., François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen, T. L., Miller, E., Bache, S. M., Müller, K., Ooms, J., Robinson, D., Seidel, D. P., Spinu, V., \ldots{} Yutani, H. (2019). Welcome to the {tidyverse}. \emph{Journal of Open Source Software}, \emph{4}(43), 1686. \url{https://doi.org/10.21105/joss.01686}

\leavevmode\hypertarget{ref-wuDietEarliest2021}{}%
Wu, Y., Tao, D., Wu, X., \& Liu, W. (2021). \emph{Diet of the earliest modern humans in {East Asia}} {[}Preprint{]}. {In Review}. \url{https://doi.org/10.21203/rs.3.rs-442096/v1}

\leavevmode\hypertarget{ref-yatesOralMicrobiome2021}{}%
Yates, J. A. F., Velsko, I. M., Aron, F., Posth, C., Hofman, C. A., Austin, R. M., Parker, C. E., Mann, A. E., Nägele, K., Arthur, K. W., Arthur, J. W., Bauer, C. C., Crevecoeur, I., Cupillard, C., Curtis, M. C., Dalén, L., Bonilla, M. D.-Z., Fernández-Lomana, J. C. D., Drucker, D. G., \ldots{} Warinner, C. (2021). The evolution and changing ecology of the {African} hominid oral microbiome. \emph{Proceedings of the National Academy of Sciences}, \emph{118}(20). \url{https://doi.org/10.1073/pnas.2021655118}

\end{CSLReferences}

\end{document}

### 2021-10-08 analysis/paper/paper.Rmd
incorporate AGH suggestions

@@
     number_sections: yes
+    keep_tex: true
+    extra_dependencies: "subfig"
+  bookdown::word_document2:
+    reference_docx: "../templates/template.docx"
   bookdown::html_document2:
@@
 abstract: |
-  Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+ Dental calculus has proven to contain a wealth of information 
+ on the dietary habits of past populations. These insights have, to a large extent,
+ been obtained by the extraction and identification of
+ starch granules contained within the mineralised dental plaque from a wide range 
+ of regions and time periods. The scope of previous studies have been limited to 
+ microfossil extraction and identification to reconstruct dietary preferences from
+ the archaeological record, and few studies have attempted to address the biases 
+ of starch retention in dental calculus. Those that have considered this problem 
+ have been limited to *in vivo* studies on modern humans and non-human primates.
+ Here, we present a multispecies oral biofilm model, which allows experimental 
+ research on  starch incorporation and retention to be conducted on *in vitro* 
+ dental calculus in a controlled laboratory setting. The biofilms were exposed 
+ to treatment solutions with known quantities of
+ dietary starches (wheat and potato) during the 25-day growth period. After this,
+ the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
+ and counted. We show that the granule counts extracted from the model dental calculus 
+ represented a low proportion (ranging from 0.064% to 0.161%) of the total number 
+ of granules exposed to the biofilms throughout the experiment.
+ Additionally, we found that the ratios of granule sizes from the extracted 
+ starch granules differed from the original treatment solutions, with large 
+ granules (>20 $\mu$m) consistently being under-represented.
+ We also found a correlation
+ between the absolute granule counts and dry-weight of the biofilm 
+ (*r* = 0.66, 90%CI[0.46,0.79]), as well
+ as between the concentration (count per mg) of granules and dry-weight 
+ (*r* = 0.30, 90%CI[0.06,0.51]).  
+ Our results reinforce previous *in vivo* studies suggesting that dental calculus 
+ presents a very small, and partly biased picture of the original dietary intake 
+ of starches, with an over-representation of plants producing granules smaller than
+ 20 $\mu$m in size. The experimental model presented here is well-suited to address
+ the need for further validation of methods and biases associated with dietary
+ research on dental calculus.
 keywords: 
@@
 <!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
-Dental calculus, mineralised dental plaque, has proven to contain a wealth of 
+Dental calculus has proven to contain a wealth of 
 dietary information in the form of plant microfossils 
@@
 and other organic residues [@buckleyDentalCalculusCooking2014]. 
-This dietary information can be preserved within calculus over many millenia,
-providing a unique window into the food-related behaviours of past populations 
+This dietary information can be preserved within the mineralised dental plaque 
+over many millennia, providing a unique window into the food-related behaviours of 
+past populations 
 [@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
@@
 In more recent years, the study of dental calculus has increased exponentially, 
-and the wealth of information contained within the mineralised matrix has been, 
-to a larger extent, acknowledged. The applications of dental calculus span a wide
+and the wealth of information contained within the mineralised matrix has largely
+been acknowledged. The use of dental calculus spans a wide
 variety of archaeological research areas, such as oral microbiome 
 characterisation (including pathogens) through the analysis of DNA and proteins 
-[@adlerSequencingAncientCalcified2013; @warinnerPathogensHostImmunity2014],
+[@adlerSequencingAncientCalcified2013; @warinnerPathogensHost2014],
 microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@@
 [@buckleyDentalCalculusCooking2014; @hendyProteomicCalculus2018], 
-and nicotine-use [@eerkensDentalCalculusSource2018]. Especially the extraction 
+and nicotine use [@eerkensDentalCalculusSource2018]. Especially the extraction 
 of starch granules has become a rich source of dietary
@@
 oral biofilm with a multifactorial aetiology and variable formation rates both 
-within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], contributing to
+within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
+contributing to
 the stochasticity of starch representation being observed in numerous studies.
@@
 of starch processing, as well as post-depositional processes that are still being 
-explored [@graneroStarchTaphonomy2020], and you have a highly unpredictable process.
+explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
+dental calculus to reconstruct diet is a highly unpredictable process.
 
@@
 the model, a limited proportion of the starch input is actually 
-retained in the calculus, as well as a shift in the size ratios of individual 
-starch granules
-that are incorporated into the calculus. We also show that the number of starch
-granules that are incorporated increases as the size of the calculus deposit 
+retained in the calculus. We also observed a shift in the size ratios of individual 
+starch granules that are incorporated into the calculus, and that the number of 
+incorporated starch granules increases as the size of the calculus deposit 
 increases.
@@
 protocol from Sissons and colleagues [-@sissonsMultistationDentalPlaque1991] and
-Shellis [-@shellisSyntheticSalivaCultural1978]. The setup comprises a polypropylene 
+Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
+with whole saliva was grown on a substrate suspended in artificial saliva, and
+fed with sugar (sucrose). After several days of growth, the biofilm was exposed
+to starch solutions. Mineralisation of the biofilm was aided by exposure to a
+calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
+collected for further analysis. The setup comprises a polypropylene 
 24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which 
@@
 for 10 days until the end of the experiment on day 24 
-(see \@ref(fig:protocol-fig) for an overview of the protocol). More detailed protocols
-are available on protocols.io<!--insert DOI-->.
+(see \@ref(fig:protocol-fig) for an overview of the protocol).
 
@@
 in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.
-For a more detailed protocol, see <!-- protocols.io DOI -->
 
@@
 (https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch).
-All protocols are available on protocols.io.
+More detailed protocols are available on protocols.io.
 
@@
 
+```{r microscope-fig, fig.cap="Microscope image of a biofilm sample exposed to the potato starch solution.
+Potato granules can be seen within a bacterial community.
+Scale bar = 20 $\\mu$m.", out.width="50%"}
+knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
+```
+
 All samples yielded sufficient biofilm growth and starch incorporation to be 
-included in the analysis, resulting in a total of 48 biofilm samples (two plates of 24),
+included in the analysis (Figure \@ref(fig:microscope-fig)), resulting in a total of 48 biofilm samples (two plates of 24),
 45 of which were used for analysis (three samples were set aside for later
@@
              caption = "The mean percentage of starches from the solutions that 
-              were incorported into the samples.")
+              were incorporated into the samples.")
 ```
@@
 There is an inverse relationship between the absolute starch count in the solutions 
-and the proportional incorporation of starches in the biofilm samples, i.e. potato 
+and the proportional incorporation of starches in the biofilm samples, i.e., potato 
 had the lowest absolute count in solutions, but the highest proportional 
 incorporation, and vice versa for the mixed treatment.
-
-<!-- separated treatments -->
 
@@
 
-```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments."}
-pl_cor
+```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments."}
+pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
 ```
@@
 archaeological remains 
-[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010]. <!-- although this study extrapolated counts to full sample content, which is
-rarely done in archaeological studies -->This could, in part, be due to the 
-activity of oral amylase, which was absent in our model.
+[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010], <!-- although this study extrapolated counts to full sample content, which is
+rarely done in archaeological studies -->which could in part be due to the 
+lack of oral amylase activity in our model.
 
+We have also shown that the size of the starch granules influences the likelihood
+of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
+diameter were underrepresented in the calculus samples compared to the original 
+starch solutions, an effect that was consistent across all three treatments, 
+while medium granules (10--20 $\mu$m) were often over-represented
+(Table \@ref(tab:sample-prop-tbl), 
+and 
+Figure \@ref(fig:ratio-plots)).
+Large potato granules were most affected, potentially because of the greater
+size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
+granules generally only reach up to 35 $\mu$m 
+[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
+Granule morphology may also play a role. Large wheat granules 
+are lenticular and have a larger surface area compared 
+to volume, whereas large potato granules are ovoid and have a larger volume
+compared to surface area 
+[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
+Another potentially important factor 
+from our results is the size of the calculus deposit. We found a 
+`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
+starch granules (Figure \@ref(fig:cor-plot));
+a result that contradicts findings from archaeological contexts
+[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
+When the concentration of starch granules
+per mg calculus is considered, the correlation is weaker, but still present
+(Figure \@ref(fig:cor-plot2)).
+While the larger deposits contain a higher absolute count, our findings also suggest
+that they contain a slightly higher concentration of starches. 
+Further research is needed to determine why this differs from previous archaeological
+findings.
 <!--
@@
 representative of individual diets than starch granules.
-
-<!-- secondary findings -->
-We have also shown that the size of the starch granules influences the likelihood
-of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
-diameter were underrepresented in the calculus samples compared to the original 
-starch solutions, an effect that was consistent across all three treatments, 
-while medium granules (10 -- 20 $\mu$m) were often over-represented
-(Table \@ref(tab:sample-prop-tbl), 
-and 
-Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).
-This was especially true for the potato starches, which can reach up to 100 $\mu$m
-in diameter, whereas wheat generally only reach up to 35 $\mu$m 
-[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004].
-The size bias is consistent with the findings by Power and colleagues
+The size bias is also consistent with the findings by Power and colleagues
 [-@powerChimpCalculus2015],
-which found that plants producing starches 10--20 $\mu$m in
-size were over-represented; although, the representation of granules larger than
-20 $\mu$m is unclear.  
-Given the differences in proportional incorporation between wheat and potato,
-especially for small and medium granules, it is also possible that the morphology
-of granules plays a role. However, the proportional count differences between large 
-wheat and potato granules were less pronounced.
-
-<!-- Since those results were obtained -->
-<!-- on a modern population of chimps that were only buried for a short period of time,  -->
-<!-- degradation was likely limited [@powerChimpCalculus2015]. -->
-
-The taphonomy of starch granules has been addressed previously, and depends on 
-burial environment (pH, temperature, moisture content, microorganisms), as well 
-as processing of the granules prior to deposition 
-[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004]. 
-The potential for detrimental diagenetic effects on starch granules increases as 
-the size and amylose content decreases
-[@francoStarchDegradation1992; @haslamDecompositionStarch2004].
+who found that plants producing starches 10--20 $\mu$m in
+size were over-represented; however, the representation of granules larger than
+20 $\mu$m in their study is unclear.
 
@@
 by $\alpha$-amylase [@kashketFoodParticles1996].
-Starch granules that are trapped in crevices and channels on the surface of the
-mineralised plaque are (at least to some extent) protected from salivary
-clearance and mechanical cleaning actions of the tongue and lips, especially once
-a new layer of plaque has covered the surface of the calculus. 
+The susceptibility of granules to hydrolysis depends on the crystallinity and size
+of the starch granule, as well as the mode of processing. Smaller and pre-processed
+(e.g., cooked) starch granules are more susceptible to enzymatic degradation, 
+while dehydrated starches will have a reduced susceptibility
+[@lingstromStarchyFood1994; @francoStarchDegradation1992; @haslamDecompositionStarch2004; 
+@bjorckStarchProcessing1984; @henryCookingStarch2009].
+Cracks on the surface of the dental calculus, as well as unmineralised islands 
+and channels may also be able to contain starch granules 
+[@powerSEMCalculus2014; @tanCalculusUltrastructure2004; @charlierSEMCalculus2010].
+Starch granules that are trapped in these pockets are (at least to some extent) 
+protected from aforementioned clearance mechanisms, especially once
+a new layer of plaque has covered the surface of the plaque/calculus. 
 The size bias against large granules (>20 $\mu$m) from both wheat and potato 
@@
 a portion of its hydrolytic activity
-[@scannapiecoSalivaryAmylase1993; @nikitkovaStarchBiofilms2013; @tanStudyBacterialViability2004; @tanCalculusUltrastructure2004].
-The susceptibility of granules to hydrolysis depends on the crystallinity and size
-of the starch granule, as well as the mode of processing. As mentioned previously,
-smaller starch granules are more susceptible to enzymatic degradation, as are 
-pre-processed starches (e.g. by cooking), while dehydrated starches will have a
-reduced susceptibility
-[@lingstromStarchyFood1994; @francoStarchDegradation1992; @haslamDecompositionStarch2004; 
-@bjorckStarchProcessing1984; @henryCookingStarch2009]. Given the lack of $\alpha$-amylase 
-activity detected in our model, hydrolysis is not the cause of the low
-incorporation rate. More details on the absence of $\alpha$-amylase 
-activity is discussed elsewhere (Bartholdy et al. in prep.).
-What we can say, is that the dietary picture we obtain from starch granules extracted 
-from dental calculus reflects a number of individual, dietary, and environmental 
-(both pre- and post-mortem environments) factors, and that this will likely be a
-somewhat random snapshot given the irregularity of plaque mineralisation within
-and between individuals [@jinSupragingivalCalculus2002; @jepsenCalculusRemoval2011]. 
-Although, certain factors may increase the likelihood of incorporation, such 
-as abundance, size, and morphology of granules.
+[@scannapiecoSalivaryAmylase1993; @nikitkovaStarchBiofilms2013; @tanBacterialViability2004; @tanCalculusUltrastructure2004].
+After the death of an individual, starches within dental calculus are susceptible 
+to further degradation by post-depositional processes, depending on burial environment 
+(pH, temperature, moisture content, microorganisms)  
+[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004; @francoStarchDegradation1992].
+Future study should explore how burial affects the recovery of starch from the
+biofilm model. <!-- I may get to it eventually, but it would be great if someone else could look at this! -->
 
@@
 A limitation of this study 
-is the absence of $\alpha$-amylase in the model. The presence of $\alpha$-amylase will likely affect 
-the total granule counts as well as the size ratios, as smaller starches may be more 
+The absence of $\alpha$-amylase in the model is a limitation of this study, as 
+the total granule counts were not subject to hydrolysis. This would likely have 
+reduced and affected the size ratios, as smaller starches may be more 
 susceptible to hydrolysis 
 [@francoStarchDegradation1992; @haslamDecompositionStarch2004]; however, 
-the absense can also allow us to directly explore
+the absence can also allow us to directly explore
 the effect of $\alpha$-amylase on starch counts in future experiments,
 where $\alpha$-amylase can be added to the model
-in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. <!-- foreshadowing...? -->
-While we are able to show the differing 
-incorporation caused by absolute counts available, this
+in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. 
+We are able to show how absolute counts in the treatments cause a difference in
+incorporation. However, this
 was merely a side-effect of the difference in the number of granules in potato and 
@@
 of the entire slide, and that the distribution of all granules on the slide is 
-relatively homogenous.   
+relatively homogeneous.   
 Finally, we only used native starches in the experimental procedure and the results 
 will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
-Based on counts obtained by Leonard
+Based on the comparatively low counts obtained by Leonard
 and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
@@
 <!-- How this engineering will facilitate future research -->
-The oral biofilm model described in this study provides a method
-to explore the incorporation and extraction of dietary compounds from dental calculus 
-in a controlled laboratory setting, as many of the variables can be adjusted. 
-The addition of known starch species in
-separate samples also removes the identification bias against the smaller
-undiagnostic starches.<!-- rephrase -->
-It can also address the call for more baseline testing of biases associated with 
-dietary research conducted on dental calculus 
+
+While we are unable to sufficiently address the 
+mechanism(s) of starch incorporation with the data obtained in this study, 
+the dental calculus model presented here is uniquely suited to explore 
+these questions and may improve interpretations of dietary practices in past 
+populations. Further analyses using this model can address the call for more 
+baseline testing of biases associated with dietary research conducted on dental calculus 
 [@lemoyneCalculusPretreatments2021].
-*In vivo* studies on humans and non-human primates will be limited by the variability 
-occurring between individuals, and difficulties in accurately documenting 
-the entirety of starch granules entering the oral cavity at any given moment,
-especially accidental intake. 
-Our experimental setup allows us a
-higher degree of control over the factors that influence starch incorporation, such as
-dietary intake, processing of starches leading to differential survivability, 
-and inter- and intra-individual variation in plaque accumulation and 
-mineralisation.
+Our high-throughput experimental setup allows us a
+higher degree of control over the factors that influence starch incorporation and
+retention, such as dietary intake, differential survivability of starches, 
+and inter- and intra-individual variation in plaque accumulation and mineralisation.
 The latter is especially difficult to control *in vivo* as it is influenced by 
-numerous factors including genetics, diet, salivary flow, tooth position and morphology 
+numerous factors including genetics, diet, salivary flow, and tooth position and
+morphology 
 [@jepsenCalculusRemoval2011; @simon-soroOralGeography2013; @proctorSpatialGradient2018; @fagernasMicrobialBiogeography2021; @haffajeeBiofilmPosition2009], 
-as well as evolutionary differences through time [@yatesOralMicrobiome2021].
+as well as evolutionary differences [@yatesOralMicrobiome2021].
 It can also facilitate training of students and researchers on methods of 
@@
 in the quantity
-of starches extracted from the dental calculus---i.e. the more starch granules
+of starches extracted from the dental calculus---i.e., the more starch granules
 entering the oral cavity, the more will be recovered from extraction---at 

### 2021-10-08 analysis/paper/paper.tex
incorporate AGH suggestions

--- 2021-10-08 analysis/paper/first_draft_rev-1/paper.tex
@@
 \linenumbers
+\usepackage{subfig}
 \ifluatex
@@
 \author{Bjørn Peare Bartholdy \and Amanda G. Henry}
-\date{2021-09-30}
+\date{2021-10-08}
 
@@
 \begin{abstract}
-Dental calculus, mineralised plaque, has proven to contain a wealth of information
-on the dietary habits of past populations. The extraction and identification of
-starch granules contained within dental calculus has already provided valuable
-dietary insights from a wide range of regions and time periods, and seem to preserve
-well in the archaeological record.
-While the majority of studies have focused on microfossil extraction and
-identification from archaeological remains, few studies have
-addressed the biases of starch retention in dental calculus, and have been
-limited to \emph{in vivo} studies on modern humans and non-human primates.
+Dental calculus has proven to contain a wealth of information
+on the dietary habits of past populations. These insights have, to a large extent,
+been obtained by the extraction and identification of
+starch granules contained within the mineralised dental plaque from a wide range
+of regions and time periods. The scope of previous studies have been limited to
+microfossil extraction and identification to reconstruct dietary preferences from
+the archaeological record, and few studies have attempted to address the biases
+of starch retention in dental calculus. Those that have considered this problem
+have been limited to \emph{in vivo} studies on modern humans and non-human primates.
 Here, we present a multispecies oral biofilm model, which allows experimental
-research to be conducted on \emph{in vitro} dental calculus in a controlled laboratory
-setting. The biofilms were exposed treatment solutions with known quantities of
+research on starch incorporation and retention to be conducted on \emph{in vitro}
+dental calculus in a controlled laboratory setting. The biofilms were exposed
+to treatment solutions with known quantities of
 dietary starches (wheat and potato) during the 25-day growth period. After this,
-the starch granules were extracted from the mature biofilm (dissolution in EDTA),
-and counted. We
-show that the granule counts extracted from the model dental calculus represented
-a low proportion of the total counts (ranging from 0.064\% to 0.161\%) exposed to
-the biofilms throughout the experiment. Additionally, we found that the ratios
-of granule sizes from the extracted starch granules differed from the original
-treatment solutions, with large granules (\textgreater20 \(\mu\)m) consistently being
-under-represented.
-Additionally, the size ratios of granules within starch species shifted between
-the input starch solutions and the granules extracted from the model calculus,
-with large granules (\textgreater20 \(\mu\)m) consistently being underrepresented.
+the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
+and counted. We show that the granule counts extracted from the model dental calculus
+represented a low proportion (ranging from 0.064\% to 0.161\%) of the total number
+of granules exposed to the biofilms throughout the experiment.
+Additionally, we found that the ratios of granule sizes from the extracted
+starch granules differed from the original treatment solutions, with large
+granules (\textgreater20 \(\mu\)m) consistently being under-represented.
 We also found a correlation
@@
 as between the concentration (count per mg) of granules and dry-weight
-(\emph{r} = 0.3, 90\%CI{[}0.06,0.51{]}).\\
+(\emph{r} = 0.30, 90\%CI{[}0.06,0.51{]}).\\
 Our results reinforce previous \emph{in vivo} studies suggesting that dental calculus
@@
 of starches, with an over-representation of plants producing granules smaller than
-20 \(\mu\) in size. The experimental model presented here is well-suited to address
+20 \(\mu\)m in size. The experimental model presented here is well-suited to address
 the need for further validation of methods and biases associated with dietary
@@
 \setstretch{2.0}
-Word count: 4634
+Word count: 4682
 
@@
 
-Dental calculus, mineralised dental plaque, has proven to contain a wealth of
+Dental calculus has proven to contain a wealth of
 dietary information in the form of plant microfossils
@@
 and other organic residues (Buckley et al., 2014).
-This dietary information can be preserved within calculus over many millenia,
-providing a unique window into the food-related behaviours of past populations
+This dietary information can be preserved within the mineralised dental plaque
+over many millennia, providing a unique window into the food-related behaviours of
+past populations
 (Henry \& Piperno, 2008; Jovanović et al., 2021; Tao et al., 2020)
@@
 In more recent years, the study of dental calculus has increased exponentially,
-and the wealth of information contained within the mineralised matrix has been,
-to a larger extent, acknowledged. The applications of dental calculus span a wide
+and the wealth of information contained within the mineralised matrix has largely
+been acknowledged. The use of dental calculus spans a wide
 variety of archaeological research areas, such as oral microbiome
@@
 of starch processing, as well as post-depositional processes that are still being
-explored (García-Granero, 2020), and you have a highly unpredictable process.
+explored (García-Granero, 2020), and it becomes clear that using
+dental calculus to reconstruct diet is a highly unpredictable process.
 
@@
 the model, a limited proportion of the starch input is actually
-retained in the calculus, as well as a shift in the size ratios of individual
-starch granules
-that are incorporated into the calculus. We also show that the number of starch
-granules that are incorporated increases as the size of the calculus deposit
+retained in the calculus. We also observed a shift in the size ratios of individual
+starch granules that are incorporated into the calculus, and that the number of
+incorporated starch granules increases as the size of the calculus deposit
 increases.
@@
 protocol from Sissons and colleagues (1991) and
-Shellis (1978). The setup comprises a polypropylene
+Shellis (1978). In brief, a biofilm inoculated
+with whole saliva was grown on a substrate suspended in artificial saliva, and
+fed with sugar (sucrose). After several days of growth, the biofilm was exposed
+to starch solutions. Mineralisation of the biofilm was aided by exposure to a
+calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
+collected for further analysis. The setup comprises a polypropylene
 24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which
@@
 for 10 days until the end of the experiment on day 24
-(see \ref{fig:protocol-fig} for an overview of the protocol). More detailed protocols
-are available on protocols.io.
+(see \ref{fig:protocol-fig} for an overview of the protocol).
 
@@
 in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.
-For a more detailed protocol, see
 
@@
 (\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}).
-All protocols are available on protocols.io.
+More detailed protocols are available on protocols.io.
 
@@
 \begin{figure}
-\centering
-\includegraphics{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar.jpg}
-\caption{\label{fig:microscope-fig}Microscope image of potato starches on a microscope slide, as well as a bacterial community. Scale bar = 20 \(\mu\)m.}
+\includegraphics[width=0.5\linewidth]{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar} \caption{Microscope image of a biofilm sample exposed to the potato starch solution.
+Potato granules can be seen within a bacterial community.
+Scale bar = 20 $\mu$m.}\label{fig:microscope-fig}
 \end{figure}
@@
 \caption{\label{tab:sample-prop-tbl}The mean percentage of starches from the solutions that 
-              were incorported into the samples.}
+              were incorporated into the samples.}
 \centering
@@
 There is an inverse relationship between the absolute starch count in the solutions
-and the proportional incorporation of starches in the biofilm samples, i.e.~potato
+and the proportional incorporation of starches in the biofilm samples, i.e., potato
 had the lowest absolute count in solutions, but the highest proportional
@@
 \includegraphics{../figures/cor-plot-1.pdf}
-\caption{\label{fig:cor-plot}Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments.}
+\caption{\label{fig:cor-plot}Scatter plot of sample weight and standardised starch count by Z-score for separated treatments.}
 \end{figure}
@@
 archaeological remains
-(Tromp et al., 2017; Tromp \& Dudgeon, 2015; Wesolowski et al., 2010). This could, in part, be due to the
-activity of oral amylase, which was absent in our model.
+(Tromp et al., 2017; Tromp \& Dudgeon, 2015; Wesolowski et al., 2010), which could in part be due to the
+lack of oral amylase activity in our model.
 

### 2021-10-08 analysis/supplementary-materials/supp-mat.Rmd
incorporate AGH suggestions

@@
 print(sessionInfo(), locale = FALSE)
+```
+
+## Experimental setup
+
+```{r plate-lid-1, fig.cap="The 24 deepwell plate and the lid with pegs (substrata)"}
+knitr::include_graphics(here("analysis/figures/plate_lid_side.jpg"))
+```
+
+```{r plate-lid-2, fig.cap="The 24 deepwell plate with the lid on (sort of...)"}
+knitr::include_graphics(here("analysis/figures/plate_lid_on.jpg"))
 ```
@@
 
-```{r img-1, echo=FALSE, fig.cap="Image of starch granules extracted from a calculus sample"}
+```{r img-1, echo=FALSE, fig.cap="Image of starch granules extracted from a potato treatment sample"}
+knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
+```
 
+```{r img-2}
+knitr::include_graphics(here("analysis/figures/SNAP-103412-0006.jpg"))
 ```
+
+```{r img-3, fig.cap="Microscope image of wheat starch from a wheat treatment sample."}
+knitr::include_graphics(here("analysis/figures/SNAP-164650-0012.jpg"))
+```

### 2021-10-11 analysis/paper/paper.Rmd
AGH second revisions to first draft

@@
 title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
-author:
-  - Bjørn Peare Bartholdy
-  - Amanda G. Henry
-date: "`r Sys.Date()`"
+author: |
+ Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
+date: |
+ $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
 output:
+  bookdown::html_document2:
+    df_print: kable
+    keep_md: true
   bookdown::pdf_document2:
-    #template: "../templates/sn-article-template/sn-article.tex"
     includes:
       in_header: "../templates/latex/pre-print/preamble.tex"
-      before_body: "../templates/latex/pre-print/before_body.tex"
+      #before_body: "../templates/latex/pre-print/before_body.tex"
     df_print: kable
@@
     keep_tex: true
-    extra_dependencies: "subfig"
+    #extra_dependencies: "subfig"
   bookdown::word_document2:
     reference_docx: "../templates/template.docx"
-  bookdown::html_document2:
-    df_print: kable
-    keep_md: true
-linestretch: 2.0 # double-spacing for review purposes
+#linestretch: 2.0 # double-spacing for review purposes
 bibliography: refs.bib
@@
   message = FALSE,
-  fig.path = "../figures/")
+  fig.path = "../figures/",
+  fig.align = "center")
 library(tidyverse)
@@
 ## Compile refs.bib using rbbt package ##
-cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
+#cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
 #cite_keys <- unlist(cite_keys, use.names = FALSE)
-rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T)
+#rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T)
 ## end refs.bib ##
@@
 ```
-
-Word count: `r wd_count`
 
@@
 
-All scripts and data used in the analysis are available on OSF 
-(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch).
-More detailed protocols are available on protocols.io.
-
 # Results
@@
 
-<!--
-
-```{r amylase-activity-tbl}
-rownames(plt1_ph1_result) <- rows
-colnames(plt1_ph1_result) <- cols
-knitr::kable(signif(plt1_ph1_result, 3), row.names = TRUE, caption = "Amylase activity in U/mL enzyme, where a U is mg maltose released from starch in six minutes at 36 degrees Celsius.
-Plate 1, photometric reading 1.
-Negative values converted to 0.")
-```
--->
-
 No $\alpha$-amylase activity was detected in any of the artificial
@@
 <!-- starch counts in the solutions -->
-
-<!-- INSERT MICROSCOPE PICTURES OF STARCHES -->
 
@@
                   paste0(round(x), " (", scales::percent(signif(x / total, 3)), ")")))
-
-# count_samp_comb_per <- count_sample_comb %>%
-#   mutate(across(c(s, m, l, total), 
-#                 function(x) 
-#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
-
-# count_samp_comb_out <- tibble(count_samp_comb_per, count_sample_comb_sd, .name_repair = "unique") %>%
-#   select(1,2,3,7,4,8,5,9,6,10) #%>%
-  #add_row(tibble_row(count_sample_comb_out), .before = 3)
 
@@
 
-```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."}
+```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.", out.width="80%"}
 #plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
@@
 ```
-
-<!--
-```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring)."}
-#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
-pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
-```
-
-```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring)."}
-#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
-pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
-```
- -->
 
@@
 cor_p <- if(count_cor$p.value > 0.001){
-  signif(count_cor$p.value, 3)
+  paste("=", signif(count_cor$p.value, 3))
 } else {
@@
 cor2_p <- if(starch_cor$p.value > 0.001){
-  signif(starch_cor$p.value, 3)
+  paste("=", signif(starch_cor$p.value, 3))
 } else {
@@
 
-```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments."}
-pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments.", out.width="80%"}
+pl_cor
+#pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
 ```
@@
 90%CI[`r paste(cor_ci[1], cor_ci[2], sep = ", ")`],
-p = `r cor_p`
+p `r cor_p`
 (Figure \@ref(fig:cor-plot)).
+
+```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus.", out.width="80%"}
+pl_cor2
+```
 
@@
 (Figure \@ref(fig:cor-plot2)).
-
-<!-- size ratios in the larger deposits vs the smaller deposits?
-  Are larger starches more proportionately represented in larger deposits?
-  -->
-
-```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
-pl_cor2
-```
 
@@
 findings.
-<!--
-Given the low proportion of starch granule retention in our dental calculus model,
-and the limited difference between treatments, we can conjecture that starch 
-availability is not a main factor in starch incorporation. While there was a
-proportional incorporation, i.e. more granules in the solution led to more granules
-in the calculus, the difference between solutions [a number/percentage] was of a
-higher magnitude than the difference between samples <!-- a number/percentage-->
 
@@
 20 $\mu$m in their study is unclear.
-
-<!-- add potential for morphological differences affecting incorporation 
- Medium potato starches had the highest rate of incorporation 
- All large starches had a consistently low rate of incorporation 
- Starch granule adhesion to bacteria and enzymes? Do they have a preference?
- Binding to select species in a biofilm could explain clustering
- Colonisation of bacteria onto starch @nagaraBacteriaNutrients2017 (although not
- clear if the bacteria in our model are even 'consuming' starch-->
-
-<!-- Combined, the effect of intra-oral starch incorporation -->
-<!-- and the bias against large granules, together with the increased effect of taphonomy -->
-<!-- on  small granules, would eliminate a large portion of the consumed starch granules, -->
-<!-- and could explain the large descrepency between our counts (unaffected by hydrolysis  -->
-<!-- and taphonomy) and the aforementioned studies.  -->
 
@@
 documents. You're awesome! -->
+We would like to thank Dr. Stephanie Schnorr for help with the amylase activity protocol.
+We also thank everyone in the general vicinity
+of the lab for enduring the smell of bacterial accumulation.
+Since we did NOT make use of Sci-Hub to access articles stuck behind a paywall, 
+we will NOT acknowledge the use of Sci-Hub in this study.
+
 This research has received funding from the European Research Council under the 
 European Union’s Horizon 2020 research and innovation program, grant agreement 
-number STG–677576 (“HARVEST”). We also thank everyone in the general vicinity
-of the lab for enduring the smell of bacterial accumulation.
-We did NOT make use of Sci-Hub to access articles stuck behind a paywall, 
-and will, therefore, NOT acknowledge the use of Sci-Hub in this study.
+number STG–677576 (“HARVEST”).
+
+# Data Availability Statement {-}
+
+All scripts and data used in the analysis are available on OSF 
+(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch),
+following the format provided by the rrtools package [@R-rrtools].
+More detailed protocols are available on OSF and protocols.io.
 

### 2021-10-11 analysis/paper/paper.tex
AGH second revisions to first draft

@@
 \usepackage{lmodern}
-\usepackage{setspace}
 \usepackage{ifxetex,ifluatex}
@@
   pdftitle={Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model},
-  pdfauthor={Bjørn Peare Bartholdy; Amanda G. Henry},
   pdfkeywords={oral biofilm, starch, alpha-amylase, dental calculus},
@@
 \linenumbers
-\usepackage{subfig}
 \ifluatex
@@
 \title{Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model}
-\author{Bjørn Peare Bartholdy \and Amanda G. Henry}
-\date{2021-10-08}
+\author{Bjørn Peare Bartholdy\(^1\), Amanda G. Henry\(^1\)}
+\date{\(^1\) Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.}
 
@@
 
-% \providecommand{\@maketitle}{}
-% \renewcommand{\@maketitle}{%
-%   \vbox{%
-%     \hsize\textwidth
-%     \linewidth\hsize
-%     \vskip 0.1in
-%     \@toptitlebar
-%     \centering
-%     {\LARGE\sc \@title\par}
-%     \@bottomtitlebar
-%     \textsc{A Preprint}\\
-%     \vskip 0.1in
-%     \def\And{%
-%       \end{tabular}\hfil\linebreak[0]\hfil%
-%       \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\ignorespaces%
-%     }
-%     \def\AND{%
-%       \end{tabular}\hfil\linebreak[4]\hfil%
-%       \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\ignorespaces%
-%     }
-%     \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\@author\end{tabular}%
-%   \vskip 0.4in \@minus 0.1in \center{\today}   \vskip 0.2in
-%   }
-% }
-%
-% \providecommand{\maketitle}{}
-% \renewcommand{\maketitle}{%
-%   \par
-%   \begingroup
-%     \renewcommand{\thefootnote}{\fnsymbol{footnote}}
-%     % for perfect author name centering
-%     \renewcommand{\@makefnmark}{\hbox to \z@{$^{\@thefnmark}$\hss}}
-%     % The footnote-mark was overlapping the footnote-text,
-%     % added the following to fix this problem               (MK)
-%     \long\def\@makefntext##1{%
-%       \parindent 1em\noindent
-%       \hbox to 1.8em{\hss $\m@th ^{\@thefnmark}$}##1
-%     }
-%     \thispagestyle{empty}
-%     \@maketitle
-%     \@thanks
-%     %\@notice
-%   \endgroup
-%   \let\maketitle\relax
-%   \let\thanks\relax
-% }
-
-\setstretch{2.0}
-Word count: 4682
+Word count: 4722
 
@@
 (Buckley et al., 2014; Hendy et al., 2018),
-and nicotine-use (Eerkens et al., 2018). Especially the extraction
+and nicotine use (Eerkens et al., 2018). Especially the extraction
 of starch granules has become a rich source of dietary
@@
 oral biofilm with a multifactorial aetiology and variable formation rates both
-within and between individuals (Haffajee et al., 2009; Jepsen et al., 2011), contributing to
+within and between individuals (Haffajee et al., 2009; Jepsen et al., 2011),
+contributing to
 the stochasticity of starch representation being observed in numerous studies.
@@
 \begin{figure}
-\centering
-\includegraphics{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview.png}
-\caption{\label{fig:protocol-fig}Overview of experiment protocol including the plate setup.}
+
+{\centering \includegraphics{/media/bjorn/hogwarts/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview}
+
+}
+
+\caption{Overview of experiment protocol including the plate setup.}\label{fig:protocol-fig}
 \end{figure}
@@
 
-All scripts and data used in the analysis are available on OSF
-(\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}).
-More detailed protocols are available on protocols.io.
-
 \hypertarget{results}{%
@@
 \begin{figure}
-\includegraphics[width=0.5\linewidth]{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar} \caption{Microscope image of a biofilm sample exposed to the potato starch solution.
-Potato granules can be seen within a bacterial community.
-Scale bar = 20 $\mu$m.}\label{fig:microscope-fig}
+
+{\centering \includegraphics[width=0.5\linewidth]{/media/bjorn/hogwarts/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar}
+
+}
+
+\caption{Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\mu$m.}\label{fig:microscope-fig}
 \end{figure}
@@
 \begin{figure}
-\centering
-\includegraphics{../figures/ratio-plots-1.pdf}
-\caption{\label{fig:ratio-plots}Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.}
+
+{\centering \includegraphics[width=0.8\linewidth]{../figures/ratio-plots-1}
+
+}
+
+\caption{Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.}\label{fig:ratio-plots}
 \end{figure}
@@
 \begin{figure}
-\centering
-\includegraphics{../figures/cor-plot-1.pdf}
-\caption{\label{fig:cor-plot}Scatter plot of sample weight and standardised starch count by Z-score for separated treatments.}
+
+{\centering \includegraphics[width=0.8\linewidth]{../figures/cor-plot-1}
+
+}
+
+\caption{Scatter plot of sample weight and standardised starch count by Z-score for separated treatments.}\label{fig:cor-plot}
 \end{figure}
@@
 90\%CI{[}0.463, 0.794{]},
-p = \textless{} 0.001
+p \textless{} 0.001
 (Figure \ref{fig:cor-plot}).
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.8\linewidth]{../figures/cor-plot2-1}
+
+}
+
+\caption{Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus.}\label{fig:cor-plot2}
+\end{figure}
 
@@
 90\%CI{[}0.0618, 0.506{]},
-p 0.0403
+p = 0.0403
 (Figure \ref{fig:cor-plot2}).
-
-\begin{figure}
-\centering
-\includegraphics{../figures/cor-plot2-1.pdf}
-\caption{\label{fig:cor-plot2}Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus.}
-\end{figure}
 
@@
 Figure \ref{fig:ratio-plots}).
-Large potato granules were most affected, and can reach up to 100 \(\mu\)m
-in diameter, whereas wheat generally only reach up to 35 \(\mu\)m
+Large potato granules were most affected, potentially because of the greater
+size-range. They can reach up to 100 \(\mu\)m in maximum length, whereas wheat
+granules generally only reach up to 35 \(\mu\)m
 (Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, pp. 174--176).
-This may also suggest that granule morphology plays a role. Large wheat granules
+Granule morphology may also play a role. Large wheat granules
 are lenticular and have a larger surface area compared
@@
 per mg calculus is considered, the correlation is weaker, but still present
-(Figure \ref{fig:cor-plot2});
-while the larger deposits contain a higher absolute count, our findings also suggest
+(Figure \ref{fig:cor-plot2}).
+While the larger deposits contain a higher absolute count, our findings also suggest
 that they contain a slightly higher concentration of starches.
@@
 who found that plants producing starches 10--20 \(\mu\)m in
-size were over-represented; although, the representation of granules larger than
-20 \(\mu\)m is unclear.
+size were over-represented; however, the representation of granules larger than
+20 \(\mu\)m in their study is unclear.
 
@@
 of the starch granule, as well as the mode of processing. Smaller and pre-processed
-(e.g.~cooking) starch granules are more susceptible to enzymatic degradation,
+(e.g., cooked) starch granules are more susceptible to enzymatic degradation,
 while dehydrated starches will have a reduced susceptibility
 (Björck et al., 1984; Franco et al., 1992; Haslam, 2004; Henry et al., 2009; Lingstrom et al., 1994).
-Cracks on the surface, as well as unmineralised islands and channels exist within
-dental calculus and may also be able to contain starch granules
+Cracks on the surface of the dental calculus, as well as unmineralised islands
+and channels may also be able to contain starch granules
 (Charlier et al., 2010; Power et al., 2014; Tan, Gillam, et al., 2004).
@@
 this incorporation pathway, as the smaller starch granules have an advantage over
-larger granules, and can be stored in larger quantities.\\
+larger granules, and can be stored in larger quantities.
 This was also suggested by Power and colleagues (2014), who
 observed clusters of starches within dental calculus, rather than an even
-distribution across the surface of the dental calculus.\\
+distribution across the surface of the dental calculus.
 Granules trapped
@@
 to further degradation by post-depositional processes, depending on burial environment
-(pH, temperature, moisture content, microorganisms)\\
+(pH, temperature, moisture content, microorganisms)
 (Franco et al., 1992; García-Granero, 2020; Haslam, 2004; Henry et al., 2009).
+Future study should explore how burial affects the recovery of starch from the
+biofilm model.
 
 A limitation of this study
-is the absence of \(\alpha\)-amylase in the model. The presence of \(\alpha\)-amylase will likely affect
-the total granule counts as well as the size ratios, as smaller starches may be more
+The absence of \(\alpha\)-amylase in the model is a limitation of this study, as
+the total granule counts were not subject to hydrolysis. This would likely have
+reduced and affected the size ratios, as smaller starches may be more
 susceptible to hydrolysis
 (Franco et al., 1992; Haslam, 2004); however,
-the absense can also allow us to directly explore
+the absence can also allow us to directly explore
 the effect of \(\alpha\)-amylase on starch counts in future experiments,
@@
 We are able to show how absolute counts in the treatments cause a difference in
-incorporation; however, this
+incorporation. However, this
 was merely a side-effect of the difference in the number of granules in potato and
@@
 of the entire slide, and that the distribution of all granules on the slide is
-relatively homogenous.\\
+relatively homogeneous.\\
 Finally, we only used native starches in the experimental procedure and the results
@@
 
-The data from this study are not sufficient to confirm or deny any proposed
-methods of starch incorporation; however, while we are unable to address the
+While we are unable to sufficiently address the
 mechanism(s) of starch incorporation with the data obtained in this study,
-the dental calculus model presented in this study is uniquely suited to explore
-these questions and may improve interpretations of dietary practices in past populations.
-It can address the call for more baseline testing of biases associated with
-dietary research conducted on dental calculus
+the dental calculus model presented here is uniquely suited to explore
+these questions and may improve interpretations of dietary practices in past
+populations. Further analyses using this model can address the call for more
+baseline testing of biases associated with dietary research conducted on dental calculus
 (Le Moyne \& Crowther, 2021).
@@
 in the quantity
-of starches extracted from the dental calculus---i.e.~the more starch granules
+of starches extracted from the dental calculus---i.e., the more starch granules
 entering the oral cavity, the more will be recovered from extraction---at
@@
 
+\hypertarget{data-availability-statement}{%
+\section*{Data Availability Statement}\label{data-availability-statement}}
+\addcontentsline{toc}{section}{Data Availability Statement}
+
+All scripts and data used in the analysis are available on OSF
+(\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}),
+following the format provided by the rrtools package (Marwick, 2019).
+More detailed protocols are available on OSF and protocols.io.
+
 \hypertarget{acknowledgements}{%
@@
 
+We would like to thank Dr.~Stephanie Schnorr for help with the amylase activity protocol.
+We also thank everyone in the general vicinity
+of the lab for enduring the smell of bacterial accumulation.
+Since we did NOT make use of Sci-Hub to access articles stuck behind a paywall,
+we will NOT acknowledge the use of Sci-Hub in this study.
+
 This research has received funding from the European Research Council under the
 European Union's Horizon 2020 research and innovation program, grant agreement
-number STG--677576 (``HARVEST''). We also thank everyone in the general vicinity
-of the lab for enduring the smell of bacterial accumulation.
-We did NOT make use of Sci-Hub to access articles stuck behind a paywall,
-and will, therefore, NOT acknowledge the use of Sci-Hub in this study.
+number STG--677576 (``HARVEST'').
 
@@
 (2014).
-Diet, {Geography} and {Drinking Water} in {Polynesia}: {Microfossil Research} from {Archaeological Human Dental Calculus}, {Rapa Nui} ({Easter Island}).
+Diet, {Geography} and {Drinking Water} in {Polynesia}: Microfossil {Research} from {Archaeological Human Dental Calculus}, {Rapa Nui} ({Easter Island}).
 \emph{International Journal of Osteoarchaeology}, \emph{24}(5), 634--648.
@@
 (2018).
-Dental calculus as a source of ancient alkaloids: {Detection} of nicotine by {LC}-{MS} in calculus samples from the {Americas}.
+Dental calculus as a source of ancient alkaloids: Detection of nicotine by {LC}-{MS} in calculus samples from the {Americas}.
 \emph{Journal of Archaeological Science: Reports}, \emph{18}, 509--515.
@@
 (1996).
-Phytolith analysis on dental calculus, enamel surface, and burial soil: {Information} about diet and paleoenvironment.
+Phytolith analysis on dental calculus, enamel surface, and burial soil: Information about diet and paleoenvironment.
 \emph{American Journal of Physical Anthropology}, \emph{101}(1), 101--113.
@@
 (2020).
-Starch taphonomy, equifinality and the importance of context: {Some} notes on the identification of food processing through starch grain analysis.
+Starch taphonomy, equifinality and the importance of context: Some notes on the identification of food processing through starch grain analysis.
 \emph{Journal of Archaeological Science}, \emph{124}, 105267.
@@
 Neanderthal medics?
-{Evidence} for food, cooking, and medicinal plants entrapped in dental calculus.
+Evidence for food, cooking, and medicinal plants entrapped in dental calculus.
 \emph{Naturwissenschaften}, \emph{99}(8), 617--626.
@@
 (2021).
-Microbotanical evidence for the spread of cereal use during the {Mesolithic}-{Neolithic} transition in the {Southeastern Europe} ({Danube Gorges}): {Data} from dental calculus analysis.
+Microbotanical evidence for the spread of cereal use during the {Mesolithic}-{Neolithic} transition in the {Southeastern Europe} ({Danube Gorges}): Data from dental calculus analysis.
 \emph{Journal of Archaeological Science}, \emph{125}, 105288.
@@
 (2021).
-Effects of chemical pre-treatments on modified starch granules: {Recommendations} for dental calculus decalcification for ancient starch research.
+Effects of chemical pre-treatments on modified starch granules: Recommendations for dental calculus decalcification for ancient starch research.
 \emph{Journal of Archaeological Science: Reports}, \emph{35}, 102762.
@@
 (2015).
-Plant microremains in dental calculus as a record of plant consumption: {A} test with {Twe} forager-horticulturalists.
+Plant microremains in dental calculus as a record of plant consumption: A test with {Twe} forager-horticulturalists.
 \emph{Journal of Archaeological Science: Reports}, \emph{2}, 449--457.
@@
 \url{https://doi.org/10.1177/00220345940730031101}
+
+\leavevmode\hypertarget{ref-R-rrtools}{}%
+Marwick, B. (2019). \emph{Rrtools: Creates a reproducible research compendium} {[}Manual{]}. \url{https://github.com/benmarwick/rrtools}
 
@@
 \leavevmode\hypertarget{ref-R-here}{}%
-Müller, K. (2020). \emph{Here: {A} simpler way to find your files} {[}Manual{]}. \url{https://CRAN.R-project.org/package=here}
+Müller, K. (2020). \emph{Here: A simpler way to find your files} {[}Manual{]}. \url{https://CRAN.R-project.org/package=here}
 
@@
 (2013).
-Taking the {Starch} out of {Oral Biofilm Formation}: {Molecular Basis} and {Functional Significance} of {Salivary} α-{Amylase Binding} to {Oral Streptococci}.
+Taking the {Starch} out of {Oral Biofilm Formation}: Molecular {Basis} and {Functional Significance} of {Salivary} α-{Amylase Binding} to {Oral Streptococci}.
 \emph{Applied and Environmental Microbiology}, \emph{79}(2), 416--423.
@@
 \leavevmode\hypertarget{ref-R-patchwork}{}%
-Pedersen, T. L. (2020). \emph{Patchwork: {The} composer of plots} {[}Manual{]}. \url{https://CRAN.R-project.org/package=patchwork}
+Pedersen, T. L. (2020). \emph{Patchwork: The composer of plots} {[}Manual{]}. \url{https://CRAN.R-project.org/package=patchwork}
 
@@
 (2020).
-\emph{R: {A} language and environment for statistical computing} {[}Manual{]}.
+\emph{R: A language and environment for statistical computing} {[}Manual{]}.
 {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing}.
@@
 (2017).
-Beyond food: {The} multiple pathways for inclusion of materials into ancient dental calculus.
+Beyond food: The multiple pathways for inclusion of materials into ancient dental calculus.
 \emph{American Journal of Physical Anthropology}, \emph{162}, 71--83.
@@
 \leavevmode\hypertarget{ref-R-broom}{}%
-Robinson, D., Hayes, A., \& Couch, S. (2021). \emph{Broom: {Convert} statistical objects into tidy tibbles} {[}Manual{]}. \url{https://CRAN.R-project.org/package=broom}
+Robinson, D., Hayes, A., \& Couch, S. (2021). \emph{Broom: Convert statistical objects into tidy tibbles} {[}Manual{]}. \url{https://CRAN.R-project.org/package=broom}
 
@@
 \leavevmode\hypertarget{ref-seidemannStarchAtlas1966}{}%
-Seidemann, J. (1966). \emph{St\{\textbackslash"a\}rke-{Atlas}: {Grundlagen} der {St}\{\textbackslash"a\}rke-{Mikroskopie} und {Beschreibung} der wichtigsten {St}\{\textbackslash"a\}rkearten}. {Parey}.
+Seidemann, J. (1966). \emph{St\{\textbackslash"a\}rke-{Atlas}: Grundlagen der {St}\{\textbackslash"a\}rke-{Mikroskopie} und {Beschreibung} der wichtigsten {St}\{\textbackslash"a\}rkearten}. {Parey}.
 
@@
 (2020).
-Investigating wheat consumption based on multiple evidences: {Stable} isotope analysis on human bone and starch grain analysis on dental calculus of humans from the {Laodaojing} cemetery, {Central Plains}, {China}.
+Investigating wheat consumption based on multiple evidences: Stable isotope analysis on human bone and starch grain analysis on dental calculus of humans from the {Laodaojing} cemetery, {Central Plains}, {China}.
 \emph{International Journal of Osteoarchaeology}, \emph{30}(5), 594--606.

### 2021-10-11 analysis/supplementary-materials/supp-mat.Rmd
AGH second revisions to first draft

@@
 library(tidyverse)
-raw_data <- read_csv(here("analysis/data/raw_data/starch_counts.csv"))
+raw_data_samples <- read_csv(here("analysis/data/raw_data/starch_counts.csv"))
+raw_data_solutions <- read_csv(here("analysis/data/raw_data/solution_counts.csv"))
 ```
@@
 
+## Raw data
+
+Raw counts from the treatment solutions before extrapolation.
+
+```{r raw-data-sol, echo=FALSE}
+raw_data_solutions
+```
+
+Raw counts from the calculus samples before extrapolation:
+
+```{r raw-data-samp, echo=FALSE}
+raw_data_samples
+```
+
+The raw data can be downloaded from GitHub:
+
+```sh
+# solution counts
+wget https://github.com/bbartholdy/byoc-starch/blob/main/analysis/data/raw_data/solution_counts.csv
+
+# sample counts
+wget https://github.com/bbartholdy/byoc-starch/blob/main/analysis/data/raw_data/starch_counts.csv
+```
+
+## Microscope images
+
+```{r img-1, echo=FALSE, fig.cap="Image of starch granules extracted from a potato treatment sample"}
+knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
+```
+
+```{r img-2}
+knitr::include_graphics(here("analysis/figures/SNAP-103412-0006.jpg"))
+```
+
+```{r img-3, fig.cap="Microscope image of wheat starch from a wheat treatment sample."}
+knitr::include_graphics(here("analysis/figures/SNAP-164650-0012.jpg"))
+```
+
 ## Amylase activity
@@
 ```{r}
-raw_data %>%
+raw_data_samples %>%
   filter(treatment == "control") %>%
@@
 $$ \text{Corrected count} = \text{raw count} \times (\text{portion of slide})^{-1} \times \frac{V_{sample}}{V_{slide}} $$
-
-## Microscope images
-
-```{r img-1, echo=FALSE, fig.cap="Image of starch granules extracted from a potato treatment sample"}
-knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
-```
-
-```{r img-2}
-knitr::include_graphics(here("analysis/figures/SNAP-103412-0006.jpg"))
-```
-
-```{r img-3, fig.cap="Microscope image of wheat starch from a wheat treatment sample."}
-knitr::include_graphics(here("analysis/figures/SNAP-164650-0012.jpg"))
-```

### 2021-10-11 analysis/scripts/amylase-assay.R

@@ -50,10 +50,6 @@ stdBMM2.ph2 <- stdcurve(plt2_ph2, cols = 10:12)
 plates <- list("plate1_phot1" = plt1_ph1, "plate1_phot2" = plt1_ph2,
                "plate2_phot1" = plt2_ph1, "plate2_phot2" = plt2_ph2)
 phot_results <- lapply(plates, sep_samples)
-# plt1_ph1.res <- sep_samples(plt1_ph1)
-# plt1_ph2.res <- sep_samples(plt1_ph2)
-# plt2_ph1.res <- sep_samples(plt2_ph1)
-# plt2_ph2.res <- sep_samples(plt2_ph2)
 
 # saliva calculation
 split_results <- lapply(phot_results, split, f = ~ solution)
@@ -62,26 +58,6 @@ saliva_samples <- list("plate1_phot1_saliva" = split_results$plate1_phot1$saliva
                        "plate2_phot1_saliva" = split_results$plate2_phot1$saliva,
                        "plate2_phot2_saliva" = split_results$plate2_phot2$saliva) %>%
   lapply(select, !solution)
-#lapply(saliva_samples, select, !solution)
-#sample_calc(split_results$plate1_phot1$saliva, std_curve = stdH2O1.ph1, dilf = 2, ml_enz = 0.075)
-
-# saliva1_ph1 <- split_results$plt1_ph1$saliva
-# saliva1_ph2 <- split_results$plt1_ph2$saliva
-# saliva2_ph1 <- split_results$plt2_ph1$saliva
-# saliva2_ph2 <- split_results$plt2_ph2$saliva
-
-# saliva1_ph1 <- plt1_ph1.res$saliva
-# saliva1_ph2 <- plt1_ph2.res$saliva
-# saliva2_ph1 <- plt2_ph1.res$saliva
-# saliva2_ph2 <- plt2_ph2.res$saliva
-
-
-# saliva_results <- lapply(
-#   saliva_samples,
-#   sample_calc,
-#   stdcurve = c(stdH2O1.ph1, stdH2O1.ph2, stdH2O2.ph1, stdH2O2.ph2),
-#   dilf = 2,
-#   ml_enz = 0.075)
 
 sal1_ph1 <- sample_calc(saliva_samples$plate1_phot1_saliva,
                         std_curve = stdH2O1.ph1,
[…]

### 2021-10-11 analysis/scripts/analysis.R

@@ -1,12 +1,80 @@
-# Analysis script
+# R Script for correcting counts according to volume and proportion of slide counted
+
+# Dependencies ------------------------------------------------------------
 
 library(here)
 library(tidyverse)
 library(broom)
 
-corr_counts <- readr::read_csv(here("analysis/data/derived_data/corr_counts.csv"))
-corr_comb <- readr::read_csv(here("analysis/data/derived_data/corr_comb.csv"))
-sol_corr <- readr::read_csv(here("analysis/data/derived_data/sol_corr.csv"))
+# Upload data -------------------------------------------------------------
+
+# starch counts in samples
+raw_counts <- readr::read_csv(here("analysis/data/raw_data", "starch_counts.csv"))
+
+# starch counts in solutions
+sol_counts <- readr::read_csv(here("analysis/data/raw_data", "solution_counts.csv"))
+
+# Functions ---------------------------------------------------------------
+
+# function to correct counts for volume and proportion of slide counted
+
+correct_count <- function(raw, prop, vol){
+  corrected <- raw * (1 / prop) * (vol / 20)
+  corrected
+}
+
+# function to extrapolate solution counts to 1 ml exposure per day
+
+extrap_count <- function(raw, prop, vol, days){
+  extrap <- raw * (1 / prop) * 100 * days
+  extrap
+}
+
+# Outlier detection and removal -------------------------------------------
+
+raw_counts %>%
+  filter(treatment != "control") %>%
+  ggplot() +
+  geom_boxplot(aes(x = starch, y = total, fill = starch))
+
+# Only lower outliers were considered, as this would suggest that there may have
+# been a problem with the sample and incorporation of starches. Given that all
+# samples were treated equally, there is no reason to consider upper outliers
+# problematic.
+
+
+# Solution counts ---------------------------------------------------------
+
+sol_corr <- sol_counts %>%
+  mutate(across(c(s, m, l, total), extrap_count, portion_slide, vol_slide, days = 16))
+readr::write_csv(sol_corr, here("analysis/data/derived_data", "sol_corr.csv"))
+
+# Apply correction --------------------------------------------------------
+
+corr_counts <- raw_counts %>%
+  #filter(treatment != "control") %>%
+  mutate(across(c(s, m, l, total), correct_count, portion_slide, vol)) %>%
+  select(!c(vol, portion_slide)) %>%
[…]

### 2021-10-11 analysis/scripts/figures.R

@@ -52,9 +52,6 @@ ratio_samp_sep <- ratio_samp_sep %>%
 
 ratio_sep <- rbind(ratio_sol_sep, ratio_samp_sep)
 ratio_sep$type <- c(rep("solution", 3), rep("sample", 3))
-#ratio_sep$ymax_wheat <- cumsum(ratio_sep$wheat)
-#ratio_sep$ymin <- c(0, head(ratio_sep$ymax_wheat, n = -1))
-
 
 # Pie charts --------------------------------------------------------------
 
@@ -63,22 +60,18 @@ ratio_sep$type <- c(rep("solution", 3), rep("sample", 3))
 pl_wheat1 <- ratio_sep %>%
   select(!contains("potato")) %>%
   ggplot(aes(x = type, y = wheat, fill = size)) +
-  #geom_rect(xmin = 2, xmax = 3, ymin = ratio_sep$ymin, ymax = ratio_sep$ymax_wheat) +
   geom_bar(stat = "identity", width = 1, colour = "white") +
   coord_polar(theta = "y") +
   theme_void() +
-  #geom_text(aes(x = 2, label = label_wheat)) +
-  scale_fill_viridis_d(name = "Size", labels = c("Large", "Medium", "Small"))
+  scale_fill_viridis_d(name = "Size", labels = c("large", "medium", "small"))
 
 pl_potato1 <- ratio_sep %>%
   select(!contains("wheat")) %>%
   ggplot(aes(x = type, y = potato, fill = size)) +
-  #geom_rect(xmin = 2, xmax = 3, ymin = ratio_sep$ymin, ymax = ratio_sep$ymax_wheat) +
   geom_bar(stat = "identity", width = 1, colour = "white") +
   coord_polar(theta = "y") +
   theme_void() +
-  #geom_text(aes(x = 2, label = label_wheat)) +
-  scale_fill_viridis_d(name = "Size", labels = c("Large", "Medium", "Small"))
+  scale_fill_viridis_d(name = "Size", labels = c("large", "medium", "small"))
 
 # transpose solution data frame
 ratio_sol_mix <- sol_sep %>%
@@ -133,31 +126,23 @@ diff_mix <- ratio_diff_mix %>%
 pl_wheat2 <- ratio_mix %>%
   select(!contains("potato")) %>%
   ggplot(aes(x = type, y = wheat, fill = size)) +
[…]

### 2021-10-13 analysis/paper/fig_tab.Rmd
AGH final comments on first draft

@@
     toc: no
+  bookdown::html_document2:
+    toc: no
 ---
@@
 ```{r setup-figs-and-tbls, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE,
+                      message = FALSE)
 library(tidyverse)
@@
 source(here("analysis/scripts/figures.R"))
-#code to include R scripts in .Rmd: r knitr::spin_child("tables.R")
 ```
@@
 <!-- ### Tables ### -->
+
+```{r anova-tbl}
+knitr::kable(anova_tbl, 
+             col.names = c("Treatment", "Mean", "SD", "Min", "Max"),
+             caption = "Summary statistics for biofilm dry-weights (in mg) by treatment.")
+```
+
+```{r sample-count}
+count_sample <- count_sample %>%
+  ungroup() %>%
+  add_row(count_sample_comb, .after = 2)
+
+# convert columns to contain mean count and percentage of total count
+count_samp_per <- count_sample %>%
+  mutate(across(c(s, m, l, total), 
+                function(x) 
+                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), ")")))
+
+count_samp_out <- tibble(count_samp_per, count_sample_sd, .name_repair = "unique") %>%
+  select(1,2,3,7,4,8,5,9,6,10) #%>%
+
+count_samp_out[1,3] <- NA # replace with NA
+  #add_row(tibble_row(count_sample_comb_out), .before = 3)
+```
+
+```{r solution-count-out}
+count_sol_out <- count_sol %>%
+  mutate(across(c(s, m, l, total), 
+                function(x) 
+                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
+
+### adding the 'both' row to the data frame ###
+count_sol_comb_out <- count_sol_comb %>%
+  mutate(across(c(s, m, l, total), 
+                function(x) 
+                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
+
+count_sol_out <- count_sol %>%
+  mutate(across(c(s, m, l, total), 
+                function(x) 
+                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")"))) %>%
+  ungroup() %>%
+  add_row(tibble_row(count_sol_comb_out), .before = 3)
+
+count_sol_out[1,3] <- NA
+```
+
+```{r solution-count-tbl}
+cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
+options(knitr.kable.NA = "")
+knitr::kable(count_sol_out, col.names = cnames, 
+             caption = "Mean starch counts from solutions, including the
+             proportional makeup of the different sizes of granules.")
+```
+
+```{r sample-count-tbl}
+cnames <- c("Treatment", "Starch", "Small (%)", "SD", "Medium (%)", "SD", 
+            "Large (%)", "SD", "Total (%)", "SD")
+knitr::kable(count_samp_out, col.names = cnames, 
+             caption = "Mean starch counts extracted from samples with standard 
+             deviation (SD), including the proportion of granule sizes of the total count.")
+```
 
@@
 
-```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring).
-l = large, m = medium, s = small."}
-#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
-pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+```{r protocol-fig, fig.cap="Overview of experiment protocol including the plate setup."}
+knitr::include_graphics(here("analysis/figures/protocol_overview.png"))
 ```
 
-```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring).
-l = large, m = medium, s = small."}
-#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
-pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+```{r microscope-fig, fig.cap="Microscope image of a biofilm sample exposed to the potato starch solution.
+Potato granules can be seen within a bacterial community.
+Scale bar = 20 $\\mu$m.", out.width="50%"}
+knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
 ```
 
-```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments."}
+```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."}
+#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
+(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+```
+
+```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments."}
 pl_cor
+#pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
 ```

### 2021-10-13 analysis/paper/paper.Rmd
AGH final comments on first draft

@@
 and non-human primates [@powerChimpCalculus2015], 
-have explored the how well microremains (phytoliths and starches)
+have explored how well microremains (phytoliths and starches)
 extracted from dental calculus represent the actual dietary intake.
@@
 collected for further analysis. The setup comprises a polypropylene 
-24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which 
-is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins.
+24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which 
+is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins. The individual pegs 
+were the substrata on which the calculus grew. Using this system allowed for easy
+transfer of the growing biofilm between saliva, feeding solutions, 
+and mineral solutions.
 
@@
 Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
-with no history of caries, who abstained from oral hygiene for 24 hours, and no 
-food was consumed two hours prior to donation. No antibiotics were taken up to six 
-months prior to donation. 
+with no history of caries, who abstained from oral hygiene for 24 hours. No 
+food was consumed two hours prior to donation and no antibiotics were taken up 
+to six months prior to donation. 
 The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15% active chlorine) 
@@
 calcium phosphate monofluorophosphate urea (CPMU) solution containing 
-20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea, 
+20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
 and (0.04 g/l MgCl)
@@
 to prevent starch and bacterial contamination. Control samples that only received
-sucrose as a treatment, were included to detect starch contamination from the
-environment, or cross-contamination from other wells in the same plate.
+sucrose as a treatment were included to detect starch contamination from the
+environment or cross-contamination from other wells in the same plate.
 
@@
 wheat starches smaller than ca. 10 $\mu$m. These were counted as wheat,
-as we assumed that the majority of the small granules were wheat. This is a
-reasonable assumption to make based on the small granule counts from the separate
-treatment solutions, where the wheat granules made up 
+as we assumed that the majority of the small granules were wheat. We make this 
+assumption based on the counts of small starches in the wheat-only and potato-only
+solutions. Of the total amount of small starches in these two solutions, 
 `r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`
-of all small granule counts in the separated solutions.
+are from wheat.
 <!-- count of small granules from wheat solution divided by the sum of small granules 
@@
 
-```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.", out.width="80%"}
+```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."}
 #plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
@@
 
-```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments.", out.width="80%"}
-pl_cor
-#pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments."}
+#pl_cor
+pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
 ```
@@
 
-```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus.", out.width="80%"}
+```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
 pl_cor2
@@
 of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
-diameter were underrepresented in the calculus samples compared to the original 
+maximum length were underrepresented in the calculus samples compared to the original 
 starch solutions, an effect that was consistent across all three treatments, 
@@
 during mastication may be retained in the dentition through attachment to 
-tooth surfaces (including plaque and dental calulus) and mucous membranes 
+tooth surfaces (including plaque and dental calculus) and mucous membranes 
 [@kashketFoodRetention1991; @doddsCarbohydrateRetention1988].
@@
 <!-- limitations of the platform show which questions would be unanswered -->
-A limitation of this study 
 The absence of $\alpha$-amylase in the model is a limitation of this study, as 

### 2021-10-13 analysis/scripts/figures.R

@@ -182,16 +182,16 @@ pl_cor <- ggplot(z_count,
        colour = "Treatment",
        shape = "Treatment") +
   scale_colour_viridis_d()
-# ## Multi-plot figure test ##
-# pl_cor <- ggplot(z_count,
-#         aes(y = weight, x = std_total, col = treatment, shape = treatment)) +
-#   geom_point(size = 2) +
-#   theme_bw() +
-#   labs(y = "Weight (mg)",
-#        x = "Total starch count (z-score)") +
-#   scale_colour_viridis_d() +
-#   theme(axis.title.x = element_text(size = 10))
-# ## End: Multi-plot figure test ##
+## Multi-plot figure test ##
+pl_cor <- ggplot(z_count,
+        aes(y = weight, x = std_total, col = treatment, shape = treatment)) +
+  geom_point(size = 2) +
+  theme_bw() +
+  labs(y = "Weight (mg)",
+       x = "Total starch count (z-score)") +
+  scale_colour_viridis_d() +
+  theme(axis.title.x = element_text(size = 10))
+## End: Multi-plot figure test ##
 
 # plot of correlation between weight and starches per mg calculus
 pl_cor2 <- starch_per_mg %>%
@@ -205,15 +205,15 @@ pl_cor2 <- starch_per_mg %>%
        shape = "Treatment") +
     scale_colour_viridis_d()
 
-# ## Multi-plot figure test ##
-# pl_cor2 <- starch_per_mg %>%
-#   filter(treatment != "control") %>%
-#   ggplot(aes(y = weight, x = std_starch_per, col = treatment, shape = treatment)) +
-#   geom_point(size = 2) +
-#   theme_bw() +
-#   labs(y = "Weight (mg)",
-#        x = "Starch count per mg of calculus (Z-score)") +
-#   scale_colour_viridis_d() +
-#   theme(axis.title.y = element_blank(),
-#         axis.text.y = element_blank(),
-#         axis.title.x = element_text(size = 10))
+## Multi-plot figure test ##
+pl_cor2 <- starch_per_mg %>%
+  filter(treatment != "control") %>%
+  ggplot(aes(y = weight, x = std_starch_per, col = treatment, shape = treatment)) +
+  geom_point(size = 2) +
+  theme_bw() +
+  labs(y = "Weight (mg)",
+       x = "Starch count per mg of calculus (Z-score)") +
+  scale_colour_viridis_d() +
+  theme(axis.title.y = element_blank(),
[…]

### 2021-10-13 analysis/scripts/tables.R

@@ -75,16 +75,6 @@ count_sample_sd <- corr_counts %>% # create a data frame with treatment count SD
   select(s, m, l, total) %>%
   signif(digits = 3)
 
-# count_sample_comb <- corr_counts %>% # create a data frame with treatment count means
-#   filter(treatment == "mix") %>%
-#   group_by(starch) %>%
-#   summarise(across(c(s, m, l, total), mean, na.rm = T)) %>%
-#   ungroup() %>%
-#   summarise(across(c(s, m, l, total), sum, na.rm = T)) %>%
-#   mutate(starch = "both",
-#          treatment = "mix") %>%
-#   select(treatment, starch, s, m, l, total)
-
 # Proportional counts -----------------------------------------------------
 
 ### for combined mix counts ###

### 2021-10-22 analysis/paper/paper.Rmd
reduce code in paper

@@
 output:
-  bookdown::html_document2:
-    df_print: kable
-    keep_md: true
   bookdown::pdf_document2:
@@
     #extra_dependencies: "subfig"
+  bookdown::html_document2:
+    df_print: kable
+    keep_md: true
   bookdown::word_document2:
@@
 slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts 
-were extrapolated (see Supplementary material for more details). For potato
+were extrapolated to the total number of granules exposed to the samples over 16
+days of treatments (see Supplementary Material for more details). For potato
 treatment samples, the whole slide was counted.
@@
 the pipette to ensure that the counted transects were representative of the 
-whole slide. Wheat and mix counts were extrapolated as described above.
+whole slide. The count from the slide was extrapolated to the whole sample
+(see Supplementary Material).
+
 Both wheat and potato granules were divided into three size categories: 
@@
 
-Pearson's r was conducted on sample weight and total starch count, as well as sample
+Pearson's *r* was conducted on sample weight and total starch count, as well as sample
 weight and starch count per mg calculus. The total count for each sample within a
@@
 ```{r solution-count-out}
-count_sol_out <- count_sol %>%
+# sol_comb_row <- sol_comb_out %>%
+#   group_by(treatment) %>%
+#   filter(treatment == "mix") %>%
+#   summarise(starch = "both",
+#             across(where(is.numeric), sum, na.rm = T))
+
+sol_tbl <- sol_comb_out %>%
+  #ungroup() %>%
+  #add_row(sol_comb_row, .after = 2) %>%
   mutate(across(c(s, m, l, total), 
@@
 
-### adding the 'both' row to the data frame ###
-count_sol_comb_out <- count_sol_comb %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
+# count_sol_out <- count_sol %>%
+#   mutate(across(c(s, m, l, total), 
+#                 function(x) 
+#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
+# 
+# ### adding the 'both' row to the data frame ###
+# count_sol_comb_out <- count_sol_comb %>%
+#   mutate(across(c(s, m, l, total), 
+#                 function(x) 
+#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
+# 
+# count_sol_out <- count_sol %>%
+#   mutate(across(c(s, m, l, total), 
+#                 function(x) 
+#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")"))) %>%
+#   ungroup() %>%
+#   add_row(tibble_row(count_sol_comb_out), .before = 3)
 
-count_sol_out <- count_sol %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")"))) %>%
-  ungroup() %>%
-  add_row(tibble_row(count_sol_comb_out), .before = 3)
-
-count_sol_out[1,3] <- NA
+#count_sol_out[1,3] <- NA
+sol_tbl[1,3] <- NA
 ```
@@
 ```{r sample-count}
-count_sample <- count_sample %>%
+samp_counts <- corr_comb_long %>%
+  pivot_wider(names_from = size, values_from = count) %>%
   ungroup() %>%
-  add_row(count_sample_comb, .after = 2)
+  add_row(select(samp_comb_row, !sample), .after = 2)
+
+samp_sd_wide <- corr_sd_long %>%
+  select(!count) %>%
+  pivot_wider(names_from = size, values_from = sd) %>%
+  ungroup() %>%
+  add_row(samp_sd_row, .after = 2) %>%
+  mutate(across(where(is.numeric), round)) %>%
+  rename(s_sd = s,
+         m_sd = m,
+         l_sd = l,
+         total_sd = total)
+
+samp_counts_per <- samp_counts %>%
+  mutate(across(c(s, m, l, total), 
+                function(x) 
+                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), 
+                         ")")))
+
+# combine samp_counts_per with samp_sd_wide
+samp_counts_tbl <- samp_counts_per %>%
+  bind_cols(select(samp_sd_wide, c(s_sd, m_sd, l_sd, total_sd)), .name_repair = "unique") %>%
+  select(treatment, starch, s, s_sd, m, m_sd, l, l_sd, total, total_sd)
+
+# count_sample <- count_sample %>%
+#   ungroup() %>%
+#   add_row(samp_sd_row, .after = 2)
 
 # convert columns to contain mean count and percentage of total count
-count_samp_per <- count_sample %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), ")")))
+# count_samp_per <- count_sample %>%
+#   mutate(across(c(s, m, l, total), 
+#                 function(x) 
+#                   paste0(round(x), " (", scales::percent(signif(x / total, 3)), ")")))
 
-count_samp_out <- tibble(count_samp_per, count_sample_sd, .name_repair = "unique") %>%
-  select(1,2,3,7,4,8,5,9,6,10) #%>%
+# count_samp_out <- tibble(count_samp_per, count_sample_sd, .name_repair = "unique") %>%
+#   select(1,2,3,7,4,8,5,9,6,10) #%>%
 
-count_samp_out[1,3] <- NA # replace with NA
+samp_counts_tbl[1,3] <- NA
+#count_samp_out[1,3] <- NA # replace with NA
   #add_row(tibble_row(count_sample_comb_out), .before = 3)
@@
 solutions. Of the total amount of small starches in these two solutions, 
-`r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`
+`r paste0(signif((sol_comb_out$s[4] / sum(sol_comb_out$s[3], sol_comb_out$s[4])) * 100, 3), "%")`
 are from wheat.
@@
 options(knitr.kable.NA = "")
-knitr::kable(count_sol_out, col.names = cnames, 
+knitr::kable(sol_tbl, col.names = cnames, 
              caption = "Mean starch counts from solutions, including the
@@
             "Large (%)", "SD", "Total (%)", "SD")
-knitr::kable(count_samp_out, col.names = cnames, 
+knitr::kable(samp_counts_tbl, col.names = cnames, 
              caption = "Mean starch counts extracted from samples with standard 
@@
 starch, with a total concentration of 0.50% (w/v).
-The mixed treatment had the highest absolute count of starch granules in solution (mean = `r signif(subset(count_sol, solution == "mix")$total, 3)`),
+The mixed treatment had the highest absolute count of starch granules in solution (mean = `r signif(sol_comb_row$total, 3)`),
 while the biofilms exposed to the wheat solution preserved the greatest number of
 granules 
-(mean = `r signif(subset(count_sample, treatment == "wheat")$total, 3)`). 
+(mean = `r signif(subset(samp_counts, treatment == "wheat")$total, 3)`). 
 The potato treatment had the lowest absolute counts in both the solution 
-(`r signif(subset(count_sol, solution == "potato")$total, 3)`)
+(`r signif(subset(sol_comb_out, treatment == "potato")$total, 3)`)
 and in the biofilm samples 
-(`r signif(subset(count_sample, treatment == "potato")$total, 3)`) 
+(`r signif(subset(samp_counts, treatment == "potato")$total, 3)`) 
 (Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
@@
 # convert proportions to percent labels
-perc_comb <- prop_comb %>% 
-  mutate(across(c(s, m, l, total), scales::percent, 0.001))
+# perc_comb <- prop_comb %>% 
+#   mutate(across(c(s, m, l, total), scales::percent, 0.001))
 
-perc <- prop %>%
-  mutate(across(c(s, m, l, total), scales::percent, 0.001)) %>% 
-  add_row(tibble_row(perc_comb), .before = 3)
+# perc <- prop %>%
+#   mutate(across(c(s, m, l, total), scales::percent, 0.001)) %>% 
+#   add_row(tibble_row(perc_comb), .before = 3)
+
+prop_wide <- prop %>%
+  ungroup() %>%
+  pivot_wider(id_cols = c(treatment, starch), 
+              names_from = size, 
+              values_from = prop) %>%
+  add_row(prop_comb_row, .after = 2)
+
+perc_tbl <- prop %>%
+  ungroup() %>%
+  select(!c(sample, solution, prop)) %>%
+  pivot_wider(names_from = size, values_from = perc) %>%
+  add_row(perc_comb_row, .after = 2)
+# samp_comb_row <- corr_comb_long %>%
+#   pivot_wider(names_from = size, values_from = count) %>%
+#   group_by(treatment) %>%
+#   filter(treatment == "mix") %>%
+#   summarise(starch = "both",
+#             across(where(is.numeric), sum, na.rm = T)) %>%
+#   mutate(sample = "sample")
+
+# perc_comb_row <- sol_comb_row %>%
+#   mutate(sample = "solution") %>%
+#   add_row(samp_comb_row) %>%
+#   pivot_longer(cols = c(s, m, l, total), names_to = "size")
+#   mutate()
+
+#perc <- prop %>%
+#  pivot_wider(id_cols = c(treatment, starch), names_from = size, values_from = perc)
 ```
@@
 cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
-knitr::kable(perc, col.names = cnames, 
+knitr::kable(perc_tbl, col.names = cnames, 
              caption = "The mean percentage of starches from the solutions that 
@@
 into the samples ranged from 
-`r paste(min(perc$total), max(perc$total), sep = " to ")`,
+`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`,
 with potato granules being more readily incorporated than wheat in both the
@@
 Wheat incorporation was most affected in the mixed-treatment samples, with only
-`r paste0(perc$total[2])` of the total available starches being incorporated into
+`r paste0(perc_tbl$total[2])` of the total available starches being incorporated into
 the sample, compared to
-`r paste0(perc$total[4])`
+`r paste0(perc_tbl$total[4])`
 in the separated wheat treatment.
@@
 
+```{r}
+sol_size_prop <- sol_comb_out %>%
+  group_by(treatment, starch) %>%
+  #select(!total) %>%
+  mutate(across(where(is.numeric), function(x) x / total)) %>%
+  mutate(type = "solution")
+
+samp_size_prop <- samp_counts %>%
+  group_by(treatment, starch) %>%
+  #select(!total) %>%
+  mutate(across(where(is.numeric), function(x) x / total)) %>%
+  mutate(type = "sample")
+
+size_diff <- sol_size_prop %>%
+  full_join(samp_size_prop) %>%
+  pivot_longer(cols = c(s,m,l,total), names_to = "size", values_to = "count") %>%
+  pivot_wider(names_from = type, values_from = count) %>%
+  mutate(diff = (sample - solution) * 100) %>%
+  select(!c(solution, sample)) %>%
+  pivot_wider(names_from = size, values_from = diff) %>%
+  select(!total)
+```
+
 Overall, medium starch granules had a higher mean rate of incorporation 
-(`r scales::percent(mean(prop$m, na.rm = T), 0.001)`) 
+(`r scales::percent(mean(prop_wide$m, na.rm = T), 0.001)`) 
 than small 
-(`r scales::percent(mean(prop$s, na.rm = T), 0.001)`) 
+(`r scales::percent(mean(prop_wide$s, na.rm = T), 0.001)`) 
 and large 
-(`r scales::percent(mean(prop$l, na.rm = T), 0.001)`) 
+(`r scales::percent(mean(prop_wide$l, na.rm = T), 0.001)`) 
 starch granules across all treatments, while large potato starches had the lowest 
@@
 Large potato granules (> 20 $\mu$m) were most affected, with a 
-`r paste0(abs(diff_sep$potato[3]), "%")` 
+`r paste0(abs(signif(size_diff$l[4]), 3), "%")` 
 decrease in relative abundance in the potato-only treatment, and a
-`r paste0(abs(diff_mix$potato[3]), "%")` 
+`r paste0(abs(signif(size_diff$l[1]), 3), "%")` 
 decrease in mixed treatments. Medium granules increased in relative abundance 
@@
 
-```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments."}
+```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by z-score for separated treatments."}
 #pl_cor
@@
 (of all sizes) present in the extracted samples were similar across all treatments
-(`r paste(min(perc$total), max(perc$total), sep = " to ")`), 
+(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`), 
 despite large differences in absolute granule counts between wheat 
-(mean = `r format(count_sol$total[4], big.mark = ",")`) 
+(mean = `r format(sol_comb_out$total[5], big.mark = ",")`) 
 and potato
-(mean = `r format(count_sol$total[3], big.mark = ",")`)
+(mean = `r format(sol_comb_out$total[4], big.mark = ",")`)
 solutions.  
@@
 Despite the low proportion of granules recovered from the model calculus 
-(`r paste(min(perc$total), max(perc$total), sep = " to ")`),
+(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`),
 the absolute counts were still substantially greater than counts recovered from 
@@
 Since we did NOT make use of Sci-Hub to access articles stuck behind a paywall, 
-we will NOT acknowledge the use of Sci-Hub in this study.
+we will NOT acknowledge the use of Sci-Hub in this study. <!-- Oops, did I just acknowledge it twice..? -->
 

### 2021-10-22 analysis/paper/paper.Rmd
reduce code in paper

@@
 Large potato granules (> 20 $\mu$m) were most affected, with a 
-`r paste0(abs(signif(size_diff$l[4]), 3), "%")` 
+`r paste0(abs(signif(size_diff$l[4], 3)), "%")` 
 decrease in relative abundance in the potato-only treatment, and a
-`r paste0(abs(signif(size_diff$l[1]), 3), "%")` 
+`r paste0(abs(signif(size_diff$l[1], 3)), "%")` 
 decrease in mixed treatments. Medium granules increased in relative abundance 
@@
 archaeological remains 
-[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010], <!-- although this study extrapolated counts to full sample content, which is
-rarely done in archaeological studies -->which could in part be due to the 
+[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010], which could in part be due to the 
 lack of oral amylase activity in our model.

### 2021-10-22 analysis/paper/paper.tex
reduce code in paper

@@
 
-Word count: 4722
-
 \hypertarget{introduction}{%
@@
 and non-human primates (Power et al., 2015),
-have explored the how well microremains (phytoliths and starches)
+have explored how well microremains (phytoliths and starches)
 extracted from dental calculus represent the actual dietary intake.
@@
 collected for further analysis. The setup comprises a polypropylene
-24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which
-is autoclaved at 120°C, 1 bar overpressure, for 20 mins.
+24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which
+is autoclaved at 120°C, 1 bar overpressure, for 20 mins. The individual pegs
+were the substrata on which the calculus grew. Using this system allowed for easy
+transfer of the growing biofilm between saliva, feeding solutions,
+and mineral solutions.
 
@@
 Fresh whole saliva (WS) for inoculation was provided by a 31-year-old male donor
-with no history of caries, who abstained from oral hygiene for 24 hours, and no
-food was consumed two hours prior to donation. No antibiotics were taken up to six
-months prior to donation.
+with no history of caries, who abstained from oral hygiene for 24 hours. No
+food was consumed two hours prior to donation and no antibiotics were taken up
+to six months prior to donation.
 The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15\% active chlorine)
@@
 calcium phosphate monofluorophosphate urea (CPMU) solution containing
-20 mmol/l CaCl\textsubscript{2}, 12 mmol/l NaH\textsubscript{2}PO\textsubscript{4}, 5 mmol/l Na\textsubscript{2}PO\textsubscript{3}F, 500 mmol/l Urea,
+20 mmol/l CaCl\textsubscript{2}, 12 mmol/l NaH\textsubscript{2}PO\textsubscript{4}, 5 mmol/l Na\textsubscript{2}PO\textsubscript{3}F, 500 mmol/l urea,
 and (0.04 g/l MgCl)
@@
 slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts
-were extrapolated (see Supplementary material for more details). For potato
+were extrapolated to the total number of granules exposed to the samples over 16
+days of treatments (see Supplementary Material for more details). For potato
 treatment samples, the whole slide was counted.
@@
 the pipette to ensure that the counted transects were representative of the
-whole slide. Wheat and mix counts were extrapolated as described above.
+whole slide. The count from the slide was extrapolated to the whole sample
+(see Supplementary Material).
+
 Both wheat and potato granules were divided into three size categories:
@@
 
-Pearson's r was conducted on sample weight and total starch count, as well as sample
+Pearson's \emph{r} was conducted on sample weight and total starch count, as well as sample
 weight and starch count per mg calculus. The total count for each sample within a
@@
 wheat starches smaller than ca. 10 \(\mu\)m. These were counted as wheat,
-as we assumed that the majority of the small granules were wheat. This is a
-reasonable assumption to make based on the small granule counts from the separate
-treatment solutions, where the wheat granules made up
-99.2\%
-of all small granule counts in the separated solutions.
+as we assumed that the majority of the small granules were wheat. We make this
+assumption based on the counts of small starches in the wheat-only and potato-only
+solutions. Of the total amount of small starches in these two solutions,
+0.653\%
+are from wheat.
 
@@
 \hline
-mix & potato &  &  & 1959 (79.6\%) & 1800 & 501 (20.40\%) & 446 & 2460 (100\%) & 2190\\
-\hline
-mix & wheat & 9515 (54.60\%) & 8860 & 6522 (37.4\%) & 6030 & 1381 (7.93\%) & 1200 & 17417 (100\%) & 15900\\
-\hline
-mix & both & 9515 (47.90\%) & 8860 & 8480 (42.7\%) & 7650 & 1882 (9.47\%) & 1600 & 19877 (100\%) & 17800\\
-\hline
-potato & potato & 351 (7.24\%) & 297 & 3565 (73.6\%) & 2400 & 930 (19.20\%) & 929 & 4846 (100\%) & 3320\\
-\hline
-wheat & wheat & 15235 (55.00\%) & 11900 & 12148 (43.9\%) & 11100 & 1953 (7.06\%) & 2020 & 27680 (100\%) & 23600\\
+mix & potato &  &  & 1959 (79.6\%) & 1801 & 501 (20.40\%) & 446 & 2460 (100\%) & 2189\\
+\hline
+mix & wheat & 9515 (54.60\%) & 8860 & 6522 (37.4\%) & 6026 & 1381 (7.93\%) & 1196 & 17417 (100\%) & 15878\\
+\hline
+mix & both & 9515 (47.90\%) & 8860 & 8480 (42.7\%) & 4929 & 1882 (9.47\%) & 989 & 19877 (100\%) & 13451\\
+\hline
+potato & potato & 351 (7.24\%) & 297 & 3565 (73.6\%) & 2402 & 930 (19.20\%) & 929 & 4846 (100\%) & 3316\\
+\hline
+wheat & wheat & 15235 (55.00\%) & 11944 & 12148 (43.9\%) & 11052 & 1953 (7.06\%) & 2016 & 27680 (100\%) & 23554\\
 \hline
@@
 starch, with a total concentration of 0.50\% (w/v).
-The mixed treatment had the highest absolute count of starch granules in solution (mean = \ensuremath{1.98\times 10^{6}}, \ensuremath{2.7\times 10^{7}}),
+The mixed treatment had the highest absolute count of starch granules in solution (mean = \ensuremath{2.9\times 10^{7}}),
 while the biofilms exposed to the wheat solution preserved the greatest number of
@@
 \hline
-mix & potato &  & 0.186\% & 0.054\% & 0.124\%\\
-\hline
-mix & wheat & 0.051\% & 0.102\% & 0.077\% & 0.064\%\\
-\hline
-mix & both & 0.051\% & 0.114\% & 0.069\% & 0.069\%\\
-\hline
-potato & potato & 0.284\% & 0.266\% & 0.060\% & 0.161\%\\
-\hline
-wheat & wheat & 0.094\% & 0.189\% & 0.069\% & 0.109\%\\
+mix & potato &  & 0.19\% & 0.05\% & 0.12\%\\
+\hline
+mix & wheat & 0.05\% & 0.10\% & 0.08\% & 0.06\%\\
+\hline
+mix & both & 0.05\% & 0.11\% & 0.07\% & 0.07\%\\
+\hline
+potato & potato & 0.28\% & 0.27\% & 0.06\% & 0.16\%\\
+\hline
+wheat & wheat & 0.09\% & 0.19\% & 0.07\% & 0.12\%\\
 \hline
@@
 into the samples ranged from
-0.064\% to 0.161\%,
+0.06\% to 0.16\%,
 with potato granules being more readily incorporated than wheat in both the
@@
 Wheat incorporation was most affected in the mixed-treatment samples, with only
-0.064\% of the total available starches being incorporated into
+0.06\% of the total available starches being incorporated into
 the sample, compared to
-0.161\%
+0.16\%
 in the separated wheat treatment.
@@
 
-{\centering \includegraphics[width=0.8\linewidth]{../figures/ratio-plots-1}
+{\centering \includegraphics{../figures/ratio-plots-1}
 
@@
 Overall, medium starch granules had a higher mean rate of incorporation
-(0.186\%)
+(0.171\%)
 than small
-(0.143\%)
+(0.120\%)
 and large
-(0.065\%)
+(0.066\%)
 starch granules across all treatments, while large potato starches had the lowest
@@
 
-{\centering \includegraphics[width=0.8\linewidth]{../figures/cor-plot-1}
+{\centering \includegraphics{../figures/cor-plot-1}
 
@@
 
-\caption{Scatter plot of sample weight and standardised starch count by Z-score for separated treatments.}\label{fig:cor-plot}
+\caption{Scatter plot of sample weight and standardised starch count by z-score for separated treatments.}\label{fig:cor-plot}
 \end{figure}
@@
 
-{\centering \includegraphics[width=0.8\linewidth]{../figures/cor-plot2-1}
+{\centering \includegraphics{../figures/cor-plot2-1}
 
@@
 (of all sizes) present in the extracted samples were similar across all treatments
-(0.064\% to 0.161\%),
+(0.06\% to 0.16\%),
 despite large differences in absolute granule counts between wheat
@@
 Despite the low proportion of granules recovered from the model calculus
-(0.064\% to 0.161\%),
+(0.06\% to 0.16\%),
 the absolute counts were still substantially greater than counts recovered from
@@
 of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
-diameter were underrepresented in the calculus samples compared to the original
+maximum length were underrepresented in the calculus samples compared to the original
 starch solutions, an effect that was consistent across all three treatments,
@@
 during mastication may be retained in the dentition through attachment to
-tooth surfaces (including plaque and dental calulus) and mucous membranes
+tooth surfaces (including plaque and dental calculus) and mucous membranes
 (Dodds \& Edgar, 1988; S. Kashket et al., 1991).
@@
 
-A limitation of this study
 The absence of \(\alpha\)-amylase in the model is a limitation of this study, as
@@
 
-\hypertarget{data-availability-statement}{%
-\section*{Data Availability Statement}\label{data-availability-statement}}
-\addcontentsline{toc}{section}{Data Availability Statement}
-
-All scripts and data used in the analysis are available on OSF
-(\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}),
-following the format provided by the rrtools package (Marwick, 2019).
-More detailed protocols are available on OSF and protocols.io.
-
 \hypertarget{acknowledgements}{%
@@
 number STG--677576 (``HARVEST'').
+
+\hypertarget{data-availability-statement}{%
+\section*{Data Availability Statement}\label{data-availability-statement}}
+\addcontentsline{toc}{section}{Data Availability Statement}
+
+All scripts and data used in the analysis are available on OSF
+(\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}),
+following the format provided by the rrtools package (Marwick, 2019).
+More detailed protocols are available on OSF and protocols.io.
 

### 2021-10-22 analysis/paper/paper.Rmd
remove redundant code from paper and R scripts

@@
 ```{r solution-count-out}
-# sol_comb_row <- sol_comb_out %>%
-#   group_by(treatment) %>%
-#   filter(treatment == "mix") %>%
-#   summarise(starch = "both",
-#             across(where(is.numeric), sum, na.rm = T))
-
 sol_tbl <- sol_comb_out %>%
-  #ungroup() %>%
-  #add_row(sol_comb_row, .after = 2) %>%
   mutate(across(c(s, m, l, total), 
@@
 
-# count_sol_out <- count_sol %>%
-#   mutate(across(c(s, m, l, total), 
-#                 function(x) 
-#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
-# 
-# ### adding the 'both' row to the data frame ###
-# count_sol_comb_out <- count_sol_comb %>%
-#   mutate(across(c(s, m, l, total), 
-#                 function(x) 
-#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
-# 
-# count_sol_out <- count_sol %>%
-#   mutate(across(c(s, m, l, total), 
-#                 function(x) 
-#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")"))) %>%
-#   ungroup() %>%
-#   add_row(tibble_row(count_sol_comb_out), .before = 3)
-
-#count_sol_out[1,3] <- NA
 sol_tbl[1,3] <- NA
@@
 
-# count_sample <- count_sample %>%
-#   ungroup() %>%
-#   add_row(samp_sd_row, .after = 2)
-
-# convert columns to contain mean count and percentage of total count
-# count_samp_per <- count_sample %>%
-#   mutate(across(c(s, m, l, total), 
-#                 function(x) 
-#                   paste0(round(x), " (", scales::percent(signif(x / total, 3)), ")")))
-
-# count_samp_out <- tibble(count_samp_per, count_sample_sd, .name_repair = "unique") %>%
-#   select(1,2,3,7,4,8,5,9,6,10) #%>%
-
 samp_counts_tbl[1,3] <- NA
-#count_samp_out[1,3] <- NA # replace with NA
-  #add_row(tibble_row(count_sample_comb_out), .before = 3)
 ```
@@
 solutions. Of the total amount of small starches in these two solutions, 
-`r paste0(signif((sol_comb_out$s[4] / sum(sol_comb_out$s[3], sol_comb_out$s[4])) * 100, 3), "%")`
+`r paste0(signif((sol_comb_out$s[5] / sum(sol_comb_out$s[4], sol_comb_out$s[5])) * 100, 3), "%")`
 are from wheat.
@@
 ```{r sample-perc}
-# convert proportions to percent labels
-# perc_comb <- prop_comb %>% 
-#   mutate(across(c(s, m, l, total), scales::percent, 0.001))
-
-# perc <- prop %>%
-#   mutate(across(c(s, m, l, total), scales::percent, 0.001)) %>% 
-#   add_row(tibble_row(perc_comb), .before = 3)
-
 prop_wide <- prop %>%
@@
   add_row(perc_comb_row, .after = 2)
-# samp_comb_row <- corr_comb_long %>%
-#   pivot_wider(names_from = size, values_from = count) %>%
-#   group_by(treatment) %>%
-#   filter(treatment == "mix") %>%
-#   summarise(starch = "both",
-#             across(where(is.numeric), sum, na.rm = T)) %>%
-#   mutate(sample = "sample")
-
-# perc_comb_row <- sol_comb_row %>%
-#   mutate(sample = "solution") %>%
-#   add_row(samp_comb_row) %>%
-#   pivot_longer(cols = c(s, m, l, total), names_to = "size")
-#   mutate()
-
-#perc <- prop %>%
-#  pivot_wider(id_cols = c(treatment, starch), names_from = size, values_from = perc)
 ```

### 2021-10-22 analysis/supplementary-materials/supp-mat.Rmd
remove redundant code from paper and R scripts

@@
 ---
-title: Supplementary Materials
+title: Supplementary Material
 output: 
@@
 library(tidyverse)
-raw_data_samples <- read_csv(here("analysis/data/raw_data/starch_counts.csv"))
-raw_data_solutions <- read_csv(here("analysis/data/raw_data/solution_counts.csv"))
+library(patchwork)
+source(here("analysis/scripts/analysis.R"))
+#raw_data_samples <- read_csv(here("analysis/data/raw_data/starch_counts.csv"))
+#raw_data_solutions <- read_csv(here("analysis/data/raw_data/solution_counts.csv"))
 ```
@@
 
-```{r plate-lid-2, fig.cap="The 24 deepwell plate with the lid on (sort of...)"}
+```{r plate-lid-2, fig.cap="The 24 deepwell plate with the lid (almost) on."}
 knitr::include_graphics(here("analysis/figures/plate_lid_on.jpg"))
@@
 
-**solution_counts**
+**solution_counts.csv**
 
@@
 ```{r raw-data-sol, echo=FALSE}
-raw_data_solutions
+sol_counts
 ```
@@
 ```{r raw-data-samp, echo=FALSE}
-raw_data_samples
+raw_counts
 ```
@@
 
-Tables containing the amylase activity results for both plates and photometric
-readings.
+Tables containing the amylase activity results for both plates and both photometric
+readings conducted on each plate. Samples (rows) were analysed in triplicates (columns).
 
@@
 source(here("analysis/scripts/amylase-assay.R"))
-cols <- c("1", "2", "3")  
+cols <- c("1", "2", "3")  # sample triplicates
 rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
@@
 ```{r}
-raw_data_samples %>%
+raw_counts %>%
   filter(treatment == "control") %>%
@@
 
-Solution counts were extrapolated to the quantity in a 1 ml solution with
-the following formula:
+A 1 mL aliquot of each of the original treatment solutions was taken, from which
+10 $\mu$L was taken and mounted on a microscope slide and mixed with 10 $\mu$L 
+20% (v/v) glycerol. Solution counts were extrapolated from a slide (10 $\mu$L) 
+to the quantity in a 1 ml solution,
+and then multipled by 16 days to achieve the total number of granules that were
+exposed to the samples:
 
-$$ \text{corrected count} = \text{raw count} \times \frac{\text{total} \space \text{transects}}{\text{counted} \space \text{transects}} \times 100 \mu l $$
+$$ \text{corrected count} = \text{raw count} \times \frac{\text{total slides}}{\text{counted slides}} \times 100 \mu L \times 16 \text{ days} $$
 
-Sample counts were extrapolated using the following formula:
+Samples were submerged in 50--100 $\mu$L EDTA, from which 20 $\mu$L was mounted 
+on a microscope slide ($V_{slide}$) and counted. Sample counts were extrapolated to the full
+volume of EDTA ($V_{sample}$) in which the sample was submerged (i.e. 50--100 $\mu$L).
 
 $$ \text{Corrected count} = \text{raw count} \times (\text{portion of slide})^{-1} \times \frac{V_{sample}}{V_{slide}} $$
+
+## Some additional plots
+
+Bar plot for the total count of granules exposed to the samples over the
+duration of the experiment,
+
+```{r solution-bar-pl}
+sol_comb_long %>%
+  filter(size == "total") %>%
+  group_by(treatment, starch) %>%
+  ggplot(aes(x = treatment, y = count, fill = treatment, col = starch)) +
+  geom_col(size = 1.5) +
+  theme(panel.background = element_rect(fill = "white"),
+        panel.grid = element_line(colour = "grey"),
+        panel.grid.major.x = element_blank(),
+        axis.title.x = element_blank()) +
+  scale_fill_viridis_d() +
+  scale_color_viridis_d(begin = 0.5)
+```
+
+and box plot with superimposed points (with added jitter) for the extrapolated
+mean counts of granules extracted from the samples.
+
+```{r sample-box-plot}
+corr_comb %>%
+  filter(treatment != "control") %>%
+  ggplot(aes(x = treatment, y = total, 
+             shape = treatment)) +
+    geom_boxplot(aes(fill = treatment), alpha = 0.5) +
+    geom_jitter(aes(col = treatment), width = 0.3, size = 2) +
+    scale_color_viridis_d() +
+    theme(panel.background = element_rect(fill = "white"),
+        panel.grid = element_line(colour = "grey"),
+        panel.grid.major.x = element_blank(),
+        axis.title.x = element_blank()) + # remove y-axis title
+    scale_fill_viridis_d()
+```
+
+Extracted-granule counts separated by size:
+
+```{r}
+corr_counts_long %>%
+  filter(size != "total",
+         treatment != "control") %>%
+  group_by(treatment, starch, size) %>%
+  summarise(sd = sd(count, na.rm = T),
+            count = mean(count, na.rm = T)) %>%
+  #mutate(percent = count / sum(count, na.rm = T) * 100) %>% 
+  ggplot(aes(x = starch, y = count, fill = size)) +
+    geom_col(position = "dodge") +
+    geom_errorbar(aes(ymin = count, ymax = count + sd), width = 0.2, position = position_dodge(0.9)) +
+    facet_wrap(~ treatment, scales = "free") +
+    scale_fill_viridis_d() +
+    theme_bw()
+```
+
+Size distribution within the solutions (top) and samples (bottom):
+
+```{r message=FALSE, warning=FALSE}
+sol_size_pl <- sol_corr %>%
+  group_by(solution, starch) %>%
+  summarise(across(c(s, m, l, total), mean, na.rm = T)) %>%
+  pivot_longer(cols = c(s,m,l, total), values_to = "count", names_to = "size") %>%
+  filter(size != "total") %>%
+  group_by(solution, starch) %>%
+  mutate(percent = count / sum(count, na.rm = T) * 100) %>% 
+  ggplot(aes(x = starch, y = percent, fill = size)) +
+    geom_col(position = "dodge") +
+    facet_wrap(~ solution, scales = "free_x") +
+    scale_fill_viridis_d() +
+    theme_bw() +
+    labs(x = "")
+
+samp_size_pl <- corr_counts_long %>%
+  filter(size != "total",
+         treatment != "control") %>%
+  group_by(treatment, starch, size) %>%
+  summarise(count = mean(count, na.rm = T)) %>%
+  mutate(percent = count / sum(count, na.rm = T) * 100) %>% 
+  ggplot(aes(x = starch, y = percent, fill = size)) +
+    geom_col(position = "dodge") +
+    facet_wrap(~ treatment, scales = "free_x") +
+    scale_fill_viridis_d() +
+    theme_bw()
+
+sol_size_pl / samp_size_pl + plot_layout(guides = "collect")
+```
+
+Size distribution within the samples:
+
+```{r message=FALSE, warning=FALSE}
+samp_size_pl <- corr_counts_long %>%
+  filter(size != "total",
+         treatment != "control") %>%
+  group_by(treatment, starch, size) %>%
+  summarise(count = mean(count, na.rm = T)) %>%
+  mutate(percent = count / sum(count, na.rm = T) * 100) %>% 
+  ggplot(aes(x = starch, y = percent, fill = size)) +
+    geom_col(position = "dodge") +
+    facet_wrap(~ treatment, scales = "free_x") +
+    scale_fill_viridis_d() +
+    theme_bw()
+```

### 2021-10-22 analysis/scripts/analysis.R

@@ -23,8 +23,8 @@ correct_count <- function(raw, prop, vol){
   corrected
 }
 
-# function to extrapolate solution counts to 1 ml exposure per day
-
+# function to extrapolate solution counts to total exposure during the experiment
+  # i.e., 1 ml per day for 16 days
 extrap_count <- function(raw, prop, vol, days){
   extrap <- raw * (1 / prop) * 100 * days
   extrap
@@ -49,6 +49,27 @@ sol_corr <- sol_counts %>%
   mutate(across(c(s, m, l, total), extrap_count, portion_slide, vol_slide, days = 16))
 readr::write_csv(sol_corr, here("analysis/data/derived_data", "sol_corr.csv"))
 
+# combining potato and wheat counts from mixed treatment solution
+
+sol_comb <- sol_corr %>%
+  filter(solution != "mix") %>%
+  group_by(solution, starch) %>%
+  summarise(across(c("s","m","l","total"), mean)) %>%
+  ungroup() %>%
+  add_row(sol_corr[5:6,c("solution", "starch", "s","m","l","total")], ) %>%
+  rename(treatment = solution) %>% # I should have done better naming the raw data...
+  mutate(sample = "solution")
+
+sol_comb_long <- sol_corr %>%
+  group_by(solution, starch) %>%
+  summarise(across(c(s,m,l, total), mean, na.rm = T)) %>%
+  pivot_longer(cols = c(s,m,l, total), values_to = "count", names_to = "size") %>%
+  filter(size != "total") %>%
+  group_by(solution, starch) %>%
+  mutate(percent = count / sum(count, na.rm = T) * 100) %>%
+  rename(treatment = solution) # rename to be consistent with sample counts
+
+
 # Apply correction --------------------------------------------------------
 
 corr_counts <- raw_counts %>%
@@ -58,7 +79,9 @@ corr_counts <- raw_counts %>%
   filter(sample != "st2C6")
 readr::write_csv(corr_counts, here("analysis/data/derived_data", "corr_counts.csv"))
 
-# Combining mixed treatment counts ----------------------------------------
+corr_counts_long <- corr_counts %>%
+  #filter(treatment != "control") %>%
+  pivot_longer(cols = c(s,m,l, total), values_to = "count", names_to = "size")
 
 # combining potato and wheat counts from mixed treatment samples
 
[…]

### 2021-10-22 analysis/scripts/figures.R

@@ -1,148 +1,69 @@
 library(tidyverse)
+source(here("analysis/scripts/analysis.R"))
+#corr_counts <- readr::read_csv(here("analysis/data/derived_data/", "corr_counts.csv"))
+#corr_comb <- readr::read_csv(here("analysis/data/derived_data/", "corr_comb.csv"))
+#sol_corr <- readr::read_csv(here("analysis/data/derived_data/", "sol_corr.csv"))
 
-corr_counts <- readr::read_csv(here("analysis/data/derived_data/", "corr_counts.csv"))
-corr_comb <- readr::read_csv(here("analysis/data/derived_data/", "corr_comb.csv"))
-sol_corr <- readr::read_csv(here("analysis/data/derived_data/", "sol_corr.csv"))
 
-# group by starch category
-# the mix treatment/solution is separated into starch types
-sol_sep <- sol_corr %>%
-  group_by(solution, starch) %>%
-  summarise(across(c(s, m, l, total), mean))
-
-samp_sep <- corr_counts %>%
+sol_comb_long_pie <- sol_comb_long %>%
   group_by(treatment, starch) %>%
-  filter(treatment != "control") %>%
-  summarise(across(c(s, m, l, total), mean, na.rm = T))
+  mutate(percent = count / sum(count, na.rm = T) * 100,
+         type = "solution")
 
-# transpose data frames to create pie charts
-# solution count data
-ratio_sol_sep <- sol_sep %>%
-  filter(solution != "mix") %>% # removed mixed treatment
-  select(!total) %>%
-  t %>%
-  as.data.frame %>%
-  rownames_to_column("size") # create size variable
-colnames(ratio_sol_sep)[2:3] <- c("potato", "wheat")
-ratio_sol_sep <- ratio_sol_sep[-c(1:2),]
-ratio_sol_sep <- ratio_sol_sep %>%
-  mutate(across(c(potato, wheat), as.character)) %>%
-  mutate(across(c(potato, wheat), as.numeric)) %>%
-  mutate(wheat = wheat / sum(wheat)) %>%
-  mutate(label_wheat = scales::percent(wheat)) %>%
-  mutate(potato = potato/ sum(potato)) %>%
-  mutate(label_potato = scales::percent(potato))
+samp_comb_long_pie <- corr_comb_long %>%
+  filter(size != "total") %>%
+  group_by(treatment, starch) %>%
+  mutate(percent = count / sum(count, na.rm = T) * 100,
+         type = "sample")
[…]

### 2021-10-22 analysis/scripts/tables.R

@@ -3,9 +3,9 @@
 library(here)
 library(tidyverse)
 
-corr_counts <- readr::read_csv(here("analysis/data/derived_data/", "corr_counts.csv"))
-corr_comb <- readr::read_csv(here("analysis/data/derived_data/", "corr_comb.csv"))
-sol_corr <- readr::read_csv(here("analysis/data/derived_data/", "sol_corr.csv"))
+#corr_counts <- readr::read_csv(here("analysis/data/derived_data/", "corr_counts.csv"))
+#corr_comb <- readr::read_csv(here("analysis/data/derived_data/", "corr_comb.csv"))
+#sol_corr <- readr::read_csv(here("analysis/data/derived_data/", "sol_corr.csv"))
 
 # Sample weights ----------------------------------------------------------
 
@@ -18,34 +18,50 @@ anova_tbl <- corr_comb %>%
             max = max(weight, na.rm = T)) %>%
   mutate(across(where(is.numeric), signif, 3))
 
-
 # Solution counts ---------------------------------------------------------
 
-count_sol <- sol_corr %>% # create a data frame with solution count means
-  group_by(solution, starch) %>%
-  summarise(across(c(s, m, l, total), mean, na.rm = T))
+sol_out <- sol_comb_long %>%
+  pivot_wider(id_cols = c(treatment, starch), names_from = size, values_from = count) %>%
+  mutate(total = sum(s,m,l, na.rm = T))
 
-# add a row containing the combined counts for the mixed treatment
-count_sol_comb <- sol_corr %>%
-  filter(solution == "mix") %>%
-  group_by(starch) %>%
-  summarise(across(c(s, m, l, total), mean, na.rm = T)) %>%
+sol_comb_row <- sol_out %>%
+  group_by(treatment) %>%
+  filter(treatment == "mix") %>%
+  summarise(starch = "both",
+            across(where(is.numeric), sum, na.rm = T))
+
+sol_comb_out <- sol_out %>%
   ungroup() %>%
-  summarise(across(c(s, m, l, total), sum, na.rm = T)) %>%
-  mutate(starch = "both",
-         solution = "mix") %>%
-  select(solution, starch, s, m, l, total)
+  add_row(sol_comb_row, .after = 2)
 
 # Sample counts -----------------------------------------------------------
 
-count_sample <- corr_counts %>% # create a data frame with treatment count means
-  group_by(treatment, starch) %>%
[…]

### 2021-10-22 analysis/paper/paper.Rmd
correct SD miscalculation

@@
 
-```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by z-score for separated treatments."}
+```{r cor-plot, warning=FALSE, fig.cap="Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus."}
 #pl_cor
@@
 ```
-
-<!-- with Z-score standardisation -->
 
@@
 p `r cor_p`
-(Figure \@ref(fig:cor-plot)).
+(Figure \@ref(fig:cor-plot)A).
 
+<!--
 ```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
@@
 ```
+-->
 
@@
 p `r cor2_p`
-(Figure \@ref(fig:cor-plot2)).
+(Figure \@ref(fig:cor-plot)B).
 
@@
 `r strength_cor` `r direct_cor` correlation between biofilm size and retained 
-starch granules (Figure \@ref(fig:cor-plot));
+starch granules (Figure \@ref(fig:cor-plot)A);
 a result that contradicts findings from archaeological contexts
@@
 per mg calculus is considered, the correlation is weaker, but still present
-(Figure \@ref(fig:cor-plot2)).
+(Figure \@ref(fig:cor-plot)B).
 While the larger deposits contain a higher absolute count, our findings also suggest

### 2021-10-22 analysis/paper/paper.tex
remove redundant code from paper and R scripts
correct SD miscalculation

@@
 solutions. Of the total amount of small starches in these two solutions,
-0.653\%
+99.2\%
 are from wheat.
@@
 \hline
-mix & both & 9515 (47.90\%) & 8860 & 8480 (42.7\%) & 4929 & 1882 (9.47\%) & 989 & 19877 (100\%) & 13451\\
+mix & both & 9515 (47.90\%) & 8860 & 8480 (42.7\%) & 7653 & 1882 (9.47\%) & 1596 & 19877 (100\%) & 17768\\
 \hline
@@
 
-\caption{Scatter plot of sample weight and standardised starch count by z-score for separated treatments.}\label{fig:cor-plot}
+\caption{Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus.}\label{fig:cor-plot}
 \end{figure}
@@
 p \textless{} 0.001
-(Figure \ref{fig:cor-plot}).
-
-\begin{figure}
-
-{\centering \includegraphics{../figures/cor-plot2-1}
-
-}
-
-\caption{Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus.}\label{fig:cor-plot2}
-\end{figure}
+(Figure \ref{fig:cor-plot}A).
 
@@
 p = 0.0403
-(Figure \ref{fig:cor-plot2}).
+(Figure \ref{fig:cor-plot}B).
 
@@
 strong positive correlation between biofilm size and retained
-starch granules (Figure \ref{fig:cor-plot});
+starch granules (Figure \ref{fig:cor-plot}A);
 a result that contradicts findings from archaeological contexts
@@
 per mg calculus is considered, the correlation is weaker, but still present
-(Figure \ref{fig:cor-plot2}).
+(Figure \ref{fig:cor-plot}B).
 While the larger deposits contain a higher absolute count, our findings also suggest

### 2021-10-22 analysis/supplementary-materials/supp-mat.Rmd
correct SD miscalculation

@@
 source(here("analysis/scripts/analysis.R"))
+source(here("analysis/scripts/figures.R"))
 #raw_data_samples <- read_csv(here("analysis/data/raw_data/starch_counts.csv"))
@@
 
-Amylase activity in U/mL enzyme, where a U is mg maltose released from starch in six minutes at 36 &deg;C.
+Amylase activity in U/mL enzyme, where U is mg maltose released from starch in six minutes at 36 &deg;C.
 
@@
 ```{r amylase-activity-tables}
-plt1_ph1_result
-plt1_ph2_result
-plt2_ph1_result
-plt2_ph2_result
+plt1_ph1_result # plate 1, photometric reading 1
+plt1_ph2_result # plate 1, photometric reading 2
+plt2_ph1_result # plate 2, photometric reading 1
+plt2_ph2_result # plate 2, photometric reading 2
 ```
@@
 ```{r solution-bar-pl}
-sol_comb_long %>%
+sol_long %>%
   filter(size == "total") %>%
@@
 
-Size distribution within the samples:
+Separated correlation plots:
 
-```{r message=FALSE, warning=FALSE}
-samp_size_pl <- corr_counts_long %>%
-  filter(size != "total",
-         treatment != "control") %>%
-  group_by(treatment, starch, size) %>%
-  summarise(count = mean(count, na.rm = T)) %>%
-  mutate(percent = count / sum(count, na.rm = T) * 100) %>% 
-  ggplot(aes(x = starch, y = percent, fill = size)) +
-    geom_col(position = "dodge") +
-    facet_wrap(~ treatment, scales = "free_x") +
-    scale_fill_viridis_d() +
-    theme_bw()
+```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by z-score for separated treatments."}
+pl_cor
 ```
+
+```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
+pl_cor2
+```

### 2021-10-22 analysis/scripts/analysis.R

@@ -51,6 +51,12 @@ readr::write_csv(sol_corr, here("analysis/data/derived_data", "sol_corr.csv"))
 
 # combining potato and wheat counts from mixed treatment solution
 
+sol_long <- sol_corr %>%
+  group_by(solution, starch) %>%
+  summarise(across(c(s,m,l, total), mean, na.rm = T)) %>%
+  pivot_longer(cols = c(s,m,l, total), values_to = "count", names_to = "size") %>%
+  rename(treatment = solution) # rename to be consistent with sample counts
+
 sol_comb <- sol_corr %>%
   filter(solution != "mix") %>%
   group_by(solution, starch) %>%
@@ -134,3 +140,14 @@ starch_per_mg <- corr_comb %>%
 
 starch_cor <- cor.test(starch_per_mg$weight, starch_per_mg$std_starch_per,
                        use = "pairwise.complete", conf.level = 0.9)
+
+# Standardise counts using Z-score
+z_count <- corr_comb %>%
+  filter(treatment != "control") %>%
+  group_by(treatment) %>%
+  mutate(std_total = (total - mean(total, na.rm = T)) / sd(total, na.rm = T))
+
+# Pearson correlation
+count_cor <- cor.test(z_count$weight, z_count$std_total,
+                      use = "pairwise.complete", conf.level = 0.9)
+

### 2021-10-22 analysis/scripts/figures.R

@@ -5,6 +5,8 @@ source(here("analysis/scripts/analysis.R"))
 #sol_corr <- readr::read_csv(here("analysis/data/derived_data/", "sol_corr.csv"))
 
 
+# Pie charts --------------------------------------------------------------
+
 sol_comb_long_pie <- sol_comb_long %>%
   group_by(treatment, starch) %>%
   mutate(percent = count / sum(count, na.rm = T) * 100,
@@ -65,33 +67,7 @@ pl_potato2 <- pie_datf %>%
 (pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
 
 
-# Scatter plot ------------------------------------------------------------
-
-# Standardise counts using Z-score
-z_count <- corr_comb %>%
-  filter(treatment != "control") %>%
-  group_by(treatment) %>%
-  mutate(std_total = (total - mean(total, na.rm = T)) / sd(total, na.rm = T))
-
-# Pearson correlation
-count_cor <- cor.test(z_count$weight, z_count$std_total,
-                      use = "pairwise.complete", conf.level = 0.9)
-cor_r <- signif(count_cor$estimate, 3)
-cor_ci <- signif(count_cor$conf.int, 3)
-
-# Convert r correlation result to text (MAY NEED TO GO BACK TO results.Rmd FILE)
-direct_cor <- ifelse(cor_r < 0, "negative", "positive")
-strength_cor <- if(cor_r >= 0.8){
-  "very strong"
-} else if(cor_r < 0.8 & cor_r >= 0.6){
-  "strong"
-} else if(cor_r < 0.6 & cor_r >= 0.4){
-  "moderate"
-} else if(cor_r < 0.4 & cor_r >= 0.2){
-  "weak"
-} else{
-  "very weak"
-}
+# Scatter plots -----------------------------------------------------------
 
 # Plot of correlation between weight and count
 pl_cor <- ggplot(z_count,

### 2021-10-22 analysis/scripts/tables.R

@@ -40,20 +40,21 @@ samp_comb_row <- corr_comb_long %>%
   pivot_wider(names_from = size, values_from = count) %>%
   group_by(treatment) %>%
   filter(treatment == "mix") %>%
-  summarise(starch = "both",
-            across(where(is.numeric), sum, na.rm = T)) %>%
-  mutate(sample = "sample")
+  summarise(across(where(is.numeric), sum, na.rm = T)) %>%
+  mutate(sample = "sample",
+         starch = "both") %>%
+  select(treatment, starch, s, m, l, total, sample)
 
 samp_sd_row <- corr_counts_long %>%
+  ungroup() %>%
   filter(treatment == "mix") %>%
+  group_by(sample, size) %>%
+  summarise(count = sum(count, na.rm = T)) %>%
   group_by(size) %>%
-  summarise(treatment = "mix",
-            starch = "both",
-            sd = sd(count, na.rm = T),
-            count = mean(count, na.rm = T)) %>%
-  select(!count) %>%
+  summarise(sd = sd(count, na.rm = T)) %>%
+  mutate(treatment = "mix",
+         starch = "both") %>%
   pivot_wider(names_from = size, values_from = sd) # discrepancy from table
-
 
 
 corr_sd_long <- corr_counts_long %>%

### 2021-10-22 analysis/paper/paper.Rmd
add table to supplementary materials

@@
 ```{r sample-count}
-samp_counts <- corr_comb_long %>%
-  pivot_wider(names_from = size, values_from = count) %>%
-  ungroup() %>%
-  add_row(select(samp_comb_row, !sample), .after = 2)
-
-samp_sd_wide <- corr_sd_long %>%
-  select(!count) %>%
-  pivot_wider(names_from = size, values_from = sd) %>%
-  ungroup() %>%
-  add_row(samp_sd_row, .after = 2) %>%
-  mutate(across(where(is.numeric), round)) %>%
-  rename(s_sd = s,
-         m_sd = m,
-         l_sd = l,
-         total_sd = total)
-
 samp_counts_per <- samp_counts %>%

### 2021-10-22 analysis/supplementary-materials/supp-mat.Rmd
add table to supplementary materials

@@
 source(here("analysis/scripts/figures.R"))
+source(here("analysis/scripts/tables.R"))
+source(here("analysis/scripts/amylase-assay.R"))
 #raw_data_samples <- read_csv(here("analysis/data/raw_data/starch_counts.csv"))
@@
 
+The raw data can be downloaded from GitHub:
+
+```sh
+# solution counts
+wget https://github.com/bbartholdy/byoc-starch/blob/main/analysis/data/raw_data/solution_counts.csv
+
+# sample counts
+wget https://github.com/bbartholdy/byoc-starch/blob/main/analysis/data/raw_data/starch_counts.csv
+```
+
 Raw counts from the treatment solutions before extrapolation.
@@
 raw_counts
-```
-
-The raw data can be downloaded from GitHub:
-
-```sh
-# solution counts
-wget https://github.com/bbartholdy/byoc-starch/blob/main/analysis/data/raw_data/solution_counts.csv
-
-# sample counts
-wget https://github.com/bbartholdy/byoc-starch/blob/main/analysis/data/raw_data/starch_counts.csv
 ```
@@
   # ...it doesn't really matter what the unit is. No activity is no activity)
-source(here("analysis/scripts/amylase-assay.R"))
 cols <- c("1", "2", "3")  # sample triplicates
@@
 ```
+
+### ...and a table
+
+Differences in size ratios (%) of granules between the solutions and the samples. 
+Negative values indicate a loss of granules from solution to sample.
+
+```{r diff-tbl}
+size_diff %>%
+  mutate(across(where(is.numeric), signif, 3))
+```

### 2021-10-22 analysis/scripts/tables.R

@@ -83,6 +83,22 @@ count_sample_comb_sd <- corr_comb %>%
   #select(s, m, l, total) %>%
   #signif(digits = 3)
 
+samp_counts <- corr_comb_long %>%
+  pivot_wider(names_from = size, values_from = count) %>%
+  ungroup() %>%
+  add_row(select(samp_comb_row, !sample), .after = 2)
+
+samp_sd_wide <- corr_sd_long %>%
+  select(!count) %>%
+  pivot_wider(names_from = size, values_from = sd) %>%
+  ungroup() %>%
+  add_row(samp_sd_row, .after = 2) %>%
+  mutate(across(where(is.numeric), round)) %>%
+  rename(s_sd = s,
+         m_sd = m,
+         l_sd = l,
+         total_sd = total)
+
 
 # Proportional counts -----------------------------------------------------
 
@@ -127,3 +143,27 @@ perc_comb_row <- prop %>%
          perc = scales::percent(prop, accuracy = 0.01)) %>%
   select(!c(prop, solution, sample)) %>%
   pivot_wider(names_from = size, values_from = perc)
+
+
+# Size ratios -------------------------------------------------------------
+
+sol_size_prop <- sol_comb_out %>%
+  group_by(treatment, starch) %>%
+  #select(!total) %>%
+  mutate(across(where(is.numeric), function(x) x / total)) %>%
+  mutate(type = "solution")
+
+samp_size_prop <- samp_counts %>%
+  group_by(treatment, starch) %>%
+  #select(!total) %>%
+  mutate(across(where(is.numeric), function(x) x / total)) %>%
+  mutate(type = "sample")
+
+size_diff <- sol_size_prop %>%
+  full_join(samp_size_prop) %>%
+  pivot_longer(cols = c(s,m,l,total), names_to = "size", values_to = "count") %>%
+  pivot_wider(names_from = type, values_from = count) %>%
+  mutate(diff = (sample - solution) * 100) %>%
+  select(!c(solution, sample)) %>%
+  pivot_wider(names_from = size, values_from = diff) %>%
+  select(!total)

### 2021-10-22 analysis/paper/paper.Rmd
fix colours for Figure 3D

@@
  and counted. We show that the granule counts extracted from the model dental calculus 
- represented a low proportion (ranging from 0.064% to 0.161%) of the total number 
+ represented a low proportion (ranging from 0.06% to 0.16%) of the total number 
  of granules exposed to the biofilms throughout the experiment.

### 2021-10-22 analysis/paper/paper.tex
fix colours for Figure 3D

@@
 and counted. We show that the granule counts extracted from the model dental calculus
-represented a low proportion (ranging from 0.064\% to 0.161\%) of the total number
+represented a low proportion (ranging from 0.06\% to 0.16\%) of the total number
 of granules exposed to the biofilms throughout the experiment.

### 2021-10-22 analysis/scripts/figures.R

@@ -1,5 +1,5 @@
 library(tidyverse)
-source(here("analysis/scripts/analysis.R"))
+#source(here("analysis/scripts/analysis.R"))
 #corr_counts <- readr::read_csv(here("analysis/data/derived_data/", "corr_counts.csv"))
 #corr_comb <- readr::read_csv(here("analysis/data/derived_data/", "corr_comb.csv"))
 #sol_corr <- readr::read_csv(here("analysis/data/derived_data/", "sol_corr.csv"))
@@ -61,7 +61,7 @@ pl_potato2 <- pie_datf %>%
   theme_void() +
   scale_fill_viridis_d(name = "",
                        labels = c("large", "medium", "small"),
-                       begin = 0.6) +
+                       begin = 0.6, end = 0.8) +
   theme(legend.position = "none")
 
 (pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")

### 2021-10-22 README.Rmd
update DESCRIPTION and README

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "Investigating Biases Associated with Dietary Starch Incorporation and Retention in an Oral Biofilm Model"
 ```
@@
 
-This compendium can be accessed in a web browser using binder...
+This compendium can be accessed in a web browser using [binder](https://mybinder.org/v2/gh/bbartholdy/byoc-starch/main?urlpath=rstudio).
 
-This research compendium has been developed using the statistical programming
-language R. To work with the compendium locally, you will need
+To work with the compendium locally, you will need
 installed on your computer the [R software](https://cloud.r-project.org/)
@@
 [DESCRIPTION](/DESCRIPTION) file).
-- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console
+- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.pdf`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console
 

### 2021-10-27 analysis/paper/paper.Rmd
preprint version

@@
 output:
+  bookdown::html_document2:
+    df_print: kable
+    keep_md: true
   bookdown::pdf_document2:
@@
       in_header: "../templates/latex/pre-print/preamble.tex"
-      #before_body: "../templates/latex/pre-print/before_body.tex"
     df_print: kable
@@
     keep_tex: true
-    #extra_dependencies: "subfig"
-  bookdown::html_document2:
-    df_print: kable
-    keep_md: true
   bookdown::word_document2:
@@
 ## Compile refs.bib using rbbt package ##
-#cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
-#cite_keys <- unlist(cite_keys, use.names = FALSE)
-#rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T)
+cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
+cite_keys <- unlist(cite_keys, use.names = FALSE)
+rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
 ## end refs.bib ##
@@
 for 10 days until the end of the experiment on day 24 
-(see \@ref(fig:protocol-fig) for an overview of the protocol).
+(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).
 
@@
 maltose released from starch by $\alpha$-amylase activity. Results are reported
-in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.
+in units (U) per mL enzyme, where 1 U releases 1 mg of maltose in 6 minutes.
 
@@
 whole slide. The count from the slide was extrapolated to the whole sample
-(see Supplementary Material).
+(see Supplementary Material for more detail).
 
@@
 <!-- R packages used -->
-A one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
-as the grouping variable (GV) was conducted to explore the effect of the different 
-starch treatments on biofilm growth.
 
-To calculate size proportions, mean counts for each treatment were taken across 
-both experimental plates for each treatment, resulting in a mean count for each
+To see if biofilm growth was differently affected by starch treatments, a 
+one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
+as the grouping variable (GV) was conducted.
+To analyse granule counts and calculate size proportions, mean counts for each 
+treatment were taken across
+both experimental plates, resulting in a mean count for each
 granule size category within each treatment.
@@
 `r paste(signif(range(positive_amylase),3), collapse = " to ")` U/mL enzyme 
-(Supplementary Material).
+(full results can be found in the Supplementary Material).
 The results are not comparable to other studies presenting $\alpha$-amylase activity
@@
 
-## Treatment type had no effect on biofilm growth
+## Treatment type had minimal effect on biofilm growth
 
@@
 While the larger deposits contain a higher absolute count, our findings also suggest
-that they contain a slightly higher concentration of starches. 
+that they contain a slightly higher concentration of starches. This may also explain
+the lower mean retention of starch granules in mixed treatments compared to
+wheat treatments.
+Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
+granule counts, wheat treatment samples had the highest mean starch retention.
 Further research is needed to determine why this differs from previous archaeological
@@
 following the format provided by the rrtools package [@R-rrtools].
-More detailed protocols are available on OSF and protocols.io.
+More detailed protocols are available on OSF (https://osf.io/akevs/) and protocols.io (https://www.protocols.io/view/biofilm-growth-with-starch-treatment-bu7jnzkn).
+Additional tables and figures are available in the Supplementary Material.
 

### 2021-10-27 analysis/paper/paper.tex
preprint version

@@
 
-{\centering \includegraphics{/media/bjorn/hogwarts/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview}
+{\centering \includegraphics{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview}
 
@@
 for 10 days until the end of the experiment on day 24
-(see \ref{fig:protocol-fig} for an overview of the protocol).
+(see Figure \ref{fig:protocol-fig} for an overview of the protocol).
 
@@
 maltose released from starch by \(\alpha\)-amylase activity. Results are reported
-in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.
+in units (U) per mL enzyme, where 1 U releases 1 mg of maltose in 6 minutes.
 
@@
 here (Müller, 2020), and patchwork (Pedersen, 2020).
-A one-way ANOVA with sample weight as the dependent variable (DV) and treatment
-as the grouping variable (GV) was conducted to explore the effect of the different
-starch treatments on biofilm growth.
-
-To calculate size proportions, mean counts for each treatment were taken across
-both experimental plates for each treatment, resulting in a mean count for each
+
+To see if biofilm growth was differently affected by starch treatments, a
+one-way ANOVA with sample weight as the dependent variable (DV) and treatment
+as the grouping variable (GV) was conducted.
+To analyse granule counts and calculate size proportions, mean counts for each
+treatment were taken across
+both experimental plates, resulting in a mean count for each
 granule size category within each treatment.
@@
 
-{\centering \includegraphics[width=0.5\linewidth]{/media/bjorn/hogwarts/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar}
+{\centering \includegraphics[width=0.5\linewidth]{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar}
 
@@
 3.4 to 10.3 U/mL enzyme
-(Supplementary Material).
+(full results can be found in the Supplementary Material).
 The results are not comparable to other studies presenting \(\alpha\)-amylase activity
@@
 
-\hypertarget{treatment-type-had-no-effect-on-biofilm-growth}{%
-\subsection{Treatment type had no effect on biofilm growth}\label{treatment-type-had-no-effect-on-biofilm-growth}}
+\hypertarget{treatment-type-had-minimal-effect-on-biofilm-growth}{%
+\subsection{Treatment type had minimal effect on biofilm growth}\label{treatment-type-had-minimal-effect-on-biofilm-growth}}
 
@@
 While the larger deposits contain a higher absolute count, our findings also suggest
-that they contain a slightly higher concentration of starches.
+that they contain a slightly higher concentration of starches. This may also explain
+the lower mean retention of starch granules in mixed treatments compared to
+wheat treatments.
+Wheat treatment samples (mean = 5.53 mg) were on average larger than mixed treatment samples (mean = 4.28 mg) (Table \ref{tab:anova-tbl}); and while mixed treatment solutions contained the highest mean overall
+granule counts, wheat treatment samples had the highest mean starch retention.
 Further research is needed to determine why this differs from previous archaeological
@@
 or accidental ingestion not related to food consumption
-(Radini et al., 2019; Radini et al., 2017).\\
+(Radini et al., 2017; Radini et al., 2019).\\
 When starch granules enter the mouth, whether through ingestion of food or accidental
@@
 following the format provided by the rrtools package (Marwick, 2019).
-More detailed protocols are available on OSF and protocols.io.
+More detailed protocols are available on OSF (\url{https://osf.io/akevs/}) and protocols.io (\url{https://www.protocols.io/view/biofilm-growth-with-starch-treatment-bu7jnzkn}).
+Additional tables and figures are available in the Supplementary Material.
 
@@
 \leavevmode\hypertarget{ref-bernfeldAmylase1955}{}%
-Bernfeld, P. (1955). Amylases, α and β. In \emph{Methods in {Enzymology}} (Vol. 1, pp. 149--158). {Academic Press}. \url{https://doi.org/10.1016/0076-6879(55)01021-5}
+Bernfeld, P. (1955). Amylases, {\(\alpha\)} and {\(\beta\)}. In \emph{Methods in {Enzymology}} (Vol. 1, pp. 149--158). {Academic Press}. \url{https://doi.org/10.1016/0076-6879(55)01021-5}
 
@@
 {CSIC Press}.
-\url{http://books.google.com?id=j66CDVfVhwEC}
 
@@
 \leavevmode\hypertarget{ref-R-rrtools}{}%
-Marwick, B. (2019). \emph{Rrtools: Creates a reproducible research compendium} {[}Manual{]}. \url{https://github.com/benmarwick/rrtools}
+Marwick, B. (2019). \emph{Rrtools: Creates a reproducible research compendium} {[}Manual{]}.
 
@@
 \leavevmode\hypertarget{ref-R-here}{}%
-Müller, K. (2020). \emph{Here: A simpler way to find your files} {[}Manual{]}. \url{https://CRAN.R-project.org/package=here}
+Müller, K. (2020). \emph{Here: A simpler way to find your files} {[}Manual{]}.
 
@@
 (2013).
-Taking the {Starch} out of {Oral Biofilm Formation}: Molecular {Basis} and {Functional Significance} of {Salivary} α-{Amylase Binding} to {Oral Streptococci}.
+Taking the {Starch} out of {Oral Biofilm Formation}: Molecular {Basis} and {Functional Significance} of {Salivary} {\(\alpha\)}-{Amylase Binding} to {Oral Streptococci}.
 \emph{Applied and Environmental Microbiology}, \emph{79}(2), 416--423.
@@
 \leavevmode\hypertarget{ref-R-patchwork}{}%
-Pedersen, T. L. (2020). \emph{Patchwork: The composer of plots} {[}Manual{]}. \url{https://CRAN.R-project.org/package=patchwork}
+Pedersen, T. L. (2020). \emph{Patchwork: The composer of plots} {[}Manual{]}.
 
@@
 \leavevmode\hypertarget{ref-R-base}{}%
-R Core Team.
-(2020).
-\emph{R: A language and environment for statistical computing} {[}Manual{]}.
-{R Foundation for Statistical Computing}; {R Foundation for Statistical Computing}.
-\url{https://www.R-project.org/}
+R Core Team. (2020). \emph{R: A language and environment for statistical computing} {[}Manual{]}. {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing}.
 
@@
 \leavevmode\hypertarget{ref-R-broom}{}%
-Robinson, D., Hayes, A., \& Couch, S. (2021). \emph{Broom: Convert statistical objects into tidy tibbles} {[}Manual{]}. \url{https://CRAN.R-project.org/package=broom}
+Robinson, D., Hayes, A., \& Couch, S. (2021). \emph{Broom: Convert statistical objects into tidy tibbles} {[}Manual{]}.
 
 \leavevmode\hypertarget{ref-scannapiecoSalivaryAmylase1993}{}%
-Scannapieco, F. A., Torres, G., \& Levine, M. J. (1993). Salivary α-amylase: Role in dental plaque and caries formation. \emph{Critical Reviews in Oral Biology \& Medicine}, \emph{4}(3), 301--307.
+Scannapieco, F.
+A., Torres, G., \& Levine, M.
+J.
+(1993).
+Salivary {\(\alpha\)}-amylase: Role in dental plaque and caries formation.
+\emph{Critical Reviews in Oral Biology \& Medicine}, \emph{4}(3), 301--307.
 

### 2021-10-27 analysis/supplementary-materials/supp-mat.Rmd
preprint version

@@
 ---
-title: Supplementary Material
+title: "Supplementary Material"
 output:
   html_document:
-    keep_md: true
+    keep_md: yes
     df_print: paged
-    toc: true
-    toc_float: true
+    toc: yes
+    toc_float: yes
+  pdf_document:
+    toc: yes
+    df_print: kable
 ---
@@
 print(sessionInfo(), locale = FALSE)
-```
-
-## Experimental setup
-
-```{r plate-lid-1, fig.cap="The 24 deepwell plate and the lid with pegs (substrata)"}
-knitr::include_graphics(here("analysis/figures/plate_lid_side.jpg"))
-```
-
-```{r plate-lid-2, fig.cap="The 24 deepwell plate with the lid (almost) on."}
-knitr::include_graphics(here("analysis/figures/plate_lid_on.jpg"))
 ```
@@
 ```
+
+## Experimental setup
+
+```{r plate-lid-1, fig.cap="The 24 deepwell plate and the lid with pegs (substrata)"}
+knitr::include_graphics(here("analysis/figures/plate_lid_side.jpg"))
+```
+
+```{r plate-lid-2, fig.cap="The 24 deepwell plate with the lid (almost) on."}
+knitr::include_graphics(here("analysis/figures/plate_lid_on.jpg"))
+```
+
+## Protocols
+
+All protocols are available on [protocols.io](https://www.protocols.io).
+
+Creating the artificial saliva: <https://www.protocols.io/view/artificial-saliva-bva9n2h6>
+
+Creating the CPMU solution: <https://www.protocols.io/view/cpmu-bv8pn9vn>
+
+Biofilm growth protocol: <https://www.protocols.io/view/biofilm-growth-with-starch-treatment-bu7jnzkn>
+
+Amylase activity assay: <https://www.protocols.io/view/amylase-activity-bw8jphun>
 
@@
 
-Extracted-granule counts separated by size:
+Extracted-granule counts separated by treatment and size, including error bars:
 
-```{r}
+```{r counts-bar, fig.cap="l = large, m = medium, s = small."}
 corr_counts_long %>%
@@
 
-Size distribution within the solutions (top) and samples (bottom):
+Size distribution (in %) within the solutions (top) and samples (bottom):
 
-```{r message=FALSE, warning=FALSE}
+```{r size-ratios-bar, fig.cap="l = large, m = medium, s = small.", message=FALSE, warning=FALSE}
 sol_size_pl <- sol_corr %>%
@@
 
-Separated correlation plots:
+Separated correlation plots. These are the same plots as in the main paper, just larger.
 

### 2021-10-27 analysis/scripts/amylase-assay.R

@@ -1,5 +1,9 @@
-#library(byocstarch)
-devtools::load_all()
+if(require(byocstarch, quietly = T)) {
+  library(byocstarch)
+} else {
+  devtools::load_all()
+}
+
 library(tidyverse)
 
 # Upload data -------------------------------------------------------------
@@ -26,12 +30,10 @@ plt2_ph2 <- readr::read_tsv(
 # Standard curve ----------------------------------------------------------
 
 # convert mL 0.2% Maltose to mg maltose
-#mg_malt <- c(3.75, 15, 30, 45, 60, 75, 150) * 0.002
+
 # two standard curves prepared: one in h2o and one in artificial saliva (BMM)
 stdH2O1.ph1 <- stdcurve(plt1_ph1, cols = 7:9)
-#stdH2O1.ph1 <- stdcurve(plt1_ph1, cols = 7:9, mg_maltose = mg_malt)
 stdBMM1.ph1 <- stdcurve(plt1_ph1, cols = 10:12)
-#stdBMM1.ph1 <- stdcurve(plt1_ph1, cols = 10:12, mg_maltose = mg_malt)
 
 stdH2O1.ph2 <- stdcurve(plt1_ph2, cols = 7:9)
 stdBMM1.ph2 <- stdcurve(plt1_ph2, cols = 10:12)

### 2021-10-27 README.Rmd
update README
add zenodo badge

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bbartholdy/byoc-starch/main?urlpath=rstudio) <!-- DOI badge: [![DOI](.svg)](repo link) -->
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bbartholdy/byoc-starch/main?urlpath=rstudio) 
+[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.5604670.svg)](https://doi.org/10.5281/zenodo.5604670)
 
-<!-- Document structure provided by rrtools package -->
+<!-- DOI badge: [![DOI](.svg)](repo link) -->
+
+<!-- Document structure provided by rrtools package
 
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> Bartholdy, B. P. & Henry, A. G. (2021). _`r Title`_. bioRxiv <https://doi.org/xxx/xxx>
 
@@
 > Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
-
+-->
 ## Contents

### 2021-10-27 analysis/scripts/amylase-assay.R
add amylase functions to amylase-assay.R

@@ -1,10 +1,113 @@
-if(require(byocstarch, quietly = T)) {
-  library(byocstarch)
-} else {
-  devtools::load_all()
-}
+# if(require(byocstarch, quietly = T)) {
+#   library(byocstarch)
+# } else {
+#   devtools::load_all()
+# }
 
 library(tidyverse)
+
+# Functions ---------------------------------------------------------------
+
+#' Standard curve calculation
+#' @param std_data Data frame with the photometric readings of the standard curve.
+#' @param cols Vector containing the column numbers containing the standard curve photometric readings. If NULL, all columns assumed to be standard curve measurements.
+#' @param ml_maltose vector of length 7 containing the volume (ml) of 0.2% maltose solution added to the assay.
+#' @importFrom magrittr %>%
+#' @importFrom dplyr slice n
+stdcurve <- function(std_data, cols = NULL, ml_maltose = c(3.75, 15, 30, 45, 60, 75, 150)){
+  if(!is.null(cols)){
+    std_data <- std_data[,cols]
+  }
+  # delta 540 calculation
+  mg_maltose <- 0.002 * ml_maltose
+  replicates <- dim(std_data)[2]
+  pht_blank <- std_data[8,] %>%
+    slice(rep(1:n(), each = 7))
+  pht_meas <- std_data[1:7,] - pht_blank
+
+  pht_meas <- c(pht_meas[, seq(1:replicates)]) %>%
+    unlist()
+  # Linear regression
+  stdCurve_data <- data.frame("d540" = pht_meas, "mgMaltose" = rep(mg_maltose, replicates))
+  std_curve <- lm(d540 ~ mgMaltose, stdCurve_data)
+  # Plot of standard curve
+  p_stdcurve <- stdCurve_data %>%
+    ggplot(aes(x = mgMaltose, y = d540)) +
+    geom_point() +
+    geom_smooth(formula = "y ~ x", method = "lm", se = F) +
+    theme_bw()
+  out <- list("std_curve" = std_curve, "plot" = p_stdcurve, "blank" = pht_blank[1,], "replicates" = replicates, "model" = stdCurve_data)
+  return(out)
+}
+
+#' Amylase activity in sample calculation
+#' @param samp_data data frame with the photometric readings of the samples.
[…]

### 2021-10-27 analysis/paper/paper.Rmd
comment out bib-genrating code

@@
 ## Compile refs.bib using rbbt package ##
-cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
-cite_keys <- unlist(cite_keys, use.names = FALSE)
-rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
-## end refs.bib ##
-
-## Word count ##
-wd_count <- wordcountaddin::word_count()
+# cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
+# cite_keys <- unlist(cite_keys, use.names = FALSE)
+# rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
+# ## end refs.bib ##
+# 
+# ## Word count ##
+# wd_count <- wordcountaddin::word_count()
 ```

### 2021-10-27 analysis/paper/paper.Rmd
add protocolsio workspace link

@@
     keep_tex: true
+  bookdown::html_document2:
+    df_print: kable
+    keep_md: true
   bookdown::word_document2:
@@
 following the format provided by the rrtools package [@R-rrtools].
-More detailed protocols are available on OSF (https://osf.io/akevs/) and protocols.io (https://www.protocols.io/view/biofilm-growth-with-starch-treatment-bu7jnzkn).
+More detailed protocols are available on OSF (https://osf.io/akevs/) and protocols.io (https://www.protocols.io/workspaces/byoc).
 Additional tables and figures are available in the Supplementary Material.

### 2021-10-27 analysis/paper/paper.tex
add protocolsio workspace link

@@
 following the format provided by the rrtools package (Marwick, 2019).
-More detailed protocols are available on OSF (\url{https://osf.io/akevs/}) and protocols.io (\url{https://www.protocols.io/view/biofilm-growth-with-starch-treatment-bu7jnzkn}).
+More detailed protocols are available on OSF (\url{https://osf.io/akevs/}) and protocols.io (\url{https://www.protocols.io/workspaces/byoc}).
 Additional tables and figures are available in the Supplementary Material.

### 2021-11-09 README.Rmd
binder not working - remove badge

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bbartholdy/byoc-starch/main?urlpath=rstudio) 
+<!--[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bbartholdy/byoc-starch/main?urlpath=rstudio)--> 
 [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.5604670.svg)](https://doi.org/10.5281/zenodo.5604670)

### 2022-02-27 R/amylase.R
remove unnecessary formats of paper

@@ -1,101 +1,0 @@
-# Functions ---------------------------------------------------------------
-
-#' Standard curve calculation
-#' @param std_data Data frame with the photometric readings of the standard curve.
-#' @param cols Vector containing the column numbers containing the standard curve photometric readings. If NULL, all columns assumed to be standard curve measurements.
-#' @param ml_maltose vector of length 7 containing the volume (ml) of 0.2% maltose solution added to the assay.
-#' @importFrom magrittr %>%
-#' @importFrom dplyr slice n
-stdcurve <- function(std_data, cols = NULL, ml_maltose = c(3.75, 15, 30, 45, 60, 75, 150)){
-  if(!is.null(cols)){
-    std_data <- std_data[,cols]
-  }
-  # delta 540 calculation
-  mg_maltose <- 0.002 * ml_maltose
-  replicates <- dim(std_data)[2]
-  pht_blank <- std_data[8,] %>%
-    slice(rep(1:n(), each = 7))
-  pht_meas <- std_data[1:7,] - pht_blank
-
-  pht_meas <- c(pht_meas[, seq(1:replicates)]) %>%
-    unlist()
-  # Linear regression
-  stdCurve_data <- data.frame("d540" = pht_meas, "mgMaltose" = rep(mg_maltose, replicates))
-  std_curve <- lm(d540 ~ mgMaltose, stdCurve_data)
-  # Plot of standard curve
-  p_stdcurve <- stdCurve_data %>%
-    ggplot(aes(x = mgMaltose, y = d540)) +
-    geom_point() +
-    geom_smooth(formula = "y ~ x", method = "lm", se = F) +
-    theme_bw()
-  out <- list("std_curve" = std_curve, "plot" = p_stdcurve, "blank" = pht_blank[1,], "replicates" = replicates, "model" = stdCurve_data)
-  return(out)
-}
-
-#' Amylase activity in sample calculation
-#' @param samp_data data frame with the photometric readings of the samples.
-#' @param cols vector containing the column numbers of the sample photometric readings. If NULL, all columns assumed to be standard curve measurements.
-#' @param std_curve an object from the `stdcurve` function output.
[…]

### 2022-02-27 analysis/paper/paper.Rmd
change default knit format to HTML
modify paper to frontiersin guidelines

@@
 title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
-author: |
- Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
-date: |
- $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
-output:
-  bookdown::pdf_document2:
-    includes:
-      in_header: "../templates/latex/pre-print/preamble.tex"
-    df_print: kable
-    toc: no
-    latex_engine: xelatex
-    fig_caption: yes
-    number_sections: yes
-    keep_tex: true
-  bookdown::html_document2:
-    df_print: kable
-    keep_md: true
-  bookdown::word_document2:
-    reference_docx: "../templates/template.docx"
+#author: |
+# Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
+# date: |
+#  $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
+#### Frontiers submission ####
+runtitle: "Biases Associated with Starch Incorporation"
+documentclass: frontiers/frontiersSCNS  # or frontiersHLTH, or frontiersFPHY
+author:
+  - name: Bjørn Peare Bartholdy
+    affiliation: '1'
+    etal: BARTHOLDY # First author's last name.
+    email: b.p.bartholdy@arch.leidenuniv.nl # Indicates corresponding Author
+    institution: Leiden University
+    street: Einsteinweg 2
+    city: Leiden
+    zip: 2333 CC
+    country: Netherlands
+  - name: Amanda G. Henry
+    affiliation: '1'
+affiliation:
+  - id: '1'
+    department: Archaeological Sciences
+    institution: Leiden University
+    city: Leiden
+    country: Netherlands
+#### End: Frontiers submission ####
 #linestretch: 2.0 # double-spacing for review purposes
 bibliography: refs.bib
-csl: apa.csl
-abstract: |
- Dental calculus has proven to contain a wealth of information 
- on the dietary habits of past populations. These insights have, to a large extent,
- been obtained by the extraction and identification of
- starch granules contained within the mineralised dental plaque from a wide range 
- of regions and time periods. The scope of previous studies have been limited to 
- microfossil extraction and identification to reconstruct dietary preferences from
- the archaeological record, and few studies have attempted to address the biases 
- of starch retention in dental calculus. Those that have considered this problem 
- have been limited to *in vivo* studies on modern humans and non-human primates.
- Here, we present a multispecies oral biofilm model, which allows experimental 
- research on  starch incorporation and retention to be conducted on *in vitro* 
- dental calculus in a controlled laboratory setting. The biofilms were exposed 
- to treatment solutions with known quantities of
- dietary starches (wheat and potato) during the 25-day growth period. After this,
- the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
- and counted. We show that the granule counts extracted from the model dental calculus 
- represented a low proportion (ranging from 0.06% to 0.16%) of the total number 
- of granules exposed to the biofilms throughout the experiment.
- Additionally, we found that the ratios of granule sizes from the extracted 
- starch granules differed from the original treatment solutions, with large 
- granules (>20 $\mu$m) consistently being under-represented.
- We also found a correlation
- between the absolute granule counts and dry-weight of the biofilm 
- (*r* = 0.66, 90%CI[0.46,0.79]), as well
- as between the concentration (count per mg) of granules and dry-weight 
- (*r* = 0.30, 90%CI[0.06,0.51]).  
- Our results reinforce previous *in vivo* studies suggesting that dental calculus 
- presents a very small, and partly biased picture of the original dietary intake 
- of starches, with an over-representation of plants producing granules smaller than
- 20 $\mu$m in size. The experimental model presented here is well-suited to address
- the need for further validation of methods and biases associated with dietary
- research on dental calculus.
+csl: frontiers/frontiers.csl
+params:
   keywords: 
-  - oral biofilm
+    - oral biofilm model
     - starch
@@
     - dental calculus
+    - paleodiet reconstruction
 ---
@@
   message = FALSE,
+  #fig.path = "frontiers/figures/",
   fig.path = "../figures/",
-  fig.align = "center")
+  fig.align = "center",
+  #fig.ext = "jpg",
+  fig.pos = "H",
+  dpi = 300) 
 library(tidyverse)
@@
 ## Compile refs.bib using rbbt package ##
-# cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
-# cite_keys <- unlist(cite_keys, use.names = FALSE)
-# rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
+#cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
+#cite_keys <- unlist(cite_keys, use.names = FALSE)
+#rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
 # ## end refs.bib ##
@@
 # wd_count <- wordcountaddin::word_count()
+cor_r <- signif(count_cor$estimate, 3)
+cor_ci <- signif(count_cor$conf.int, 3)
+cor2_r <- signif(count_cor$estimate, 3)
+cor2_ci <- signif(count_cor$conf.int, 3)
 ```
+
+\begin{abstract}
+
+Dental calculus has proven to contain a wealth of information on the
+dietary habits of past populations. These insights have, to a large
+extent, been obtained by the extraction and identification of starch
+granules contained within the mineralised dental plaque from a wide
+range of regions and time periods. The scope of previous studies have
+been limited to microfossil extraction and identification to reconstruct
+dietary preferences from the archaeological record, and few studies have
+attempted to address the biases of starch retention in dental calculus.
+Those that have considered this problem have been limited to \emph{in
+vivo} studies on modern humans and non-human primates. Here, we present
+a multispecies oral biofilm model, which allows experimental research on
+starch incorporation and retention to be conducted on \emph{in vitro}
+dental calculus in a controlled laboratory setting. The biofilms were
+exposed to treatment solutions with known quantities of dietary starches
+(wheat and potato) during the 25-day growth period. After this, the
+starch granules were extracted from the mature biofilm (by dissolution
+in EDTA), and counted. We show that the granule counts extracted from
+the model dental calculus represented a low proportion (ranging from
+0.06\% to 0.16\%) of the total number of granules exposed to the
+biofilms throughout the experiment. Additionally, we found that the
+ratios of granule sizes from the extracted starch granules differed from
+the original treatment solutions, with large granules (\textgreater20
+\(\mu\)m) consistently being under-represented. We also found a
+correlation between the absolute granule counts and dry-weight of the
+biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
+the concentration (count per mg) of granules and dry-weight (\emph{r} =
+0.3, 90\%CI{[}0.0618, 0.506{]}).\\
+Our results reinforce previous \emph{in vivo} studies suggesting that
+dental calculus presents a very small, and partly biased picture of the
+original dietary intake of starches, with an over-representation of
+plants producing granules smaller than 20 \(\mu\)m in size. The
+experimental model presented here is well-suited to address the need for
+further validation of methods and biases associated with dietary
+research on dental calculus.
+
+\tiny
+ \keyFont{ \section{Keywords:} `r params$keywords`}
+
+\end{abstract}
 
@@
 
-<!-- insert figure of the setup -->
-
 In this study we employ a multispecies oral biofilm model following a modified
@@
 and 1 mmol/l arginine [@sissonsMultistationDentalPlaque1991].
-
-```{r protocol-fig, fig.cap="Overview of experiment protocol including the plate setup."}
-knitr::include_graphics(here("analysis/figures/protocol_overview.png"))
-```
 
@@
 Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
-The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
-(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
+The protocol is a modified version of an Enzymatic Assay of $\alpha$-Amylase
+(<https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase>) [@bernfeldAmylase1955], which measures the amount of
 maltose released from starch by $\alpha$-amylase activity. Results are reported
-in units (U) per mL enzyme, where 1 U releases 1 mg of maltose in 6 minutes.
+in units (U) per mL enzyme, where 1 U releases 1 $\mu$mole of maltose in 6 minutes.
+The detailed protocol can be found here: <https://www.protocols.io/view/amylase-activity-bw8jphun>.
 
@@
 
-```{r microscope-fig, fig.cap="Microscope image of a biofilm sample exposed to the potato starch solution.
-Potato granules can be seen within a bacterial community.
-Scale bar = 20 $\\mu$m.", out.width="50%"}
-knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
-```
-
 All samples yielded sufficient biofilm growth and starch incorporation to be 
@@
 saliva samples from any of the days that were sampled. Only positive controls
+(saliva)
 contained amylase activity that could be detected in the assay, ranging from
@@
 levels in humans, as the unit definition may differ; 
-however, they are sufficient to show that there is no activity in the system.
+however, they are sufficient to show that there is no activity in the model.
 
 ## Treatment type had minimal effect on biofilm growth
-
-```{r anova-tbl}
-knitr::kable(anova_tbl, 
-             col.names = c("Treatment", "Mean", "SD", "Min", "Max"),
-             caption = "Summary statistics for biofilm dry-weights (in mg) by treatment.")
-```
 
@@
 
-```{r solution-count-tbl}
-cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
-options(knitr.kable.NA = "")
-knitr::kable(sol_tbl, col.names = cnames, 
-             caption = "Mean starch counts from solutions, including the
-             proportional makeup of the different sizes of granules.")
-```
-
-```{r sample-count-tbl}
-cnames <- c("Treatment", "Starch", "Small (%)", "SD", "Medium (%)", "SD", 
-            "Large (%)", "SD", "Total (%)", "SD")
-knitr::kable(samp_counts_tbl, col.names = cnames, 
-             caption = "Mean starch counts extracted from samples with standard 
-             deviation (SD), including the proportion of granule sizes of the total count.")
-```
-
 <!-- absolute solution counts compared to absolute sample counts. does exposure
@@
   add_row(prop_comb_row, .after = 2)
-
-perc_tbl <- prop %>%
-  ungroup() %>%
-  select(!c(sample, solution, prop)) %>%
-  pivot_wider(names_from = size, values_from = perc) %>%
-  add_row(perc_comb_row, .after = 2)
-```
-
-```{r sample-prop-tbl}
-cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
-knitr::kable(perc_tbl, col.names = cnames, 
-             caption = "The mean percentage of starches from the solutions that 
-              were incorporated into the samples.")
 ```
@@
 
-```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."}
-#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
-(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
-```
-
 <!-- overall size trends -->
 
-```{r}
+```{r size-trends}
 sol_size_prop <- sol_comb_out %>%
@@
 
-```{r cor-plot, warning=FALSE, fig.cap="Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus."}
-#pl_cor
-pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
-```
-
 Pearson's *r* suggests a 
@@
 (Figure \@ref(fig:cor-plot)A).
-
-<!--
-```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
-pl_cor2
-```
--->
 
@@
 susceptible to hydrolysis 
-[@francoStarchDegradation1992; @haslamDecompositionStarch2004]; however, 
-the absence can also allow us to directly explore
-the effect of $\alpha$-amylase on starch counts in future experiments,
-where $\alpha$-amylase can be added to the model
-in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. 
+[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. The absence may
+also affect biofilm growth due to the lack of amylase-bacterium interactions
+[@nikitkovaStarchBiofilms2013]. Conversely, the model may benefit from the
+absence of $\alpha$-amylase, because it can allow us to directly explore
+its effect on starch counts in future experiments, where $\alpha$-amylase can be 
+added to the model in concentrations similar to those found in the oral cavity
+[@scannapiecoSalivaryAmylase1993]. 
 We are able to show how absolute counts in the treatments cause a difference in
@@
 
+# Open Science Statement {-}
+
+All scripts and data used in the analysis are available on OSF 
+(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch),
+following the format provided by the rrtools package [@R-rrtools].
+More detailed protocols are available on OSF (https://osf.io/akevs/)
+and protocols.io (https://www.protocols.io/workspaces/byoc).
+Additional tables and figures are available in the Supplementary Material
+(https://osf.io/ucxsv/).
+
+# Author Contributions {-}
+
+AGH acquired the funding for the analysis. BPB and AGH conceptualised and designed
+the study. BPB conducted the experiments and analysed the data.
+BPB wrote the manuscript. AGH reviewed and edited the manuscript.
+
+# Funding {-}
+
+This research has received funding from the European Research Council under the 
+European Union’s Horizon 2020 research and innovation program, grant agreement 
+number STG–677576 (“HARVEST”).
+
+# Conflict of Interest {-}
+
+The authors declare that the research was conducted in the absence of any
+commercial or financial relationships that could be construed as a potential
+conflict of interest.
+
 # Acknowledgements {-}
@@
 
-This research has received funding from the European Research Council under the 
-European Union’s Horizon 2020 research and innovation program, grant agreement 
-number STG–677576 (“HARVEST”).
+# References {-}
 
-# Data Availability Statement {-}
+::: {#refs}
+:::
 
-All scripts and data used in the analysis are available on OSF 
-(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch),
-following the format provided by the rrtools package [@R-rrtools].
-More detailed protocols are available on OSF (https://osf.io/akevs/) and protocols.io (https://www.protocols.io/workspaces/byoc).
-Additional tables and figures are available in the Supplementary Material.
+\clearpage
 
-# References cited
+# Tables
+
+```{r anova-tbl}
+knitr::kable(anova_tbl, 
+             col.names = c("Treatment", "Mean", "SD", "Min", "Max"),
+             caption = "Summary statistics for biofilm dry-weights (in mg) by treatment.")
+```
+
+```{r solution-count-tbl}
+cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
+options(knitr.kable.NA = "")
+knitr::kable(sol_tbl, col.names = cnames, 
+             caption = "Mean starch counts from solutions, including the
+             proportional makeup of the different sizes of granules.")
+```
+
+```{r sample-count-tbl}
+cnames <- c("Treatment", "Starch", "Small (%)", "SD", "Medium (%)", "SD", 
+            "Large (%)", "SD", "Total (%)", "SD")
+knitr::kable(samp_counts_tbl, col.names = cnames, 
+             caption = "Mean starch counts extracted from samples with standard 
+             deviation (SD), including the proportion of granule sizes of the total count.")
+```
+
+```{r sample-prop-tbl}
+cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
+knitr::kable(perc_tbl, col.names = cnames, 
+             caption = "The mean percentage of starches from the solutions that 
+              were incorporated into the samples.")
+```
+
+# Figure Captions
+
+```{r protocol-fig, fig.cap="Overview of experiment protocol including the plate setup.", out.width="70%"}
+knitr::include_graphics(here("analysis/figures/protocol_overview.png"))
+```
+
+```{r microscope-fig, fig.cap="Microscope image of a biofilm sample exposed to the potato starch solution.
+Potato granules can be seen within a bacterial community.
+Scale bar = 20 $\\mu$m.", out.width="50%"}
+knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
+```
+
+```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."}
+#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
+(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+```
+
+```{r cor-plot, warning=FALSE, fig.cap="Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus."}
+#pl_cor
+pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+#ggsave("frontiers/figures/cor-plot.png", dpi = 300)
+```
+
+<!--
+Figure 1: Overview of experiment protocol including the plate setup.
+
+Figure 2: Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\mu$m.
+
+Figure 3: Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.
+
+Figure 4: Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus.
+-->

### 2022-02-27 analysis/supplementary-materials/supp-mat.Rmd
update supp mat

@@
 output:
-  html_document:
-    keep_md: yes
+  bookdown::html_document2:
     df_print: paged
@@
     toc_float: yes
-  pdf_document:
+  bookdown::pdf_document2:
     toc: yes
@@
 
-## R Session info
+# R Session info
 
@@
 print(sessionInfo(), locale = FALSE)
+```
+
+# Experimental setup
+
+```{r plate-lid-1, fig.cap="The 24 deepwell plate and the lid with pegs (substrata)"}
+knitr::include_graphics(here("analysis/supplementary-materials/figures/plate_lid_side.jpg"))
+```
+
+```{r plate-lid-2, fig.cap="The 24 deepwell plate with the lid (almost) on."}
+knitr::include_graphics(here("analysis/supplementary-materials/figures/plate_lid_on.jpg"))
+```
+
+# Protocols
+
+All protocols are available on [protocols.io](https://www.protocols.io).
+
+Creating the artificial saliva: <https://www.protocols.io/view/artificial-saliva-bva9n2h6>
+
+Creating the CPMU solution: <https://www.protocols.io/view/cpmu-bv8pn9vn>
+
+Biofilm growth protocol: <https://www.protocols.io/view/biofilm-growth-with-starch-treatment-bu7jnzkn>
+
+Amylase activity assay: <https://www.protocols.io/view/amylase-activity-bw8jphun>
+
+# Raw data
+
+The raw data can be downloaded from OSF:
+
+Solution counts: <https://osf.io/kz3b2/>  
+Sample counts: <https://osf.io/kz3b2/>
+
+```sh
+# solution counts
+wget https://osf.io/kz3b2/download -O solution_counts.csv
+
+# sample counts
+wget https://osf.io/kz3b2/download -O starch_counts.csv
 ```
@@
 
-## Raw data
-
-The raw data can be downloaded from GitHub:
-
-```sh
-# solution counts
-wget https://github.com/bbartholdy/byoc-starch/blob/main/analysis/data/raw_data/solution_counts.csv
-
-# sample counts
-wget https://github.com/bbartholdy/byoc-starch/blob/main/analysis/data/raw_data/starch_counts.csv
-```
-
-Raw counts from the treatment solutions before extrapolation.
-
-```{r raw-data-sol, echo=FALSE}
-sol_counts
-```
-
-Raw counts from the calculus samples before extrapolation:
-
-```{r raw-data-samp, echo=FALSE}
-raw_counts
-```
-
-## Experimental setup
-
-```{r plate-lid-1, fig.cap="The 24 deepwell plate and the lid with pegs (substrata)"}
-knitr::include_graphics(here("analysis/figures/plate_lid_side.jpg"))
-```
-
-```{r plate-lid-2, fig.cap="The 24 deepwell plate with the lid (almost) on."}
-knitr::include_graphics(here("analysis/figures/plate_lid_on.jpg"))
-```
-
-## Protocols
-
-All protocols are available on [protocols.io](https://www.protocols.io).
-
-Creating the artificial saliva: <https://www.protocols.io/view/artificial-saliva-bva9n2h6>
-
-Creating the CPMU solution: <https://www.protocols.io/view/cpmu-bv8pn9vn>
-
-Biofilm growth protocol: <https://www.protocols.io/view/biofilm-growth-with-starch-treatment-bu7jnzkn>
-
-Amylase activity assay: <https://www.protocols.io/view/amylase-activity-bw8jphun>
-
-## Microscope images
+# Microscope images
 
 ```{r img-1, echo=FALSE, fig.cap="Image of starch granules extracted from a potato treatment sample"}
-knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
+knitr::include_graphics(here("analysis/supplementary-materials/figures/starches_w_bar.jpg"))
 ```
@@
 ```{r img-2}
-knitr::include_graphics(here("analysis/figures/SNAP-103412-0006.jpg"))
+knitr::include_graphics(here("analysis/supplementary-materials/figures/SNAP-103412-0006.jpg"))
 ```
@@
 ```{r img-3, fig.cap="Microscope image of wheat starch from a wheat treatment sample."}
-knitr::include_graphics(here("analysis/figures/SNAP-164650-0012.jpg"))
+knitr::include_graphics(here("analysis/supplementary-materials/figures/SNAP-164650-0012.jpg"))
 ```
 
-## Amylase activity
+# Amylase activity
 
-Amylase activity in U/mL enzyme, where U is mg maltose released from starch in six minutes at 36 &deg;C.
+Amylase activity in U/mL enzyme, where U is the amount of enzyme needed to
+release 1 $\mu$mole maltose from starch in six minutes at 36 &deg;C.
 
-Tables containing the amylase activity results for both plates and both photometric
-readings conducted on each plate. Samples (rows) were analysed in triplicates (columns).
+Tables containing the amylase activity results for both plates and both
+photometric readings conducted on each plate. Samples (rows) were analysed in
+triplicates (columns).
 
@@
 
-```{r amylase-activity-tables}
-plt1_ph1_result # plate 1, photometric reading 1
-plt1_ph2_result # plate 1, photometric reading 2
-plt2_ph1_result # plate 2, photometric reading 1
-plt2_ph2_result # plate 2, photometric reading 2
+```{r amylase-plt1-read1}
+plt1_ph1_result %>%
+  knitr::kable("Amylase activity in U/mL for plate 1, photometric read 1.")
 ```
 
-## Control samples
+```{r amylase-plt1-read2}
+plt1_ph2_result %>%
+  knitr::kable("Amylase activity in U/mL for plate 1, photometric read 2.")
+```
+
+```{r amylase-plt2-read1}
+plt2_ph1_result %>%
+  knitr::kable("Amylase activity in U/mL for plate 2, photometric read 1.")
+```
+
+```{r amylase-plt2-read2}
+plt2_ph2_result %>%
+  knitr::kable("Amylase activity in U/mL for plate 2, photometric read 2.")
+```
+
+# Control samples
 
@@
 
-## Count corrections
+# Count corrections
 
@@
 
-## Some additional plots
+# Some additional plots
 
@@
 
-### ...and a table
+## ...and a table
 

### 2022-02-28 analysis/paper/paper.Rmd
minor changes to paper

@@
 despite meticulous documentation and observation, due to unknown variables and 
-uncertainty involved when studying living organisms. Dental calculus is a complex 
+uncertainty involved in this kind of *in vivo* research. Dental calculus is a complex 
 oral biofilm with a multifactorial aetiology and variable formation rates both 
@@
 An $\alpha$-amylase (EC 3.2.1.1) activity
-assay was conducted to confirm that no amylase was present in the system before
+assay was conducted to confirm that no amylase was present in the model before
 starch treatments started. Starch treatments replaced sucrose treatments, occurring twice per day
@@
 H~2~O, while artificial saliva samples were compared against a standard curve
-containing sterile artificial saliva (due to the colour of artificial saliva).
+containing stock AS (due to the colour of artificial saliva).
 Two photometric readings were conducted for each plate with a 540 nm filter on a 
@@
 assumption based on the counts of small starches in the wheat-only and potato-only
-solutions. Of the total amount of small starches in these two solutions, 
+solutions. Of the combined amount of small starches in these two solutions, 
 `r paste0(signif((sol_comb_out$s[5] / sum(sol_comb_out$s[4], sol_comb_out$s[5])) * 100, 3), "%")`
@@
 lack of oral amylase activity in our model. 
+<!-- Scientific findings as extension of previous work -->
+Previous research conducted on dental calculus from contemporary humans and non-human
+primates suggest a high level of stochasticity involved in the retention of 
+starch granules in dental calculus, and that starch granules extracted from dental 
+calculus are underrepresented with regard
+to actual starch intake, which is consistent with our findings (illustrated by high
+standard deviations and low proportional incorporation).
+Leonard and colleagues [-@leonardDentalCalculus2015] found individual
+calculus samples to be a poor predictor of diet in a population, as many of the 
+consumed plants were missing from some individual samples, but were present in others.   
+Power and colleagues [-@powerChimpCalculus2015] 
+presented similar findings in non-human primates, where phytoliths were more
+representative of individual diets than starch granules.
+The size bias is also consistent with the findings by Power and colleagues
+[-@powerChimpCalculus2015],
+who found that plants producing starches 10--20 $\mu$m in
+size were over-represented; however, the representation of granules larger than
+20 $\mu$m in their study is unclear.
 
@@
 maximum length were underrepresented in the calculus samples compared to the original 
-starch solutions, an effect that was consistent across all three treatments, 
-while medium granules (10--20 $\mu$m) were often over-represented
+starch solutions, an effect that was consistent across all three treatments. Medium
+granules (10--20 $\mu$m) were often over-represented
 (Table \@ref(tab:sample-prop-tbl), 
@@
 findings.
-
-<!-- Scientific findings as extension of previous work -->
-Previous research conducted on dental calculus from contemporary humans and non-human
-primates suggest a high level of stochasticity involved in the retention of 
-starch granules in dental calculus, and that starch granules extracted from dental 
-calculus are underrepresented with regard
-to actual starch intake, which is consistent with our findings (illustrated by high
-standard deviations and low proportional incorporation).
-Leonard and colleagues [-@leonardDentalCalculus2015] found individual
-calculus samples to be a poor predictor of diet in a population, as many of the 
-consumed plants were missing from some individual samples, but were present in others.   
-Power and colleagues [-@powerChimpCalculus2015] 
-presented similar findings in non-human primates, where phytoliths were more
-representative of individual diets than starch granules.
-The size bias is also consistent with the findings by Power and colleagues
-[-@powerChimpCalculus2015],
-who found that plants producing starches 10--20 $\mu$m in
-size were over-represented; however, the representation of granules larger than
-20 $\mu$m in their study is unclear.
 

### 2022-02-28 analysis/scripts/amylase-assay.R
change units for amylase activity

@@ -1,9 +1,3 @@
-# if(require(byocstarch, quietly = T)) {
-#   library(byocstarch)
-# } else {
-#   devtools::load_all()
-# }
-
 library(tidyverse)
 
 # Functions ---------------------------------------------------------------
@@ -14,12 +8,13 @@ library(tidyverse)
 #' @param ml_maltose vector of length 7 containing the volume (ml) of 0.2% maltose solution added to the assay.
 #' @importFrom magrittr %>%
 #' @importFrom dplyr slice n
-stdcurve <- function(std_data, cols = NULL, ml_maltose = c(3.75, 15, 30, 45, 60, 75, 150)){
+stdcurve <- function(std_data, cols = NULL, ml_maltose = c(0.00375, 0.015, 0.030, 0.045, 0.060, 0.075, 0.150)){
   if(!is.null(cols)){
     std_data <- std_data[,cols]
   }
   # delta 540 calculation
-  mg_maltose <- 0.002 * ml_maltose
+  mg_maltose <- 2 * ml_maltose # DOUBLE CHECK THIS
+  umol_maltose <- (mg_maltose * 1000) / 342.3 # calculate maltose in umoles
   replicates <- dim(std_data)[2]
   pht_blank <- std_data[8,] %>%
     slice(rep(1:n(), each = 7))
@@ -28,14 +23,15 @@ stdcurve <- function(std_data, cols = NULL, ml_maltose = c(3.75, 15, 30, 45, 60,
   pht_meas <- c(pht_meas[, seq(1:replicates)]) %>%
     unlist()
   # Linear regression
-  stdCurve_data <- data.frame("d540" = pht_meas, "mgMaltose" = rep(mg_maltose, replicates))
-  std_curve <- lm(d540 ~ mgMaltose, stdCurve_data)
+  stdCurve_data <- data.frame("d540" = pht_meas, "umol_maltose" = rep(umol_maltose, replicates))
+  std_curve <- lm(d540 ~ umol_maltose, stdCurve_data)
   # Plot of standard curve
   p_stdcurve <- stdCurve_data %>%
-    ggplot(aes(x = mgMaltose, y = d540)) +
+    ggplot(aes(x = umol_maltose, y = d540)) +
     geom_point() +
     geom_smooth(formula = "y ~ x", method = "lm", se = F) +
-    theme_bw()
+    theme_bw() +
+    labs(x = "umoles Maltose", y = "Absorbance (d540)")
[…]

### 2022-02-28 analysis/scripts/analysis.R
update .gitignore

@@ -91,13 +91,16 @@ corr_counts_long <- corr_counts %>%
 
 # combining potato and wheat counts from mixed treatment samples
 
-# merge two adjacent rows and add variables s, m , l, total
-
+# merge Row C rows with the sum of counts
 corr_comb <- corr_counts %>%
-  select(!starch) %>%
-  pivot_wider(names_sep = "", values_from = c(s, m, l, total),
-              values_fill = NA,
-              values_fn = function(x) sum(x, na.rm = T))
+  group_by(sample, plate, row, treatment, weight) %>%
+  summarise(across(c(s,m,l,total), sum, na.rm = T))
+
+# corr_comb <- corr_counts %>%
+#   select(!starch) %>%
+#   pivot_wider(names_sep = "", values_from = c(s, m, l, total),
+#               values_fill = NA,
+#               values_fn = function(x) sum(x, na.rm = T))
 readr::write_csv(corr_comb, here("analysis/data/derived_data", "corr_comb.csv"))
 
 corr_comb_long <- corr_counts_long %>%

### 2022-02-28 analysis/scripts/tables.R

@@ -145,6 +145,13 @@ perc_comb_row <- prop %>%
   pivot_wider(names_from = size, values_from = perc)
 
 
+perc_tbl <- prop %>%
+  ungroup() %>%
+  select(!c(sample, solution, prop)) %>%
+  pivot_wider(names_from = size, values_from = perc) %>%
+  add_row(perc_comb_row, .after = 2)
+
+
 # Size ratios -------------------------------------------------------------
 
 sol_size_prop <- sol_comb_out %>%

### 2022-02-28 analysis/paper/abstract.Rmd
add S prefix to all tables and figs in supp mat

--- 2021-08-19 analysis/paper/abstract.Rmd
@@
+---
+output:
+  bookdown::pdf_document2:
+    keep_tex: true
+---
+
 # Abstract
+
+```{r include=FALSE}
+library(here)
+library(patchwork)
+source(here("analysis/scripts/analysis.R"))
+source(here("analysis/scripts/tables.R"))
+perc_tbl <- prop %>%
+  ungroup() %>%
+  select(!c(sample, solution, prop)) %>%
+  pivot_wider(names_from = size, values_from = perc) %>%
+  add_row(perc_comb_row, .after = 2)
+# prepare cor output for reporting
+cor_r <- signif(count_cor$estimate, 3)
+cor_ci <- signif(count_cor$conf.int, 3)
+cor2_r <- signif(starch_cor$estimate, 3)
+cor2_ci <- signif(starch_cor$conf.int, 3)
+
+#w_count <- wordcountaddin::word_count()
+```
 
@@
 <!-- Two or three sentences to provide a broader perspective, readily comprehensible to a scientist in any discipline, may be included in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced by their inclusion.
-Under these circumstances, the length of the paragraph can be yp to 300 words.
-(This example is 190 words without the final section, and 250 words with it).
+-->
+
+Dental calculus has proven to contain a wealth of information 
+on the dietary habits of past populations. These insights have, to a large extent,
+been obtained by the extraction and identification of
+starch granules contained within the mineralised dental plaque from a wide range 
+of regions and time periods. The scope of previous studies have been limited to 
+microfossil extraction and identification to reconstruct dietary preferences from
+the archaeological record, and few studies have attempted to address the biases 
+of starch retention in dental calculus. Those that have considered this problem 
+have been limited to *in vivo* studies on modern humans and non-human primates.
+Here, we present a multispecies oral biofilm model, which allows experimental 
+research on  starch incorporation and retention to be conducted on *in vitro* 
+dental calculus in a controlled laboratory setting. The biofilms were exposed 
+to treatment solutions with known quantities of
+dietary starches (wheat and potato) during the 25-day growth period. After this,
+the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
+and counted. We show that the granule counts extracted from the model dental calculus 
+represented a low proportion (ranging from `r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`) of the total number 
+of granules exposed to the biofilms throughout the experiment.
+Additionally, we found that the ratios of granule sizes from the extracted 
+starch granules differed from the original treatment solutions, with large 
+granules (>20 $\mu$m) consistently being under-represented.
+We also found a correlation
+between the absolute granule counts and dry-weight of the biofilm 
+(*r* = `r cor_r`, 90%CI[`r paste(cor_ci[1], cor_ci[2], sep = ", ")`]), as well
+as between the concentration (count per mg) of granules and dry-weight 
+(*r* = `r cor2_r`, 90%CI[`r paste(cor2_ci[1], cor2_ci[2], sep = ", ")`]).  
+Our results reinforce previous *in vivo* studies suggesting that dental calculus 
+presents a very small, and partly biased picture of the original dietary intake 
+of starches, with an over-representation of plants producing granules smaller than
+20 $\mu$m in size. The experimental model presented here is well-suited to address
+the need for further validation of methods and biases associated with dietary
+research on dental calculus.

### 2022-02-28 analysis/paper/fig_tab.Rmd
add S prefix to all tables and figs in supp mat

@@
 
-## Tables
-<!-- ### Tables ### -->
+# Tables
 
@@
 
-```{r sample-count}
-count_sample <- count_sample %>%
-  ungroup() %>%
-  add_row(count_sample_comb, .after = 2)
-
-# convert columns to contain mean count and percentage of total count
-count_samp_per <- count_sample %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), ")")))
-
-count_samp_out <- tibble(count_samp_per, count_sample_sd, .name_repair = "unique") %>%
-  select(1,2,3,7,4,8,5,9,6,10) #%>%
-
-count_samp_out[1,3] <- NA # replace with NA
-  #add_row(tibble_row(count_sample_comb_out), .before = 3)
-```
-
-```{r solution-count-out}
-count_sol_out <- count_sol %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
-
-### adding the 'both' row to the data frame ###
-count_sol_comb_out <- count_sol_comb %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
-
-count_sol_out <- count_sol %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")"))) %>%
-  ungroup() %>%
-  add_row(tibble_row(count_sol_comb_out), .before = 3)
-
-count_sol_out[1,3] <- NA
-```
-
 ```{r solution-count-tbl}
@@
 
-```{r sample-perc}
-# convert proportions to percent labels
-perc_comb <- prop_comb %>% 
-  mutate(across(c(s, m, l, total), scales::percent, 0.001))
-
-perc <- prop %>%
-  mutate(across(c(s, m, l, total), scales::percent, 0.001)) %>% 
-  add_row(tibble_row(perc_comb), .before = 3)
-```
-
 ```{r sample-prop-tbl}
 cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
-knitr::kable(perc, col.names = cnames, 
+knitr::kable(perc_tbl, col.names = cnames, 
              caption = "The mean percentage of starches from the solutions that 
@@
 
-## Figures
-<!-- ### Figures ### -->
+# Figure Captions
 
@@
 
-```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments."}
-pl_cor
-#pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+```{r cor-plot, warning=FALSE, fig.cap="Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus."}
+#pl_cor
+pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+#ggsave("frontiers/figures/cor-plot.png", dpi = 300)
 ```
-
-```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
-pl_cor2
-```

### 2022-02-28 analysis/paper/paper.tex
add S prefix to all tables and figs in supp mat

@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
+% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.
+%                                              %
+%                                                                                                                                                         %
+% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%% Version 3.4 Generated 2018/06/15 %%%
+%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
+%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%
+
+\documentclass[utf8]{frontiers/frontiersSCNS}
+
+%\setcitestyle{square} % for Physics and Applied Mathematics and Statistics articles
+\usepackage{url,hyperref,lineno,microtype,subcaption}
+\usepackage[onehalfspacing]{setspace}
+
+\linenumbers
+
+% BELOW TAKEN FROM rticles plos template
 %
-\documentclass[
-]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model},
-  pdfkeywords={oral biofilm, starch, alpha-amylase, dental calculus},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
+% amsmath package, useful for mathematical formulas
+\usepackage{amsmath}
+% amssymb package, useful for mathematical symbols
+\usepackage{amssymb}
+
+% hyperref package, useful for hyperlinks
+\usepackage{hyperref}
+
+% graphicx package, useful for including eps and pdf graphics
+% include graphics with the command \includegraphics
+\usepackage{graphicx}
+
+% Sweave(-like)
+\usepackage{fancyvrb}
+\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
+\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
+\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
+\newenvironment{Schunk}{}{}
+\DefineVerbatimEnvironment{Code}{Verbatim}{}
+\DefineVerbatimEnvironment{CodeInput}{Verbatim}{fontshape=sl}
+\DefineVerbatimEnvironment{CodeOutput}{Verbatim}{}
+\newenvironment{CodeChunk}{}{}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+\usepackage{color}
+
+% Below is from frontiers
+%
+\bibliographystyle{frontiersinSCNS}
+% Use doublespacing - comment out for single spacing
+%\usepackage{setspace}
+%\doublespacing
+
+% Leave a blank line between paragraphs instead of using \\
+
+\def\keyFont{\fontsize{8}{11}\helveticabold }
+
+%% ** EDIT HERE **
+%% PLEASE INCLUDE ALL MACROS BELOW
+
+%% END MACROS SECTION
+
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% From pandoc table feature
 \usepackage{longtable,booktabs,array}
@@
 \makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-\usepackage{fancyhdr}
-\fancyhf{}
-\pagestyle{fancy}
-\renewcommand{\headrulewidth}{0pt}
-\fancyheadoffset{0pt}
-\rhead{\scshape A preprint - \today}
-\cfoot{\thepage}
-
-%Handling Keywords
-\def\keywordname{{\bfseries \emph Keywords}}%
-\def\keywords#1{\par\addvspace\medskipamount{\rightskip=0pt plus1cm
-\def\and{\ifhmode\unskip\nobreak\fi\ $\cdot$
-}\noindent\keywordname\enspace\ignorespaces#1\par}}
-
-% create title
-% \providecommand{\maketitle}{}
-% \renewcommand{\maketitle}{%
-%   \par
-%   \begingroup
-%     \renewcommand{\thefootnote}{\fnsymbol{footnote}}
-%     % for perfect author name centering
-%     \renewcommand{\@makefnmark}{\hbox to \z@{$^{\@thefnmark}$\hss}}
-%     % The footnote-mark was overlapping the footnote-text,
-%     % added the following to fix this problem               (MK)
-%     \long\def\@makefntext##1{%
-%       \parindent 1em\noindent
-%       \hbox to 1.8em{\hss $\m@th ^{\@thefnmark}$}##1
-%     }
-%     \thispagestyle{empty}
-%     \@maketitle
-%     \@thanks
-%     %\@notice
-%   \endgroup
-%   \let\maketitle\relax
-%   \let\thanks\relax
-% }
-
-% rules for title box at top of first page
-% \newcommand{\@toptitlebar}{
-%   \hrule height 2\p@
-%   \vskip 0.25in
-%   \vskip -\parskip%
-% }
-% \newcommand{\@bottomtitlebar}{
-%   \vskip 0.29in
-%   \vskip -\parskip
-%   \hrule height 2\p@
-%   \vskip 0.09in%
-% }
-
-% create title (includes both anonymized and non-anonymized versions)
-% \providecommand{\@maketitle}{}
-% \renewcommand{\@maketitle}{%
-%   \vbox{%
-%     \hsize\textwidth
-%     \linewidth\hsize
-%     \vskip 0.1in
-%     \@toptitlebar
-%     \centering
-%     {\LARGE\sc \@title\par}
-%     \@bottomtitlebar
-%     \textsc{A Preprint}\\
-%     \vskip 0.1in
-%     \def\And{%
-%       \end{tabular}\hfil\linebreak[0]\hfil%
-%       \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\ignorespaces%
-%     }
-%     \def\AND{%
-%       \end{tabular}\hfil\linebreak[4]\hfil%
-%       \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\ignorespaces%
-%     }
-%     \begin{tabular}[t]{c}\bf\rule{\z@}{24\p@}\@author\end{tabular}%
-%   \vskip 0.4in \@minus 0.1in \center{\today}   \vskip 0.2in
-%   }
-% }
-
-% add conference notice to bottom of first page
-% \newcommand{\ftype@noticebox}{8}
-% \newcommand{\@notice}{%
-%   % give a bit of extra room back to authors on first page
-%   \enlargethispage{2\baselineskip}%
-%   \@float{noticebox}[b]%
-%     \footnotesize\@noticestring%
-%   \end@float%
-% }
-
-% abstract styling
-\renewenvironment{abstract}
-{
-  \centerline
-  {\large \bfseries \scshape Abstract}
-  \begin{quote}
-}
-{
-  \end{quote}
-}
-
-\usepackage{lineno}
-\linenumbers
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
+
+% Pandoc citation processing
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
 \newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
@@
   % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
   % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
+  \setlength{\parskip}{#2\cslentryspacingunit}
  }%
@@
 
-\title{Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model}
-\author{Bjørn Peare Bartholdy\(^1\), Amanda G. Henry\(^1\)}
-\date{\(^1\) Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.}
+\usepackage{float}
+\let\origfigure\figure
+\let\endorigfigure\endfigure
+\renewenvironment{figure}[1][2] {
+    \expandafter\origfigure\expandafter[H]
+} {
+    \endorigfigure
+}
+\usepackage{float}
+
+\def\Authors{
+  Bjørn Peare Bartholdy\,\textsuperscript{1*},
+  Amanda G. Henry\,\textsuperscript{1}}
+
+\def\Address{
+
+  \textsuperscript{1} Archaeological Sciences, Leiden University,  Leiden,   Netherlands
+  }
+
+  \def\corrAuthor{Bjørn Peare Bartholdy}\def\corrAddress{Leiden University\\Einsteinweg 2\\Leiden, 2333 CC Netherlands}\def\corrEmail{\href{mailto:b.p.bartholdy@arch.leidenuniv.nl}{\nolinkurl{b.p.bartholdy@arch.leidenuniv.nl}}}
+  \def\firstAuthorLast{BARTHOLDY {et~al.}}
 
 \begin{document}
+
+\onecolumn
+\firstpage{1}
+
+\title[Biases Associated with Starch Incorporation]{Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model}
+\author[\firstAuthorLast]{\Authors}
+\address{} %This field will be automatically populated
+\correspondance{} %This field will be automatically populated
+
+\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
+%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}
+
 \maketitle
+
 \begin{abstract}
-Dental calculus has proven to contain a wealth of information
-on the dietary habits of past populations. These insights have, to a large extent,
-been obtained by the extraction and identification of
-starch granules contained within the mineralised dental plaque from a wide range
-of regions and time periods. The scope of previous studies have been limited to
-microfossil extraction and identification to reconstruct dietary preferences from
-the archaeological record, and few studies have attempted to address the biases
-of starch retention in dental calculus. Those that have considered this problem
-have been limited to \emph{in vivo} studies on modern humans and non-human primates.
-Here, we present a multispecies oral biofilm model, which allows experimental
-research on starch incorporation and retention to be conducted on \emph{in vitro}
-dental calculus in a controlled laboratory setting. The biofilms were exposed
-to treatment solutions with known quantities of
-dietary starches (wheat and potato) during the 25-day growth period. After this,
-the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
-and counted. We show that the granule counts extracted from the model dental calculus
-represented a low proportion (ranging from 0.06\% to 0.16\%) of the total number
-of granules exposed to the biofilms throughout the experiment.
-Additionally, we found that the ratios of granule sizes from the extracted
-starch granules differed from the original treatment solutions, with large
-granules (\textgreater20 \(\mu\)m) consistently being under-represented.
-We also found a correlation
-between the absolute granule counts and dry-weight of the biofilm
-(\emph{r} = 0.66, 90\%CI{[}0.46,0.79{]}), as well
-as between the concentration (count per mg) of granules and dry-weight
-(\emph{r} = 0.30, 90\%CI{[}0.06,0.51{]}).\\
-Our results reinforce previous \emph{in vivo} studies suggesting that dental calculus
-presents a very small, and partly biased picture of the original dietary intake
-of starches, with an over-representation of plants producing granules smaller than
-20 \(\mu\)m in size. The experimental model presented here is well-suited to address
-the need for further validation of methods and biases associated with dietary
+
+Dental calculus has proven to contain a wealth of information on the
+dietary habits of past populations. These insights have, to a large
+extent, been obtained by the extraction and identification of starch
+granules contained within the mineralised dental plaque from a wide
+range of regions and time periods. The scope of previous studies have
+been limited to microfossil extraction and identification to reconstruct
+dietary preferences from the archaeological record, and few studies have
+attempted to address the biases of starch retention in dental calculus.
+Those that have considered this problem have been limited to \emph{in
+vivo} studies on modern humans and non-human primates. Here, we present
+a multispecies oral biofilm model, which allows experimental research on
+starch incorporation and retention to be conducted on \emph{in vitro}
+dental calculus in a controlled laboratory setting. The biofilms were
+exposed to treatment solutions with known quantities of dietary starches
+(wheat and potato) during the 25-day growth period. After this, the
+starch granules were extracted from the mature biofilm (by dissolution
+in EDTA), and counted. We show that the granule counts extracted from
+the model dental calculus represented a low proportion (ranging from
+0.06\% to 0.16\%) of the total number of granules exposed to the
+biofilms throughout the experiment. Additionally, we found that the
+ratios of granule sizes from the extracted starch granules differed from
+the original treatment solutions, with large granules (\textgreater20
+\(\mu\)m) consistently being under-represented. We also found a
+correlation between the absolute granule counts and dry-weight of the
+biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
+the concentration (count per mg) of granules and dry-weight (\emph{r} =
+0.3, 90\%CI{[}0.0618, 0.506{]}).\\
+Our results reinforce previous \emph{in vivo} studies suggesting that
+dental calculus presents a very small, and partly biased picture of the
+original dietary intake of starches, with an over-representation of
+plants producing granules smaller than 20 \(\mu\)m in size. The
+experimental model presented here is well-suited to address the need for
+further validation of methods and biases associated with dietary
 research on dental calculus.
+
+\tiny
+ \keyFont{ \section{Keywords:} oral biofilm model, starch, alpha-amylase, dental calculus, paleodiet reconstruction}
+
 \end{abstract}
@@
 dietary information in the form of plant microfossils
-(Hardy et al., 2009; Henry \& Piperno, 2008),
-proteins (Hendy et al., 2018; Warinner, Hendy, et al., 2014),
+(Henry and Piperno, 2008; Hardy et al., 2009),
+proteins (Hendy et al., 2018; Warinner et al., 2014a),
 and other organic residues (Buckley et al., 2014).
@@
 past populations
-(Henry \& Piperno, 2008; Jovanović et al., 2021; Tao et al., 2020)
-and extinct species (Hardy et al., 2012; Henry et al., 2014).
+(Henry and Piperno, 2008; Jovanović et al., 2021; Tao et al., 2020)
+and extinct species (Henry et al., 2014; Hardy et al., 2012).
 
@@
 characterisation (including pathogens) through the analysis of DNA and proteins
-(Adler et al., 2013; Warinner, Rodrigues, et al., 2014),
-microbotanical remains (Hardy et al., 2009; Henry \& Piperno, 2008; Mickleburgh \& Pagán-Jiménez, 2012),
+(Adler et al., 2013; Warinner et al., 2014b),
+microbotanical remains (Henry and Piperno, 2008; Hardy et al., 2009; Mickleburgh and Pagán-Jiménez, 2012),
 other organic residues and proteins from dietary compounds
@@
 over a variety of geographical and temporal ranges
-(Henry et al., 2014; Jovanović et al., 2021; Piperno \& Dillehay, 2008; Tao et al., 2020).
+(Piperno and Dillehay, 2008; Henry et al., 2014; Tao et al., 2020; Jovanović et al., 2021).
 
@@
 oral biofilm with a multifactorial aetiology and variable formation rates both
-within and between individuals (Haffajee et al., 2009; Jepsen et al., 2011),
+within and between individuals (Jepsen et al., 2011; Haffajee et al., 2009),
 contributing to
@@
 and 1 mmol/l arginine (Sissons et al., 1991).
-
-\begin{figure}
-
-{\centering \includegraphics{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview}
-
-}
-
-\caption{Overview of experiment protocol including the plate setup.}\label{fig:protocol-fig}
-\end{figure}
 
@@
 and (0.04 g/l MgCl)
-(Pearce \& Sissons, 1987; Sissons et al., 1991).
+(Pearce and Sissons, 1987; Sissons et al., 1991).
 The substrata were submerged in 1 ml/well CPMU for six minutes, five times
@@
 Multiskan FC Microplate Photometer (Thermo Scientific 51119000).
-The protocol is a slightly modified version of an Enzymatic Assay of \(\alpha\)-Amylase
+The protocol is a modified version of an Enzymatic Assay of \(\alpha\)-Amylase
 (\url{https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase}) (Bernfeld, 1955), which measures the amount of
 maltose released from starch by \(\alpha\)-amylase activity. Results are reported
-in units (U) per mL enzyme, where 1 U releases 1 mg of maltose in 6 minutes.
+in units (U) per mL enzyme, where 1 U releases 1 \(\mu\)mole of maltose in 6 minutes.
+The detailed protocol can be found here: \url{https://www.protocols.io/view/amylase-activity-bw8jphun}.
 
@@
 calculus in 0.5 \(\tiny{M}\) ethylenediaminetetraacetic acid (EDTA)
-(Le Moyne \& Crowther, 2021; Modi et al., 2020; Tromp et al., 2017),
+(Tromp et al., 2017; Le Moyne and Crowther, 2021; Modi et al., 2020),
 and vortexing for 3 days until the sample was completely dissolved.
@@
 
-Statistical analysis was conducted in R version 4.1.1 (2021-08-10) (R Core Team, 2020) and
+Statistical analysis was conducted in R version 4.1.2 (2021-11-01) (R Core Team, 2020) and
 the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=0.5\linewidth]{/home/bjorn/Documents/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar}
-
-}
-
-\caption{Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\mu$m.}\label{fig:microscope-fig}
-\end{figure}
-
 All samples yielded sufficient biofilm growth and starch incorporation to be
@@
 saliva samples from any of the days that were sampled. Only positive controls
+(saliva)
 contained amylase activity that could be detected in the assay, ranging from
-3.4 to 10.3 U/mL enzyme
+9.93 to 30.2 U/mL enzyme
 (full results can be found in the Supplementary Material).
@@
 levels in humans, as the unit definition may differ;
-however, they are sufficient to show that there is no activity in the system.
+however, they are sufficient to show that there is no activity in the model.
 
@@
 \subsection{Treatment type had minimal effect on biofilm growth}\label{treatment-type-had-minimal-effect-on-biofilm-growth}}
-
-\begin{table}
-
-\caption{\label{tab:anova-tbl}Summary statistics for biofilm dry-weights (in mg) by treatment.}
-\centering
-\begin{tabular}[t]{l|r|r|r|r}
-\hline
-Treatment & Mean & SD & Min & Max\\
-\hline
-control & 5.44 & 2.45 & 1.67 & 11.20\\
-\hline
-mix & 4.28 & 1.95 & 1.50 & 8.44\\
-\hline
-potato & 6.25 & 2.07 & 2.54 & 8.92\\
-\hline
-wheat & 5.53 & 3.45 & 0.56 & 9.80\\
-\hline
-\end{tabular}
-\end{table}
 
@@
 are from wheat.
-
-\begin{table}
-
-\caption{\label{tab:solution-count-tbl}Mean starch counts from solutions, including the
-             proportional makeup of the different sizes of granules.}
-\centering
-\begin{tabular}[t]{l|l|l|l|l|l}
-\hline
-Solution & Starch & Small (\%) & Medium (\%) & Large (\%) & Total (\%)\\
-\hline
-mix & potato &  & 1051733 (53.1\%) & 928000 (46.9\%) & 1979733 (100.0\%)\\
-\hline
-mix & wheat & 18838400 (69.7\%) & 6403200 (23.7\%) & 1794133 (6.6\%) & 27035733 (100.0\%)\\
-\hline
-mix & both & 18838400 (64.9\%) & 7454933 (25.7\%) & 2722133 (9.4\%) & 29015467 (100.0\%)\\
-\hline
-potato & potato & 123733 (4.1\%) & 1337867 (44.4\%) & 1554400 (51.5\%) & 3016000 (100.0\%)\\
-\hline
-wheat & wheat & 16139467 (63.5\%) & 6434133 (25.3\%) & 2830400 (11.1\%) & 25404000 (100.0\%)\\
-\hline
-\end{tabular}
-\end{table}
-
-\begin{table}
-
-\caption{\label{tab:sample-count-tbl}Mean starch counts extracted from samples with standard 
-             deviation (SD), including the proportion of granule sizes of the total count.}
-\centering
-\begin{tabular}[t]{l|l|l|r|l|r|l|r|l|r}
-\hline
-Treatment & Starch & Small (\%) & SD & Medium (\%) & SD & Large (\%) & SD & Total (\%) & SD\\
-\hline
-mix & potato &  &  & 1959 (79.6\%) & 1801 & 501 (20.40\%) & 446 & 2460 (100\%) & 2189\\
-\hline
-mix & wheat & 9515 (54.60\%) & 8860 & 6522 (37.4\%) & 6026 & 1381 (7.93\%) & 1196 & 17417 (100\%) & 15878\\
-\hline
-mix & both & 9515 (47.90\%) & 8860 & 8480 (42.7\%) & 7653 & 1882 (9.47\%) & 1596 & 19877 (100\%) & 17768\\
-\hline
-potato & potato & 351 (7.24\%) & 297 & 3565 (73.6\%) & 2402 & 930 (19.20\%) & 929 & 4846 (100\%) & 3316\\
-\hline
-wheat & wheat & 15235 (55.00\%) & 11944 & 12148 (43.9\%) & 11052 & 1953 (7.06\%) & 2016 & 27680 (100\%) & 23554\\
-\hline
-\end{tabular}
-\end{table}
 
@@
 
-\begin{table}
-
-\caption{\label{tab:sample-prop-tbl}The mean percentage of starches from the solutions that 
-              were incorporated into the samples.}
-\centering
-\begin{tabular}[t]{l|l|l|l|l|l}
-\hline
-Treatment & Starch & Small & Medium & Large & Total\\
-\hline
-mix & potato &  & 0.19\% & 0.05\% & 0.12\%\\
-\hline
-mix & wheat & 0.05\% & 0.10\% & 0.08\% & 0.06\%\\
-\hline
-mix & both & 0.05\% & 0.11\% & 0.07\% & 0.07\%\\
-\hline
-potato & potato & 0.28\% & 0.27\% & 0.06\% & 0.16\%\\
-\hline
-wheat & wheat & 0.09\% & 0.19\% & 0.07\% & 0.12\%\\
-\hline
-\end{tabular}
-\end{table}
-
 The proportion of total starches from the solutions that were incorporated
@@
 
-\begin{figure}
-
-{\centering \includegraphics{../figures/ratio-plots-1}
-
-}
-
-\caption{Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.}\label{fig:ratio-plots}
-\end{figure}
-
 Overall, medium starch granules had a higher mean rate of incorporation
@@
 \subsubsection{Biofilm weight correlated positively with extracted starch counts}\label{biofilm-weight-correlated-positively-with-extracted-starch-counts}}
-
-\begin{figure}
-
-{\centering \includegraphics{../figures/cor-plot-1}
-
-}
-
-\caption{Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus.}\label{fig:cor-plot}
-\end{figure}
 
@@
 archaeological remains
-(Tromp et al., 2017; Tromp \& Dudgeon, 2015; Wesolowski et al., 2010), which could in part be due to the
+(Tromp and Dudgeon, 2015; Tromp et al., 2017; Wesolowski et al., 2010), which could in part be due to the
 lack of oral amylase activity in our model.
@@
 granules generally only reach up to 35 \(\mu\)m
-(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, pp. 174--176).
+(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, 174--176).
 Granule morphology may also play a role. Large wheat granules
@@
 compared to surface area
-(Jane et al., 1994; Reichert, 1913, pp. 364--365; Seidemann, 1966, pp. 174--176; van de Velde et al., 2002).
+(van de Velde et al., 2002; Jane et al., 1994; Reichert, 1913, 364--365; Seidemann, 1966, 174--176).
 Another potentially important factor
@@
 a result that contradicts findings from archaeological contexts
-(Dudgeon \& Tromp, 2014; Wesolowski et al., 2010).
+(Wesolowski et al., 2010; Dudgeon and Tromp, 2014).
 When the concentration of starch granules
@@
 archaeological samples of considerable age
-(Buckley et al., 2014; Henry et al., 2014; Wu et al., 2021).
+(Henry et al., 2014; Buckley et al., 2014; Wu et al., 2021).
 Studies have also shown that not all starch granules come from a dietary source.
@@
 as palm phytoliths adhering to the skin of sweet potatoes
-(Tromp \& Dudgeon, 2015),
+(Tromp and Dudgeon, 2015),
 or accidental ingestion not related to food consumption
@@
 tooth surfaces (including plaque and dental calculus) and mucous membranes
-(Dodds \& Edgar, 1988; S. Kashket et al., 1991).
+(Kashket et al., 1991; Dodds and Edgar, 1988).
 Bacteria also have the ability to adhere to starch granules
@@
 susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
-by \(\alpha\)-amylase (S. Kashket et al., 1996).
+by \(\alpha\)-amylase (Kashket et al., 1996).
 The susceptibility of granules to hydrolysis depends on the crystallinity and size
@@
 while dehydrated starches will have a reduced susceptibility
-(Björck et al., 1984; Franco et al., 1992; Haslam, 2004; Henry et al., 2009; Lingstrom et al., 1994).
+(Lingstrom et al., 1994; Franco et al., 1992; Haslam, 2004; Björck et al., 1984; Henry et al., 2009).
 Cracks on the surface of the dental calculus, as well as unmineralised islands
 and channels may also be able to contain starch granules
-(Charlier et al., 2010; Power et al., 2014; Tan, Gillam, et al., 2004).
+(Power et al., 2014; Tan et al., 2004a; Charlier et al., 2010).
 Starch granules that are trapped in these pockets are (at least to some extent)
@@
 a portion of its hydrolytic activity
-(Nikitkova et al., 2013; Scannapieco et al., 1993; Tan, Mordan, et al., 2004; Tan, Gillam, et al., 2004).
+(Scannapieco et al., 1993; Nikitkova et al., 2013; Tan et al., 2004b, 2004a).
 After the death of an individual, starches within dental calculus are susceptible
@@
 (pH, temperature, moisture content, microorganisms)
-(Franco et al., 1992; García-Granero, 2020; Haslam, 2004; Henry et al., 2009).
+(Henry et al., 2009; García-Granero, 2020; Haslam, 2004; Franco et al., 1992).
 Future study should explore how burial affects the recovery of starch from the
@@
 susceptible to hydrolysis
-(Franco et al., 1992; Haslam, 2004); however,
-the absence can also allow us to directly explore
-the effect of \(\alpha\)-amylase on starch counts in future experiments,
-where \(\alpha\)-amylase can be added to the model
-in concentrations similar to those found in the oral cavity (Scannapieco et al., 1993).
+(Franco et al., 1992; Haslam, 2004). The absence may
+also affect biofilm growth due to the lack of amylase-bacterium interactions
+(Nikitkova et al., 2013). Conversely, the model may benefit from the
+absence of \(\alpha\)-amylase, because it can allow us to directly explore
+its effect on starch counts in future experiments, where \(\alpha\)-amylase can be
+added to the model in concentrations similar to those found in the oral cavity
+(Scannapieco et al., 1993).
 We are able to show how absolute counts in the treatments cause a difference in
@@
 negligible morphological changes caused by exposure to EDTA
-(Le Moyne \& Crowther, 2021; Modi et al., 2020; Tromp et al., 2017),
+(Tromp et al., 2017; Le Moyne and Crowther, 2021; Modi et al., 2020),
 these studies have not considered changes to separate size
@@
 baseline testing of biases associated with dietary research conducted on dental calculus
-(Le Moyne \& Crowther, 2021).
+(Le Moyne and Crowther, 2021).
 Our high-throughput experimental setup allows us a
@@
 morphology
-(Fagernäs et al., 2021; Haffajee et al., 2009; Jepsen et al., 2011; Proctor et al., 2018; Simón-Soro et al., 2013),
-as well as evolutionary differences (Yates et al., 2021).
+(Jepsen et al., 2011; Simón-Soro et al., 2013; Proctor et al., 2018; Fagernäs et al., 2021; Haffajee et al., 2009),
+as well as evolutionary differences (Fellows Yates et al., 2021).
 It can also facilitate training of students and researchers on methods of
@@
 
+\hypertarget{open-science-statement}{%
+\section*{Open Science Statement}\label{open-science-statement}}
+\addcontentsline{toc}{section}{Open Science Statement}
+
+All scripts and data used in the analysis are available on OSF
+(\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}),
+following the format provided by the rrtools package (Marwick, 2019).
+More detailed protocols are available on OSF (\url{https://osf.io/akevs/})
+and protocols.io (\url{https://www.protocols.io/workspaces/byoc}).
+Additional tables and figures are available in the Supplementary Material
+(\url{https://osf.io/ucxsv/}).
+
+\hypertarget{author-contributions}{%
+\section*{Author Contributions}\label{author-contributions}}
+\addcontentsline{toc}{section}{Author Contributions}
+
+AGH acquired the funding for the analysis. BPB and AGH conceptualised and designed
+the study. BPB conducted the experiments and analysed the data.
+BPB wrote the manuscript. AGH reviewed and edited the manuscript.
+
+\hypertarget{funding}{%
+\section*{Funding}\label{funding}}
+\addcontentsline{toc}{section}{Funding}
+
+This research has received funding from the European Research Council under the
+European Union's Horizon 2020 research and innovation program, grant agreement
+number STG--677576 (``HARVEST'').
+
+\hypertarget{conflict-of-interest}{%
+\section*{Conflict of Interest}\label{conflict-of-interest}}
+\addcontentsline{toc}{section}{Conflict of Interest}
+
+The authors declare that the research was conducted in the absence of any
+commercial or financial relationships that could be construed as a potential
+conflict of interest.
+
 \hypertarget{acknowledgements}{%
@@
 
-This research has received funding from the European Research Council under the
-European Union's Horizon 2020 research and innovation program, grant agreement
-number STG--677576 (``HARVEST'').
-
-\hypertarget{data-availability-statement}{%
-\section*{Data Availability Statement}\label{data-availability-statement}}
-\addcontentsline{toc}{section}{Data Availability Statement}
-
-All scripts and data used in the analysis are available on OSF
-(\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}),
-following the format provided by the rrtools package (Marwick, 2019).
-More detailed protocols are available on OSF (\url{https://osf.io/akevs/}) and protocols.io (\url{https://www.protocols.io/workspaces/byoc}).
-Additional tables and figures are available in the Supplementary Material.
-
-\hypertarget{references-cited}{%
-\section*{References cited}\label{references-cited}}
-\addcontentsline{toc}{section}{References cited}
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
 
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-adlerSequencingAncientCalcified2013}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-adlerSequencingAncientCalcified2013}{}}%
 Adler, C.
@@
 J., Townsend, G., Soltysiak, A., Alt, K.
-W., Parkhill, J., \& Cooper, A.
+W., et al.
 (2013).
 Sequencing ancient calcified dental plaque shows changes in oral microbiota with dietary shifts of the {Neolithic} and {Industrial} revolutions.
-\emph{Nature Genetics}, \emph{45}(4), 450--455, 455e1.
-\url{https://doi.org/10.1038/ng.2536}
-
-\leavevmode\hypertarget{ref-armitageExtractionIdentificationOpal1975}{}%
-Armitage, P. L. (1975). The {Extraction} and {Identification} of {Opal Phytoliths} from the {Teeth} of {Ungulates}. \emph{Journal of Archaeological Science}, \emph{2}, 187--197.
-
-\leavevmode\hypertarget{ref-bernfeldAmylase1955}{}%
-Bernfeld, P. (1955). Amylases, {\(\alpha\)} and {\(\beta\)}. In \emph{Methods in {Enzymology}} (Vol. 1, pp. 149--158). {Academic Press}. \url{https://doi.org/10.1016/0076-6879(55)01021-5}
-
-\leavevmode\hypertarget{ref-bjorckStarchProcessing1984}{}%
-Björck, I., Asp, N.-G., Birkhed, D., Eliasson, A.-C., Sjöberg, L.-B., \& Lundquist, I.
+\emph{Nature Genetics} 45, 450--5, 455e1.
+doi:\href{https://doi.org/10.1038/ng.2536}{10.1038/ng.2536}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-armitageExtractionIdentificationOpal1975}{}}%
+Armitage, P. L. (1975). The {Extraction} and {Identification} of {Opal Phytoliths} from the {Teeth} of {Ungulates}. \emph{Journal of Archaeological Science} 2, 187--197.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bernfeldAmylase1955}{}}%
+Bernfeld, P.
+(1955).
+{``Amylases, {\(\alpha\)} and {\(\beta\)},''} in \emph{Methods in {Enzymology}} ({Academic Press}), 149--158.
+doi:\href{https://doi.org/10.1016/0076-6879(55)01021-5}{10.1016/0076-6879(55)01021-5}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bjorckStarchProcessing1984}{}}%
+Björck, I., Asp, N.-G., Birkhed, D., Eliasson, A.-C., Sjöberg, L.-B., and Lundquist, I.
 (1984).
@@
 {II}.
-{Drum}-drying of wheat flour.
-\emph{Journal of Cereal Science}, \emph{2}(3), 165--178.
-\url{https://doi.org/10.1016/S0733-5210(84)80030-2}
-
-\leavevmode\hypertarget{ref-buckleyDentalCalculusCooking2014}{}%
-Buckley, S., Usai, D., Jakob, T., Radini, A., \& Hardy, K.
+{Drum-drying} of wheat flour.
+\emph{Journal of Cereal Science} 2, 165--178.
+doi:\href{https://doi.org/10.1016/S0733-5210(84)80030-2}{10.1016/S0733-5210(84)80030-2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-buckleyDentalCalculusCooking2014}{}}%
+Buckley, S., Usai, D., Jakob, T., Radini, A., and Hardy, K.
 (2014).
 Dental {Calculus Reveals Unique Insights} into {Food Items}, {Cooking} and {Plant Processing} in {Prehistoric Central Sudan}.
-\emph{PLOS ONE}, \emph{9}(7), e100808.
-\url{https://doi.org/10.1371/journal.pone.0100808}
-
-\leavevmode\hypertarget{ref-charlierSEMCalculus2010}{}%
-Charlier, P., Huynh-Charlier, I., Munoz, O., Billard, M., Brun, L., \& Grandmaison, G.
+\emph{PLOS ONE} 9, e100808.
+doi:\href{https://doi.org/10.1371/journal.pone.0100808}{10.1371/journal.pone.0100808}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-charlierSEMCalculus2010}{}}%
+Charlier, P., Huynh-Charlier, I., Munoz, O., Billard, M., Brun, L., and Grandmaison, G.
 L.
-de la.
-(2010).
+de la (2010).
 The microscopic (optical and {SEM}) examination of dental calculus deposits ({DCD}).
 {Potential} interest in forensic anthropology of a bio-archaeological method.
-\emph{Legal Medicine}, \emph{12}(4), 163--171.
-\url{https://doi.org/10.1016/j.legalmed.2010.03.003}
-
-\leavevmode\hypertarget{ref-cummingsMayanCalculus1997}{}%
+\emph{Legal Medicine} 12, 163--171.
+doi:\href{https://doi.org/10.1016/j.legalmed.2010.03.003}{10.1016/j.legalmed.2010.03.003}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cummingsMayanCalculus1997}{}}%
 Cummings, L.
-S., \& Magennis, A.
+S., and Magennis, A.
 (1997).
-A phytolith and starch record of food and grit in {Mayan} human tooth tartar.
-In A.
+{``A phytolith and starch record of food and grit in {Mayan} human tooth tartar,''} in \emph{The {State-of-the-Art} of {Phytoliths} in {Soils} and {Plants}}, eds.
+A.
 Pinilla, J.
-Juan-Tresserras, \& M.
+Juan-Tresserras, and M.
 J.
-Machado (Eds.), \emph{The {State}-of-the-{Art} of {Phytoliths} in {Soils} and {Plants}}.
-{CSIC Press}.
-
-\leavevmode\hypertarget{ref-doddsCarbohydrateRetention1988}{}%
+Machado ({Spain}: {CSIC Press}).
+
+\leavevmode\vadjust pre{\hypertarget{ref-doddsCarbohydrateRetention1988}{}}%
 Dodds, M.
 W.
-J., \& Edgar, W.
+J., and Edgar, W.
 M.
@@
 The {Relationship Between Plaque pH}, {Plaque Acid Anion Profiles}, and {Oral Carbohydrate Retention After Ingestion} of {Several} '{Reference Foods}' by {Human Subjects}.
-\emph{Journal of Dental Research}, \emph{67}(5), 861--865.
-\url{https://doi.org/10.1177/00220345880670051301}
-
-\leavevmode\hypertarget{ref-dudgeonDietGeographyDrinking2014}{}%
+\emph{Journal of Dental Research} 67, 861--865.
+doi:\href{https://doi.org/10.1177/00220345880670051301}{10.1177/00220345880670051301}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dudgeonDietGeographyDrinking2014}{}}%
 Dudgeon, J.
-V., \& Tromp, M.
+V., and Tromp, M.
 (2014).
-Diet, {Geography} and {Drinking Water} in {Polynesia}: Microfossil {Research} from {Archaeological Human Dental Calculus}, {Rapa Nui} ({Easter Island}).
-\emph{International Journal of Osteoarchaeology}, \emph{24}(5), 634--648.
-\url{https://doi.org/10.1002/oa.2249}
-
-\leavevmode\hypertarget{ref-eerkensDentalCalculusSource2018}{}%
+Diet, {Geography} and {Drinking Water} in {Polynesia}: {Microfossil Research} from {Archaeological Human Dental Calculus}, {Rapa Nui} ({Easter Island}).
+\emph{International Journal of Osteoarchaeology} 24, 634--648.
+doi:\href{https://doi.org/10.1002/oa.2249}{10.1002/oa.2249}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-eerkensDentalCalculusSource2018}{}}%
 Eerkens, J.
@@
 J., Garibay, R., Perez, K., Murga, E., Kaijankoski, P., Rosenthal, J.
-S., \& Gang, D.
+S., and Gang, D.
 R.
 (2018).
-Dental calculus as a source of ancient alkaloids: Detection of nicotine by {LC}-{MS} in calculus samples from the {Americas}.
-\emph{Journal of Archaeological Science: Reports}, \emph{18}, 509--515.
-\url{https://doi.org/10.1016/j.jasrep.2018.02.004}
-
-\leavevmode\hypertarget{ref-fagernasMicrobialBiogeography2021}{}%
+Dental calculus as a source of ancient alkaloids: {Detection} of nicotine by {LC-MS} in calculus samples from the {Americas}.
+\emph{Journal of Archaeological Science: Reports} 18, 509--515.
+doi:\href{https://doi.org/10.1016/j.jasrep.2018.02.004}{10.1016/j.jasrep.2018.02.004}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-fagernasMicrobialBiogeography2021}{}}%
 Fagernäs, Z., Salazar-García, D.
@@
 L., Ozga, A., Velsko, I.
-M., \& Warinner, C.
+M., and Warinner, C.
 (2021).
@@
 \emph{bioRxiv}, 2021.08.16.456492.
-\url{https://doi.org/10.1101/2021.08.16.456492}
-
-\leavevmode\hypertarget{ref-foxPhytolithCalculus1996}{}%
+doi:\href{https://doi.org/10.1101/2021.08.16.456492}{10.1101/2021.08.16.456492}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-yatesOralMicrobiome2021}{}}%
+Fellows Yates, J.
+A., Velsko, I.
+M., Aron, F., Posth, C., Hofman, C.
+A., Austin, R.
+M., Parker, C.
+E., Mann, A.
+E., Nägele, K., Arthur, K.
+W., et al.
+(2021).
+The evolution and changing ecology of the {African} hominid oral microbiome.
+\emph{Proceedings of the National Academy of Sciences} 118.
+doi:\href{https://doi.org/10.1073/pnas.2021655118}{10.1073/pnas.2021655118}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-foxPhytolithCalculus1996}{}}%
 Fox, C.
-L., Juan, J., \& Albert, R.
+L., Juan, J., and Albert, R.
 M.
 (1996).
-Phytolith analysis on dental calculus, enamel surface, and burial soil: Information about diet and paleoenvironment.
-\emph{American Journal of Physical Anthropology}, \emph{101}(1), 101--113.
-\url{https://doi.org/10.1002/(SICI)1096-8644(199609)101:1\%3C101::AID-AJPA7\%3E3.0.CO;2-Y}
-
-\leavevmode\hypertarget{ref-francoStarchDegradation1992}{}%
+Phytolith analysis on dental calculus, enamel surface, and burial soil: {Information} about diet and paleoenvironment.
+\emph{American Journal of Physical Anthropology} 101, 101--113.
+doi:\href{https://doi.org/10.1002/(SICI)1096-8644(199609)101:1\%3C101::AID-AJPA7\%3E3.0.CO;2-Y}{10.1002/(SICI)1096-8644(199609)101:1\textless101::AID-AJPA7\textgreater3.0.CO;2-Y}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-francoStarchDegradation1992}{}}%
 Franco, C.
@@
 J.
-do R., \& Ciacco, C.
+do R., and Ciacco, C.
 F.
@@
 Factors that {Affect} the {Enzymatic Degradation} of {Natural Starch Granules} -{Effect} of the {Size} of the {Granules}.
-\emph{Starch - Stärke}, \emph{44}(11), 422--426.
-\url{https://doi.org/10.1002/star.19920441106}
-
-\leavevmode\hypertarget{ref-froehlichEffectOral1987}{}%
+\emph{Starch - Stärke} 44, 422--426.
+doi:\href{https://doi.org/10.1002/star.19920441106}{10.1002/star.19920441106}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-froehlichEffectOral1987}{}}%
 Froehlich, D.
 A., Pangborn, R.
-M., \& Whitaker, J.
+M., and Whitaker, J.
 R.
@@
 The effect of oral stimulation on human parotid salivary flow rate and alpha-amylase secretion.
-\emph{Physiology \& Behavior}, \emph{41}(3), 209--217.
-\url{https://doi.org/10.1016/0031-9384(87)90355-6}
-
-\leavevmode\hypertarget{ref-graneroStarchTaphonomy2020}{}%
+\emph{Physiology \& Behavior} 41, 209--217.
+doi:\href{https://doi.org/10.1016/0031-9384(87)90355-6}{10.1016/0031-9384(87)90355-6}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-graneroStarchTaphonomy2020}{}}%
 García-Granero, J.
@@
 (2020).
-Starch taphonomy, equifinality and the importance of context: Some notes on the identification of food processing through starch grain analysis.
-\emph{Journal of Archaeological Science}, \emph{124}, 105267.
-\url{https://doi.org/10.1016/j.jas.2020.105267}
-
-\leavevmode\hypertarget{ref-gismondiStarchGranulesData2019}{}%
-Gismondi, A., D'Agostino, A., Canuti, L., Di Marco, G., Basoli, F., \& Canini, A.
+Starch taphonomy, equifinality and the importance of context: {Some} notes on the identification of food processing through starch grain analysis.
+\emph{Journal of Archaeological Science} 124, 105267.
+doi:\href{https://doi.org/10.1016/j.jas.2020.105267}{10.1016/j.jas.2020.105267}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gismondiStarchGranulesData2019}{}}%
+Gismondi, A., D'Agostino, A., Canuti, L., Di Marco, G., Basoli, F., and Canini, A.
 (2019).
 Starch granules: A data collection of 40 food species.
-\emph{Plant Biosystems - An International Journal Dealing with All Aspects of Plant Biology}, \emph{153}(2), 273--279.
-\url{https://doi.org/10.1080/11263504.2018.1473523}
-
-\leavevmode\hypertarget{ref-haffajeeBiofilmPosition2009}{}%
+\emph{Plant Biosystems - An International Journal Dealing with all Aspects of Plant Biology} 153, 273--279.
+doi:\href{https://doi.org/10.1080/11263504.2018.1473523}{10.1080/11263504.2018.1473523}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-haffajeeBiofilmPosition2009}{}}%
 Haffajee, A.
@@
 P., Patel, M.
-R., Song, X., Yaskell, T., \& Socransky, S.
+R., Song, X., Yaskell, T., and Socransky, S.
 S.
@@
 {Tooth} position.
-\emph{Journal of Periodontal Research}, \emph{44}(4), 520--528.
-\url{https://doi.org/10.1111/j.1600-0765.2008.01155.x}
-
-\leavevmode\hypertarget{ref-hardyStarchGranulesDental2009}{}%
-Hardy, K., Blakeney, T., Copeland, L., Kirkham, J., Wrangham, R., \& Collins, M.
+\emph{Journal of Periodontal Research} 44, 520--528.
+doi:\href{https://doi.org/10.1111/j.1600-0765.2008.01155.x}{10.1111/j.1600-0765.2008.01155.x}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hardyStarchGranulesDental2009}{}}%
+Hardy, K., Blakeney, T., Copeland, L., Kirkham, J., Wrangham, R., and Collins, M.
 (2009).
 Starch granules, dental calculus and new perspectives on ancient diet.
-\emph{Journal of Archaeological Science}, \emph{36}(2), 248--255.
-\url{https://doi.org/10.1016/j.jas.2008.09.015}
-
-\leavevmode\hypertarget{ref-hardyNeanderthalMedics2012}{}%
+\emph{Journal of Archaeological Science} 36, 248--255.
+doi:\href{https://doi.org/10.1016/j.jas.2008.09.015}{10.1016/j.jas.2008.09.015}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hardyNeanderthalMedics2012}{}}%
 Hardy, K., Buckley, S., Collins, M.
-J., Estalrrich, A., Brothwell, D., Copeland, L., García-Tabernero, A., García-Vargas, S., de la Rasilla, M., Lalueza-Fox, C., Huguet, R., Bastir, M., Santamaría, D., Madella, M., Wilson, J., Cortés, Á.
-F., \& Rosas, A.
+J., Estalrrich, A., Brothwell, D., Copeland, L., García-Tabernero, A., García-Vargas, S., de la Rasilla, M., Lalueza-Fox, C., et al.
 (2012).
 Neanderthal medics?
-Evidence for food, cooking, and medicinal plants entrapped in dental calculus.
-\emph{Naturwissenschaften}, \emph{99}(8), 617--626.
-\url{https://doi.org/10.1007/s00114-012-0942-0}
-
-\leavevmode\hypertarget{ref-haslamDecompositionStarch2004}{}%
+{Evidence} for food, cooking, and medicinal plants entrapped in dental calculus.
+\emph{Naturwissenschaften} 99, 617--626.
+doi:\href{https://doi.org/10.1007/s00114-012-0942-0}{10.1007/s00114-012-0942-0}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-haslamDecompositionStarch2004}{}}%
 Haslam, M.
@@
 The decomposition of starch grains in soils: Implications for archaeological residue analyses.
-\emph{Journal of Archaeological Science}, \emph{31}(12), 1715--1734.
-\url{https://doi.org/10.1016/j.jas.2004.05.006}
-
-\leavevmode\hypertarget{ref-hendyProteomicCalculus2018}{}%
+\emph{Journal of Archaeological Science} 31, 1715--1734.
+doi:\href{https://doi.org/10.1016/j.jas.2004.05.006}{10.1016/j.jas.2004.05.006}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hendyProteomicCalculus2018}{}}%
 Hendy, J., Warinner, C., Bouwman, A., Collins, M.
 J., Fiddyment, S., Fischer, R., Hagan, R., Hofman, C.
-A., Holst, M., Chaves, E., Klaus, L., Larson, G., Mackie, M., McGrath, K., Mundorff, A.
-Z., Radini, A., Rao, H., Trachsel, C., Velsko, I.
-M., \& Speller, C.
-F.
+A., Holst, M., Chaves, E., et al.
 (2018).
@@
 \emph{Proceedings.
-Biological Sciences}, \emph{285}(1883), 20180977.
-\url{https://doi.org/10.1098/rspb.2018.0977}
-
-\leavevmode\hypertarget{ref-henryNeanderthalCalculus2014}{}%
+Biological Sciences} 285, 20180977.
+doi:\href{https://doi.org/10.1098/rspb.2018.0977}{10.1098/rspb.2018.0977}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-henryNeanderthalCalculus2014}{}}%
 Henry, A.
 G., Brooks, A.
-S., \& Piperno, D.
+S., and Piperno, D.
 R.
@@
 Plant foods and the dietary ecology of {Neanderthals} and early modern humans.
-\emph{Journal of Human Evolution}, \emph{69}, 44--54.
-\url{https://doi.org/10.1016/j.jhevol.2013.12.014}
-
-\leavevmode\hypertarget{ref-henryCookingStarch2009}{}%
+\emph{Journal of Human Evolution} 69, 44--54.
+doi:\href{https://doi.org/10.1016/j.jhevol.2013.12.014}{10.1016/j.jhevol.2013.12.014}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-henryCookingStarch2009}{}}%
 Henry, A.
 G., Hudson, H.
-F., \& Piperno, D.
+F., and Piperno, D.
 R.
@@
 Changes in starch grain morphologies from cooking.
-\emph{Journal of Archaeological Science}, \emph{36}(3), 915--922.
-\url{https://doi.org/10.1016/j.jas.2008.11.008}
-
-\leavevmode\hypertarget{ref-henryCalculusSyria2008}{}%
+\emph{Journal of Archaeological Science} 36, 915--922.
+doi:\href{https://doi.org/10.1016/j.jas.2008.11.008}{10.1016/j.jas.2008.11.008}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-henryCalculusSyria2008}{}}%
 Henry, A.
-G., \& Piperno, D.
+G., and Piperno, D.
 R.
@@
 Using plant microfossils from dental calculus to recover human diet: A case study from {Tell} al-{Raqā}'i, {Syria}.
-\emph{Journal of Archaeological Science}, \emph{35}(7), 1943--1950.
-\url{https://doi.org/10.1016/j.jas.2007.12.005}
-
-\leavevmode\hypertarget{ref-janeAnthologyStarch1994}{}%
-Jane, J.-L., Kasemsuwan, T., Leas, S., Zobel, H., \& Robyt, J.
+\emph{Journal of Archaeological Science} 35, 1943--1950.
+doi:\href{https://doi.org/10.1016/j.jas.2007.12.005}{10.1016/j.jas.2007.12.005}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-janeAnthologyStarch1994}{}}%
+Jane, J.-L., Kasemsuwan, T., Leas, S., Zobel, H., and Robyt, J.
 F.
@@
 Anthology of {Starch Granule Morphology} by {Scanning Electron Microscopy}.
-\emph{Starch - Stärke}, \emph{46}(4), 121--129.
-\url{https://doi.org/10.1002/star.19940460402}
-
-\leavevmode\hypertarget{ref-jepsenCalculusRemoval2011}{}%
-Jepsen, S., Deschner, J., Braun, A., Schwarz, F., \& Eberhard, J.
+\emph{Starch - Stärke} 46, 121--129.
+doi:\href{https://doi.org/10.1002/star.19940460402}{10.1002/star.19940460402}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-jepsenCalculusRemoval2011}{}}%
+Jepsen, S., Deschner, J., Braun, A., Schwarz, F., and Eberhard, J.
 (2011).
 Calculus removal and the prevention of its formation.
-\emph{Periodontology 2000}, \emph{55}(1), 167--188.
-\url{https://doi.org/10.1111/j.1600-0757.2010.00382.x}
-
-\leavevmode\hypertarget{ref-jovanovicNeolithicCalculus2021}{}%
+\emph{Periodontology 2000} 55, 167--188.
+doi:\href{https://doi.org/10.1111/j.1600-0757.2010.00382.x}{10.1111/j.1600-0757.2010.00382.x}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-jovanovicNeolithicCalculus2021}{}}%
 Jovanović, J., Power, R.
-C., de Becdelièvre, C., Goude, G., \& Stefanović, S.
+C., de Becdelièvre, C., Goude, G., and Stefanović, S.
 (2021).
-Microbotanical evidence for the spread of cereal use during the {Mesolithic}-{Neolithic} transition in the {Southeastern Europe} ({Danube Gorges}): Data from dental calculus analysis.
-\emph{Journal of Archaeological Science}, \emph{125}, 105288.
-\url{https://doi.org/10.1016/j.jas.2020.105288}
-
-\leavevmode\hypertarget{ref-kashketFoodRetention1991}{}%
+Microbotanical evidence for the spread of cereal use during the {Mesolithic-Neolithic} transition in the {Southeastern Europe} ({Danube Gorges}): {Data} from dental calculus analysis.
+\emph{Journal of Archaeological Science} 125, 105288.
+doi:\href{https://doi.org/10.1016/j.jas.2020.105288}{10.1016/j.jas.2020.105288}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kashketFoodRetention1991}{}}%
 Kashket, S., Van Houte, J., Lopez, L.
-R., \& Stocks, S.
+R., and Stocks, S.
 (1991).
 Lack of {Correlation Between Food Retention} on the {Human Dentition} and {Consumer Perception} of {Food Stickiness}.
-\emph{Journal of Dental Research}, \emph{70}(10), 1314--1319.
-\url{https://doi.org/10.1177/00220345910700100101}
-
-\leavevmode\hypertarget{ref-kashketFoodParticles1996}{}%
-Kashket, S., Zhang, J., \& Houte, J.
+\emph{Journal of Dental Research} 70, 1314--1319.
+doi:\href{https://doi.org/10.1177/00220345910700100101}{10.1177/00220345910700100101}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kashketFoodParticles1996}{}}%
+Kashket, S., Zhang, J., and Houte, J.
 V.
@@
 
-\leavevmode\hypertarget{ref-lemoyneCalculusPretreatments2021}{}%
-Le Moyne, C., \& Crowther, A.
+\leavevmode\vadjust pre{\hypertarget{ref-lemoyneCalculusPretreatments2021}{}}%
+Le Moyne, C., and Crowther, A.
 (2021).
-Effects of chemical pre-treatments on modified starch granules: Recommendations for dental calculus decalcification for ancient starch research.
-\emph{Journal of Archaeological Science: Reports}, \emph{35}, 102762.
-\url{https://doi.org/10.1016/j.jasrep.2020.102762}
-
-\leavevmode\hypertarget{ref-leonardDentalCalculus2015}{}%
+Effects of chemical pre-treatments on modified starch granules: {Recommendations} for dental calculus decalcification for ancient starch research.
+\emph{Journal of Archaeological Science: Reports} 35, 102762.
+doi:\href{https://doi.org/10.1016/j.jasrep.2020.102762}{10.1016/j.jasrep.2020.102762}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-leonardDentalCalculus2015}{}}%
 Leonard, C., Vashro, L., O'Connell, J.
-F., \& Henry, A.
+F., and Henry, A.
 G.
 (2015).
-Plant microremains in dental calculus as a record of plant consumption: A test with {Twe} forager-horticulturalists.
-\emph{Journal of Archaeological Science: Reports}, \emph{2}, 449--457.
-\url{https://doi.org/10.1016/j.jasrep.2015.03.009}
-
-\leavevmode\hypertarget{ref-lingstromStarchyFood1994}{}%
-Lingstrom, P., Birkhed, D., Ruben, J., \& Arends, J.
+Plant microremains in dental calculus as a record of plant consumption: {A} test with {Twe} forager-horticulturalists.
+\emph{Journal of Archaeological Science: Reports} 2, 449--457.
+doi:\href{https://doi.org/10.1016/j.jasrep.2015.03.009}{10.1016/j.jasrep.2015.03.009}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lingstromStarchyFood1994}{}}%
+Lingstrom, P., Birkhed, D., Ruben, J., and Arends, J.
 (1994).
 Effect of {Frequent Consumption} of {Starchy Food Items} on {Enamel} and {Dentin Demineralization} and on {Plaque pH} in situ.
-\emph{Journal of Dental Research}, \emph{73}(3), 652--660.
-\url{https://doi.org/10.1177/00220345940730031101}
-
-\leavevmode\hypertarget{ref-R-rrtools}{}%
-Marwick, B. (2019). \emph{Rrtools: Creates a reproducible research compendium} {[}Manual{]}.
-
-\leavevmode\hypertarget{ref-mickleburghNewInsightsConsumption2012}{}%
+\emph{Journal of Dental Research} 73, 652--660.
+doi:\href{https://doi.org/10.1177/00220345940730031101}{10.1177/00220345940730031101}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-R-rrtools}{}}%
+Marwick, B. (2019). \emph{Rrtools: {Creates} a reproducible research compendium}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mickleburghNewInsightsConsumption2012}{}}%
 Mickleburgh, H.
-L., \& Pagán-Jiménez, J.
+L., and Pagán-Jiménez, J.
 R.
@@
 New insights into the consumption of maize and other food plants in the pre-{Columbian Caribbean} from starch grains trapped in human dental calculus.
-\emph{Journal of Archaeological Science}, \emph{39}(7), 2468--2478.
-\url{https://doi.org/10.1016/j.jas.2012.02.020}
-
-\leavevmode\hypertarget{ref-middletonExtractionOpalPhytoliths1994}{}%
+\emph{Journal of Archaeological Science} 39, 2468--2478.
+doi:\href{https://doi.org/10.1016/j.jas.2012.02.020}{10.1016/j.jas.2012.02.020}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-middletonExtractionOpalPhytoliths1994}{}}%
 Middleton, W.
-D., \& Rovner, I.
+D., and Rovner, I.
 (1994).
 Extraction of {Opal Phytoliths} from {Herbivore Dental Calculus}.
-\emph{Journal of Archaeological Science}, \emph{21}(4), 469--473.
-\url{https://doi.org/10.1006/jasc.1994.1046}
-
-\leavevmode\hypertarget{ref-modiCalculusMethodologies2020}{}%
-Modi, A., Pisaneschi, L., Zaro, V., Vai, S., Vergata, C., Casalone, E., Caramelli, D., Moggi-Cecchi, J., Mariotti Lippi, M., \& Lari, M.
+\emph{Journal of Archaeological Science} 21, 469--473.
+doi:\href{https://doi.org/10.1006/jasc.1994.1046}{10.1006/jasc.1994.1046}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-modiCalculusMethodologies2020}{}}%
+Modi, A., Pisaneschi, L., Zaro, V., Vai, S., Vergata, C., Casalone, E., Caramelli, D., Moggi-Cecchi, J., Mariotti Lippi, M., and Lari, M.
 (2020).
 Combined methodologies for gaining much information from ancient dental calculus: Testing experimental strategies for simultaneously analysing {DNA} and food residues.
-\emph{Archaeological and Anthropological Sciences}, \emph{12}(1), 10.
-\url{https://doi.org/10.1007/s12520-019-00983-5}
-
-\leavevmode\hypertarget{ref-R-here}{}%
-Müller, K. (2020). \emph{Here: A simpler way to find your files} {[}Manual{]}.
-
-\leavevmode\hypertarget{ref-naterHumanAmylase2005}{}%
+\emph{Archaeological and Anthropological Sciences} 12, 10.
+doi:\href{https://doi.org/10.1007/s12520-019-00983-5}{10.1007/s12520-019-00983-5}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-R-here}{}}%
+Müller, K. (2020). \emph{Here: {A} simpler way to find your files}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-naterHumanAmylase2005}{}}%
 Nater, U.
-M., Rohleder, N., Gaab, J., Berger, S., Jud, A., Kirschbaum, C., \& Ehlert, U.
+M., Rohleder, N., Gaab, J., Berger, S., Jud, A., Kirschbaum, C., and Ehlert, U.
 (2005).
 Human salivary alpha-amylase reactivity in a psychosocial stress paradigm.
-\emph{International Journal of Psychophysiology}, \emph{55}(3), 333--342.
-\url{https://doi.org/10.1016/j.ijpsycho.2004.09.009}
-
-\leavevmode\hypertarget{ref-nikitkovaStarchBiofilms2013}{}%
+\emph{International Journal of Psychophysiology} 55, 333--342.
+doi:\href{https://doi.org/10.1016/j.ijpsycho.2004.09.009}{10.1016/j.ijpsycho.2004.09.009}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nikitkovaStarchBiofilms2013}{}}%
 Nikitkova, A.
 E., Haase, E.
-M., \& Scannapieco, F.
+M., and Scannapieco, F.
 A.
 (2013).
-Taking the {Starch} out of {Oral Biofilm Formation}: Molecular {Basis} and {Functional Significance} of {Salivary} {\(\alpha\)}-{Amylase Binding} to {Oral Streptococci}.
-\emph{Applied and Environmental Microbiology}, \emph{79}(2), 416--423.
-\url{https://doi.org/10.1128/AEM.02581-12}
-
-\leavevmode\hypertarget{ref-pearceConcomitantDepositionStrontium1987}{}%
+Taking the {Starch} out of {Oral Biofilm Formation}: {Molecular Basis} and {Functional Significance} of {Salivary} {\(\alpha\)}-{Amylase Binding} to {Oral Streptococci}.
+\emph{Applied and Environmental Microbiology} 79, 416--423.
+doi:\href{https://doi.org/10.1128/AEM.02581-12}{10.1128/AEM.02581-12}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pearceConcomitantDepositionStrontium1987}{}}%
 Pearce, E.
 I.
-F., \& Sissons, C.
+F., and Sissons, C.
 H.
@@
 The {Concomitant Deposition} of {Strontium} and {Fluoride} in {Dental Plaque}.
-\emph{Journal of Dental Research}, \emph{66}(10), 1518--1522.
-\url{https://doi.org/10.1177/00220345870660100101}
-
-\leavevmode\hypertarget{ref-R-patchwork}{}%
-Pedersen, T. L. (2020). \emph{Patchwork: The composer of plots} {[}Manual{]}.
-
-\leavevmode\hypertarget{ref-pipernoStarchGrains2008}{}%
+\emph{Journal of Dental Research} 66, 1518--1522.
+doi:\href{https://doi.org/10.1177/00220345870660100101}{10.1177/00220345870660100101}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-R-patchwork}{}}%
+Pedersen, T. L. (2020). \emph{Patchwork: {The} composer of plots}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pipernoStarchGrains2008}{}}%
 Piperno, D.
-R., \& Dillehay, T.
+R., and Dillehay, T.
 D.
@@
 Starch grains on human teeth reveal early broad crop diet in northern {Peru}.
-\emph{Proceedings of the National Academy of Sciences}, \emph{105}(50), 19622--19627.
-\url{https://doi.org/10.1073/pnas.0808752105}
-
-\leavevmode\hypertarget{ref-powerChimpCalculus2015}{}%
+\emph{Proceedings of the National Academy of Sciences} 105, 19622--19627.
+doi:\href{https://doi.org/10.1073/pnas.0808752105}{10.1073/pnas.0808752105}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-powerChimpCalculus2015}{}}%
 Power, R.
@@
 C., Wittig, R.
-M., Freiberg, M., \& Henry, A.
+M., Freiberg, M., and Henry, A.
 G.
@@
 Dental calculus evidence of {Tai Forest Chimpanzee} plant consumption and life history transitions.
-\emph{Scientific Reports}, \emph{5}, 15161.
-\url{https://doi.org/10.1038/srep15161}
-
-\leavevmode\hypertarget{ref-powerSEMCalculus2014}{}%
+\emph{Scientific Reports} 5, 15161.
+doi:\href{https://doi.org/10.1038/srep15161}{10.1038/srep15161}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-powerSEMCalculus2014}{}}%
 Power, R.
@@
 C., Wittig, R.
-M., \& Henry, A.
+M., and Henry, A.
 G.
@@
 Assessing use and suitability of scanning electron microscopy in the analysis of micro remains in dental calculus.
-\emph{Journal of Archaeological Science}, \emph{49}, 160--169.
-\url{https://doi.org/10.1016/j.jas.2014.04.016}
-
-\leavevmode\hypertarget{ref-proctorSpatialGradient2018}{}%
+\emph{Journal of Archaeological Science} 49, 160--169.
+doi:\href{https://doi.org/10.1016/j.jas.2014.04.016}{10.1016/j.jas.2014.04.016}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-proctorSpatialGradient2018}{}}%
 Proctor, D.
@@
 I., Holmes, S.
-P., \& Relman, D.
+P., and Relman, D.
 A.
@@
 A spatial gradient of bacterial diversity in the human oral cavity shaped by salivary flow.
-\emph{Nature Communications}, \emph{9}(1), 681.
-\url{https://doi.org/10.1038/s41467-018-02900-1}
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2020). \emph{R: A language and environment for statistical computing} {[}Manual{]}. {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing}.
-
-\leavevmode\hypertarget{ref-radiniFoodMultiplePathways2017}{}%
-Radini, A., Nikita, E., Buckley, S., Copeland, L., \& Hardy, K.
+\emph{Nature Communications} 9, 681.
+doi:\href{https://doi.org/10.1038/s41467-018-02900-1}{10.1038/s41467-018-02900-1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-R-base}{}}%
+R Core Team (2020). \emph{R: {A} language and environment for statistical computing}. {Vienna, Austria}: {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-radiniFoodMultiplePathways2017}{}}%
+Radini, A., Nikita, E., Buckley, S., Copeland, L., and Hardy, K.
 (2017).
-Beyond food: The multiple pathways for inclusion of materials into ancient dental calculus.
-\emph{American Journal of Physical Anthropology}, \emph{162}, 71--83.
-\url{https://doi.org/10.1002/ajpa.23147}
-
-\leavevmode\hypertarget{ref-radiniMedievalWomenEarly2019}{}%
+Beyond food: {The} multiple pathways for inclusion of materials into ancient dental calculus.
+\emph{American Journal of Physical Anthropology} 162, 71--83.
+doi:\href{https://doi.org/10.1002/ajpa.23147}{10.1002/ajpa.23147}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-radiniMedievalWomenEarly2019}{}}%
 Radini, A., Tromp, M., Beach, A., Tong, E., Speller, C., McCormick, M., Dudgeon, J.
 V., Collins, M.
-J., Rühli, F., Kröger, R., \& Warinner, C.
+J., Rühli, F., Kröger, R., et al.
 (2019).
 Medieval women's early involvement in manuscript production suggested by lapis lazuli identification in dental calculus.
-\emph{Science Advances}, \emph{5}(1), eaau7126.
-\url{https://doi.org/10.1126/sciadv.aau7126}
-
-\leavevmode\hypertarget{ref-reichertStarchBible1913b}{}%
+\emph{Science Advances} 5, eaau7126.
+doi:\href{https://doi.org/10.1126/sciadv.aau7126}{10.1126/sciadv.aau7126}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-reichertStarchBible1913b}{}}%
 Reichert, E.
@@
 (1913).
-\emph{The differentiation and specificity of starches in relation to genera, species, etc: Stereochemistry applied to protoplasmic processes and products, and as a strictly scientific basis for the classification of plants and animals} (Vol.
-2).
-{Carnegie institution of Washington}.
-
-\leavevmode\hypertarget{ref-R-broom}{}%
-Robinson, D., Hayes, A., \& Couch, S. (2021). \emph{Broom: Convert statistical objects into tidy tibbles} {[}Manual{]}.
-
-\leavevmode\hypertarget{ref-scannapiecoSalivaryAmylase1993}{}%
-Scannapieco, F.
-A., Torres, G., \& Levine, M.
-J.
-(1993).
-Salivary {\(\alpha\)}-amylase: Role in dental plaque and caries formation.
-\emph{Critical Reviews in Oral Biology \& Medicine}, \emph{4}(3), 301--307.
-
-\leavevmode\hypertarget{ref-seidemannStarchAtlas1966}{}%
-Seidemann, J. (1966). \emph{St\{\textbackslash"a\}rke-{Atlas}: Grundlagen der {St}\{\textbackslash"a\}rke-{Mikroskopie} und {Beschreibung} der wichtigsten {St}\{\textbackslash"a\}rkearten}. {Parey}.
-
-\leavevmode\hypertarget{ref-shellisSyntheticSalivaCultural1978}{}%
-Shellis, R. P. (1978). A synthetic saliva for cultural studies of dental plaque. \emph{Archives of Oral Biology}, \emph{23}(6), 485--489. \url{https://doi.org/10.1016/0003-9969(78)90081-X}
-
-\leavevmode\hypertarget{ref-simon-soroOralGeography2013}{}%
+\emph{The differentiation and specificity of starches in relation to genera, species, etc: Stereochemistry applied to protoplasmic processes and products, and as a strictly scientific basis for the classification of plants and animals}.
+{Washington, D.C.}: {Carnegie institution of Washington}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-R-broom}{}}%
+Robinson, D., Hayes, A., and Couch, S. (2021). \emph{Broom: {Convert} statistical objects into tidy tibbles}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-scannapiecoSalivaryAmylase1993}{}}%
+Scannapieco, F. A., Torres, G., and Levine, M. J. (1993). Salivary {\(\alpha\)}-amylase: Role in dental plaque and caries formation. \emph{Critical Reviews in Oral Biology \& Medicine} 4, 301--307.
+
+\leavevmode\vadjust pre{\hypertarget{ref-seidemannStarchAtlas1966}{}}%
+Seidemann, J.
+(1966).
+\emph{St\{\textbackslash"a\}rke-{Atlas}: {Grundlagen} der {St}\{\textbackslash"a\}rke-{Mikroskopie} und {Beschreibung} der wichtigsten {St}\{\textbackslash"a\}rkearten}.
+{Berlin}: {Parey}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-shellisSyntheticSalivaCultural1978}{}}%
+Shellis, R.
+P.
+(1978).
+A synthetic saliva for cultural studies of dental plaque.
+\emph{Archives of Oral Biology} 23, 485--489.
+doi:\href{https://doi.org/10.1016/0003-9969(78)90081-X}{10.1016/0003-9969(78)90081-X}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-simon-soroOralGeography2013}{}}%
 Simón-Soro, A., Tomás, I., Cabrera-Rubio, R., Catalan, M.
-D., Nyvad, B., \& Mira, A.
+D., Nyvad, B., and Mira, A.
 (2013).
 Microbial geography of the oral cavity.
-\emph{Journal of Dental Research}, \emph{92}(7), 616--621.
-\url{https://doi.org/10.1177/0022034513488119}
-
-\leavevmode\hypertarget{ref-sissonsMultistationDentalPlaque1991}{}%
+\emph{Journal of Dental Research} 92, 616--621.
+doi:\href{https://doi.org/10.1177/0022034513488119}{10.1177/0022034513488119}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sissonsMultistationDentalPlaque1991}{}}%
 Sissons, C.
@@
 W., Hoffman, M.
-P., \& Wakefield, J.
+P., and Wakefield, J.
 S.
@@
 (1991).
-A {Multi}-station {Dental Plaque Microcosm} ({Artificial Mouth}) for the {Study} of {Plaque Growth}, {Metabolism}, {pH}, and {Mineralization}: \emph{Journal of Dental Research}.
-\url{https://doi.org/10.1177/00220345910700110301}
-
-\leavevmode\hypertarget{ref-tanCalculusUltrastructure2004}{}%
+A {Multi-station Dental Plaque Microcosm} ({Artificial Mouth}) for the {Study} of {Plaque Growth}, {Metabolism}, {pH}, and {Mineralization}: \emph{Journal of Dental Research}.
+doi:\href{https://doi.org/10.1177/00220345910700110301}{10.1177/00220345910700110301}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tanCalculusUltrastructure2004}{}}%
 Tan, B.
@@
 G., Mordan, N.
-J., \& Galgut, P.
+J., and Galgut, P.
 N.
-(2004).
+(2004a).
 A preliminary investigation into the ultrastructure of dental calculus and associated bacteria.
-\emph{Journal of Clinical Periodontology}, \emph{31}(5), 364--369.
-\url{https://doi.org/10.1111/j.1600-051X.2004.00484.x}
-
-\leavevmode\hypertarget{ref-tanBacterialViability2004}{}%
+\emph{Journal of Clinical Periodontology} 31, 364--369.
+doi:\href{https://doi.org/10.1111/j.1600-051X.2004.00484.x}{10.1111/j.1600-051X.2004.00484.x}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tanBacterialViability2004}{}}%
 Tan, B.
@@
 K., Mordan, N.
-J., Embleton, J., Pratten, J., \& Galgut, P.
+J., Embleton, J., Pratten, J., and Galgut, P.
 N.
-(2004).
+(2004b).
 Study of {Bacterial Viability} within {Human Supragingival Dental Calculus}.
-\emph{Journal of Periodontology}, \emph{75}(1), 23--29.
-\url{https://doi.org/10.1902/jop.2004.75.1.23}
-
-\leavevmode\hypertarget{ref-taoWheatCalculus2020}{}%
-Tao, D., Zhang, G., Zhou, Y., \& Zhao, H.
+\emph{Journal of Periodontology} 75, 23--29.
+doi:\href{https://doi.org/10.1902/jop.2004.75.1.23}{10.1902/jop.2004.75.1.23}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-taoWheatCalculus2020}{}}%
+Tao, D., Zhang, G., Zhou, Y., and Zhao, H.
 (2020).
-Investigating wheat consumption based on multiple evidences: Stable isotope analysis on human bone and starch grain analysis on dental calculus of humans from the {Laodaojing} cemetery, {Central Plains}, {China}.
-\emph{International Journal of Osteoarchaeology}, \emph{30}(5), 594--606.
-\url{https://doi.org/10.1002/oa.2884}
-
-\leavevmode\hypertarget{ref-toppingResistantStarch2003}{}%
+Investigating wheat consumption based on multiple evidences: {Stable} isotope analysis on human bone and starch grain analysis on dental calculus of humans from the {Laodaojing} cemetery, {Central Plains}, {China}.
+\emph{International Journal of Osteoarchaeology} 30, 594--606.
+doi:\href{https://doi.org/10.1002/oa.2884}{10.1002/oa.2884}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-toppingResistantStarch2003}{}}%
 Topping, D.
-L., Fukushima, M., \& Bird, A.
+L., Fukushima, M., and Bird, A.
 R.
@@
 Resistant starch as a prebiotic and synbiotic: State of the art.
-\emph{Proceedings of the Nutrition Society}, \emph{62}(1), 171--176.
-\url{https://doi.org/10.1079/PNS2002224}
-
-\leavevmode\hypertarget{ref-trompEDTACalculus2017}{}%
-Tromp, M., Buckley, H., Geber, J., \& Matisoo-Smith, E.
+\emph{Proceedings of the Nutrition Society} 62, 171--176.
+doi:\href{https://doi.org/10.1079/PNS2002224}{10.1079/PNS2002224}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-trompEDTACalculus2017}{}}%
+Tromp, M., Buckley, H., Geber, J., and Matisoo-Smith, E.
 (2017).
 {EDTA} decalcification of dental calculus as an alternate means of microparticle extraction from archaeological samples.
-\emph{Journal of Archaeological Science: Reports}, \emph{14}, 461--466.
-\url{https://doi.org/10.1016/j.jasrep.2017.06.035}
-
-\leavevmode\hypertarget{ref-trompDietaryNondietary2015}{}%
-Tromp, M., \& Dudgeon, J.
+\emph{Journal of Archaeological Science: Reports} 14, 461--466.
+doi:\href{https://doi.org/10.1016/j.jasrep.2017.06.035}{10.1016/j.jasrep.2017.06.035}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-trompDietaryNondietary2015}{}}%
+Tromp, M., and Dudgeon, J.
 V.
@@
 Differentiating dietary and non-dietary microfossils extracted from human dental calculus: The importance of sweet potato to ancient diet on {Rapa Nui}.
-\emph{Journal of Archaeological Science}, \emph{54}, 54--63.
-\url{https://doi.org/10.1016/j.jas.2014.11.024}
-
-\leavevmode\hypertarget{ref-vandeveldeStarchMorphology2002}{}%
-van de Velde, F., van Riel, J., \& Tromp, R.
+\emph{Journal of Archaeological Science} 54, 54--63.
+doi:\href{https://doi.org/10.1016/j.jas.2014.11.024}{10.1016/j.jas.2014.11.024}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-vandeveldeStarchMorphology2002}{}}%
+van de Velde, F., van Riel, J., and Tromp, R.
 H.
@@
 Visualisation of starch granule morphologies using confocal scanning laser microscopy ({CSLM}).
-\emph{Journal of the Science of Food and Agriculture}, \emph{82}(13), 1528--1536.
-\url{https://doi.org/10.1002/jsfa.1165}
-
-\leavevmode\hypertarget{ref-warinnerDirectEvidenceMilk2014}{}%
+\emph{Journal of the Science of Food and Agriculture} 82, 1528--1536.
+doi:\href{https://doi.org/10.1002/jsfa.1165}{10.1002/jsfa.1165}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-warinnerDirectEvidenceMilk2014}{}}%
 Warinner, C., Hendy, J., Speller, C., Cappellini, E., Fischer, R., Trachsel, C., Arneborg, J., Lynnerup, N., Craig, O.
 E., Swallow, D.
-M., Fotakis, A., Christensen, R.
-J., Olsen, J.
-V., Liebert, A., Montalva, N., Fiddyment, S., Charlton, S., Mackie, M., Canci, A., \ldots{} Collins, M.
-J.
-(2014).
+M., et al.
+(2014a).
 Direct evidence of milk consumption from ancient human dental calculus.
-\emph{Scientific Reports}, \emph{4}, 7104.
-\url{https://doi.org/10.1038/srep07104}
-
-\leavevmode\hypertarget{ref-warinnerPathogensHost2014}{}%
+\emph{Scientific Reports} 4, 7104.
+doi:\href{https://doi.org/10.1038/srep07104}{10.1038/srep07104}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-warinnerPathogensHost2014}{}}%
 Warinner, C., Rodrigues, J.
 F., Vyas, R., Trachsel, C., Shved, N., Grossmann, J., Radini, A., Hancock, Y., Tito, R.
-Y., Fiddyment, S., Speller, C., Hendy, J., Charlton, S., Luder, H.
-U., Salazar-Garcia, D.
-C., Eppler, E., Seiler, R., Hansen, L.
-H., Castruita, J.
-A., \ldots{} Cappellini, E.
-(2014).
+Y., Fiddyment, S., et al.
+(2014b).
 Pathogens and host immunity in the ancient human oral cavity.
-\emph{Nature Genetics}, \emph{46}(4), 336--344.
-\url{https://doi.org/10.1038/ng.2906}
-
-\leavevmode\hypertarget{ref-wesolowskiEvaluatingMicrofossilContent2010}{}%
+\emph{Nature Genetics} 46, 336--44.
+doi:\href{https://doi.org/10.1038/ng.2906}{10.1038/ng.2906}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wesolowskiEvaluatingMicrofossilContent2010}{}}%
 Wesolowski, V., Ferraz Mendonça de Souza, S.
 M., Reinhard, K.
-J., \& Ceccantini, G.
+J., and Ceccantini, G.
 (2010).
 Evaluating microfossil content of dental calculus from {Brazilian} sambaquis.
-\emph{Journal of Archaeological Science}, \emph{37}(6), 1326--1338.
-\url{https://doi.org/10.1016/j.jas.2009.12.037}
-
-\leavevmode\hypertarget{ref-tidyverse2019}{}%
+\emph{Journal of Archaeological Science} 37, 1326--1338.
+doi:\href{https://doi.org/10.1016/j.jas.2009.12.037}{10.1016/j.jas.2009.12.037}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tidyverse2019}{}}%
 Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L.
-D., François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., Kuhn, M., Pedersen, T.
-L., Miller, E., Bache, S.
-M., Müller, K., Ooms, J., Robinson, D., Seidel, D.
-P., Spinu, V., \ldots{} Yutani, H.
+D., François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., et al.
 (2019).
 Welcome to the {tidyverse}.
-\emph{Journal of Open Source Software}, \emph{4}(43), 1686.
-\url{https://doi.org/10.21105/joss.01686}
-
-\leavevmode\hypertarget{ref-wuDietEarliest2021}{}%
-Wu, Y., Tao, D., Wu, X., \& Liu, W. (2021). \emph{Diet of the earliest modern humans in {East Asia}} {[}Preprint{]}. {In Review}. \url{https://doi.org/10.21203/rs.3.rs-442096/v1}
-
-\leavevmode\hypertarget{ref-yatesOralMicrobiome2021}{}%
-Yates, J.
-A.
-F., Velsko, I.
-M., Aron, F., Posth, C., Hofman, C.
-A., Austin, R.
-M., Parker, C.
-E., Mann, A.
-E., Nägele, K., Arthur, K.
-W., Arthur, J.
-W., Bauer, C.
-C., Crevecoeur, I., Cupillard, C., Curtis, M.
-C., Dalén, L., Bonilla, M.
-D.-Z., Fernández-Lomana, J.
-C.
-D., Drucker, D.
-G., \ldots{} Warinner, C.
-(2021).
-The evolution and changing ecology of the {African} hominid oral microbiome.
-\emph{Proceedings of the National Academy of Sciences}, \emph{118}(20).
-\url{https://doi.org/10.1073/pnas.2021655118}
+\emph{Journal of Open Source Software} 4, 1686.
+doi:\href{https://doi.org/10.21105/joss.01686}{10.21105/joss.01686}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wuDietEarliest2021}{}}%
+Wu, Y., Tao, D., Wu, X., and Liu, W. (2021). Diet of the earliest modern humans in {East Asia}. {In Review} doi:\href{https://doi.org/10.21203/rs.3.rs-442096/v1}{10.21203/rs.3.rs-442096/v1}.
 
@@
 
+\hypertarget{tables}{%
+\section{Tables}\label{tables}}
+
+\begin{table}
+
+\caption{\label{tab:anova-tbl}Summary statistics for biofilm dry-weights (in mg) by treatment.}
+\centering
+\begin{tabular}[t]{l|r|r|r|r}
+\hline
+Treatment & Mean & SD & Min & Max\\
+\hline
+control & 5.44 & 2.45 & 1.67 & 11.20\\
+\hline
+mix & 4.28 & 1.95 & 1.50 & 8.44\\
+\hline
+potato & 6.25 & 2.07 & 2.54 & 8.92\\
+\hline
+wheat & 5.53 & 3.45 & 0.56 & 9.80\\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}
+
+\caption{\label{tab:solution-count-tbl}Mean starch counts from solutions, including the
+             proportional makeup of the different sizes of granules.}
+\centering
+\begin{tabular}[t]{l|l|l|l|l|l}
+\hline
+Solution & Starch & Small (\%) & Medium (\%) & Large (\%) & Total (\%)\\
+\hline
+mix & potato &  & 1051733 (53.1\%) & 928000 (46.9\%) & 1979733 (100.0\%)\\
+\hline
+mix & wheat & 18838400 (69.7\%) & 6403200 (23.7\%) & 1794133 (6.6\%) & 27035733 (100.0\%)\\
+\hline
+mix & both & 18838400 (64.9\%) & 7454933 (25.7\%) & 2722133 (9.4\%) & 29015467 (100.0\%)\\
+\hline
+potato & potato & 123733 (4.1\%) & 1337867 (44.4\%) & 1554400 (51.5\%) & 3016000 (100.0\%)\\
+\hline
+wheat & wheat & 16139467 (63.5\%) & 6434133 (25.3\%) & 2830400 (11.1\%) & 25404000 (100.0\%)\\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}
+
+\caption{\label{tab:sample-count-tbl}Mean starch counts extracted from samples with standard 
+             deviation (SD), including the proportion of granule sizes of the total count.}
+\centering
+\begin{tabular}[t]{l|l|l|r|l|r|l|r|l|r}
+\hline
+Treatment & Starch & Small (\%) & SD & Medium (\%) & SD & Large (\%) & SD & Total (\%) & SD\\
+\hline
+mix & potato &  &  & 1959 (79.6\%) & 1801 & 501 (20.40\%) & 446 & 2460 (100\%) & 2189\\
+\hline
+mix & wheat & 9515 (54.60\%) & 8860 & 6522 (37.4\%) & 6026 & 1381 (7.93\%) & 1196 & 17417 (100\%) & 15878\\
+\hline
+mix & both & 9515 (47.90\%) & 8860 & 8480 (42.7\%) & 7653 & 1882 (9.47\%) & 1596 & 19877 (100\%) & 17768\\
+\hline
+potato & potato & 351 (7.24\%) & 297 & 3565 (73.6\%) & 2402 & 930 (19.20\%) & 929 & 4846 (100\%) & 3316\\
+\hline
+wheat & wheat & 15235 (55.00\%) & 11944 & 12148 (43.9\%) & 11052 & 1953 (7.06\%) & 2016 & 27680 (100\%) & 23554\\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}
+
+\caption{\label{tab:sample-prop-tbl}The mean percentage of starches from the solutions that 
+              were incorporated into the samples.}
+\centering
+\begin{tabular}[t]{l|l|l|l|l|l}
+\hline
+Treatment & Starch & Small & Medium & Large & Total\\
+\hline
+mix & potato &  & 0.19\% & 0.05\% & 0.12\%\\
+\hline
+mix & wheat & 0.05\% & 0.10\% & 0.08\% & 0.06\%\\
+\hline
+mix & both & 0.05\% & 0.11\% & 0.07\% & 0.07\%\\
+\hline
+potato & potato & 0.28\% & 0.27\% & 0.06\% & 0.16\%\\
+\hline
+wheat & wheat & 0.09\% & 0.19\% & 0.07\% & 0.12\%\\
+\hline
+\end{tabular}
+\end{table}
+
+\hypertarget{figure-captions}{%
+\section{Figure Captions}\label{figure-captions}}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.7\linewidth]{/mnt/hogwarts/NextCloud/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview}
+
+}
+
+\caption{Overview of experiment protocol including the plate setup.}\label{fig:protocol-fig}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=0.5\linewidth]{/mnt/hogwarts/NextCloud/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar}
+
+}
+
+\caption{Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\mu$m.}\label{fig:microscope-fig}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{../figures/ratio-plots-1}
+
+}
+
+\caption{Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.}\label{fig:ratio-plots}
+\end{figure}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{../figures/cor-plot-1}
+
+}
+
+\caption{Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus.}\label{fig:cor-plot}
+\end{figure}
+
 \end{document}

### 2022-02-28 R/versioning.R

@@ -1,0 +1,23 @@
+# versioning function
+
+# copy all files to a directory
+
+#' Versioning
+#'
+#' Function to copy documents to a separate directory
+#' @param version name of version
+#' @importFrom here here
+versionr <- function(version, ...){
+  if(!isTRUE(dir.exists(here("analysis/paper/_archive/", version)))){
+    dir.create(here("analysis/paper/_archive", version))
+  }
+  vers_dir <- here("analysis/paper/_archive", version)
+  files_cp <- list.files(here("analysis/paper"), pattern = "[.][a-z,A-Z]")
+  # copy all files to _archive/"$version"
+  file.copy(here("analysis/paper/", files_cp), vers_dir)
+  if(all(file.exists(here(vers_dir, files_cp)) == TRUE)){
+    return(TRUE)
+  } else {
+    warning("One or more of the files did not copy")
+  }
+}

### 2022-02-28 analysis/paper/paper.Rmd
add S prefix to all tables and figs in supp mat
submission to journal

@@
 ## Compile refs.bib using rbbt package ##
-#cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
-#cite_keys <- unlist(cite_keys, use.names = FALSE)
-#rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
+cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
+cite_keys <- unlist(cite_keys, use.names = FALSE)
+rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
 # ## end refs.bib ##
@@
 
-# Tables
+# Tables {-}
 
@@
 
-# Figure Captions
+# Figures {-}
 

### 2022-02-28 analysis/paper/paper.tex
submission to journal

@@
 
-\usepackage{float}
-\let\origfigure\figure
-\let\endorigfigure\endfigure
-\renewenvironment{figure}[1][2] {
-    \expandafter\origfigure\expandafter[H]
-} {
-    \endorigfigure
-}
 \usepackage{float}
@@
 despite meticulous documentation and observation, due to unknown variables and
-uncertainty involved when studying living organisms. Dental calculus is a complex
+uncertainty involved in this kind of \emph{in vivo} research. Dental calculus is a complex
 oral biofilm with a multifactorial aetiology and variable formation rates both
@@
 An \(\alpha\)-amylase (EC 3.2.1.1) activity
-assay was conducted to confirm that no amylase was present in the system before
+assay was conducted to confirm that no amylase was present in the model before
 starch treatments started. Starch treatments replaced sucrose treatments, occurring twice per day
@@
 H\textsubscript{2}O, while artificial saliva samples were compared against a standard curve
-containing sterile artificial saliva (due to the colour of artificial saliva).
+containing stock AS (due to the colour of artificial saliva).
 Two photometric readings were conducted for each plate with a 540 nm filter on a
@@
 assumption based on the counts of small starches in the wheat-only and potato-only
-solutions. Of the total amount of small starches in these two solutions,
+solutions. Of the combined amount of small starches in these two solutions,
 99.2\%
@@
 lack of oral amylase activity in our model.
+Previous research conducted on dental calculus from contemporary humans and non-human
+primates suggest a high level of stochasticity involved in the retention of
+starch granules in dental calculus, and that starch granules extracted from dental
+calculus are underrepresented with regard
+to actual starch intake, which is consistent with our findings (illustrated by high
+standard deviations and low proportional incorporation).
+Leonard and colleagues (2015) found individual
+calculus samples to be a poor predictor of diet in a population, as many of the
+consumed plants were missing from some individual samples, but were present in others.\\
+Power and colleagues (2015)
+presented similar findings in non-human primates, where phytoliths were more
+representative of individual diets than starch granules.
+The size bias is also consistent with the findings by Power and colleagues
+(2015),
+who found that plants producing starches 10--20 \(\mu\)m in
+size were over-represented; however, the representation of granules larger than
+20 \(\mu\)m in their study is unclear.
 
@@
 maximum length were underrepresented in the calculus samples compared to the original
-starch solutions, an effect that was consistent across all three treatments,
-while medium granules (10--20 \(\mu\)m) were often over-represented
+starch solutions, an effect that was consistent across all three treatments. Medium
+granules (10--20 \(\mu\)m) were often over-represented
 (Table \ref{tab:sample-prop-tbl},
@@
 findings.
-
-Previous research conducted on dental calculus from contemporary humans and non-human
-primates suggest a high level of stochasticity involved in the retention of
-starch granules in dental calculus, and that starch granules extracted from dental
-calculus are underrepresented with regard
-to actual starch intake, which is consistent with our findings (illustrated by high
-standard deviations and low proportional incorporation).
-Leonard and colleagues (2015) found individual
-calculus samples to be a poor predictor of diet in a population, as many of the
-consumed plants were missing from some individual samples, but were present in others.\\
-Power and colleagues (2015)
-presented similar findings in non-human primates, where phytoliths were more
-representative of individual diets than starch granules.
-The size bias is also consistent with the findings by Power and colleagues
-(2015),
-who found that plants producing starches 10--20 \(\mu\)m in
-size were over-represented; however, the representation of granules larger than
-20 \(\mu\)m in their study is unclear.
 
@@
 
+\clearpage
+
 \hypertarget{tables}{%
-\section{Tables}\label{tables}}
+\section*{Tables}\label{tables}}
+\addcontentsline{toc}{section}{Tables}
 
@@
 
-\hypertarget{figure-captions}{%
-\section{Figure Captions}\label{figure-captions}}
+\hypertarget{figures}{%
+\section*{Figures}\label{figures}}
+\addcontentsline{toc}{section}{Figures}
 

### 2022-02-28 analysis/supplementary-materials/supp-mat.Rmd
submission to journal

@@
 output:
+  bookdown::pdf_document2:
+    toc: yes
+    df_print: kable
+    extra_dependencies: ["float"]
   bookdown::html_document2:
@@
     toc_float: yes
-  bookdown::pdf_document2:
-    toc: yes
-    df_print: kable
 ---
@@
 ```{r setup, include=FALSE}
+knitr::opts_chunk$set(warning = FALSE, 
+                      message = FALSE,
+                      fig.pos = "H",
+                      dpi = 300)
 library(here)
@@
 
-```{r plate-lid-1, fig.cap="The 24 deepwell plate and the lid with pegs (substrata)"}
+```{r plate-lid-1, fig.cap="The 24 deepwell plate and the lid with pegs (substrata)", echo=FALSE}
 knitr::include_graphics(here("analysis/supplementary-materials/figures/plate_lid_side.jpg"))
@@
 
-```{r plate-lid-2, fig.cap="The 24 deepwell plate with the lid (almost) on."}
+```{r plate-lid-2, fig.cap="The 24 deepwell plate with the lid (almost) on.", echo=FALSE}
 knitr::include_graphics(here("analysis/supplementary-materials/figures/plate_lid_on.jpg"))
@@
 
-```{r img-1, echo=FALSE, fig.cap="Image of starch granules extracted from a potato treatment sample"}
-knitr::include_graphics(here("analysis/supplementary-materials/figures/starches_w_bar.jpg"))
-```
-
-```{r img-2}
+```{r img-2, fig.cap="Microscope image of wheat starch from a Potato treatment sample.", echo=FALSE, out.width="80%"}
 knitr::include_graphics(here("analysis/supplementary-materials/figures/SNAP-103412-0006.jpg"))
@@
 
-```{r img-3, fig.cap="Microscope image of wheat starch from a wheat treatment sample."}
+```{r img-3, fig.cap="Microscope image of wheat starch from a wheat treatment sample.", echo=FALSE, out.width="80%"}
 knitr::include_graphics(here("analysis/supplementary-materials/figures/SNAP-164650-0012.jpg"))
@@
 
-```{r amylase-plt1-read1}
+```{r amylase-plt1-read1, echo=FALSE}
 plt1_ph1_result %>%
-  knitr::kable("Amylase activity in U/mL for plate 1, photometric read 1.")
+  knitr::kable(caption = "Amylase activity in U/mL for plate 1, photometric read 1.")
 ```
 
-```{r amylase-plt1-read2}
+```{r amylase-plt1-read2, echo=FALSE}
 plt1_ph2_result %>%
-  knitr::kable("Amylase activity in U/mL for plate 1, photometric read 2.")
+  knitr::kable(caption = "Amylase activity in U/mL for plate 1, photometric read 2.")
 ```
 
-```{r amylase-plt2-read1}
+```{r amylase-plt2-read1, echo=FALSE}
 plt2_ph1_result %>%
-  knitr::kable("Amylase activity in U/mL for plate 2, photometric read 1.")
+  knitr::kable(caption = "Amylase activity in U/mL for plate 2, photometric read 1.")
 ```
 
-```{r amylase-plt2-read2}
+```{r amylase-plt2-read2, echo=FALSE}
 plt2_ph2_result %>%
-  knitr::kable("Amylase activity in U/mL for plate 2, photometric read 2.")
+  knitr::kable(caption = "Amylase activity in U/mL for plate 2, photometric read 2.")
 ```
@@
 
-```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by z-score for separated treatments."}
+```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by z-score for separated treatments.", echo=FALSE}
 pl_cor
@@
 
-```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
+```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus.", echo=FALSE}
 pl_cor2

### 2022-04-24 analysis/paper/paper.Rmd
fix reference keys

@@
 [@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
-proteins [@hendyProteomicCalculus2018; @warinnerDirectEvidenceMilk2014],
-and other organic residues [@buckleyDentalCalculusCooking2014]. 
+proteins [@hendyProteomicCalculus2018; @warinnerEvidenceMilk2014],
+and other organic residues [@buckleyDentalCalculus2014]. 
 This dietary information can be preserved within the mineralised dental plaque 
@@
 from the dental calculus of archaeological fauna to investigate diet was first 
-noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
-and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
+noted by Armitage [-@armitageExtractionIdentification1975], 
+and later by Middleton and Rovner [-@middletonOpalPhytoliths1994],
 and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
@@
 characterisation (including pathogens) through the analysis of DNA and proteins 
-[@adlerSequencingAncientCalcified2013; @warinnerPathogensHost2014],
+[@adlerSequencingAncient2013; @warinnerPathogensHost2014],
 microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
-@mickleburghNewInsightsConsumption2012], 
+@mickleburghNewInsights2012], 
 other organic residues and proteins from dietary compounds 
-[@buckleyDentalCalculusCooking2014; @hendyProteomicCalculus2018], 
-and nicotine use [@eerkensDentalCalculusSource2018]. Especially the extraction 
+[@buckleyDentalCalculus2014; @hendyProteomicCalculus2018], 
+and nicotine use [@eerkensDentalCalculus2018]. Especially the extraction 
 of starch granules has become a rich source of dietary
@@
 Despite this, our knowledge of dental calculus and the incorporation pathways of 
-the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
+the various markers is limited [@radiniFoodPathways2017], as is our 
 knowledge of information-loss caused by these pathways. Additionally, the methods
@@
 
-Statistical analysis was conducted in `r version$version.string`  [@R-base] and
+Statistical analysis was conducted in `r version$version.string`  [@Rbase] and
 the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
@@
 a result that contradicts findings from archaeological contexts
-[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
+[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeography2014].
 When the concentration of starch granules
@@
 archaeological samples of considerable age 
-[@henryNeanderthalCalculus2014; @buckleyDentalCalculusCooking2014; @wuDietEarliest2021]. 
+[@henryNeanderthalCalculus2014; @buckleyDentalCalculus2014; @wuDietEarliest2021]. 
 Studies have also shown that not all starch granules come from a dietary source. 
@@
 or accidental ingestion not related to food consumption 
-[@radiniFoodMultiplePathways2017; @radiniMedievalWomenEarly2019].  
+[@radiniFoodPathways2017; @radiniMedievalWomenEarly2019].  
 When starch granules enter the mouth, whether through ingestion of food or accidental

### 2022-04-24 analysis/paper/paper.tex
fix reference keys

@@
 
-Statistical analysis was conducted in R version 4.1.2 (2021-11-01) (R Core Team, 2020) and
+Statistical analysis was conducted in R version 4.1.3 (2022-03-10) (R Core Team, 2020) and
 the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-adlerSequencingAncientCalcified2013}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-adlerSequencingAncient2013}{}}%
 Adler, C.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-armitageExtractionIdentificationOpal1975}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-armitageExtractionIdentification1975}{}}%
 Armitage, P. L. (1975). The {Extraction} and {Identification} of {Opal Phytoliths} from the {Teeth} of {Ungulates}. \emph{Journal of Archaeological Science} 2, 187--197.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-buckleyDentalCalculusCooking2014}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-buckleyDentalCalculus2014}{}}%
 Buckley, S., Usai, D., Jakob, T., Radini, A., and Hardy, K.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-dudgeonDietGeographyDrinking2014}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-dudgeonDietGeography2014}{}}%
 Dudgeon, J.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-eerkensDentalCalculusSource2018}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-eerkensDentalCalculus2018}{}}%
 Eerkens, J.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-mickleburghNewInsightsConsumption2012}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-mickleburghNewInsights2012}{}}%
 Mickleburgh, H.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-middletonExtractionOpalPhytoliths1994}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-middletonOpalPhytoliths1994}{}}%
 Middleton, W.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-R-base}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-Rbase}{}}%
 R Core Team (2020). \emph{R: {A} language and environment for statistical computing}. {Vienna, Austria}: {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-radiniFoodMultiplePathways2017}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-radiniFoodPathways2017}{}}%
 Radini, A., Nikita, E., Buckley, S., Copeland, L., and Hardy, K.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-warinnerDirectEvidenceMilk2014}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-warinnerEvidenceMilk2014}{}}%
 Warinner, C., Hendy, J., Speller, C., Cappellini, E., Fischer, R., Trachsel, C., Arneborg, J., Lynnerup, N., Craig, O.
@@
 
-{\centering \includegraphics[width=0.7\linewidth]{/mnt/hogwarts/NextCloud/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview}
+{\centering \includegraphics[width=0.7\linewidth]{/mnt/hogwarts/Insync/publications/PhD/byocstarch/analysis/figures/protocol_overview}
 
@@
 
-{\centering \includegraphics[width=0.5\linewidth]{/mnt/hogwarts/NextCloud/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar}
+{\centering \includegraphics[width=0.5\linewidth]{/mnt/hogwarts/Insync/publications/PhD/byocstarch/analysis/figures/starches_w_bar}
 

### 2022-05-17 analysis/paper/paper.Rmd
fix citation key

@@
 In this study we employ a multispecies oral biofilm model following a modified
-protocol from Sissons and colleagues [-@sissonsMultistationDentalPlaque1991] and
+protocol from Sissons and colleagues [-@sissonsMultistationPlaque1991] and
 Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
@@
 The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
-described by Sissons and colleagues [-@sissonsMultistationDentalPlaque1991]. 
+described by Sissons and colleagues [-@sissonsMultistationPlaque1991]. 
 It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
@@
 5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl~2~, 
-5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationDentalPlaque1991], 6.4 mmol/l NaHCO~3~ 
+5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationPlaque1991], 6.4 mmol/l NaHCO~3~ 
 [@shellisSyntheticSalivaCultural1978], 2.5 mg/l haemin. This is subsequently 
@@
 1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
-and 1 mmol/l arginine [@sissonsMultistationDentalPlaque1991].
+and 1 mmol/l arginine [@sissonsMultistationPlaque1991].
 
@@
 and (0.04 g/l MgCl)
-[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991].
+[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationPlaque1991].
 The substrata were submerged in 1 ml/well CPMU for six minutes, five times 

### 2022-05-17 analysis/paper/abstract.Rmd
revise project structure to comply with dissertation

--- 2021-08-19 analysis/paper/abstract.Rmd
@@
-# Abstract
-
-<!-- One or two sentences providing a basic introduction to the field, comprehensible to a scientist in any discipline -->
-
-<!-- Two to three sentences of more detailed background, comprehensible to scientists in related disciplines -->
-
-<!-- One sentence clearly stating the general problem being addressed by this particular study -->
-
-<!-- One sentence summarising the main result (with the words "here we show" or their equivalent) -->
-
-<!-- Two or three sentences explaining what the main result reveals in direct comparison to what was thought to be the case previously, or how the main result adds to previous knowledge -->
-
-<!-- One or two sentences to put the results into a more general context -->
-
-<!-- Two or three sentences to provide a broader perspective, readily comprehensible to a scientist in any discipline, may be included in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced by their inclusion.
-Under these circumstances, the length of the paragraph can be yp to 300 words.
-(This example is 190 words without the final section, and 250 words with it).
+Dental calculus has proven to contain a wealth of information on the
+dietary habits of past populations. These insights have, to a large
+extent, been obtained by the extraction and identification of starch
+granules contained within the mineralised dental plaque from a wide
+range of regions and time periods. The scope of previous studies have
+been limited to microfossil extraction and identification to reconstruct
+dietary preferences from the archaeological record, and few studies have
+attempted to address the biases of starch retention in dental calculus.
+Those that have considered this problem have been limited to \emph{in
+vivo} studies on modern humans and non-human primates. Here, we present
+a multispecies oral biofilm model, which allows experimental research on
+starch incorporation and retention to be conducted on \emph{in vitro}
+dental calculus in a controlled laboratory setting. The biofilms were
+exposed to treatment solutions with known quantities of dietary starches
+(wheat and potato) during the 25-day growth period. After this, the
+starch granules were extracted from the mature biofilm (by dissolution
+in EDTA), and counted. We show that the granule counts extracted from
+the model dental calculus represented a low proportion (ranging from
+0.06\% to 0.16\%) of the total number of granules exposed to the
+biofilms throughout the experiment. Additionally, we found that the
+ratios of granule sizes from the extracted starch granules differed from
+the original treatment solutions, with large granules (\textgreater20
+\(\mu\)m) consistently being under-represented. We also found a
+correlation between the absolute granule counts and dry-weight of the
+biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
+the concentration (count per mg) of granules and dry-weight (\emph{r} =
+0.3, 90\%CI{[}0.0618, 0.506{]}).\\
+Our results reinforce previous \emph{in vivo} studies suggesting that
+dental calculus presents a very small, and partly biased picture of the
+original dietary intake of starches, with an over-representation of
+plants producing granules smaller than 20 \(\mu\)m in size. The
+experimental model presented here is well-suited to address the need for
+further validation of methods and biases associated with dietary
+research on dental calculus.

### 2022-05-17 analysis/paper/conclusion.Rmd
push manuscript files to github

--- 2022-05-17 analysis/paper/paper.Rmd
@@
----
-title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
-#author: |
-# Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
-# date: |
-#  $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
-#### Frontiers submission ####
-runtitle: "Biases Associated with Starch Incorporation"
-documentclass: frontiers/frontiersSCNS  # or frontiersHLTH, or frontiersFPHY
-author:
-  - name: Bjørn Peare Bartholdy
-    affiliation: '1'
-    etal: BARTHOLDY # First author's last name.
-    email: b.p.bartholdy@arch.leidenuniv.nl # Indicates corresponding Author
-    institution: Leiden University
-    street: Einsteinweg 2
-    city: Leiden
-    zip: 2333 CC
-    country: Netherlands
-  - name: Amanda G. Henry
-    affiliation: '1'
-affiliation:
-  - id: '1'
-    department: Archaeological Sciences
-    institution: Leiden University
-    city: Leiden
-    country: Netherlands
-#### End: Frontiers submission ####
-#linestretch: 2.0 # double-spacing for review purposes
-bibliography: refs.bib
-csl: frontiers/frontiers.csl
-params:
-  keywords: 
-    - oral biofilm model
-    - starch
-    - alpha-amylase
-    - dental calculus
-    - paleodiet reconstruction
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(
-  echo = FALSE,
-  warning = FALSE,
-  message = FALSE,
-  #fig.path = "frontiers/figures/",
-  fig.path = "../figures/",
-  fig.align = "center",
-  #fig.ext = "jpg",
-  fig.pos = "H",
-  dpi = 300) 
-library(tidyverse)
-library(broom)
-library(here)
-library(patchwork)
-# load R scripts
-source(here("analysis/scripts/analysis.R"))
-source(here("analysis/scripts/tables.R"))
-source(here("analysis/scripts/figures.R"))
-source(here("analysis/scripts/amylase-assay.R"))
-## Compile refs.bib using rbbt package ##
-cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
-cite_keys <- unlist(cite_keys, use.names = FALSE)
-rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
-# ## end refs.bib ##
-# 
-# ## Word count ##
-# wd_count <- wordcountaddin::word_count()
-cor_r <- signif(count_cor$estimate, 3)
-cor_ci <- signif(count_cor$conf.int, 3)
-cor2_r <- signif(count_cor$estimate, 3)
-cor2_ci <- signif(count_cor$conf.int, 3)
-```
-
-\begin{abstract}
-
-Dental calculus has proven to contain a wealth of information on the
-dietary habits of past populations. These insights have, to a large
-extent, been obtained by the extraction and identification of starch
-granules contained within the mineralised dental plaque from a wide
-range of regions and time periods. The scope of previous studies have
-been limited to microfossil extraction and identification to reconstruct
-dietary preferences from the archaeological record, and few studies have
-attempted to address the biases of starch retention in dental calculus.
-Those that have considered this problem have been limited to \emph{in
-vivo} studies on modern humans and non-human primates. Here, we present
-a multispecies oral biofilm model, which allows experimental research on
-starch incorporation and retention to be conducted on \emph{in vitro}
-dental calculus in a controlled laboratory setting. The biofilms were
-exposed to treatment solutions with known quantities of dietary starches
-(wheat and potato) during the 25-day growth period. After this, the
-starch granules were extracted from the mature biofilm (by dissolution
-in EDTA), and counted. We show that the granule counts extracted from
-the model dental calculus represented a low proportion (ranging from
-0.06\% to 0.16\%) of the total number of granules exposed to the
-biofilms throughout the experiment. Additionally, we found that the
-ratios of granule sizes from the extracted starch granules differed from
-the original treatment solutions, with large granules (\textgreater20
-\(\mu\)m) consistently being under-represented. We also found a
-correlation between the absolute granule counts and dry-weight of the
-biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
-the concentration (count per mg) of granules and dry-weight (\emph{r} =
-0.3, 90\%CI{[}0.0618, 0.506{]}).\\
-Our results reinforce previous \emph{in vivo} studies suggesting that
-dental calculus presents a very small, and partly biased picture of the
-original dietary intake of starches, with an over-representation of
-plants producing granules smaller than 20 \(\mu\)m in size. The
-experimental model presented here is well-suited to address the need for
-further validation of methods and biases associated with dietary
-research on dental calculus.
-
-\tiny
- \keyFont{ \section{Keywords:} `r params$keywords`}
-
-\end{abstract}
-
-# Introduction
-
-<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
-<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
-Dental calculus has proven to contain a wealth of 
-dietary information in the form of plant microfossils 
-[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
-proteins [@hendyProteomicCalculus2018; @warinnerEvidenceMilk2014],
-and other organic residues [@buckleyDentalCalculus2014]. 
-This dietary information can be preserved within the mineralised dental plaque 
-over many millennia, providing a unique window into the food-related behaviours of 
-past populations 
-[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
-and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].
-
-<!-- more detailed background, comprehensible to scientists in related disciplines -->
-Until recently, only a few studies directly investigated the presence of plant
-microremains
-in the dental calculus of archaeological remains. The ability to extract phytoliths
-from the dental calculus of archaeological fauna to investigate diet was first 
-noted by Armitage [-@armitageExtractionIdentification1975], 
-and later by Middleton and Rovner [-@middletonOpalPhytoliths1994],
-and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
-extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
-In more recent years, the study of dental calculus has increased exponentially, 
-and the wealth of information contained within the mineralised matrix has largely
-been acknowledged. The use of dental calculus spans a wide
-variety of archaeological research areas, such as oral microbiome 
-characterisation (including pathogens) through the analysis of DNA and proteins 
-[@adlerSequencingAncient2013; @warinnerPathogensHost2014],
-microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
-@mickleburghNewInsights2012], 
-other organic residues and proteins from dietary compounds 
-[@buckleyDentalCalculus2014; @hendyProteomicCalculus2018], 
-and nicotine use [@eerkensDentalCalculus2018]. Especially the extraction 
-of starch granules has become a rich source of dietary
-information, as starch granules have proven to preserve well within dental calculus
-over a variety of geographical and temporal ranges 
-[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
-@jovanovicNeolithicCalculus2021].
-
-<!-- clearly stating the general problem being addressed by this particular study -->
-Despite this, our knowledge of dental calculus and the incorporation pathways of 
-the various markers is limited [@radiniFoodPathways2017], as is our 
-knowledge of information-loss caused by these pathways. Additionally, the methods
-we use to extract and analyse dental calculus, and make inferences on past diets
-represent another potential source of bias. Studies on both archaeological and
-modern individuals have explored these biases in more detail.
-Extraction methods were tested 
-by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
-decalcification using HCl or EDTA. 
-The authors found significantly more starches with the EDTA extraction method 
-than the HCl extraction method; however, as noted by the authors, comparisons 
-involving archaeological calculus are problematic due to variability between and 
-within individuals. 
-Studies conducted on modern humans [@leonardDentalCalculus2015] 
-and non-human primates [@powerChimpCalculus2015], 
-have explored how well microremains (phytoliths and starches)
-extracted from dental calculus represent the actual dietary intake.
-These studies are justifiably limited, 
-despite meticulous documentation and observation, due to unknown variables and 
-uncertainty involved in this kind of *in vivo* research. Dental calculus is a complex 
-oral biofilm with a multifactorial aetiology and variable formation rates both 
-within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
-contributing to
-the stochasticity of starch representation being observed in numerous studies.
-Additionally, the concentration of oral $\alpha$-amylase differs both between and
-within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
-causing different rates of hydrolysis of the starch granules present in the oral
-cavity. Add to this the effects of the many different methods
-of starch processing, as well as post-depositional processes that are still being 
-explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
-dental calculus to reconstruct diet is a highly unpredictable process.
-
-In this exploratory study, we use an oral biofilm model to investigate the 
-retention of starch granules within dental calculus in a controlled laboratory 
-setting, allowing us full control over dietary input. Our main questions concern
-the representation of granules extracted
-from the calculus compared to the actual intake. How much of the original diet is 
-incorporated into the calculus, and how much is recovered? 
-Is there differential loss of information from specific dietary markers that affects 
-the obtained dietary information, and how does this affect the representation of 
-diet from extracted microremains?  
-We find that, despite the absence of $\alpha$-amylase in
-the model, a limited proportion of the starch input is actually 
-retained in the calculus. We also observed a shift in the size ratios of individual 
-starch granules that are incorporated into the calculus, and that the number of 
-incorporated starch granules increases as the size of the calculus deposit 
-increases.
-
-# Materials and Methods
-
-## Biofilm formation
-
-In this study we employ a multispecies oral biofilm model following a modified
-protocol from Sissons and colleagues [-@sissonsMultistationPlaque1991] and
-Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
-with whole saliva was grown on a substrate suspended in artificial saliva, and
-fed with sugar (sucrose). After several days of growth, the biofilm was exposed
-to starch solutions. Mineralisation of the biofilm was aided by exposure to a
-calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
-collected for further analysis. The setup comprises a polypropylene 
-24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which 
-is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins. The individual pegs 
-were the substrata on which the calculus grew. Using this system allowed for easy
-transfer of the growing biofilm between saliva, feeding solutions, 
-and mineral solutions.
-
-The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
-described by Sissons and colleagues [-@sissonsMultistationPlaque1991]. 
-It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
-5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
-5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl~2~, 
-5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationPlaque1991], 6.4 mmol/l NaHCO~3~ 
-[@shellisSyntheticSalivaCultural1978], 2.5 mg/l haemin. This is subsequently 
-adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120&deg;C, 
-1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
-and 1 mmol/l arginine [@sissonsMultistationPlaque1991].
-
-Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
-with no history of caries, who abstained from oral hygiene for 24 hours. No 
-food was consumed two hours prior to donation and no antibiotics were taken up 
-to six months prior to donation. 
-The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15% active chlorine) 
-nylon cloth to remove particulates.
-Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
-20% (v/v) glycerine for four hours at 36&deg;C, to allow attachment of the
-salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
-substrata were transferred to a new plate containing 1 ml/well AS and incubated 
-in a shaking incubator (Infors HT Ecotron) at 36&deg;C, 30 rpm. 
-The inoculation process was repeated on days 3 and 5. 
-AS was partially refreshed once per day and fully refreshed every three days, 
-throughout the experiment, by transferring the substrata to a new plate containing 
-stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
-5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
-(days 0, 3, and 5), where the samples only received one sucrose treatment after
-inoculation.
-
-Starch treatments were initiated on day 9 to avoid starch granule counts being 
-affected by $\alpha$-amylase hydrolysis from saliva inoculation days. 
-An $\alpha$-amylase (EC 3.2.1.1) activity
-assay was conducted to confirm that no amylase was present in the model before
-starch treatments started. Starch treatments replaced sucrose treatments, occurring twice per day
-for six minutes. The starch treatments involved transferring the substrata to a 
-new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
-concentrations (w/v) wheat and potato. All starch treatments were created in dH~2~O
-with 5% (w/v) sucrose. Before transferring biofilm samples to the starch treatment
-plate, the plates were agitated to keep the starches in suspension in the
-solutions. During treatments, the rpm was increased to 60 to facilitate contact
-between starch granules and biofilms.
-
-After 15 days, mineralisation was encouraged with a 
-calcium phosphate monofluorophosphate urea (CPMU) solution containing 
-20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
-and (0.04 g/l MgCl)
-[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationPlaque1991].
-The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
-daily, in a two-hour cycle. During the mineralisation period, starch treatments
-were reduced to once per day after the five CPMU treatments. This process was repeated 
-for 10 days until the end of the experiment on day 24 
-(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).
-
-All laboratory work was conducted in sterile conditions under a laminar flow hood 
-to prevent starch and bacterial contamination. Control samples that only received
-sucrose as a treatment were included to detect starch contamination from the
-environment or cross-contamination from other wells in the same plate.
-
-## Amylase activity detection
-
-An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
-saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14. 
-Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
-Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
-All samples and standard curves were run in triplicates on two separate plates.
-Positive control saliva samples were compared against a standard curve containing 
-H~2~O, while artificial saliva samples were compared against a standard curve
-containing stock AS (due to the colour of artificial saliva).
-Two photometric readings were conducted for each plate with a 540 nm filter on a 
-Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
-The protocol is a modified version of an Enzymatic Assay of $\alpha$-Amylase
-(<https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase>) [@bernfeldAmylase1955], which measures the amount of
-maltose released from starch by $\alpha$-amylase activity. Results are reported
-in units (U) per mL enzyme, where 1 U releases 1 $\mu$mole of maltose in 6 minutes.
-The detailed protocol can be found here: <https://www.protocols.io/view/amylase-activity-bw8jphun>.
-
-## Treatment solutions
-
-A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
-on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
-(Zeiss Axioscope A1). For wheat and mixed treatment samples, we counted three 
-slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts 
-were extrapolated to the total number of granules exposed to the samples over 16
-days of treatments (see Supplementary Material for more details). For potato
-treatment samples, the whole slide was counted.
-
-## Extraction method
-
-Extraction of starches from the calculus samples was performed by dissolving the
-calculus in 0.5 $\tiny{M}$ ethylenediaminetetraacetic acid (EDTA) 
-[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
-and vortexing for 3 days until the sample was completely dissolved. 
-Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
-When transferring the sample to the slide, the sample was homogenised using 
-the pipette to ensure that the counted transects were representative of the 
-whole slide. The count from the slide was extrapolated to the whole sample
-(see Supplementary Material for more detail).
-
-Both wheat and potato granules were divided into three size categories: 
-small (<10 $\mu$m), medium (10 -- 20 $\mu$m), and large (>20 $\mu$m).
-
-## Statistical analysis
-
-Statistical analysis was conducted in `r version$version.string`  [@Rbase] and
-the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
-here [@R-here], and patchwork [@R-patchwork].
-<!-- R packages used -->
-
-To see if biofilm growth was differently affected by starch treatments, a 
-one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
-as the grouping variable (GV) was conducted.
-To analyse granule counts and calculate size proportions, mean counts for each 
-treatment were taken across
-both experimental plates, resulting in a mean count for each
-granule size category within each treatment.
-
-Pearson's *r* was conducted on sample weight and total starch count, as well as sample
-weight and starch count per mg calculus. The total count for each sample within a
-treatment was standardised by z-score to account for the differences in magnitude
-between the potato and wheat counts.
-This was applied to total biofilm weight and starch count per mg 
-calculus (also z-score standardised) to account for differences in starch 
-concentration in the calculus [as per 
-@wesolowskiEvaluatingMicrofossilContent2010].
-
-# Results
-
-All samples yielded sufficient biofilm growth and starch incorporation to be 
-included in the analysis (Figure \@ref(fig:microscope-fig)), resulting in a total of 48 biofilm samples (two plates of 24),
-45 of which were used for analysis (three samples were set aside for later
- analysis).
-Most control samples contained no starch granules, while some contained negligible
-quantities (see Supplementary Material).
-
-## No amylase activity detected in the model
-
-```{r amylase-activity}
-# table of results reported in units amylase per mL enzyme (but let's be honest,
-  # ...it doesn't really matter what the unit is. No activity is no activity)
-cols <- c("1", "2", "3")  
-rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
-plt1_ph1_result <- rbind(sal1_ph1, bmm1_ph1)
-  plt1_ph1_result[plt1_ph1_result < 0] <- 0
-
-amylase_results <- rbind(sal1_ph1, sal1_ph2, sal2_ph1, sal2_ph2, bmm1_ph1,
-                         bmm1_ph2, bmm2_ph1, bmm2_ph2)
-amylase_results[amylase_results < 0] <- NA
-positive_amylase <- na.omit(amylase_results)
-```
-
-No $\alpha$-amylase activity was detected in any of the artificial
-saliva samples from any of the days that were sampled. Only positive controls
-(saliva)
-contained amylase activity that could be detected in the assay, ranging from
-`r paste(signif(range(positive_amylase),3), collapse = " to ")` U/mL enzyme 
-(full results can be found in the Supplementary Material).
-The results are not comparable to other studies presenting $\alpha$-amylase activity
-levels in humans, as the unit definition may differ; 
-however, they are sufficient to show that there is no activity in the model.
-
-## Treatment type had minimal effect on biofilm growth
-
-A one-way ANOVA suggests that the type of starch used during the biofilm 
-growth period had a minimal effect on the growth of the biofilm 
-(expressed as total dry weight of the sample), F(`r paste(treat_df[1], treat_df[2], sep = ", ")`)
-= `r treat_f.stat`, p = `r treat_p`.
-A summary of sample weights is available in Table \@ref(tab:anova-tbl).
-
-## Starch counts
-
-<!-- starch counts in the solutions -->
-
-```{r solution-count-out}
-sol_tbl <- sol_comb_out %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
-
-sol_tbl[1,3] <- NA
-```
-
-<!-- starch counts in the samples -->
-
-```{r sample-count}
-samp_counts_per <- samp_counts %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), 
-                         ")")))
-
-# combine samp_counts_per with samp_sd_wide
-samp_counts_tbl <- samp_counts_per %>%
-  bind_cols(select(samp_sd_wide, c(s_sd, m_sd, l_sd, total_sd)), .name_repair = "unique") %>%
-  select(treatment, starch, s, s_sd, m, m_sd, l, l_sd, total, total_sd)
-
-samp_counts_tbl[1,3] <- NA
-```
-
-It was not possible to differentiate between potato and 
-wheat starches smaller than ca. 10 $\mu$m. These were counted as wheat,
-as we assumed that the majority of the small granules were wheat. We make this 
-assumption based on the counts of small starches in the wheat-only and potato-only
-solutions. Of the combined amount of small starches in these two solutions, 
-`r paste0(signif((sol_comb_out$s[5] / sum(sol_comb_out$s[4], sol_comb_out$s[5])) * 100, 3), "%")`
-are from wheat.
-<!-- count of small granules from wheat solution divided by the sum of small granules 
-from the potato and wheat solutions -->
-
-<!-- absolute solution counts compared to absolute sample counts. does exposure
-to higher quantities of starches mean more incorporated starches? -->
-The separate wheat and potato solutions were made with a 0.25% (w/v) starch
-concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
-starch, with a total concentration of 0.50% (w/v).
-The mixed treatment had the highest absolute count of starch granules in solution (mean = `r signif(sol_comb_row$total, 3)`),
-while the biofilms exposed to the wheat solution preserved the greatest number of
-granules 
-(mean = `r signif(subset(samp_counts, treatment == "wheat")$total, 3)`). 
-The potato treatment had the lowest absolute counts in both the solution 
-(`r signif(subset(sol_comb_out, treatment == "potato")$total, 3)`)
-and in the biofilm samples 
-(`r signif(subset(samp_counts, treatment == "potato")$total, 3)`) 
-(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
-
-### Proportion of available starches incorporated in samples
-
-```{r sample-perc}
-prop_wide <- prop %>%
-  ungroup() %>%
-  pivot_wider(id_cols = c(treatment, starch), 
-              names_from = size, 
-              values_from = prop) %>%
-  add_row(prop_comb_row, .after = 2)
-```
-
-The proportion of total starches from the solutions that were incorporated 
-into the samples ranged from 
-`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`,
-with potato granules being more readily incorporated than wheat in both the
-separated- and mixed-treatment samples
-(Table \@ref(tab:sample-prop-tbl)). 
-There is an inverse relationship between the absolute starch count in the solutions 
-and the proportional incorporation of starches in the biofilm samples, i.e., potato 
-had the lowest absolute count in solutions, but the highest proportional 
-incorporation, and vice versa for the mixed treatment.
-
-<!-- mixed treatments -->
-Wheat incorporation was most affected in the mixed-treatment samples, with only
-`r paste0(perc_tbl$total[2])` of the total available starches being incorporated into
-the sample, compared to
-`r paste0(perc_tbl$total[4])`
-in the separated wheat treatment.
-
-### Size ratios differ between solutions and samples
-
-<!-- size proportions -->
-<!-- pie charts -->
-
-<!-- overall size trends -->
-
-```{r size-trends}
-sol_size_prop <- sol_comb_out %>%
-  group_by(treatment, starch) %>%
-  #select(!total) %>%
-  mutate(across(where(is.numeric), function(x) x / total)) %>%
-  mutate(type = "solution")
-
-samp_size_prop <- samp_counts %>%
-  group_by(treatment, starch) %>%
-  #select(!total) %>%
-  mutate(across(where(is.numeric), function(x) x / total)) %>%
-  mutate(type = "sample")
-
-size_diff <- sol_size_prop %>%
-  full_join(samp_size_prop) %>%
-  pivot_longer(cols = c(s,m,l,total), names_to = "size", values_to = "count") %>%
-  pivot_wider(names_from = type, values_from = count) %>%
-  mutate(diff = (sample - solution) * 100) %>%
-  select(!c(solution, sample)) %>%
-  pivot_wider(names_from = size, values_from = diff) %>%
-  select(!total)
-```
-
-Overall, medium starch granules had a higher mean rate of incorporation 
-(`r scales::percent(mean(prop_wide$m, na.rm = T), 0.001)`) 
-than small 
-(`r scales::percent(mean(prop_wide$s, na.rm = T), 0.001)`) 
-and large 
-(`r scales::percent(mean(prop_wide$l, na.rm = T), 0.001)`) 
-starch granules across all treatments, while large potato starches had the lowest 
-rate of incorporation across all treatments.
-
-<!-- separated and mixed treatments -->
-The difference in incorporation between the size categories resulted in a change
-in size ratios between the original starch solutions and the extracted samples.
-Large potato granules (> 20 $\mu$m) were most affected, with a 
-`r paste0(abs(signif(size_diff$l[4], 3)), "%")` 
-decrease in relative abundance in the potato-only treatment, and a
-`r paste0(abs(signif(size_diff$l[1], 3)), "%")` 
-decrease in mixed treatments. Medium granules increased in relative abundance 
-across all samples, while small granules decreased in wheat treatments and 
-increased in potato treatments 
-(Figure \@ref(fig:ratio-plots)).
-
-<!-- Proportion and size ratios are the most important points to be made -->
-
-### Biofilm weight correlated positively with extracted starch counts
-
-```{r cor, echo=FALSE}
-# prepare cor output for reporting
-cor_r <- signif(count_cor$estimate, 3)
-cor_ci <- signif(count_cor$conf.int, 3)
-cor_p <- if(count_cor$p.value > 0.001){
-  paste("=", signif(count_cor$p.value, 3))
-} else {
-  paste("< 0.001")
-}
-
-cor2_r <- signif(starch_cor$estimate, 3)
-cor2_ci <- signif(starch_cor$conf.int, 3)
-cor2_p <- if(starch_cor$p.value > 0.001){
-  paste("=", signif(starch_cor$p.value, 3))
-} else {
-  paste("< 0.001")
-}
-# Convert r correlation result to text
-direct_cor <- ifelse(cor_r < 0, "negative", "positive")
-strength_cor <- if(cor_r >= 0.8){
-  "very strong"
-} else if(cor_r < 0.8 & cor_r >= 0.6){
-  "strong"
-} else if(cor_r < 0.6 & cor_r >= 0.4){
-  "moderate"
-} else if(cor_r < 0.4 & cor_r >= 0.2){
-  "weak"
-} else{
-  "very weak"
-}
-```
-
-Pearson's *r* suggests a 
-`r strength_cor` `r direct_cor` 
-correlation between the total weight of the biofilms and the total starch count 
-(standardised by z-score) extracted from the samples across treatments, 
-*r* = `r cor_r`, 
-90%CI[`r paste(cor_ci[1], cor_ci[2], sep = ", ")`],
-p `r cor_p`
-(Figure \@ref(fig:cor-plot)A).
-
-The same test was applied to total biofilm weight and starch count per mg 
-calculus (also standardised by z-score), resulting in a weak positive correlation, 
-*r* = `r cor2_r`, 
-90%CI[`r paste(cor2_ci[1], cor2_ci[2], sep = ", ")`],
-p `r cor2_p`
-(Figure \@ref(fig:cor-plot)B).
-
-# Discussion
-
-<!-- Big summary/important statement for the science -->
-Here, we have provided a method for exploring the incorporation of dietary 
-starches into the mineral matrix of a dental calculus biofilm model. Our results show
-that a very low proportion of the starches exposed to the biofilm during growth are
-retained in the mineral matrix, and that the size of the starch granules
-may affect the likelihood of incorporation. The proportions of starch granules 
-(of all sizes) present in the extracted samples were similar across all treatments
-(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`), 
-despite large differences in absolute granule counts between wheat 
-(mean = `r format(sol_comb_out$total[5], big.mark = ",")`) 
-and potato
-(mean = `r format(sol_comb_out$total[4], big.mark = ",")`)
-solutions.  
-The absolute counts, however, differed more visibly between treatments and was
-proportional with the total count of granules in the treatment solutions. Wheat 
-and mixed solutions had the highest absolute mean count of starch granules, and 
-also had the highest absolute mean count of starch granules extracted from the 
-dental calculus
-(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
-This suggests that the starches that are more frequently consumed will be present
-in higher quantities in the dental calculus, at least prior to inhumation and 
-degradation in the burial environment. 
-Despite the low proportion of granules recovered from the model calculus 
-(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`),
-the absolute counts were still substantially greater than counts recovered from 
-archaeological remains 
-[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010], which could in part be due to the 
-lack of oral amylase activity in our model. 
-<!-- Scientific findings as extension of previous work -->
-Previous research conducted on dental calculus from contemporary humans and non-human
-primates suggest a high level of stochasticity involved in the retention of 
-starch granules in dental calculus, and that starch granules extracted from dental 
-calculus are underrepresented with regard
-to actual starch intake, which is consistent with our findings (illustrated by high
-standard deviations and low proportional incorporation).
-Leonard and colleagues [-@leonardDentalCalculus2015] found individual
-calculus samples to be a poor predictor of diet in a population, as many of the 
-consumed plants were missing from some individual samples, but were present in others.   
-Power and colleagues [-@powerChimpCalculus2015] 
-presented similar findings in non-human primates, where phytoliths were more
-representative of individual diets than starch granules.
-The size bias is also consistent with the findings by Power and colleagues
-[-@powerChimpCalculus2015],
-who found that plants producing starches 10--20 $\mu$m in
-size were over-represented; however, the representation of granules larger than
-20 $\mu$m in their study is unclear.
-
-We have also shown that the size of the starch granules influences the likelihood
-of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
-maximum length were underrepresented in the calculus samples compared to the original 
-starch solutions, an effect that was consistent across all three treatments. Medium
-granules (10--20 $\mu$m) were often over-represented
-(Table \@ref(tab:sample-prop-tbl), 
-and 
-Figure \@ref(fig:ratio-plots)).
-Large potato granules were most affected, potentially because of the greater
-size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
-granules generally only reach up to 35 $\mu$m 
-[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
-Granule morphology may also play a role. Large wheat granules 
-are lenticular and have a larger surface area compared 
-to volume, whereas large potato granules are ovoid and have a larger volume
-compared to surface area 
-[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
-Another potentially important factor 
-from our results is the size of the calculus deposit. We found a 
-`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
-starch granules (Figure \@ref(fig:cor-plot)A);
-a result that contradicts findings from archaeological contexts
-[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeography2014].
-When the concentration of starch granules
-per mg calculus is considered, the correlation is weaker, but still present
-(Figure \@ref(fig:cor-plot)B).
-While the larger deposits contain a higher absolute count, our findings also suggest
-that they contain a slightly higher concentration of starches. This may also explain
-the lower mean retention of starch granules in mixed treatments compared to
-wheat treatments.
-Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
-granule counts, wheat treatment samples had the highest mean starch retention.
-Further research is needed to determine why this differs from previous archaeological
-findings.
-
-<!-- How are starch granules trapped in dental calculus? Does anyone actually know? 
--->
-The mechanism by which starch granules are incorporated into plaque and calculus 
-remains largely unknown, and few studies have directly investigated potential
-mechanisms. We know that a proportion of the starch granules entering 
-the mouth can become trapped in the plaque/calculus, and can be recovered from
-archaeological samples of considerable age 
-[@henryNeanderthalCalculus2014; @buckleyDentalCalculus2014; @wuDietEarliest2021]. 
-Studies have also shown that not all starch granules come from a dietary source. 
-Other pathways include cross-contamination from plant interactions in soil, such 
-as palm phytoliths adhering to the skin of sweet potatoes 
-[@trompDietaryNondietary2015], 
-or accidental ingestion not related to food consumption 
-[@radiniFoodPathways2017; @radiniMedievalWomenEarly2019].  
-When starch granules enter the mouth, whether through ingestion of food or accidental
-intake, they immediately encounter multiple obstacles. It is likely
-that the bulk of starch granules are swallowed along with the food, and are 
-only briefly present in the oral cavity. Other granules that are broken off
-during mastication may be retained in the dentition through attachment to 
-tooth surfaces (including plaque and dental calculus) and mucous membranes 
-[@kashketFoodRetention1991; @doddsCarbohydrateRetention1988].
-Bacteria also have the ability to adhere to starch granules 
-[@toppingResistantStarch2003],
-which would allow starches to attach to bacterial communities within the biofilm.
-These granules are then
-susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
-by $\alpha$-amylase [@kashketFoodParticles1996].
-The susceptibility of granules to hydrolysis depends on the crystallinity and size
-of the starch granule, as well as the mode of processing. Smaller and pre-processed
-(e.g., cooked) starch granules are more susceptible to enzymatic degradation, 
-while dehydrated starches will have a reduced susceptibility
-[@lingstromStarchyFood1994; @francoStarchDegradation1992; @haslamDecompositionStarch2004; 
-@bjorckStarchProcessing1984; @henryCookingStarch2009].
-Cracks on the surface of the dental calculus, as well as unmineralised islands 
-and channels may also be able to contain starch granules 
-[@powerSEMCalculus2014; @tanCalculusUltrastructure2004; @charlierSEMCalculus2010].
-Starch granules that are trapped in these pockets are (at least to some extent) 
-protected from aforementioned clearance mechanisms, especially once
-a new layer of plaque has covered the surface of the plaque/calculus. 
-The size bias against large granules (>20 $\mu$m) from both wheat and potato 
-(Table \@ref(tab:sample-prop-tbl)) may give further credence to
-this incorporation pathway, as the smaller starch granules have an advantage over
-larger granules, and can be stored in larger quantities.
-This was also suggested by Power and colleagues [-@powerSEMCalculus2014], who
-observed clusters of starches within dental calculus, rather than an even 
-distribution across the surface of the dental calculus.
-Granules trapped 
-in plaque/calculus may still be susceptible to hydrolysis, as $\alpha$-amylase has
-the ability to bind to both tooth enamel and bacteria within a biofilm and retain
-a portion of its hydrolytic activity
-[@scannapiecoSalivaryAmylase1993; @nikitkovaStarchBiofilms2013; @tanBacterialViability2004; @tanCalculusUltrastructure2004].
-After the death of an individual, starches within dental calculus are susceptible 
-to further degradation by post-depositional processes, depending on burial environment 
-(pH, temperature, moisture content, microorganisms)
-[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004; @francoStarchDegradation1992].
-Future study should explore how burial affects the recovery of starch from the
-biofilm model. <!-- I may get to it eventually, but it would be great if someone else could look at this! -->
-
-<!-- in summary: we know very little about the mechanism behind how starches 
-become trapped in dental calculus, but it would be nice to find out more. -->
-
-<!-- limitations of the platform show which questions would be unanswered -->
-The absence of $\alpha$-amylase in the model is a limitation of this study, as 
-the total granule counts were not subject to hydrolysis. This would likely have 
-reduced and affected the size ratios, as smaller starches may be more 
-susceptible to hydrolysis 
-[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. The absence may
-also affect biofilm growth due to the lack of amylase-bacterium interactions
-[@nikitkovaStarchBiofilms2013]. Conversely, the model may benefit from the
-absence of $\alpha$-amylase, because it can allow us to directly explore
-its effect on starch counts in future experiments, where $\alpha$-amylase can be 
-added to the model in concentrations similar to those found in the oral cavity
-[@scannapiecoSalivaryAmylase1993]. 
-We are able to show how absolute counts in the treatments cause a difference in
-incorporation. However, this
-was merely a side-effect of the difference in the number of granules in potato and 
-wheat solutions of the same concentration (w/v). Further research should test 
-multiple differing concentrations of the same starch type.
-The use of EDTA may also have affected counts. While previous studies have shown 
-negligible morphological changes caused by exposure to EDTA 
-[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
-these studies have not considered changes to separate size
-categories within starch types, and whether shifts in size ratios occur due to
-exposure to the pre-treatment chemicals.
-The total number of granules on a slide often exceeded a number that 
-was feasible to count in a reasonable time period, so we calculated the total 
-counts by extrapolating from three slide transects. 
-Thus, we reasonably assume that the three transects are a good representation 
-of the entire slide, and that the distribution of all granules on the slide is 
-relatively homogeneous.   
-Finally, we only used native starches in the experimental procedure and the results 
-will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
-Based on the comparatively low counts obtained by Leonard
-and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
-may have a substantial effect on starch granule retention in the oral cavity.
-
-<!-- How this engineering will facilitate future research -->
-
-While we are unable to sufficiently address the 
-mechanism(s) of starch incorporation with the data obtained in this study, 
-the dental calculus model presented here is uniquely suited to explore 
-these questions and may improve interpretations of dietary practices in past 
-populations. Further analyses using this model can address the call for more 
-baseline testing of biases associated with dietary research conducted on dental calculus 
-[@lemoyneCalculusPretreatments2021].
-Our high-throughput experimental setup allows us a
-higher degree of control over the factors that influence starch incorporation and
-retention, such as dietary intake, differential survivability of starches, 
-and inter- and intra-individual variation in plaque accumulation and mineralisation.
-The latter is especially difficult to control *in vivo* as it is influenced by 
-numerous factors including genetics, diet, salivary flow, and tooth position and
-morphology 
-[@jepsenCalculusRemoval2011; @simon-soroOralGeography2013; @proctorSpatialGradient2018; @fagernasMicrobialBiogeography2021; @haffajeeBiofilmPosition2009], 
-as well as evolutionary differences [@yatesOralMicrobiome2021].
-It can also facilitate training of students and researchers on methods of 
-dental calculus analysis, such as starch and phytolith extraction and 
-identification, where it can replace the use of finite archaeological resources.
-
-<!-- we still need to replicate results and do more science... but pretty cool, right? -->
-
-# Conclusions
-
 This preliminary study shows that a very small proportion of the input starch
@@
 with dietary research conducted on archaeological dental calculus.
-
-# Open Science Statement {-}
-
-All scripts and data used in the analysis are available on OSF 
-(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch),
-following the format provided by the rrtools package [@R-rrtools].
-More detailed protocols are available on OSF (https://osf.io/akevs/)
-and protocols.io (https://www.protocols.io/workspaces/byoc).
-Additional tables and figures are available in the Supplementary Material
-(https://osf.io/ucxsv/).
-
-# Author Contributions {-}
-
-AGH acquired the funding for the analysis. BPB and AGH conceptualised and designed
-the study. BPB conducted the experiments and analysed the data.
-BPB wrote the manuscript. AGH reviewed and edited the manuscript.
-
-# Funding {-}
-
-This research has received funding from the European Research Council under the 
-European Union’s Horizon 2020 research and innovation program, grant agreement 
-number STG–677576 (“HARVEST”).
-
-# Conflict of Interest {-}
-
-The authors declare that the research was conducted in the absence of any
-commercial or financial relationships that could be construed as a potential
-conflict of interest.
-
-# Acknowledgements {-}
-
-<!-- If you're reading this, thank you for taking the time to read the source 
-documents. You're awesome! -->
-We would like to thank Dr. Stephanie Schnorr for help with the amylase activity protocol.
-We also thank everyone in the general vicinity
-of the lab for enduring the smell of bacterial accumulation.
-Since we did NOT make use of Sci-Hub to access articles stuck behind a paywall, 
-we will NOT acknowledge the use of Sci-Hub in this study. <!-- Oops, did I just acknowledge it twice..? -->
-
-# References {-}
-
-::: {#refs}
-:::
-
-\clearpage
-
-# Tables {-}
-
-```{r anova-tbl}
-knitr::kable(anova_tbl, 
-             col.names = c("Treatment", "Mean", "SD", "Min", "Max"),
-             caption = "Summary statistics for biofilm dry-weights (in mg) by treatment.")
-```
-
-```{r solution-count-tbl}
-cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
-options(knitr.kable.NA = "")
-knitr::kable(sol_tbl, col.names = cnames, 
-             caption = "Mean starch counts from solutions, including the
-             proportional makeup of the different sizes of granules.")
-```
-
-```{r sample-count-tbl}
-cnames <- c("Treatment", "Starch", "Small (%)", "SD", "Medium (%)", "SD", 
-            "Large (%)", "SD", "Total (%)", "SD")
-knitr::kable(samp_counts_tbl, col.names = cnames, 
-             caption = "Mean starch counts extracted from samples with standard 
-             deviation (SD), including the proportion of granule sizes of the total count.")
-```
-
-```{r sample-prop-tbl}
-cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
-knitr::kable(perc_tbl, col.names = cnames, 
-             caption = "The mean percentage of starches from the solutions that 
-              were incorporated into the samples.")
-```
-
-# Figures {-}
-
-```{r protocol-fig, fig.cap="Overview of experiment protocol including the plate setup.", out.width="70%"}
-knitr::include_graphics(here("analysis/figures/protocol_overview.png"))
-```
-
-```{r microscope-fig, fig.cap="Microscope image of a biofilm sample exposed to the potato starch solution.
-Potato granules can be seen within a bacterial community.
-Scale bar = 20 $\\mu$m.", out.width="50%"}
-knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
-```
-
-```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."}
-#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
-(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
-```
-
-```{r cor-plot, warning=FALSE, fig.cap="Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus."}
-#pl_cor
-pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
-#ggsave("frontiers/figures/cor-plot.png", dpi = 300)
-```
-
-<!--
-Figure 1: Overview of experiment protocol including the plate setup.
-
-Figure 2: Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\mu$m.
-
-Figure 3: Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.
-
-Figure 4: Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus.
--->

### 2022-05-17 analysis/paper/discussion.Rmd
push manuscript files to github

--- 2022-05-17 analysis/paper/paper.Rmd
@@
----
-title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
-#author: |
-# Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
-# date: |
-#  $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
-#### Frontiers submission ####
-runtitle: "Biases Associated with Starch Incorporation"
-documentclass: frontiers/frontiersSCNS  # or frontiersHLTH, or frontiersFPHY
-author:
-  - name: Bjørn Peare Bartholdy
-    affiliation: '1'
-    etal: BARTHOLDY # First author's last name.
-    email: b.p.bartholdy@arch.leidenuniv.nl # Indicates corresponding Author
-    institution: Leiden University
-    street: Einsteinweg 2
-    city: Leiden
-    zip: 2333 CC
-    country: Netherlands
-  - name: Amanda G. Henry
-    affiliation: '1'
-affiliation:
-  - id: '1'
-    department: Archaeological Sciences
-    institution: Leiden University
-    city: Leiden
-    country: Netherlands
-#### End: Frontiers submission ####
-#linestretch: 2.0 # double-spacing for review purposes
-bibliography: refs.bib
-csl: frontiers/frontiers.csl
-params:
-  keywords: 
-    - oral biofilm model
-    - starch
-    - alpha-amylase
-    - dental calculus
-    - paleodiet reconstruction
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(
-  echo = FALSE,
-  warning = FALSE,
-  message = FALSE,
-  #fig.path = "frontiers/figures/",
-  fig.path = "../figures/",
-  fig.align = "center",
-  #fig.ext = "jpg",
-  fig.pos = "H",
-  dpi = 300) 
-library(tidyverse)
-library(broom)
-library(here)
-library(patchwork)
-# load R scripts
-source(here("analysis/scripts/analysis.R"))
-source(here("analysis/scripts/tables.R"))
-source(here("analysis/scripts/figures.R"))
-source(here("analysis/scripts/amylase-assay.R"))
-## Compile refs.bib using rbbt package ##
-cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
-cite_keys <- unlist(cite_keys, use.names = FALSE)
-rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
-# ## end refs.bib ##
-# 
-# ## Word count ##
-# wd_count <- wordcountaddin::word_count()
-cor_r <- signif(count_cor$estimate, 3)
-cor_ci <- signif(count_cor$conf.int, 3)
-cor2_r <- signif(count_cor$estimate, 3)
-cor2_ci <- signif(count_cor$conf.int, 3)
-```
-
-\begin{abstract}
-
-Dental calculus has proven to contain a wealth of information on the
-dietary habits of past populations. These insights have, to a large
-extent, been obtained by the extraction and identification of starch
-granules contained within the mineralised dental plaque from a wide
-range of regions and time periods. The scope of previous studies have
-been limited to microfossil extraction and identification to reconstruct
-dietary preferences from the archaeological record, and few studies have
-attempted to address the biases of starch retention in dental calculus.
-Those that have considered this problem have been limited to \emph{in
-vivo} studies on modern humans and non-human primates. Here, we present
-a multispecies oral biofilm model, which allows experimental research on
-starch incorporation and retention to be conducted on \emph{in vitro}
-dental calculus in a controlled laboratory setting. The biofilms were
-exposed to treatment solutions with known quantities of dietary starches
-(wheat and potato) during the 25-day growth period. After this, the
-starch granules were extracted from the mature biofilm (by dissolution
-in EDTA), and counted. We show that the granule counts extracted from
-the model dental calculus represented a low proportion (ranging from
-0.06\% to 0.16\%) of the total number of granules exposed to the
-biofilms throughout the experiment. Additionally, we found that the
-ratios of granule sizes from the extracted starch granules differed from
-the original treatment solutions, with large granules (\textgreater20
-\(\mu\)m) consistently being under-represented. We also found a
-correlation between the absolute granule counts and dry-weight of the
-biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
-the concentration (count per mg) of granules and dry-weight (\emph{r} =
-0.3, 90\%CI{[}0.0618, 0.506{]}).\\
-Our results reinforce previous \emph{in vivo} studies suggesting that
-dental calculus presents a very small, and partly biased picture of the
-original dietary intake of starches, with an over-representation of
-plants producing granules smaller than 20 \(\mu\)m in size. The
-experimental model presented here is well-suited to address the need for
-further validation of methods and biases associated with dietary
-research on dental calculus.
-
-\tiny
- \keyFont{ \section{Keywords:} `r params$keywords`}
-
-\end{abstract}
-
-# Introduction
-
-<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
-<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
-Dental calculus has proven to contain a wealth of 
-dietary information in the form of plant microfossils 
-[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
-proteins [@hendyProteomicCalculus2018; @warinnerEvidenceMilk2014],
-and other organic residues [@buckleyDentalCalculus2014]. 
-This dietary information can be preserved within the mineralised dental plaque 
-over many millennia, providing a unique window into the food-related behaviours of 
-past populations 
-[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
-and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].
-
-<!-- more detailed background, comprehensible to scientists in related disciplines -->
-Until recently, only a few studies directly investigated the presence of plant
-microremains
-in the dental calculus of archaeological remains. The ability to extract phytoliths
-from the dental calculus of archaeological fauna to investigate diet was first 
-noted by Armitage [-@armitageExtractionIdentification1975], 
-and later by Middleton and Rovner [-@middletonOpalPhytoliths1994],
-and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
-extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
-In more recent years, the study of dental calculus has increased exponentially, 
-and the wealth of information contained within the mineralised matrix has largely
-been acknowledged. The use of dental calculus spans a wide
-variety of archaeological research areas, such as oral microbiome 
-characterisation (including pathogens) through the analysis of DNA and proteins 
-[@adlerSequencingAncient2013; @warinnerPathogensHost2014],
-microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
-@mickleburghNewInsights2012], 
-other organic residues and proteins from dietary compounds 
-[@buckleyDentalCalculus2014; @hendyProteomicCalculus2018], 
-and nicotine use [@eerkensDentalCalculus2018]. Especially the extraction 
-of starch granules has become a rich source of dietary
-information, as starch granules have proven to preserve well within dental calculus
-over a variety of geographical and temporal ranges 
-[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
-@jovanovicNeolithicCalculus2021].
-
-<!-- clearly stating the general problem being addressed by this particular study -->
-Despite this, our knowledge of dental calculus and the incorporation pathways of 
-the various markers is limited [@radiniFoodPathways2017], as is our 
-knowledge of information-loss caused by these pathways. Additionally, the methods
-we use to extract and analyse dental calculus, and make inferences on past diets
-represent another potential source of bias. Studies on both archaeological and
-modern individuals have explored these biases in more detail.
-Extraction methods were tested 
-by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
-decalcification using HCl or EDTA. 
-The authors found significantly more starches with the EDTA extraction method 
-than the HCl extraction method; however, as noted by the authors, comparisons 
-involving archaeological calculus are problematic due to variability between and 
-within individuals. 
-Studies conducted on modern humans [@leonardDentalCalculus2015] 
-and non-human primates [@powerChimpCalculus2015], 
-have explored how well microremains (phytoliths and starches)
-extracted from dental calculus represent the actual dietary intake.
-These studies are justifiably limited, 
-despite meticulous documentation and observation, due to unknown variables and 
-uncertainty involved in this kind of *in vivo* research. Dental calculus is a complex 
-oral biofilm with a multifactorial aetiology and variable formation rates both 
-within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
-contributing to
-the stochasticity of starch representation being observed in numerous studies.
-Additionally, the concentration of oral $\alpha$-amylase differs both between and
-within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
-causing different rates of hydrolysis of the starch granules present in the oral
-cavity. Add to this the effects of the many different methods
-of starch processing, as well as post-depositional processes that are still being 
-explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
-dental calculus to reconstruct diet is a highly unpredictable process.
-
-In this exploratory study, we use an oral biofilm model to investigate the 
-retention of starch granules within dental calculus in a controlled laboratory 
-setting, allowing us full control over dietary input. Our main questions concern
-the representation of granules extracted
-from the calculus compared to the actual intake. How much of the original diet is 
-incorporated into the calculus, and how much is recovered? 
-Is there differential loss of information from specific dietary markers that affects 
-the obtained dietary information, and how does this affect the representation of 
-diet from extracted microremains?  
-We find that, despite the absence of $\alpha$-amylase in
-the model, a limited proportion of the starch input is actually 
-retained in the calculus. We also observed a shift in the size ratios of individual 
-starch granules that are incorporated into the calculus, and that the number of 
-incorporated starch granules increases as the size of the calculus deposit 
-increases.
-
-# Materials and Methods
-
-## Biofilm formation
-
-In this study we employ a multispecies oral biofilm model following a modified
-protocol from Sissons and colleagues [-@sissonsMultistationPlaque1991] and
-Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
-with whole saliva was grown on a substrate suspended in artificial saliva, and
-fed with sugar (sucrose). After several days of growth, the biofilm was exposed
-to starch solutions. Mineralisation of the biofilm was aided by exposure to a
-calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
-collected for further analysis. The setup comprises a polypropylene 
-24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which 
-is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins. The individual pegs 
-were the substrata on which the calculus grew. Using this system allowed for easy
-transfer of the growing biofilm between saliva, feeding solutions, 
-and mineral solutions.
-
-The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
-described by Sissons and colleagues [-@sissonsMultistationPlaque1991]. 
-It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
-5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
-5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl~2~, 
-5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationPlaque1991], 6.4 mmol/l NaHCO~3~ 
-[@shellisSyntheticSalivaCultural1978], 2.5 mg/l haemin. This is subsequently 
-adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120&deg;C, 
-1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
-and 1 mmol/l arginine [@sissonsMultistationPlaque1991].
-
-Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
-with no history of caries, who abstained from oral hygiene for 24 hours. No 
-food was consumed two hours prior to donation and no antibiotics were taken up 
-to six months prior to donation. 
-The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15% active chlorine) 
-nylon cloth to remove particulates.
-Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
-20% (v/v) glycerine for four hours at 36&deg;C, to allow attachment of the
-salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
-substrata were transferred to a new plate containing 1 ml/well AS and incubated 
-in a shaking incubator (Infors HT Ecotron) at 36&deg;C, 30 rpm. 
-The inoculation process was repeated on days 3 and 5. 
-AS was partially refreshed once per day and fully refreshed every three days, 
-throughout the experiment, by transferring the substrata to a new plate containing 
-stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
-5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
-(days 0, 3, and 5), where the samples only received one sucrose treatment after
-inoculation.
-
-Starch treatments were initiated on day 9 to avoid starch granule counts being 
-affected by $\alpha$-amylase hydrolysis from saliva inoculation days. 
-An $\alpha$-amylase (EC 3.2.1.1) activity
-assay was conducted to confirm that no amylase was present in the model before
-starch treatments started. Starch treatments replaced sucrose treatments, occurring twice per day
-for six minutes. The starch treatments involved transferring the substrata to a 
-new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
-concentrations (w/v) wheat and potato. All starch treatments were created in dH~2~O
-with 5% (w/v) sucrose. Before transferring biofilm samples to the starch treatment
-plate, the plates were agitated to keep the starches in suspension in the
-solutions. During treatments, the rpm was increased to 60 to facilitate contact
-between starch granules and biofilms.
-
-After 15 days, mineralisation was encouraged with a 
-calcium phosphate monofluorophosphate urea (CPMU) solution containing 
-20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
-and (0.04 g/l MgCl)
-[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationPlaque1991].
-The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
-daily, in a two-hour cycle. During the mineralisation period, starch treatments
-were reduced to once per day after the five CPMU treatments. This process was repeated 
-for 10 days until the end of the experiment on day 24 
-(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).
-
-All laboratory work was conducted in sterile conditions under a laminar flow hood 
-to prevent starch and bacterial contamination. Control samples that only received
-sucrose as a treatment were included to detect starch contamination from the
-environment or cross-contamination from other wells in the same plate.
-
-## Amylase activity detection
-
-An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
-saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14. 
-Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
-Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
-All samples and standard curves were run in triplicates on two separate plates.
-Positive control saliva samples were compared against a standard curve containing 
-H~2~O, while artificial saliva samples were compared against a standard curve
-containing stock AS (due to the colour of artificial saliva).
-Two photometric readings were conducted for each plate with a 540 nm filter on a 
-Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
-The protocol is a modified version of an Enzymatic Assay of $\alpha$-Amylase
-(<https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase>) [@bernfeldAmylase1955], which measures the amount of
-maltose released from starch by $\alpha$-amylase activity. Results are reported
-in units (U) per mL enzyme, where 1 U releases 1 $\mu$mole of maltose in 6 minutes.
-The detailed protocol can be found here: <https://www.protocols.io/view/amylase-activity-bw8jphun>.
-
-## Treatment solutions
-
-A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
-on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
-(Zeiss Axioscope A1). For wheat and mixed treatment samples, we counted three 
-slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts 
-were extrapolated to the total number of granules exposed to the samples over 16
-days of treatments (see Supplementary Material for more details). For potato
-treatment samples, the whole slide was counted.
-
-## Extraction method
-
-Extraction of starches from the calculus samples was performed by dissolving the
-calculus in 0.5 $\tiny{M}$ ethylenediaminetetraacetic acid (EDTA) 
-[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
-and vortexing for 3 days until the sample was completely dissolved. 
-Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
-When transferring the sample to the slide, the sample was homogenised using 
-the pipette to ensure that the counted transects were representative of the 
-whole slide. The count from the slide was extrapolated to the whole sample
-(see Supplementary Material for more detail).
-
-Both wheat and potato granules were divided into three size categories: 
-small (<10 $\mu$m), medium (10 -- 20 $\mu$m), and large (>20 $\mu$m).
-
-## Statistical analysis
-
-Statistical analysis was conducted in `r version$version.string`  [@Rbase] and
-the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
-here [@R-here], and patchwork [@R-patchwork].
-<!-- R packages used -->
-
-To see if biofilm growth was differently affected by starch treatments, a 
-one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
-as the grouping variable (GV) was conducted.
-To analyse granule counts and calculate size proportions, mean counts for each 
-treatment were taken across
-both experimental plates, resulting in a mean count for each
-granule size category within each treatment.
-
-Pearson's *r* was conducted on sample weight and total starch count, as well as sample
-weight and starch count per mg calculus. The total count for each sample within a
-treatment was standardised by z-score to account for the differences in magnitude
-between the potato and wheat counts.
-This was applied to total biofilm weight and starch count per mg 
-calculus (also z-score standardised) to account for differences in starch 
-concentration in the calculus [as per 
-@wesolowskiEvaluatingMicrofossilContent2010].
-
-# Results
-
-All samples yielded sufficient biofilm growth and starch incorporation to be 
-included in the analysis (Figure \@ref(fig:microscope-fig)), resulting in a total of 48 biofilm samples (two plates of 24),
-45 of which were used for analysis (three samples were set aside for later
- analysis).
-Most control samples contained no starch granules, while some contained negligible
-quantities (see Supplementary Material).
-
-## No amylase activity detected in the model
-
-```{r amylase-activity}
-# table of results reported in units amylase per mL enzyme (but let's be honest,
-  # ...it doesn't really matter what the unit is. No activity is no activity)
-cols <- c("1", "2", "3")  
-rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
-plt1_ph1_result <- rbind(sal1_ph1, bmm1_ph1)
-  plt1_ph1_result[plt1_ph1_result < 0] <- 0
-
-amylase_results <- rbind(sal1_ph1, sal1_ph2, sal2_ph1, sal2_ph2, bmm1_ph1,
-                         bmm1_ph2, bmm2_ph1, bmm2_ph2)
-amylase_results[amylase_results < 0] <- NA
-positive_amylase <- na.omit(amylase_results)
-```
-
-No $\alpha$-amylase activity was detected in any of the artificial
-saliva samples from any of the days that were sampled. Only positive controls
-(saliva)
-contained amylase activity that could be detected in the assay, ranging from
-`r paste(signif(range(positive_amylase),3), collapse = " to ")` U/mL enzyme 
-(full results can be found in the Supplementary Material).
-The results are not comparable to other studies presenting $\alpha$-amylase activity
-levels in humans, as the unit definition may differ; 
-however, they are sufficient to show that there is no activity in the model.
-
-## Treatment type had minimal effect on biofilm growth
-
-A one-way ANOVA suggests that the type of starch used during the biofilm 
-growth period had a minimal effect on the growth of the biofilm 
-(expressed as total dry weight of the sample), F(`r paste(treat_df[1], treat_df[2], sep = ", ")`)
-= `r treat_f.stat`, p = `r treat_p`.
-A summary of sample weights is available in Table \@ref(tab:anova-tbl).
-
-## Starch counts
-
-<!-- starch counts in the solutions -->
-
-```{r solution-count-out}
-sol_tbl <- sol_comb_out %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
-
-sol_tbl[1,3] <- NA
-```
-
-<!-- starch counts in the samples -->
-
-```{r sample-count}
-samp_counts_per <- samp_counts %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), 
-                         ")")))
-
-# combine samp_counts_per with samp_sd_wide
-samp_counts_tbl <- samp_counts_per %>%
-  bind_cols(select(samp_sd_wide, c(s_sd, m_sd, l_sd, total_sd)), .name_repair = "unique") %>%
-  select(treatment, starch, s, s_sd, m, m_sd, l, l_sd, total, total_sd)
-
-samp_counts_tbl[1,3] <- NA
-```
-
-It was not possible to differentiate between potato and 
-wheat starches smaller than ca. 10 $\mu$m. These were counted as wheat,
-as we assumed that the majority of the small granules were wheat. We make this 
-assumption based on the counts of small starches in the wheat-only and potato-only
-solutions. Of the combined amount of small starches in these two solutions, 
-`r paste0(signif((sol_comb_out$s[5] / sum(sol_comb_out$s[4], sol_comb_out$s[5])) * 100, 3), "%")`
-are from wheat.
-<!-- count of small granules from wheat solution divided by the sum of small granules 
-from the potato and wheat solutions -->
-
-<!-- absolute solution counts compared to absolute sample counts. does exposure
-to higher quantities of starches mean more incorporated starches? -->
-The separate wheat and potato solutions were made with a 0.25% (w/v) starch
-concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
-starch, with a total concentration of 0.50% (w/v).
-The mixed treatment had the highest absolute count of starch granules in solution (mean = `r signif(sol_comb_row$total, 3)`),
-while the biofilms exposed to the wheat solution preserved the greatest number of
-granules 
-(mean = `r signif(subset(samp_counts, treatment == "wheat")$total, 3)`). 
-The potato treatment had the lowest absolute counts in both the solution 
-(`r signif(subset(sol_comb_out, treatment == "potato")$total, 3)`)
-and in the biofilm samples 
-(`r signif(subset(samp_counts, treatment == "potato")$total, 3)`) 
-(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
-
-### Proportion of available starches incorporated in samples
-
-```{r sample-perc}
-prop_wide <- prop %>%
-  ungroup() %>%
-  pivot_wider(id_cols = c(treatment, starch), 
-              names_from = size, 
-              values_from = prop) %>%
-  add_row(prop_comb_row, .after = 2)
-```
-
-The proportion of total starches from the solutions that were incorporated 
-into the samples ranged from 
-`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`,
-with potato granules being more readily incorporated than wheat in both the
-separated- and mixed-treatment samples
-(Table \@ref(tab:sample-prop-tbl)). 
-There is an inverse relationship between the absolute starch count in the solutions 
-and the proportional incorporation of starches in the biofilm samples, i.e., potato 
-had the lowest absolute count in solutions, but the highest proportional 
-incorporation, and vice versa for the mixed treatment.
-
-<!-- mixed treatments -->
-Wheat incorporation was most affected in the mixed-treatment samples, with only
-`r paste0(perc_tbl$total[2])` of the total available starches being incorporated into
-the sample, compared to
-`r paste0(perc_tbl$total[4])`
-in the separated wheat treatment.
-
-### Size ratios differ between solutions and samples
-
-<!-- size proportions -->
-<!-- pie charts -->
-
-<!-- overall size trends -->
-
-```{r size-trends}
-sol_size_prop <- sol_comb_out %>%
-  group_by(treatment, starch) %>%
-  #select(!total) %>%
-  mutate(across(where(is.numeric), function(x) x / total)) %>%
-  mutate(type = "solution")
-
-samp_size_prop <- samp_counts %>%
-  group_by(treatment, starch) %>%
-  #select(!total) %>%
-  mutate(across(where(is.numeric), function(x) x / total)) %>%
-  mutate(type = "sample")
-
-size_diff <- sol_size_prop %>%
-  full_join(samp_size_prop) %>%
-  pivot_longer(cols = c(s,m,l,total), names_to = "size", values_to = "count") %>%
-  pivot_wider(names_from = type, values_from = count) %>%
-  mutate(diff = (sample - solution) * 100) %>%
-  select(!c(solution, sample)) %>%
-  pivot_wider(names_from = size, values_from = diff) %>%
-  select(!total)
-```
-
-Overall, medium starch granules had a higher mean rate of incorporation 
-(`r scales::percent(mean(prop_wide$m, na.rm = T), 0.001)`) 
-than small 
-(`r scales::percent(mean(prop_wide$s, na.rm = T), 0.001)`) 
-and large 
-(`r scales::percent(mean(prop_wide$l, na.rm = T), 0.001)`) 
-starch granules across all treatments, while large potato starches had the lowest 
-rate of incorporation across all treatments.
-
-<!-- separated and mixed treatments -->
-The difference in incorporation between the size categories resulted in a change
-in size ratios between the original starch solutions and the extracted samples.
-Large potato granules (> 20 $\mu$m) were most affected, with a 
-`r paste0(abs(signif(size_diff$l[4], 3)), "%")` 
-decrease in relative abundance in the potato-only treatment, and a
-`r paste0(abs(signif(size_diff$l[1], 3)), "%")` 
-decrease in mixed treatments. Medium granules increased in relative abundance 
-across all samples, while small granules decreased in wheat treatments and 
-increased in potato treatments 
-(Figure \@ref(fig:ratio-plots)).
-
-<!-- Proportion and size ratios are the most important points to be made -->
-
-### Biofilm weight correlated positively with extracted starch counts
-
-```{r cor, echo=FALSE}
-# prepare cor output for reporting
-cor_r <- signif(count_cor$estimate, 3)
-cor_ci <- signif(count_cor$conf.int, 3)
-cor_p <- if(count_cor$p.value > 0.001){
-  paste("=", signif(count_cor$p.value, 3))
-} else {
-  paste("< 0.001")
-}
-
-cor2_r <- signif(starch_cor$estimate, 3)
-cor2_ci <- signif(starch_cor$conf.int, 3)
-cor2_p <- if(starch_cor$p.value > 0.001){
-  paste("=", signif(starch_cor$p.value, 3))
-} else {
-  paste("< 0.001")
-}
-# Convert r correlation result to text
-direct_cor <- ifelse(cor_r < 0, "negative", "positive")
-strength_cor <- if(cor_r >= 0.8){
-  "very strong"
-} else if(cor_r < 0.8 & cor_r >= 0.6){
-  "strong"
-} else if(cor_r < 0.6 & cor_r >= 0.4){
-  "moderate"
-} else if(cor_r < 0.4 & cor_r >= 0.2){
-  "weak"
-} else{
-  "very weak"
-}
-```
-
-Pearson's *r* suggests a 
-`r strength_cor` `r direct_cor` 
-correlation between the total weight of the biofilms and the total starch count 
-(standardised by z-score) extracted from the samples across treatments, 
-*r* = `r cor_r`, 
-90%CI[`r paste(cor_ci[1], cor_ci[2], sep = ", ")`],
-p `r cor_p`
-(Figure \@ref(fig:cor-plot)A).
-
-The same test was applied to total biofilm weight and starch count per mg 
-calculus (also standardised by z-score), resulting in a weak positive correlation, 
-*r* = `r cor2_r`, 
-90%CI[`r paste(cor2_ci[1], cor2_ci[2], sep = ", ")`],
-p `r cor2_p`
-(Figure \@ref(fig:cor-plot)B).
-
-# Discussion
-
 <!-- Big summary/important statement for the science -->
@@
 <!-- we still need to replicate results and do more science... but pretty cool, right? -->
-
-# Conclusions
-
-This preliminary study shows that a very small proportion of the input starch
-granules are retained in a dental calculus model. This and previous studies 
-have shown that calculus has a low capacity for retention of starch granules, 
-an effect that is compounded by diagenetic effects in archaeological remains, 
-resulting in low overall counts of extracted granules. 
-The proportion of starches consumed will in many cases be reflected 
-in the quantity
-of starches extracted from the dental calculus---i.e., the more starch granules
-entering the oral cavity, the more will be recovered from extraction---at 
-least in modern calculus samples unaffected by diagenesis and hydrolysis.
-Whether or not this also applies to archaeological samples remains to be tested.
-Additionally, we have
-shown that the size of granules will influence the likelihood of incorporation, 
-as large (>20 $\mu$m) starches have a decreased incorporation rate, medium 
-(10--20 $\mu$m)
-starches an increased rate, and small (<10 $\mu$m) granules remained somewhat
-constant. The size of calculus deposit also seems to influence the capacity of 
-granule incorporation; as the size of the deposit increases, so does the
-absolute count of incorporated granules.  
-While we have shown multiple factors that influence the likelihood 
-of incorporation, the process still appears to be somewhat stochastic. Further
-research is needed to make sense of the contributing factors, and to explore the
-mechanisms of intra-oral starch incorporation and retention in dental calculus.
-The oral biofilm model described in this study provides a method
-to explore the incorporation and extraction of dietary compounds from dental calculus 
-in a controlled laboratory setting, and unearth the potential biases associated 
-with dietary research conducted on archaeological dental calculus.
-
-# Open Science Statement {-}
-
-All scripts and data used in the analysis are available on OSF 
-(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch),
-following the format provided by the rrtools package [@R-rrtools].
-More detailed protocols are available on OSF (https://osf.io/akevs/)
-and protocols.io (https://www.protocols.io/workspaces/byoc).
-Additional tables and figures are available in the Supplementary Material
-(https://osf.io/ucxsv/).
-
-# Author Contributions {-}
-
-AGH acquired the funding for the analysis. BPB and AGH conceptualised and designed
-the study. BPB conducted the experiments and analysed the data.
-BPB wrote the manuscript. AGH reviewed and edited the manuscript.
-
-# Funding {-}
-
-This research has received funding from the European Research Council under the 
-European Union’s Horizon 2020 research and innovation program, grant agreement 
-number STG–677576 (“HARVEST”).
-
-# Conflict of Interest {-}
-
-The authors declare that the research was conducted in the absence of any
-commercial or financial relationships that could be construed as a potential
-conflict of interest.
-
-# Acknowledgements {-}
-
-<!-- If you're reading this, thank you for taking the time to read the source 
-documents. You're awesome! -->
-We would like to thank Dr. Stephanie Schnorr for help with the amylase activity protocol.
-We also thank everyone in the general vicinity
-of the lab for enduring the smell of bacterial accumulation.
-Since we did NOT make use of Sci-Hub to access articles stuck behind a paywall, 
-we will NOT acknowledge the use of Sci-Hub in this study. <!-- Oops, did I just acknowledge it twice..? -->
-
-# References {-}
-
-::: {#refs}
-:::
-
-\clearpage
-
-# Tables {-}
-
-```{r anova-tbl}
-knitr::kable(anova_tbl, 
-             col.names = c("Treatment", "Mean", "SD", "Min", "Max"),
-             caption = "Summary statistics for biofilm dry-weights (in mg) by treatment.")
-```
-
-```{r solution-count-tbl}
-cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
-options(knitr.kable.NA = "")
-knitr::kable(sol_tbl, col.names = cnames, 
-             caption = "Mean starch counts from solutions, including the
-             proportional makeup of the different sizes of granules.")
-```
-
-```{r sample-count-tbl}
-cnames <- c("Treatment", "Starch", "Small (%)", "SD", "Medium (%)", "SD", 
-            "Large (%)", "SD", "Total (%)", "SD")
-knitr::kable(samp_counts_tbl, col.names = cnames, 
-             caption = "Mean starch counts extracted from samples with standard 
-             deviation (SD), including the proportion of granule sizes of the total count.")
-```
-
-```{r sample-prop-tbl}
-cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
-knitr::kable(perc_tbl, col.names = cnames, 
-             caption = "The mean percentage of starches from the solutions that 
-              were incorporated into the samples.")
-```
-
-# Figures {-}
-
-```{r protocol-fig, fig.cap="Overview of experiment protocol including the plate setup.", out.width="70%"}
-knitr::include_graphics(here("analysis/figures/protocol_overview.png"))
-```
-
-```{r microscope-fig, fig.cap="Microscope image of a biofilm sample exposed to the potato starch solution.
-Potato granules can be seen within a bacterial community.
-Scale bar = 20 $\\mu$m.", out.width="50%"}
-knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
-```
-
-```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."}
-#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
-(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
-```
-
-```{r cor-plot, warning=FALSE, fig.cap="Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus."}
-#pl_cor
-pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
-#ggsave("frontiers/figures/cor-plot.png", dpi = 300)
-```
-
-<!--
-Figure 1: Overview of experiment protocol including the plate setup.
-
-Figure 2: Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\mu$m.
-
-Figure 3: Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.
-
-Figure 4: Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus.
--->

### 2022-05-17 analysis/paper/index.Rmd
push manuscript files to github

--- 2022-05-17 analysis/paper/paper.Rmd
@@
 ## Compile refs.bib using rbbt package ##
-cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
-cite_keys <- unlist(cite_keys, use.names = FALSE)
+cite_keys <- rbbt::bbt_detect_citations(list.files(pattern = ".Rmd"))
+#cite_keys <- unlist(cite_keys, use.names = FALSE)
 rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
@@
 
-Dental calculus has proven to contain a wealth of information on the
-dietary habits of past populations. These insights have, to a large
-extent, been obtained by the extraction and identification of starch
-granules contained within the mineralised dental plaque from a wide
-range of regions and time periods. The scope of previous studies have
-been limited to microfossil extraction and identification to reconstruct
-dietary preferences from the archaeological record, and few studies have
-attempted to address the biases of starch retention in dental calculus.
-Those that have considered this problem have been limited to \emph{in
-vivo} studies on modern humans and non-human primates. Here, we present
-a multispecies oral biofilm model, which allows experimental research on
-starch incorporation and retention to be conducted on \emph{in vitro}
-dental calculus in a controlled laboratory setting. The biofilms were
-exposed to treatment solutions with known quantities of dietary starches
-(wheat and potato) during the 25-day growth period. After this, the
-starch granules were extracted from the mature biofilm (by dissolution
-in EDTA), and counted. We show that the granule counts extracted from
-the model dental calculus represented a low proportion (ranging from
-0.06\% to 0.16\%) of the total number of granules exposed to the
-biofilms throughout the experiment. Additionally, we found that the
-ratios of granule sizes from the extracted starch granules differed from
-the original treatment solutions, with large granules (\textgreater20
-\(\mu\)m) consistently being under-represented. We also found a
-correlation between the absolute granule counts and dry-weight of the
-biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
-the concentration (count per mg) of granules and dry-weight (\emph{r} =
-0.3, 90\%CI{[}0.0618, 0.506{]}).\\
-Our results reinforce previous \emph{in vivo} studies suggesting that
-dental calculus presents a very small, and partly biased picture of the
-original dietary intake of starches, with an over-representation of
-plants producing granules smaller than 20 \(\mu\)m in size. The
-experimental model presented here is well-suited to address the need for
-further validation of methods and biases associated with dietary
-research on dental calculus.
+```{r child="abstract.Rmd"}
+
+```
 
@@
 
-<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
-<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
-Dental calculus has proven to contain a wealth of 
-dietary information in the form of plant microfossils 
-[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
-proteins [@hendyProteomicCalculus2018; @warinnerEvidenceMilk2014],
-and other organic residues [@buckleyDentalCalculus2014]. 
-This dietary information can be preserved within the mineralised dental plaque 
-over many millennia, providing a unique window into the food-related behaviours of 
-past populations 
-[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
-and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].
+```{r child="intro.Rmd"}
 
-<!-- more detailed background, comprehensible to scientists in related disciplines -->
-Until recently, only a few studies directly investigated the presence of plant
-microremains
-in the dental calculus of archaeological remains. The ability to extract phytoliths
-from the dental calculus of archaeological fauna to investigate diet was first 
-noted by Armitage [-@armitageExtractionIdentification1975], 
-and later by Middleton and Rovner [-@middletonOpalPhytoliths1994],
-and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
-extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
-In more recent years, the study of dental calculus has increased exponentially, 
-and the wealth of information contained within the mineralised matrix has largely
-been acknowledged. The use of dental calculus spans a wide
-variety of archaeological research areas, such as oral microbiome 
-characterisation (including pathogens) through the analysis of DNA and proteins 
-[@adlerSequencingAncient2013; @warinnerPathogensHost2014],
-microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
-@mickleburghNewInsights2012], 
-other organic residues and proteins from dietary compounds 
-[@buckleyDentalCalculus2014; @hendyProteomicCalculus2018], 
-and nicotine use [@eerkensDentalCalculus2018]. Especially the extraction 
-of starch granules has become a rich source of dietary
-information, as starch granules have proven to preserve well within dental calculus
-over a variety of geographical and temporal ranges 
-[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
-@jovanovicNeolithicCalculus2021].
-
-<!-- clearly stating the general problem being addressed by this particular study -->
-Despite this, our knowledge of dental calculus and the incorporation pathways of 
-the various markers is limited [@radiniFoodPathways2017], as is our 
-knowledge of information-loss caused by these pathways. Additionally, the methods
-we use to extract and analyse dental calculus, and make inferences on past diets
-represent another potential source of bias. Studies on both archaeological and
-modern individuals have explored these biases in more detail.
-Extraction methods were tested 
-by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
-decalcification using HCl or EDTA. 
-The authors found significantly more starches with the EDTA extraction method 
-than the HCl extraction method; however, as noted by the authors, comparisons 
-involving archaeological calculus are problematic due to variability between and 
-within individuals. 
-Studies conducted on modern humans [@leonardDentalCalculus2015] 
-and non-human primates [@powerChimpCalculus2015], 
-have explored how well microremains (phytoliths and starches)
-extracted from dental calculus represent the actual dietary intake.
-These studies are justifiably limited, 
-despite meticulous documentation and observation, due to unknown variables and 
-uncertainty involved in this kind of *in vivo* research. Dental calculus is a complex 
-oral biofilm with a multifactorial aetiology and variable formation rates both 
-within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
-contributing to
-the stochasticity of starch representation being observed in numerous studies.
-Additionally, the concentration of oral $\alpha$-amylase differs both between and
-within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
-causing different rates of hydrolysis of the starch granules present in the oral
-cavity. Add to this the effects of the many different methods
-of starch processing, as well as post-depositional processes that are still being 
-explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
-dental calculus to reconstruct diet is a highly unpredictable process.
-
-In this exploratory study, we use an oral biofilm model to investigate the 
-retention of starch granules within dental calculus in a controlled laboratory 
-setting, allowing us full control over dietary input. Our main questions concern
-the representation of granules extracted
-from the calculus compared to the actual intake. How much of the original diet is 
-incorporated into the calculus, and how much is recovered? 
-Is there differential loss of information from specific dietary markers that affects 
-the obtained dietary information, and how does this affect the representation of 
-diet from extracted microremains?  
-We find that, despite the absence of $\alpha$-amylase in
-the model, a limited proportion of the starch input is actually 
-retained in the calculus. We also observed a shift in the size ratios of individual 
-starch granules that are incorporated into the calculus, and that the number of 
-incorporated starch granules increases as the size of the calculus deposit 
-increases.
+```
 
@@
 
-## Biofilm formation
+```{r child="matmet.Rmd"}
 
-In this study we employ a multispecies oral biofilm model following a modified
-protocol from Sissons and colleagues [-@sissonsMultistationPlaque1991] and
-Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
-with whole saliva was grown on a substrate suspended in artificial saliva, and
-fed with sugar (sucrose). After several days of growth, the biofilm was exposed
-to starch solutions. Mineralisation of the biofilm was aided by exposure to a
-calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
-collected for further analysis. The setup comprises a polypropylene 
-24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which 
-is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins. The individual pegs 
-were the substrata on which the calculus grew. Using this system allowed for easy
-transfer of the growing biofilm between saliva, feeding solutions, 
-and mineral solutions.
-
-The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
-described by Sissons and colleagues [-@sissonsMultistationPlaque1991]. 
-It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
-5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
-5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl~2~, 
-5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationPlaque1991], 6.4 mmol/l NaHCO~3~ 
-[@shellisSyntheticSalivaCultural1978], 2.5 mg/l haemin. This is subsequently 
-adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120&deg;C, 
-1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
-and 1 mmol/l arginine [@sissonsMultistationPlaque1991].
-
-Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
-with no history of caries, who abstained from oral hygiene for 24 hours. No 
-food was consumed two hours prior to donation and no antibiotics were taken up 
-to six months prior to donation. 
-The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15% active chlorine) 
-nylon cloth to remove particulates.
-Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
-20% (v/v) glycerine for four hours at 36&deg;C, to allow attachment of the
-salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
-substrata were transferred to a new plate containing 1 ml/well AS and incubated 
-in a shaking incubator (Infors HT Ecotron) at 36&deg;C, 30 rpm. 
-The inoculation process was repeated on days 3 and 5. 
-AS was partially refreshed once per day and fully refreshed every three days, 
-throughout the experiment, by transferring the substrata to a new plate containing 
-stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
-5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
-(days 0, 3, and 5), where the samples only received one sucrose treatment after
-inoculation.
-
-Starch treatments were initiated on day 9 to avoid starch granule counts being 
-affected by $\alpha$-amylase hydrolysis from saliva inoculation days. 
-An $\alpha$-amylase (EC 3.2.1.1) activity
-assay was conducted to confirm that no amylase was present in the model before
-starch treatments started. Starch treatments replaced sucrose treatments, occurring twice per day
-for six minutes. The starch treatments involved transferring the substrata to a 
-new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
-concentrations (w/v) wheat and potato. All starch treatments were created in dH~2~O
-with 5% (w/v) sucrose. Before transferring biofilm samples to the starch treatment
-plate, the plates were agitated to keep the starches in suspension in the
-solutions. During treatments, the rpm was increased to 60 to facilitate contact
-between starch granules and biofilms.
-
-After 15 days, mineralisation was encouraged with a 
-calcium phosphate monofluorophosphate urea (CPMU) solution containing 
-20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
-and (0.04 g/l MgCl)
-[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationPlaque1991].
-The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
-daily, in a two-hour cycle. During the mineralisation period, starch treatments
-were reduced to once per day after the five CPMU treatments. This process was repeated 
-for 10 days until the end of the experiment on day 24 
-(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).
-
-All laboratory work was conducted in sterile conditions under a laminar flow hood 
-to prevent starch and bacterial contamination. Control samples that only received
-sucrose as a treatment were included to detect starch contamination from the
-environment or cross-contamination from other wells in the same plate.
-
-## Amylase activity detection
-
-An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
-saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14. 
-Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
-Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
-All samples and standard curves were run in triplicates on two separate plates.
-Positive control saliva samples were compared against a standard curve containing 
-H~2~O, while artificial saliva samples were compared against a standard curve
-containing stock AS (due to the colour of artificial saliva).
-Two photometric readings were conducted for each plate with a 540 nm filter on a 
-Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
-The protocol is a modified version of an Enzymatic Assay of $\alpha$-Amylase
-(<https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase>) [@bernfeldAmylase1955], which measures the amount of
-maltose released from starch by $\alpha$-amylase activity. Results are reported
-in units (U) per mL enzyme, where 1 U releases 1 $\mu$mole of maltose in 6 minutes.
-The detailed protocol can be found here: <https://www.protocols.io/view/amylase-activity-bw8jphun>.
-
-## Treatment solutions
-
-A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
-on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
-(Zeiss Axioscope A1). For wheat and mixed treatment samples, we counted three 
-slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts 
-were extrapolated to the total number of granules exposed to the samples over 16
-days of treatments (see Supplementary Material for more details). For potato
-treatment samples, the whole slide was counted.
-
-## Extraction method
-
-Extraction of starches from the calculus samples was performed by dissolving the
-calculus in 0.5 $\tiny{M}$ ethylenediaminetetraacetic acid (EDTA) 
-[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
-and vortexing for 3 days until the sample was completely dissolved. 
-Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
-When transferring the sample to the slide, the sample was homogenised using 
-the pipette to ensure that the counted transects were representative of the 
-whole slide. The count from the slide was extrapolated to the whole sample
-(see Supplementary Material for more detail).
-
-Both wheat and potato granules were divided into three size categories: 
-small (<10 $\mu$m), medium (10 -- 20 $\mu$m), and large (>20 $\mu$m).
-
-## Statistical analysis
-
-Statistical analysis was conducted in `r version$version.string`  [@Rbase] and
-the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
-here [@R-here], and patchwork [@R-patchwork].
-<!-- R packages used -->
-
-To see if biofilm growth was differently affected by starch treatments, a 
-one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
-as the grouping variable (GV) was conducted.
-To analyse granule counts and calculate size proportions, mean counts for each 
-treatment were taken across
-both experimental plates, resulting in a mean count for each
-granule size category within each treatment.
-
-Pearson's *r* was conducted on sample weight and total starch count, as well as sample
-weight and starch count per mg calculus. The total count for each sample within a
-treatment was standardised by z-score to account for the differences in magnitude
-between the potato and wheat counts.
-This was applied to total biofilm weight and starch count per mg 
-calculus (also z-score standardised) to account for differences in starch 
-concentration in the calculus [as per 
-@wesolowskiEvaluatingMicrofossilContent2010].
+```
 
@@
 
-All samples yielded sufficient biofilm growth and starch incorporation to be 
-included in the analysis (Figure \@ref(fig:microscope-fig)), resulting in a total of 48 biofilm samples (two plates of 24),
-45 of which were used for analysis (three samples were set aside for later
- analysis).
-Most control samples contained no starch granules, while some contained negligible
-quantities (see Supplementary Material).
+```{r child="results.Rmd"}
 
-## No amylase activity detected in the model
-
-```{r amylase-activity}
-# table of results reported in units amylase per mL enzyme (but let's be honest,
-  # ...it doesn't really matter what the unit is. No activity is no activity)
-cols <- c("1", "2", "3")  
-rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
-plt1_ph1_result <- rbind(sal1_ph1, bmm1_ph1)
-  plt1_ph1_result[plt1_ph1_result < 0] <- 0
-
-amylase_results <- rbind(sal1_ph1, sal1_ph2, sal2_ph1, sal2_ph2, bmm1_ph1,
-                         bmm1_ph2, bmm2_ph1, bmm2_ph2)
-amylase_results[amylase_results < 0] <- NA
-positive_amylase <- na.omit(amylase_results)
 ```
-
-No $\alpha$-amylase activity was detected in any of the artificial
-saliva samples from any of the days that were sampled. Only positive controls
-(saliva)
-contained amylase activity that could be detected in the assay, ranging from
-`r paste(signif(range(positive_amylase),3), collapse = " to ")` U/mL enzyme 
-(full results can be found in the Supplementary Material).
-The results are not comparable to other studies presenting $\alpha$-amylase activity
-levels in humans, as the unit definition may differ; 
-however, they are sufficient to show that there is no activity in the model.
-
-## Treatment type had minimal effect on biofilm growth
-
-A one-way ANOVA suggests that the type of starch used during the biofilm 
-growth period had a minimal effect on the growth of the biofilm 
-(expressed as total dry weight of the sample), F(`r paste(treat_df[1], treat_df[2], sep = ", ")`)
-= `r treat_f.stat`, p = `r treat_p`.
-A summary of sample weights is available in Table \@ref(tab:anova-tbl).
-
-## Starch counts
-
-<!-- starch counts in the solutions -->
-
-```{r solution-count-out}
-sol_tbl <- sol_comb_out %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
-
-sol_tbl[1,3] <- NA
-```
-
-<!-- starch counts in the samples -->
-
-```{r sample-count}
-samp_counts_per <- samp_counts %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), 
-                         ")")))
-
-# combine samp_counts_per with samp_sd_wide
-samp_counts_tbl <- samp_counts_per %>%
-  bind_cols(select(samp_sd_wide, c(s_sd, m_sd, l_sd, total_sd)), .name_repair = "unique") %>%
-  select(treatment, starch, s, s_sd, m, m_sd, l, l_sd, total, total_sd)
-
-samp_counts_tbl[1,3] <- NA
-```
-
-It was not possible to differentiate between potato and 
-wheat starches smaller than ca. 10 $\mu$m. These were counted as wheat,
-as we assumed that the majority of the small granules were wheat. We make this 
-assumption based on the counts of small starches in the wheat-only and potato-only
-solutions. Of the combined amount of small starches in these two solutions, 
-`r paste0(signif((sol_comb_out$s[5] / sum(sol_comb_out$s[4], sol_comb_out$s[5])) * 100, 3), "%")`
-are from wheat.
-<!-- count of small granules from wheat solution divided by the sum of small granules 
-from the potato and wheat solutions -->
-
-<!-- absolute solution counts compared to absolute sample counts. does exposure
-to higher quantities of starches mean more incorporated starches? -->
-The separate wheat and potato solutions were made with a 0.25% (w/v) starch
-concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
-starch, with a total concentration of 0.50% (w/v).
-The mixed treatment had the highest absolute count of starch granules in solution (mean = `r signif(sol_comb_row$total, 3)`),
-while the biofilms exposed to the wheat solution preserved the greatest number of
-granules 
-(mean = `r signif(subset(samp_counts, treatment == "wheat")$total, 3)`). 
-The potato treatment had the lowest absolute counts in both the solution 
-(`r signif(subset(sol_comb_out, treatment == "potato")$total, 3)`)
-and in the biofilm samples 
-(`r signif(subset(samp_counts, treatment == "potato")$total, 3)`) 
-(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
-
-### Proportion of available starches incorporated in samples
-
-```{r sample-perc}
-prop_wide <- prop %>%
-  ungroup() %>%
-  pivot_wider(id_cols = c(treatment, starch), 
-              names_from = size, 
-              values_from = prop) %>%
-  add_row(prop_comb_row, .after = 2)
-```
-
-The proportion of total starches from the solutions that were incorporated 
-into the samples ranged from 
-`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`,
-with potato granules being more readily incorporated than wheat in both the
-separated- and mixed-treatment samples
-(Table \@ref(tab:sample-prop-tbl)). 
-There is an inverse relationship between the absolute starch count in the solutions 
-and the proportional incorporation of starches in the biofilm samples, i.e., potato 
-had the lowest absolute count in solutions, but the highest proportional 
-incorporation, and vice versa for the mixed treatment.
-
-<!-- mixed treatments -->
-Wheat incorporation was most affected in the mixed-treatment samples, with only
-`r paste0(perc_tbl$total[2])` of the total available starches being incorporated into
-the sample, compared to
-`r paste0(perc_tbl$total[4])`
-in the separated wheat treatment.
-
-### Size ratios differ between solutions and samples
-
-<!-- size proportions -->
-<!-- pie charts -->
-
-<!-- overall size trends -->
-
-```{r size-trends}
-sol_size_prop <- sol_comb_out %>%
-  group_by(treatment, starch) %>%
-  #select(!total) %>%
-  mutate(across(where(is.numeric), function(x) x / total)) %>%
-  mutate(type = "solution")
-
-samp_size_prop <- samp_counts %>%
-  group_by(treatment, starch) %>%
-  #select(!total) %>%
-  mutate(across(where(is.numeric), function(x) x / total)) %>%
-  mutate(type = "sample")
-
-size_diff <- sol_size_prop %>%
-  full_join(samp_size_prop) %>%
-  pivot_longer(cols = c(s,m,l,total), names_to = "size", values_to = "count") %>%
-  pivot_wider(names_from = type, values_from = count) %>%
-  mutate(diff = (sample - solution) * 100) %>%
-  select(!c(solution, sample)) %>%
-  pivot_wider(names_from = size, values_from = diff) %>%
-  select(!total)
-```
-
-Overall, medium starch granules had a higher mean rate of incorporation 
-(`r scales::percent(mean(prop_wide$m, na.rm = T), 0.001)`) 
-than small 
-(`r scales::percent(mean(prop_wide$s, na.rm = T), 0.001)`) 
-and large 
-(`r scales::percent(mean(prop_wide$l, na.rm = T), 0.001)`) 
-starch granules across all treatments, while large potato starches had the lowest 
-rate of incorporation across all treatments.
-
-<!-- separated and mixed treatments -->
-The difference in incorporation between the size categories resulted in a change
-in size ratios between the original starch solutions and the extracted samples.
-Large potato granules (> 20 $\mu$m) were most affected, with a 
-`r paste0(abs(signif(size_diff$l[4], 3)), "%")` 
-decrease in relative abundance in the potato-only treatment, and a
-`r paste0(abs(signif(size_diff$l[1], 3)), "%")` 
-decrease in mixed treatments. Medium granules increased in relative abundance 
-across all samples, while small granules decreased in wheat treatments and 
-increased in potato treatments 
-(Figure \@ref(fig:ratio-plots)).
-
-<!-- Proportion and size ratios are the most important points to be made -->
-
-### Biofilm weight correlated positively with extracted starch counts
-
-```{r cor, echo=FALSE}
-# prepare cor output for reporting
-cor_r <- signif(count_cor$estimate, 3)
-cor_ci <- signif(count_cor$conf.int, 3)
-cor_p <- if(count_cor$p.value > 0.001){
-  paste("=", signif(count_cor$p.value, 3))
-} else {
-  paste("< 0.001")
-}
-
-cor2_r <- signif(starch_cor$estimate, 3)
-cor2_ci <- signif(starch_cor$conf.int, 3)
-cor2_p <- if(starch_cor$p.value > 0.001){
-  paste("=", signif(starch_cor$p.value, 3))
-} else {
-  paste("< 0.001")
-}
-# Convert r correlation result to text
-direct_cor <- ifelse(cor_r < 0, "negative", "positive")
-strength_cor <- if(cor_r >= 0.8){
-  "very strong"
-} else if(cor_r < 0.8 & cor_r >= 0.6){
-  "strong"
-} else if(cor_r < 0.6 & cor_r >= 0.4){
-  "moderate"
-} else if(cor_r < 0.4 & cor_r >= 0.2){
-  "weak"
-} else{
-  "very weak"
-}
-```
-
-Pearson's *r* suggests a 
-`r strength_cor` `r direct_cor` 
-correlation between the total weight of the biofilms and the total starch count 
-(standardised by z-score) extracted from the samples across treatments, 
-*r* = `r cor_r`, 
-90%CI[`r paste(cor_ci[1], cor_ci[2], sep = ", ")`],
-p `r cor_p`
-(Figure \@ref(fig:cor-plot)A).
-
-The same test was applied to total biofilm weight and starch count per mg 
-calculus (also standardised by z-score), resulting in a weak positive correlation, 
-*r* = `r cor2_r`, 
-90%CI[`r paste(cor2_ci[1], cor2_ci[2], sep = ", ")`],
-p `r cor2_p`
-(Figure \@ref(fig:cor-plot)B).
 
@@
 
-<!-- Big summary/important statement for the science -->
-Here, we have provided a method for exploring the incorporation of dietary 
-starches into the mineral matrix of a dental calculus biofilm model. Our results show
-that a very low proportion of the starches exposed to the biofilm during growth are
-retained in the mineral matrix, and that the size of the starch granules
-may affect the likelihood of incorporation. The proportions of starch granules 
-(of all sizes) present in the extracted samples were similar across all treatments
-(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`), 
-despite large differences in absolute granule counts between wheat 
-(mean = `r format(sol_comb_out$total[5], big.mark = ",")`) 
-and potato
-(mean = `r format(sol_comb_out$total[4], big.mark = ",")`)
-solutions.  
-The absolute counts, however, differed more visibly between treatments and was
-proportional with the total count of granules in the treatment solutions. Wheat 
-and mixed solutions had the highest absolute mean count of starch granules, and 
-also had the highest absolute mean count of starch granules extracted from the 
-dental calculus
-(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
-This suggests that the starches that are more frequently consumed will be present
-in higher quantities in the dental calculus, at least prior to inhumation and 
-degradation in the burial environment. 
-Despite the low proportion of granules recovered from the model calculus 
-(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`),
-the absolute counts were still substantially greater than counts recovered from 
-archaeological remains 
-[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010], which could in part be due to the 
-lack of oral amylase activity in our model. 
-<!-- Scientific findings as extension of previous work -->
-Previous research conducted on dental calculus from contemporary humans and non-human
-primates suggest a high level of stochasticity involved in the retention of 
-starch granules in dental calculus, and that starch granules extracted from dental 
-calculus are underrepresented with regard
-to actual starch intake, which is consistent with our findings (illustrated by high
-standard deviations and low proportional incorporation).
-Leonard and colleagues [-@leonardDentalCalculus2015] found individual
-calculus samples to be a poor predictor of diet in a population, as many of the 
-consumed plants were missing from some individual samples, but were present in others.   
-Power and colleagues [-@powerChimpCalculus2015] 
-presented similar findings in non-human primates, where phytoliths were more
-representative of individual diets than starch granules.
-The size bias is also consistent with the findings by Power and colleagues
-[-@powerChimpCalculus2015],
-who found that plants producing starches 10--20 $\mu$m in
-size were over-represented; however, the representation of granules larger than
-20 $\mu$m in their study is unclear.
+```{r child="discussion.Rmd"}
 
-We have also shown that the size of the starch granules influences the likelihood
-of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
-maximum length were underrepresented in the calculus samples compared to the original 
-starch solutions, an effect that was consistent across all three treatments. Medium
-granules (10--20 $\mu$m) were often over-represented
-(Table \@ref(tab:sample-prop-tbl), 
-and 
-Figure \@ref(fig:ratio-plots)).
-Large potato granules were most affected, potentially because of the greater
-size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
-granules generally only reach up to 35 $\mu$m 
-[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
-Granule morphology may also play a role. Large wheat granules 
-are lenticular and have a larger surface area compared 
-to volume, whereas large potato granules are ovoid and have a larger volume
-compared to surface area 
-[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
-Another potentially important factor 
-from our results is the size of the calculus deposit. We found a 
-`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
-starch granules (Figure \@ref(fig:cor-plot)A);
-a result that contradicts findings from archaeological contexts
-[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeography2014].
-When the concentration of starch granules
-per mg calculus is considered, the correlation is weaker, but still present
-(Figure \@ref(fig:cor-plot)B).
-While the larger deposits contain a higher absolute count, our findings also suggest
-that they contain a slightly higher concentration of starches. This may also explain
-the lower mean retention of starch granules in mixed treatments compared to
-wheat treatments.
-Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
-granule counts, wheat treatment samples had the highest mean starch retention.
-Further research is needed to determine why this differs from previous archaeological
-findings.
-
-<!-- How are starch granules trapped in dental calculus? Does anyone actually know? 
--->
-The mechanism by which starch granules are incorporated into plaque and calculus 
-remains largely unknown, and few studies have directly investigated potential
-mechanisms. We know that a proportion of the starch granules entering 
-the mouth can become trapped in the plaque/calculus, and can be recovered from
-archaeological samples of considerable age 
-[@henryNeanderthalCalculus2014; @buckleyDentalCalculus2014; @wuDietEarliest2021]. 
-Studies have also shown that not all starch granules come from a dietary source. 
-Other pathways include cross-contamination from plant interactions in soil, such 
-as palm phytoliths adhering to the skin of sweet potatoes 
-[@trompDietaryNondietary2015], 
-or accidental ingestion not related to food consumption 
-[@radiniFoodPathways2017; @radiniMedievalWomenEarly2019].  
-When starch granules enter the mouth, whether through ingestion of food or accidental
-intake, they immediately encounter multiple obstacles. It is likely
-that the bulk of starch granules are swallowed along with the food, and are 
-only briefly present in the oral cavity. Other granules that are broken off
-during mastication may be retained in the dentition through attachment to 
-tooth surfaces (including plaque and dental calculus) and mucous membranes 
-[@kashketFoodRetention1991; @doddsCarbohydrateRetention1988].
-Bacteria also have the ability to adhere to starch granules 
-[@toppingResistantStarch2003],
-which would allow starches to attach to bacterial communities within the biofilm.
-These granules are then
-susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
-by $\alpha$-amylase [@kashketFoodParticles1996].
-The susceptibility of granules to hydrolysis depends on the crystallinity and size
-of the starch granule, as well as the mode of processing. Smaller and pre-processed
-(e.g., cooked) starch granules are more susceptible to enzymatic degradation, 
-while dehydrated starches will have a reduced susceptibility
-[@lingstromStarchyFood1994; @francoStarchDegradation1992; @haslamDecompositionStarch2004; 
-@bjorckStarchProcessing1984; @henryCookingStarch2009].
-Cracks on the surface of the dental calculus, as well as unmineralised islands 
-and channels may also be able to contain starch granules 
-[@powerSEMCalculus2014; @tanCalculusUltrastructure2004; @charlierSEMCalculus2010].
-Starch granules that are trapped in these pockets are (at least to some extent) 
-protected from aforementioned clearance mechanisms, especially once
-a new layer of plaque has covered the surface of the plaque/calculus. 
-The size bias against large granules (>20 $\mu$m) from both wheat and potato 
-(Table \@ref(tab:sample-prop-tbl)) may give further credence to
-this incorporation pathway, as the smaller starch granules have an advantage over
-larger granules, and can be stored in larger quantities.
-This was also suggested by Power and colleagues [-@powerSEMCalculus2014], who
-observed clusters of starches within dental calculus, rather than an even 
-distribution across the surface of the dental calculus.
-Granules trapped 
-in plaque/calculus may still be susceptible to hydrolysis, as $\alpha$-amylase has
-the ability to bind to both tooth enamel and bacteria within a biofilm and retain
-a portion of its hydrolytic activity
-[@scannapiecoSalivaryAmylase1993; @nikitkovaStarchBiofilms2013; @tanBacterialViability2004; @tanCalculusUltrastructure2004].
-After the death of an individual, starches within dental calculus are susceptible 
-to further degradation by post-depositional processes, depending on burial environment 
-(pH, temperature, moisture content, microorganisms)
-[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004; @francoStarchDegradation1992].
-Future study should explore how burial affects the recovery of starch from the
-biofilm model. <!-- I may get to it eventually, but it would be great if someone else could look at this! -->
-
-<!-- in summary: we know very little about the mechanism behind how starches 
-become trapped in dental calculus, but it would be nice to find out more. -->
-
-<!-- limitations of the platform show which questions would be unanswered -->
-The absence of $\alpha$-amylase in the model is a limitation of this study, as 
-the total granule counts were not subject to hydrolysis. This would likely have 
-reduced and affected the size ratios, as smaller starches may be more 
-susceptible to hydrolysis 
-[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. The absence may
-also affect biofilm growth due to the lack of amylase-bacterium interactions
-[@nikitkovaStarchBiofilms2013]. Conversely, the model may benefit from the
-absence of $\alpha$-amylase, because it can allow us to directly explore
-its effect on starch counts in future experiments, where $\alpha$-amylase can be 
-added to the model in concentrations similar to those found in the oral cavity
-[@scannapiecoSalivaryAmylase1993]. 
-We are able to show how absolute counts in the treatments cause a difference in
-incorporation. However, this
-was merely a side-effect of the difference in the number of granules in potato and 
-wheat solutions of the same concentration (w/v). Further research should test 
-multiple differing concentrations of the same starch type.
-The use of EDTA may also have affected counts. While previous studies have shown 
-negligible morphological changes caused by exposure to EDTA 
-[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
-these studies have not considered changes to separate size
-categories within starch types, and whether shifts in size ratios occur due to
-exposure to the pre-treatment chemicals.
-The total number of granules on a slide often exceeded a number that 
-was feasible to count in a reasonable time period, so we calculated the total 
-counts by extrapolating from three slide transects. 
-Thus, we reasonably assume that the three transects are a good representation 
-of the entire slide, and that the distribution of all granules on the slide is 
-relatively homogeneous.   
-Finally, we only used native starches in the experimental procedure and the results 
-will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
-Based on the comparatively low counts obtained by Leonard
-and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
-may have a substantial effect on starch granule retention in the oral cavity.
-
-<!-- How this engineering will facilitate future research -->
-
-While we are unable to sufficiently address the 
-mechanism(s) of starch incorporation with the data obtained in this study, 
-the dental calculus model presented here is uniquely suited to explore 
-these questions and may improve interpretations of dietary practices in past 
-populations. Further analyses using this model can address the call for more 
-baseline testing of biases associated with dietary research conducted on dental calculus 
-[@lemoyneCalculusPretreatments2021].
-Our high-throughput experimental setup allows us a
-higher degree of control over the factors that influence starch incorporation and
-retention, such as dietary intake, differential survivability of starches, 
-and inter- and intra-individual variation in plaque accumulation and mineralisation.
-The latter is especially difficult to control *in vivo* as it is influenced by 
-numerous factors including genetics, diet, salivary flow, and tooth position and
-morphology 
-[@jepsenCalculusRemoval2011; @simon-soroOralGeography2013; @proctorSpatialGradient2018; @fagernasMicrobialBiogeography2021; @haffajeeBiofilmPosition2009], 
-as well as evolutionary differences [@yatesOralMicrobiome2021].
-It can also facilitate training of students and researchers on methods of 
-dental calculus analysis, such as starch and phytolith extraction and 
-identification, where it can replace the use of finite archaeological resources.
-
-<!-- we still need to replicate results and do more science... but pretty cool, right? -->
+```
 
@@
 
-This preliminary study shows that a very small proportion of the input starch
-granules are retained in a dental calculus model. This and previous studies 
-have shown that calculus has a low capacity for retention of starch granules, 
-an effect that is compounded by diagenetic effects in archaeological remains, 
-resulting in low overall counts of extracted granules. 
-The proportion of starches consumed will in many cases be reflected 
-in the quantity
-of starches extracted from the dental calculus---i.e., the more starch granules
-entering the oral cavity, the more will be recovered from extraction---at 
-least in modern calculus samples unaffected by diagenesis and hydrolysis.
-Whether or not this also applies to archaeological samples remains to be tested.
-Additionally, we have
-shown that the size of granules will influence the likelihood of incorporation, 
-as large (>20 $\mu$m) starches have a decreased incorporation rate, medium 
-(10--20 $\mu$m)
-starches an increased rate, and small (<10 $\mu$m) granules remained somewhat
-constant. The size of calculus deposit also seems to influence the capacity of 
-granule incorporation; as the size of the deposit increases, so does the
-absolute count of incorporated granules.  
-While we have shown multiple factors that influence the likelihood 
-of incorporation, the process still appears to be somewhat stochastic. Further
-research is needed to make sense of the contributing factors, and to explore the
-mechanisms of intra-oral starch incorporation and retention in dental calculus.
-The oral biofilm model described in this study provides a method
-to explore the incorporation and extraction of dietary compounds from dental calculus 
-in a controlled laboratory setting, and unearth the potential biases associated 
-with dietary research conducted on archaeological dental calculus.
+```{r child="conclusion.Rmd"}
+
+```
 

### 2022-05-17 analysis/paper/intro.Rmd
further development of manuscript
push manuscript files to github

@@
-# Introduction
-
 <!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
 <!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
-Dental calculus, mineralised dental plaque, has proven to contain a wealth of 
+Dental calculus has proven to contain a wealth of 
 dietary information in the form of plant microfossils 
 [@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
-proteins [@hendyProteomicCalculus2018; @warinnerDirectEvidenceMilk2014],
-and other organic residues [@buckleyDentalCalculusCooking2014]. 
-It has also proven to be able to retain and preserve dietary information for millenea,
-providing detailed dietary information on past populations 
+proteins [@hendyProteomicCalculus2018; @warinnerEvidenceMilk2014],
+and other organic residues [@buckleyDentalCalculus2014]. 
+This dietary information can be preserved within the mineralised dental plaque 
+over many millennia, providing a unique window into the food-related behaviours of 
+past populations 
 [@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
-and species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].
+and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].
 
 <!-- more detailed background, comprehensible to scientists in related disciplines -->
-
-Until recently, few studies directly investigated the presence of plant microremains
-from the dental calculus of archaeological remains. The ability to extract phytoliths
+Until recently, only a few studies directly investigated the presence of plant
+microremains
+in the dental calculus of archaeological remains. The ability to extract phytoliths
 from the dental calculus of archaeological fauna to investigate diet was first 
-noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
-and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
-and Fox and colleagues [-@foxPhytolithCalculus1996]. The ability to detect coca
-consumption by looking at human dental calculus was explored by Klepinger and Kuhn
-[-@klepingerCalculusCoca1977], and starches and phytoliths were extracted from 
-human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].
-
+noted by Armitage [-@armitageExtractionIdentification1975], 
+and later by Middleton and Rovner [-@middletonOpalPhytoliths1994],
+and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
+extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
 In more recent years, the study of dental calculus has increased exponentially, 
-and the wealth of information contained within the mineralised matrix has been, 
-to a larger extent, acknowledged. The applications of dental calculus span a wide
+and the wealth of information contained within the mineralised matrix has largely
+been acknowledged. The use of dental calculus spans a wide
 variety of archaeological research areas, such as oral microbiome 
 characterisation (including pathogens) through the analysis of DNA and proteins 
-[@adlerSequencingAncientCalcified2013; @warinnerPathogensHostImmunity2014],
+[@adlerSequencingAncient2013; @warinnerPathogensHost2014],
 microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
-@mickleburghNewInsightsConsumption2012], 
+@mickleburghNewInsights2012], 
 other organic residues and proteins from dietary compounds 
-[@buckleyDentalCalculusCooking2014; @hendyProteomicCalculus2018], 
-and nicotine-use [@eerkensDentalCalculusSource2018]. Especially the extraction 
+[@buckleyDentalCalculus2014; @hendyProteomicCalculus2018], 
+and nicotine use [@eerkensDentalCalculus2018]. Especially the extraction 
 of starch granules has become a rich source of dietary
@@
 
-<!-- starches from dental calculus -->
+<!-- clearly stating the general problem being addressed by this particular study -->
 Despite this, our knowledge of dental calculus and the incorporation pathways of 
-the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
-knowledge of information-loss caused by these pathways and the methods we use to 
-extract and analyse dental calculus and make inferences on past diets. 
+the various markers is limited [@radiniFoodPathways2017], as is our 
+knowledge of information-loss caused by these pathways. Additionally, the methods
+we use to extract and analyse dental calculus, and make inferences on past diets
+represent another potential source of bias. Studies on both archaeological and
+modern individuals have explored these biases in more detail.
 Extraction methods were tested 
@@
 involving archaeological calculus are problematic due to variability between and 
-within individuals. Other studies have been conducted on modern populations 
-[@leonardDentalCalculus2015] and primates [@powerChimpCalculus2015], both with
-well-documented dietary intake, to reconcile our interpretations on the recovery 
-of microremains (both phytoliths and starches) from dental calculus and how well
-they represent actual dietary intake. These studies are justifiably limited, 
+within individuals. 
+Studies conducted on modern humans [@leonardDentalCalculus2015] 
+and non-human primates [@powerChimpCalculus2015], 
+have explored how well microremains (phytoliths and starches)
+extracted from dental calculus represent the actual dietary intake.
+These studies are justifiably limited, 
 despite meticulous documentation and observation, due to unknown variables and 
-uncertainty involved when studying living organisms. Dental calculus is a complex 
-oral biofilm with a
-multifactorial aetiology and variable formation rates both within and between
-individuals [@jepsenCalculusRemoval2011].
-Additionally, the effects of the many different methods
-of starch processing, as well as post-depositional processes, are still being 
-explored [@graneroStarchTaphonomy2020].
+uncertainty involved in this kind of *in vivo* research. Dental calculus is a complex 
+oral biofilm with a multifactorial aetiology and variable formation rates both 
+within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
+contributing to
+the stochasticity of starch representation being observed in numerous studies.
+Additionally, the concentration of oral $\alpha$-amylase differs both between and
+within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
+causing different rates of hydrolysis of the starch granules present in the oral
+cavity. Add to this the effects of the many different methods
+of starch processing, as well as post-depositional processes that are still being 
+explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
+dental calculus to reconstruct diet is a highly unpredictable process.
 
-<!-- clearly stating the general problem being addressed by this particular study -->
-<!-- Dental calculus research has its limitations, all of which I will solve with this
-single method. -->
 In this exploratory study, we use an oral biofilm model to investigate the 
-retention of starch granules within dental calculus, in a controlled laboratory 
-setting. Our main questions concern the representation of granules extracted
+retention of starch granules within dental calculus in a controlled laboratory 
+setting, allowing us full control over dietary input. Our main questions concern
+the representation of granules extracted
 from the calculus compared to the actual intake. How much of the original diet is 
@@
 Is there differential loss of information from specific dietary markers that affects 
-the obtained dietary information, and how does this affect the representation of diet from
-extracted microremains? We find that a limited proportion of the starch input is actually 
-retained in the calculus, as well as a shift in the size ratios of individual starch granules
-that are incoporated into the calculus. We also show that the number of starch
-granules that are incorporated increases as the size of the calculus deposit 
+the obtained dietary information, and how does this affect the representation of 
+diet from extracted microremains?  
+We find that, despite the absence of $\alpha$-amylase in
+the model, a limited proportion of the starch input is actually 
+retained in the calculus. We also observed a shift in the size ratios of individual 
+starch granules that are incorporated into the calculus, and that the number of 
+incorporated starch granules increases as the size of the calculus deposit 
 increases.

### 2022-05-17 analysis/paper/matmet.Rmd
further development of manuscript
push manuscript files to github

@@
-# Materials and Methods
-
-```{r protocol-fig, fig.cap="Overview of experiment protocol"}
-knitr::include_graphics(here("analysis/figures", "ExpL_protocol.png"))
-```
-
-<!-- remove sampling from protocol picture -->
-
 ## Biofilm formation
@@
 In this study we employ a multispecies oral biofilm model following a modified
-protocol from Sissons and colleagues [-@sissonsMultistationDentalPlaque1991] and
-Shellis [-@shellisSyntheticSalivaCultural1978]. The setup comprises a polypropylene 
-24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which 
-is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins.
+protocol from Sissons and colleagues [-@sissonsMultistationPlaque1991] and
+Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
+with whole saliva was grown on a substrate suspended in artificial saliva, and
+fed with sugar (sucrose). After several days of growth, the biofilm was exposed
+to starch solutions. Mineralisation of the biofilm was aided by exposure to a
+calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
+collected for further analysis. The setup comprises a polypropylene 
+24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which 
+is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins. The individual pegs 
+were the substrata on which the calculus grew. Using this system allowed for easy
+transfer of the growing biofilm between saliva, feeding solutions, 
+and mineral solutions.
 
 The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
-described by Sissons and colleagues [-@sissonsMultistationDentalPlaque1991]. 
+described by Sissons and colleagues [-@sissonsMultistationPlaque1991]. 
 It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
@@
 1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
-and 1 mmol/l arginine [@sissonsMultistationDentalPlaque1991].
+and 1 mmol/l arginine [@sissonsMultistationPlaque1991].
 
-<!-- add to protocols.io -->
-Fresh whole saliva (WS) for inoculation was provided by a 31-year-old male donor 
-with no history of caries, who abstained from oral hygiene for 24 hours, and no 
-food was consumed two hours prior to donation. No antibiotics were taken up to six 
-months prior to donation. 
-The saliva was filtered through a sterilised (with bleach) nylon cloth to remove
-particulates.
+Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
+with no history of caries, who abstained from oral hygiene for 24 hours. No 
+food was consumed two hours prior to donation and no antibiotics were taken up 
+to six months prior to donation. 
+The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15% active chlorine) 
+nylon cloth to remove particulates.
 Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
-20% glycerine for four hours at 36&deg;C, to allow attachment of the
+20% (v/v) glycerine for four hours at 36&deg;C, to allow attachment of the
 salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
 substrata were transferred to a new plate containing 1 ml/well AS and incubated 
-at 36&deg;C, 30 rpm. The inoculation process was repeated on days 3 and 5. 
+in a shaking incubator (Infors HT Ecotron) at 36&deg;C, 30 rpm. 
+The inoculation process was repeated on days 3 and 5. 
 AS was partially refreshed once per day and fully refreshed every three days, 
 throughout the experiment, by transferring the substrata to a new plate containing 
-AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
+stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
 5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
@@
 Starch treatments were initiated on day 9 to avoid starch granule counts being 
-affected by $\alpha$-amylase hydrolysis. An $\alpha$-amylase (EC 3.2.1.1) activity
-assay was conducted to confirm that no amylase was present in the system (see 
-Bartholdy et al. ???? <!-- ## [bartholdy] ## --> for more details). Starch treatments replaced sucrose treatments, occurring twice per day
+affected by $\alpha$-amylase hydrolysis from saliva inoculation days. 
+An $\alpha$-amylase (EC 3.2.1.1) activity
+assay was conducted to confirm that no amylase was present in the model before
+starch treatments started. Starch treatments replaced sucrose treatments, occurring twice per day
 for six minutes. The starch treatments involved transferring the substrata to a 
 new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
-concentrations (w/v) wheat and potato. All starch solutions were created in a 5%
-(w/v) sucrose solution. Before transferring biofilm samples to the starch treatments, 
-the starch plates were agitated to keep the starches in suspension in the solutions, 
-and during treatments, the rpm was increased to 60.
+concentrations (w/v) wheat and potato. All starch treatments were created in dH~2~O
+with 5% (w/v) sucrose. Before transferring biofilm samples to the starch treatment
+plate, the plates were agitated to keep the starches in suspension in the
+solutions. During treatments, the rpm was increased to 60 to facilitate contact
+between starch granules and biofilms.
 
@@
 calcium phosphate monofluorophosphate urea (CPMU) solution containing 
-20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea
-[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991], 
-and (0.04 g/l MgCl). The substrata were submerged in 1 ml/well CPMU five times daily, 
-every two hours, for six minutes. During the mineralisation period, starch treatments
-were reduced to once per day after the five CPMU treatments. This cycle was repeated 
+20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
+and (0.04 g/l MgCl)
+[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationPlaque1991].
+The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
+daily, in a two-hour cycle. During the mineralisation period, starch treatments
+were reduced to once per day after the five CPMU treatments. This process was repeated 
 for 10 days until the end of the experiment on day 24 
-(see \@ref(fig:protocol-fig) for an overview of protocol). More detailed protocols
-are available on protocols.io<!--insert DOI-->.
+(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).
 
 All laboratory work was conducted in sterile conditions under a laminar flow hood 
-to prevent starch and bacterial contamination. Control samples were included to 
-detect starch contamination.
+to prevent starch and bacterial contamination. Control samples that only received
+sucrose as a treatment were included to detect starch contamination from the
+environment or cross-contamination from other wells in the same plate.
 
-## Starch counts
+## Amylase activity detection
 
-Starch granule counts were divided into three size categories: small (<10 $\mu$m),
-medium (10 -- 20 $\mu$m), and large (>20 $\mu$m). Counts for small wheat and
-potato starches were combined for the mixed treatment samples, as it is not 
-possible to distinguish the small starch granules between the two species.
+An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
+saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14. 
+Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
+Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
+All samples and standard curves were run in triplicates on two separate plates.
+Positive control saliva samples were compared against a standard curve containing 
+H~2~O, while artificial saliva samples were compared against a standard curve
+containing stock AS (due to the colour of artificial saliva).
+Two photometric readings were conducted for each plate with a 540 nm filter on a 
+Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
+The protocol is a modified version of an Enzymatic Assay of $\alpha$-Amylase
+(<https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase>) [@bernfeldAmylase1955], which measures the amount of
+maltose released from starch by $\alpha$-amylase activity. Results are reported
+in units (U) per mL enzyme, where 1 U releases 1 $\mu$mole of maltose in 6 minutes.
+The detailed protocol can be found here: <https://www.protocols.io/view/amylase-activity-bw8jphun>.
 
-### Treatment solutions
+## Treatment solutions
 
@@
 on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
-(Zeiss Axioscope A1). Counting starches on a full slide was not feasible, so three slide transects were 
-counted (at 1/3, 1/2, and 2/3 of the slide), and the sample counts were extrapolated 
-(see Supplementary material for more details).
+(Zeiss Axioscope A1). For wheat and mixed treatment samples, we counted three 
+slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts 
+were extrapolated to the total number of granules exposed to the samples over 16
+days of treatments (see Supplementary Material for more details). For potato
+treatment samples, the whole slide was counted.
 
-### Extraction from samples
+## Extraction method
 
 Extraction of starches from the calculus samples was performed by dissolving the
-calculus in ethylenediaminetetraacetic acid (EDTA), as it has been shown to
-have little to no effect on starch granules (especially native granules)
+calculus in 0.5 $\tiny{M}$ ethylenediaminetetraacetic acid (EDTA) 
 [@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
@@
 Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
-When dispersing the liquid (EDTA) sample on the slide, the sample was homogenised using the pipette. 
-It was not possible to count all wheat starches
-on a slide, so wheat and mix counts were extrapolated as described for the treatment
-solutions.
+When transferring the sample to the slide, the sample was homogenised using 
+the pipette to ensure that the counted transects were representative of the 
+whole slide. The count from the slide was extrapolated to the whole sample
+(see Supplementary Material for more detail).
+
+Both wheat and potato granules were divided into three size categories: 
+small (<10 $\mu$m), medium (10 -- 20 $\mu$m), and large (>20 $\mu$m).
 
@@
 
-Statistical analysis was conducted in R versions 4.0.5 ("Shake and Throw") 
-and 4.0.3 ("Bunny-Wunnies Freak Out") [@R-base].
+Statistical analysis was conducted in `r version$version.string`  [@Rbase] and
+the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
+here [@R-here], and patchwork [@R-patchwork].
 <!-- R packages used -->
-A one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
-as the grouping variable (GV) was conducted to explore the effect of the different 
-starch treatments on biofilm growth.
 
-To calculate size proportions, mean counts for each treatment were taken across 
-both experimental plates for each treatment, resulting in a mean count for each
+To see if biofilm growth was differently affected by starch treatments, a 
+one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
+as the grouping variable (GV) was conducted.
+To analyse granule counts and calculate size proportions, mean counts for each 
+treatment were taken across
+both experimental plates, resulting in a mean count for each
 granule size category within each treatment.
 
-Pearson's r was conducted on sample weight and total starch count, as well as sample
+Pearson's *r* was conducted on sample weight and total starch count, as well as sample
 weight and starch count per mg calculus. The total count for each sample within a
@@
 between the potato and wheat counts.
-
-Scripts and data will be made available on osf.io upon publication.
-Until then, they can be found on the first author's GitHub repository
-(https://github.com/bbartholdy/byocStarch).
-
-<!-- script and data availability: provide link (or doi?) to osf.io repository -->
+This was applied to total biofilm weight and starch count per mg 
+calculus (also z-score standardised) to account for differences in starch 
+concentration in the calculus [as per 
+@wesolowskiEvaluatingMicrofossilContent2010].

### 2022-05-17 analysis/paper/results.Rmd
push manuscript files to github

@@
-# Results
-
 All samples yielded sufficient biofilm growth and starch incorporation to be 
-included in the analysis, resulting in a total of 48 biofilm samples (two plates of 24),
-45 of which were used for analysis (three samples were set aside for separate analysis).
+included in the analysis (Figure \@ref(fig:microscope-fig)), resulting in a total of 48 biofilm samples (two plates of 24),
+45 of which were used for analysis (three samples were set aside for later
+ analysis).
 Most control samples contained no starch granules, while some contained negligible
-quantities (supp. mat).
+quantities (see Supplementary Material).
 
@@
 
-<!-- insert amylase results -->
+```{r amylase-activity}
+# table of results reported in units amylase per mL enzyme (but let's be honest,
+  # ...it doesn't really matter what the unit is. No activity is no activity)
+cols <- c("1", "2", "3")  
+rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
+plt1_ph1_result <- rbind(sal1_ph1, bmm1_ph1)
+  plt1_ph1_result[plt1_ph1_result < 0] <- 0
 
-## Treatment type had no effect on biofilm growth
+amylase_results <- rbind(sal1_ph1, sal1_ph2, sal2_ph1, sal2_ph2, bmm1_ph1,
+                         bmm1_ph2, bmm2_ph1, bmm2_ph2)
+amylase_results[amylase_results < 0] <- NA
+positive_amylase <- na.omit(amylase_results)
+```
 
-```{r anova-tbl}
-knitr::kable(anova_tbl, 
-             col.names = c("Treatment", "Mean", "SD", "Min", "Max"),
-             caption = "Summary statistics for biofilm dry-weights (in mg) by treatment.")
-```
+No $\alpha$-amylase activity was detected in any of the artificial
+saliva samples from any of the days that were sampled. Only positive controls
+(saliva)
+contained amylase activity that could be detected in the assay, ranging from
+`r paste(signif(range(positive_amylase),3), collapse = " to ")` U/mL enzyme 
+(full results can be found in the Supplementary Material).
+The results are not comparable to other studies presenting $\alpha$-amylase activity
+levels in humans, as the unit definition may differ; 
+however, they are sufficient to show that there is no activity in the model.
+
+## Treatment type had minimal effect on biofilm growth
 
@@
 ```{r solution-count-out}
-count_sol_out <- count_sol %>%
+sol_tbl <- sol_comb_out %>%
   mutate(across(c(s, m, l, total), 
@@
 
-### adding the 'both' row to the data frame ###
-count_sol_comb_out <- count_sol_comb %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
-
-count_sol_out <- count_sol %>%
-  mutate(across(c(s, m, l, total), 
-                function(x) 
-                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")"))) %>%
-  ungroup() %>%
-  add_row(tibble_row(count_sol_comb_out), .before = 3)
-```
-
-```{r solution-count-tbl}
-cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
-knitr::kable(count_sol_out, col.names = cnames, 
-             caption = "Mean starch counts from solutions, including the
-             proportional makeup of the different sizes of granules.")
+sol_tbl[1,3] <- NA
 ```
@@
 ```{r sample-count}
-count_sample <- count_sample %>%
-  ungroup() %>%
-  add_row(count_sample_comb, .after = 2)
-
-# convert columns to contain mean count and percentage of total count
-count_samp_per <- count_sample %>%
+samp_counts_per <- samp_counts %>%
   mutate(across(c(s, m, l, total), 
                 function(x) 
-                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), ")")))
+                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), 
+                         ")")))
 
-# count_samp_comb_per <- count_sample_comb %>%
-#   mutate(across(c(s, m, l, total), 
-#                 function(x) 
-#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
+# combine samp_counts_per with samp_sd_wide
+samp_counts_tbl <- samp_counts_per %>%
+  bind_cols(select(samp_sd_wide, c(s_sd, m_sd, l_sd, total_sd)), .name_repair = "unique") %>%
+  select(treatment, starch, s, s_sd, m, m_sd, l, l_sd, total, total_sd)
 
-# count_samp_comb_out <- tibble(count_samp_comb_per, count_sample_comb_sd, .name_repair = "unique") %>%
-#   select(1,2,3,7,4,8,5,9,6,10) #%>%
-  #add_row(tibble_row(count_sample_comb_out), .before = 3)
-
-count_samp_out <- tibble(count_samp_per, count_sample_sd, .name_repair = "unique") %>%
-  select(1,2,3,7,4,8,5,9,6,10) #%>%
-  #add_row(tibble_row(count_sample_comb_out), .before = 3)
-```
-
-```{r sample-count-tbl}
-cnames <- c("Treatment", "Starch", "Small (%)", "SD", "Medium (%)", "SD", 
-            "Large (%)", "SD", "Total (%)", "SD")
-knitr::kable(count_samp_out, col.names = cnames, 
-             caption = "Mean starch counts extracted from samples with standard 
-             deviation (SD), including the proportion of granule sizes of the total count.")
+samp_counts_tbl[1,3] <- NA
 ```
@@
 It was not possible to differentiate between potato and 
-wheat starches smaller than ca. 10 $\mu$m,
-small potato starches were counted as wheat starches in the mixed-treatment samples. This is reasonable given that the small potato granules make up an insignificant proportion of the 
-total count of small granules within mixed-treatment solutions which are predominantly
-wheat granules
-(`r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`).
+wheat starches smaller than ca. 10 $\mu$m. These were counted as wheat,
+as we assumed that the majority of the small granules were wheat. We make this 
+assumption based on the counts of small starches in the wheat-only and potato-only
+solutions. Of the combined amount of small starches in these two solutions, 
+`r paste0(signif((sol_comb_out$s[5] / sum(sol_comb_out$s[4], sol_comb_out$s[5])) * 100, 3), "%")`
+are from wheat.
 <!-- count of small granules from wheat solution divided by the sum of small granules 
@@
 starch, with a total concentration of 0.50% (w/v).
-The mixed-treatment had the highest absolute count of starch granules in solution,
+The mixed treatment had the highest absolute count of starch granules in solution (mean = `r signif(sol_comb_row$total, 3)`),
 while the biofilms exposed to the wheat solution preserved the greatest number of
-granules. The potato treatment had the lowest absolute counts in both the solution 
-and in the biofilm samples <!-- insert counts? -->
+granules 
+(mean = `r signif(subset(samp_counts, treatment == "wheat")$total, 3)`). 
+The potato treatment had the lowest absolute counts in both the solution 
+(`r signif(subset(sol_comb_out, treatment == "potato")$total, 3)`)
+and in the biofilm samples 
+(`r signif(subset(samp_counts, treatment == "potato")$total, 3)`) 
 (Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
-
-<!-- look at proportions of incorporated starches in each sample before looking at overall means -->
 
@@
 ```{r sample-perc}
-# convert proportions to percent labels
-perc_comb <- prop_comb %>% 
-  mutate(across(c(s, m, l, total), scales::percent, 0.001))
-
-perc <- prop %>%
-  mutate(across(c(s, m, l, total), scales::percent, 0.001)) %>% 
-  add_row(tibble_row(perc_comb), .before = 3)
-```
-
-```{r sample-prop-tbl}
-cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
-knitr::kable(perc, col.names = cnames, 
-             caption = "The mean percentage of starches from the solutions that 
-              were incorported into the samples.")
+prop_wide <- prop %>%
+  ungroup() %>%
+  pivot_wider(id_cols = c(treatment, starch), 
+              names_from = size, 
+              values_from = prop) %>%
+  add_row(prop_comb_row, .after = 2)
 ```
@@
 There is an inverse relationship between the absolute starch count in the solutions 
-and the proportional incorporation of starches in the biofilm samples, i.e. potato 
+and the proportional incorporation of starches in the biofilm samples, i.e., potato 
 had the lowest absolute count in solutions, but the highest proportional 
@@
 
-<!-- separated treatments -->
-
 <!-- mixed treatments -->
 Wheat incorporation was most affected in the mixed-treatment samples, with only
-`r paste0(perc$total[2])` of the total available starches being incorporated into
+`r paste0(perc_tbl$total[2])` of the total available starches being incorporated into
 the sample, compared to
-`r paste0(perc$total[4])`
+`r paste0(perc_tbl$total[4])`
 in the separated wheat treatment.
@@
 
-```{r sep-plot, fig.cap="Proportion (%) of sizes of (A) starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring).
-l = large, m = medium, s = small."}
-#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
-pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+<!-- overall size trends -->
+
+```{r size-trends}
+sol_size_prop <- sol_comb_out %>%
+  group_by(treatment, starch) %>%
+  #select(!total) %>%
+  mutate(across(where(is.numeric), function(x) x / total)) %>%
+  mutate(type = "solution")
+
+samp_size_prop <- samp_counts %>%
+  group_by(treatment, starch) %>%
+  #select(!total) %>%
+  mutate(across(where(is.numeric), function(x) x / total)) %>%
+  mutate(type = "sample")
+
+size_diff <- sol_size_prop %>%
+  full_join(samp_size_prop) %>%
+  pivot_longer(cols = c(s,m,l,total), names_to = "size", values_to = "count") %>%
+  pivot_wider(names_from = type, values_from = count) %>%
+  mutate(diff = (sample - solution) * 100) %>%
+  select(!c(solution, sample)) %>%
+  pivot_wider(names_from = size, values_from = diff) %>%
+  select(!total)
 ```
 
-<!-- change colour of medium starches in D to match C -->
-```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring).
-l = large, m = medium, s = small."}
-#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
-pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
-```
-<!-- Combine to a single figure? -->
-
-<!-- overall size trends -->
-
 Overall, medium starch granules had a higher mean rate of incorporation 
-(`r scales::percent(mean(prop$m, na.rm = T), 0.001)`) 
+(`r scales::percent(mean(prop_wide$m, na.rm = T), 0.001)`) 
 than small 
-(`r scales::percent(mean(prop$s, na.rm = T), 0.001)`) 
+(`r scales::percent(mean(prop_wide$s, na.rm = T), 0.001)`) 
 and large 
-(`r scales::percent(mean(prop$l, na.rm = T), 0.001)`) 
+(`r scales::percent(mean(prop_wide$l, na.rm = T), 0.001)`) 
 starch granules across all treatments, while large potato starches had the lowest 
@@
 Large potato granules (> 20 $\mu$m) were most affected, with a 
-`r paste0(abs(diff_sep$potato[3]), "%")` 
+`r paste0(abs(signif(size_diff$l[4], 3)), "%")` 
 decrease in relative abundance in the potato-only treatment, and a
-`r paste0(abs(diff_mix$potato[3]), "%")` 
+`r paste0(abs(signif(size_diff$l[1], 3)), "%")` 
 decrease in mixed treatments. Medium granules increased in relative abundance 
@@
 
-```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments."}
-pl_cor
-```
-<!-- with Z-score standardisation -->
-
 Pearson's *r* suggests a 
@@
 90%CI[`r paste(cor_ci[1], cor_ci[2], sep = ", ")`],
-p = `r cor_p`
-(Figure \@ref(fig:cor-plot)).
+p `r cor_p`
+(Figure \@ref(fig:cor-plot)A).
 
@@
 p `r cor2_p`
-(Figure \@ref(fig:cor-plot2)).
-
-<!-- size ratios in the larger deposits vs the smaller deposits?
-  Are larger starches more proportionately represented in larger deposits?
-  -->
-
-```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
-pl_cor2
-```
+(Figure \@ref(fig:cor-plot)B).

### 2022-05-17 README.Rmd
Update README.Rmd

@@
     for manuscript. Includes code to reproduce the figures and tables
-    generated by the analysis. It also has a rendered version,
-    `paper.docx`, suitable for reading (the code is replaced by figures
+    generated by the analysis. It also has a [rendered version](https://github.com/bbartholdy/byoc-starch/blob/main/analysis/paper/index.pdf)
+    suitable for reading (the code is replaced by figures
     and tables in this file)

### 2022-05-17 README.Rmd
update README

--- 2021-11-09 README.Rmd
@@
 
-<!-- Document structure provided by rrtools package
+<!-- Document structure provided by rrtools package -->
 
@@
 
-> Bartholdy, B. P. & Henry, A. G. (2021). _`r Title`_. bioRxiv <https://doi.org/xxx/xxx>
+>  _`r Title`_.
+Bjørn Peare Bartholdy, Amanda G. Henry,
+bioRxiv 2021.10.27.466104; doi: <https://doi.org/10.1101/2021.10.27.466104>
 
+<!--
 Our pre-print is online here:
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
--->
+> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx> -->
+
 ## Contents

### 2022-05-17 README.Rmd
update README badge

@@
 [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.5604670.svg)](https://doi.org/10.5281/zenodo.5604670)
-
+![Current version](https://img.shields.io/github/v/release/bbartholdy/byoc-starch?label=version)
 <!-- DOI badge: [![DOI](.svg)](repo link) -->

### 2022-05-19 analysis/paper/matmet.Rmd
fix citekeys and keep md

@@
 Statistical analysis was conducted in `r version$version.string`  [@Rbase] and
-the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
-here [@R-here], and patchwork [@R-patchwork].
+the following packages: tidyverse [@tidyverse2019], broom [@Rbroom], 
+here [@Rhere], and patchwork [@Rpatchwork].
 <!-- R packages used -->

### 2022-05-19 analysis/paper/paper.Rmd
fix citekeys and keep md

@@
 (https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch),
-following the format provided by the rrtools package [@R-rrtools].
+following the format provided by the rrtools package [@Rrrtools].
 More detailed protocols are available on OSF (https://osf.io/akevs/)

### 2022-05-31 analysis/paper/abstract.Rmd
incorporate reviewer comments

@@
 \(\mu\)m) consistently being under-represented. We also found a
-correlation between the absolute granule counts and dry-weight of the
-biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
+positive correlation between the absolute granule counts and dry-weight of the
+biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), meaning the absolute
+quantity of starch granules will increase as the size of the calculus deposit
+increases. A similar, but weaker correlation was found between
 the concentration (count per mg) of granules and dry-weight (\emph{r} =

### 2022-05-31 analysis/paper/discussion.Rmd
incorporate reviewer comments

@@
 from our results is the size of the calculus deposit. We found a 
-`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
-starch granules (Figure \@ref(fig:cor-plot)A);
+`r strength_cor` `r direct_cor` correlation between size of biofilm deposit and
+retained starch granules (Figure \@ref(fig:cor-plot)A), meaning larger calculus
+deposits contain a higher quantity of granules;
 a result that contradicts findings from archaeological contexts

### 2022-05-31 analysis/paper/intro.Rmd
incorporate reviewer comments

@@
 Studies conducted on modern humans [@leonardDentalCalculus2015] 
-and non-human primates [@powerChimpCalculus2015], 
+and non-human primates [@powerChimpCalculus2015; @powerRepresentativenessDental2021] 
 have explored how well microremains (phytoliths and starches)
@@
 cavity. Add to this the effects of the many different methods
-of starch processing, as well as post-depositional processes that are still being 
-explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
+of starch processing [@hardyRecoveringInformation2018], as well as post-depositional
+processes that are still being explored
+[@graneroStarchTaphonomy2020; @mercaderExaggeratedExpectations2018],
+and it becomes clear that using 
 dental calculus to reconstruct diet is a highly unpredictable process.

### 2022-05-31 analysis/paper/paper.Rmd
incorporate reviewer comments

@@
 Potato granules can be seen within a bacterial community.
-Scale bar = 20 $\\mu$m.", out.width="50%"}
-knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
+Scale bar = 20 $\\mu$m.", fig.show='hold', out.width="40%"}
+knitr::include_graphics(c(here("analysis/figures/starches_w_bar.jpg"), here("analysis/supplementary-materials/figures/SNAP-103412-0006.jpg")), )
 ```

### 2022-06-02 analysis/paper/abstract.Rmd
incorporate reviewer comments and resubmit

@@
 0.3, 90\%CI{[}0.0618, 0.506{]}).\\
-Our results reinforce previous \emph{in vivo} studies suggesting that
-dental calculus presents a very small, and partly biased picture of the
+Our results complement and reinforce previous \emph{in vivo} studies suggesting
+that dental calculus presents a very small, and partly biased picture of the
 original dietary intake of starches, with an over-representation of

### 2022-06-02 analysis/paper/conclusion.Rmd
incorporate reviewer comments and resubmit

@@
 to explore the incorporation and extraction of dietary compounds from dental calculus 
-in a controlled laboratory setting, and unearth the potential biases associated 
+in a controlled laboratory setting. We do not expect our model to replace *in vivo*
+methods; instead, it can provide a complementary means to address the limitations
+of *in vivo* studies, and unearth the potential biases associated 
 with dietary research conducted on archaeological dental calculus.

### 2022-06-02 analysis/paper/paper.Rmd
incorporate reviewer comments and resubmit

--- 2022-05-19 analysis/paper/paper.Rmd
@@
 #cite_keys <- unlist(cite_keys, use.names = FALSE)
+try(
   rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
+  #rbbt::bbt_update_bib(here("analysis/paper/paper.Rmd"), here("analysis/paper/refs.bib"))
+)
 # ## end refs.bib ##
@@
 
-```{r microscope-fig, fig.cap="Microscope image of a biofilm sample exposed to the potato starch solution.
-Potato granules can be seen within a bacterial community.
-Scale bar = 20 $\\mu$m.", out.width="50%"}
-knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
+```{r microscope-fig, fig.cap="Microscope images of biofilm samples that were exposed to the starch solutions.
+Starch granules can be seen within bacterial communities and isolated.
+Scale bar = 20 $\\mu$m.", fig.show='hold', out.width="40%"}
+knitr::include_graphics(c(here("analysis/figures/starches_w_bar.jpg"), here("analysis/figures/st2C3.2-mix.jpg"), here("analysis/figures/st1B4-wheat.jpg"), here("analysis/figures/2D2-potato.jpg")))
 ```

### 2022-06-02 analysis/paper/paper.tex
incorporate reviewer comments and resubmit

@@
 \(\mu\)m) consistently being under-represented. We also found a
-correlation between the absolute granule counts and dry-weight of the
-biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
+positive correlation between the absolute granule counts and dry-weight of the
+biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), meaning the absolute
+quantity of starch granules will increase as the size of the calculus deposit
+increases. A similar, but weaker correlation was found between
 the concentration (count per mg) of granules and dry-weight (\emph{r} =
 0.3, 90\%CI{[}0.0618, 0.506{]}).\\
-Our results reinforce previous \emph{in vivo} studies suggesting that
-dental calculus presents a very small, and partly biased picture of the
+Our results complement and reinforce previous \emph{in vivo} studies suggesting
+that dental calculus presents a very small, and partly biased picture of the
 original dietary intake of starches, with an over-representation of
@@
 Studies conducted on modern humans (Leonard et al., 2015)
-and non-human primates (Power et al., 2015),
+and non-human primates (Power et al., 2015; Power et al., 2021)
 have explored how well microremains (phytoliths and starches)
@@
 cavity. Add to this the effects of the many different methods
-of starch processing, as well as post-depositional processes that are still being
-explored (García-Granero, 2020), and it becomes clear that using
+of starch processing (Hardy et al., 2018), as well as post-depositional
+processes that are still being explored
+(García-Granero, 2020; Mercader et al., 2018),
+and it becomes clear that using
 dental calculus to reconstruct diet is a highly unpredictable process.
@@
 
-Statistical analysis was conducted in R version 4.1.3 (2022-03-10) (R Core Team, 2020) and
+Statistical analysis was conducted in R version 4.2.0 (2022-04-22) (R Core Team, 2020) and
 the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
@@
 from our results is the size of the calculus deposit. We found a
-strong positive correlation between biofilm size and retained
-starch granules (Figure \ref{fig:cor-plot}A);
+strong positive correlation between size of biofilm deposit and
+retained starch granules (Figure \ref{fig:cor-plot}A), meaning larger calculus
+deposits contain a higher quantity of granules;
 a result that contradicts findings from archaeological contexts
@@
 (Jepsen et al., 2011; Simón-Soro et al., 2013; Proctor et al., 2018; Fagernäs et al., 2021; Haffajee et al., 2009),
-as well as evolutionary differences (Fellows Yates et al., 2021).
+as well as evolutionary differences (Fellows Yates et al., 2021). The set of
+limitations for our model differ from \emph{in vivo} methods and, as such, we expect
+our model to complement the results and interpretations of existing and new
+\emph{in vivo} studies.
 It can also facilitate training of students and researchers on methods of
@@
 to explore the incorporation and extraction of dietary compounds from dental calculus
-in a controlled laboratory setting, and unearth the potential biases associated
+in a controlled laboratory setting. We do not expect our model to replace \emph{in vivo}
+methods; instead, it can provide a complementary means to address the limitations
+of \emph{in vivo} studies, and unearth the potential biases associated
 with dietary research conducted on archaeological dental calculus.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-hardyRecoveringInformation2018}{}}%
+Hardy, K., Buckley, S., and Copeland, L.
+(2018).
+Pleistocene dental calculus: {Recovering} information on {Paleolithic} food items, medicines, paleoenvironment and microbes.
+\emph{Evolutionary Anthropology: Issues, News, and Reviews} 27, 234--246.
+doi:\href{https://doi.org/10.1002/evan.21718}{10.1002/evan.21718}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-haslamDecompositionStarch2004}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-R-rrtools}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-Rrrtools}{}}%
 Marwick, B. (2019). \emph{Rrtools: {Creates} a reproducible research compendium}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mercaderExaggeratedExpectations2018}{}}%
+Mercader, J., Akeju, T., Brown, M., Bundala, M., Collins, M.
+J., Copeland, L., Crowther, A., Dunfield, P., Henry, A., Inwood, J., et al.
+(2018).
+Exaggerated expectations in ancient starch research and the need for new taphonomic and authenticity criteria.
+\emph{FACETS} 3, 777--798.
+doi:\href{https://doi.org/10.1139/facets-2017-0126}{10.1139/facets-2017-0126}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-R-here}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-Rhere}{}}%
 Müller, K. (2020). \emph{Here: {A} simpler way to find your files}.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-R-patchwork}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-Rpatchwork}{}}%
 Pedersen, T. L. (2020). \emph{Patchwork: {The} composer of plots}.
@@
 doi:\href{https://doi.org/10.1016/j.jas.2014.04.016}{10.1016/j.jas.2014.04.016}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-powerRepresentativenessDental2021}{}}%
+Power, R.
+C., Wittig, R.
+M., Stone, J.
+R., Kupczik, K., and Schulz-Kornas, E.
+(2021).
+The representativeness of the dental calculus dietary record: Insights from {Taï} chimpanzee faecal phytoliths.
+\emph{Archaeological and Anthropological Sciences} 13, 104.
+doi:\href{https://doi.org/10.1007/s12520-021-01342-z}{10.1007/s12520-021-01342-z}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-R-broom}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-Rbroom}{}}%
 Robinson, D., Hayes, A., and Couch, S. (2021). \emph{Broom: {Convert} statistical objects into tidy tibbles}.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sissonsMultistationDentalPlaque1991}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-sissonsMultistationPlaque1991}{}}%
 Sissons, C.
@@
 
-{\centering \includegraphics[width=0.7\linewidth]{/mnt/hogwarts/Insync/publications/PhD/byocstarch/analysis/figures/protocol_overview}
+{\centering \includegraphics[width=0.7\linewidth]{../figures/protocol_overview}
 
@@
 
-{\centering \includegraphics[width=0.5\linewidth]{/mnt/hogwarts/Insync/publications/PhD/byocstarch/analysis/figures/starches_w_bar}
+{\centering \includegraphics[width=0.4\linewidth]{../figures/starches_w_bar} \includegraphics[width=0.4\linewidth]{../figures/st2C3.2-mix} \includegraphics[width=0.4\linewidth]{../figures/st1B4-wheat} \includegraphics[width=0.4\linewidth]{../figures/2D2-potato}
 
@@
 
-\caption{Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\mu$m.}\label{fig:microscope-fig}
+\caption{Microscope images of biofilm samples that were exposed to the starch solutions.
+Starch granules can be seen within bacterial communities and isolated.
+Scale bar = 20 $\mu$m.}\label{fig:microscope-fig}
 \end{figure}

### 2022-06-02 analysis/supplementary-materials/supp-mat.Rmd
incorporate reviewer comments and resubmit

@@
 
-```{r img-2, fig.cap="Microscope image of wheat starch from a Potato treatment sample.", echo=FALSE, out.width="80%"}
+```{r img-2, fig.cap="Microscope image of potato starch from a potato treatment sample.", echo=FALSE, out.width="80%"}
 knitr::include_graphics(here("analysis/supplementary-materials/figures/SNAP-103412-0006.jpg"))

### 2022-06-03 analysis/scripts/amylase-assay.R
test changed filepaths for dissertation

@@ -1,4 +1,22 @@
-library(tidyverse)
+library(dplyr)
+library(ggplot2)
+
+# Upload data -------------------------------------------------------------
+
+plt1_ph1 <- readr::read_tsv(
+  "analysis/data/raw_data/ExpM_plate1_photometric1.csv",
+  col_names = F)
+plt1_ph2 <- readr::read_tsv(
+  "analysis/data/raw_data/ExpM_plate1_photometric2.csv",
+  col_names = F)
+
+plt2_ph1 <- readr::read_tsv(
+  "analysis/data/raw_data/ExpM_plate2_photometric1.csv",
+  col_names = F)
+plt2_ph2 <- readr::read_tsv(
+  "analysis/data/raw_data/ExpM_plate2_photometric2.csv",
+  col_names = F)
+
 
 # Functions ---------------------------------------------------------------
 
@@ -103,24 +121,6 @@ sep_samples <- function(data){
 #   out <- list("saliva" = saliva, "bmm" = bmm)
 #   return(out)
 # }
-
-
-# Upload data -------------------------------------------------------------
-
-plt1_ph1 <- readr::read_tsv(
-  here::here("analysis/data/raw_data/ExpM_plate1_photometric1.csv"),
-  col_names = F)
-plt1_ph2 <- readr::read_tsv(
-  here::here("analysis/data/raw_data/ExpM_plate1_photometric2.csv"),
-  col_names = F)
-
-plt2_ph1 <- readr::read_tsv(
-  here::here("analysis/data/raw_data/ExpM_plate2_photometric1.csv"),
-  col_names = F)
-plt2_ph2 <- readr::read_tsv(
-  here::here("analysis/data/raw_data/ExpM_plate2_photometric2.csv"),
-  col_names = F)
-
 
 # Data prep ---------------------------------------------------------------
 

### 2022-06-03 analysis/scripts/analysis.R

@@ -3,16 +3,16 @@
 # Dependencies ------------------------------------------------------------
 
 library(here)
-library(tidyverse)
+library(dplyr)
 library(broom)
 
 # Upload data -------------------------------------------------------------
 
 # starch counts in samples
-raw_counts <- readr::read_csv(here("analysis/data/raw_data", "starch_counts.csv"))
+raw_counts <- readr::read_csv("../data/raw_data/starch_counts.csv")
 
 # starch counts in solutions
-sol_counts <- readr::read_csv(here("analysis/data/raw_data", "solution_counts.csv"))
+sol_counts <- readr::read_csv("../data/raw_data/solution_counts.csv")
 
 # Functions ---------------------------------------------------------------
 
@@ -47,7 +47,7 @@ raw_counts %>%
 
 sol_corr <- sol_counts %>%
   mutate(across(c(s, m, l, total), extrap_count, portion_slide, vol_slide, days = 16))
-readr::write_csv(sol_corr, here("analysis/data/derived_data", "sol_corr.csv"))
+readr::write_csv(sol_corr, "analysis/data/derived_data/sol_corr.csv")
 
 # combining potato and wheat counts from mixed treatment solution
 
@@ -83,7 +83,7 @@ corr_counts <- raw_counts %>%
   mutate(across(c(s, m, l, total), correct_count, portion_slide, vol)) %>%
   select(!c(vol, portion_slide)) %>%
   filter(sample != "st2C6")
-readr::write_csv(corr_counts, here("analysis/data/derived_data", "corr_counts.csv"))
+readr::write_csv(corr_counts, "analysis/data/derived_data/corr_counts.csv")
 
 corr_counts_long <- corr_counts %>%
   #filter(treatment != "control") %>%
@@ -101,7 +101,7 @@ corr_comb <- corr_counts %>%
 #   pivot_wider(names_sep = "", values_from = c(s, m, l, total),
 #               values_fill = NA,
 #               values_fn = function(x) sum(x, na.rm = T))
-readr::write_csv(corr_comb, here("analysis/data/derived_data", "corr_comb.csv"))
+readr::write_csv(corr_comb, "analysis/data/derived_data/corr_comb.csv")
 
 corr_comb_long <- corr_counts_long %>%
   filter(treatment != "control") %>%

### 2022-06-03 analysis/scripts/figures.R
clean up dependencies

@@ -1,4 +1,5 @@
-library(tidyverse)
+library(dplyr)
+library(ggplot2)
 #source(here("analysis/scripts/analysis.R"))
 #corr_counts <- readr::read_csv(here("analysis/data/derived_data/", "corr_counts.csv"))
 #corr_comb <- readr::read_csv(here("analysis/data/derived_data/", "corr_comb.csv"))

### 2022-06-03 analysis/scripts/script1.R

@@ -1,73 +1,0 @@
-# R Script for correcting counts according to volume and proportion of slide counted
-
-# Dependencies ------------------------------------------------------------
-
-library(magrittr)
-library(dplyr)
-library(here)
-library(readr)
-
-# Upload data -------------------------------------------------------------
-
- # starch counts in samples
-raw_counts <- readr::read_csv(here("analysis/data/derived_data/", "starch_counts.csv"))
-
- # starch counts in solutions
-sol_counts <- readr::read_csv(here("analysis/data/derived_data/", "solution_counts.csv"))
-
-# Functions ---------------------------------------------------------------
-
- # function to correct counts for volume and proportion of slide counted
-
-correct_count <- function(raw, prop, vol){
-  corrected <- raw * (1 / prop) * (vol / 20)
-  corrected
-}
-
- # function to extrapolate solution counts to 1 ml exposure per day
-
-extrap_count <- function(raw, prop, vol, days){
-  extrap <- raw * (1 / prop) * 100 * days
-  extrap
-}
-
-# Outlier detection and removal -------------------------------------------
-
-raw_counts %>%
-  filter(treatment != "control") %>%
-  ggplot() +
-  geom_boxplot(aes(x = starch, y = total, fill = starch))
-
-# Only lower outliers were considered, as this would suggest that there may have
-  # been a problem with the sample and incorporation of starches. Given that all
-  # samples were treated equally, there is no reason to consider upper outliers
-  # problematic.
-
-
-# Solution counts ---------------------------------------------------------
-
-sol_corr <- sol_counts %>%
-  mutate(across(c(s, m, l, total), extrap_count, portion_slide, vol_slide, days = 16))
-readr::write_csv(sol_corr, here("data", "sol_corr.csv"))
-
-# Apply correction --------------------------------------------------------
-
-corr_counts <- raw_counts %>%
-  #filter(treatment != "control") %>%
-  mutate(across(c(s, m, l, total), correct_count, portion_slide, vol)) %>%
-  select(!c(vol, portion_slide)) %>%
-  filter(sample != "st2C6")
-readr::write_csv(corr_counts, here("data", "corr_counts.csv"))
-
-# Combining mixed treatment counts ----------------------------------------
-
-  # combining potato and wheat counts from mixed treatment samples
-
-# merge two adjacent rows and add variables s, m , l, total
-
-corr_comb <- corr_counts %>%
-  select(!starch) %>%
[…]

### 2022-06-03 analysis/scripts/tables.R

@@ -1,7 +1,7 @@
 # Script for tables
 
 library(here)
-library(tidyverse)
+library(dplyr)
 
 #corr_counts <- readr::read_csv(here("analysis/data/derived_data/", "corr_counts.csv"))
 #corr_comb <- readr::read_csv(here("analysis/data/derived_data/", "corr_comb.csv"))

### 2022-06-03 analysis/paper/paper.Rmd
move all filepaths from scripts to index.Rmd

@@
   dpi = 300) 
-library(tidyverse)
+library(dplyr)
+library(ggplot2)
 library(broom)
@@
 library(patchwork)
+plt1_ph1 <- readr::read_tsv(
+  here("analysis/data/raw_data/ExpM_plate1_photometric1.csv"),
+  col_names = F)
+plt1_ph2 <- readr::read_tsv(
+  here("analysis/data/raw_data/ExpM_plate1_photometric2.csv"),
+  col_names = F)
+
+plt2_ph1 <- readr::read_tsv(
+  here("analysis/data/raw_data/ExpM_plate2_photometric1.csv"),
+  col_names = F)
+plt2_ph2 <- readr::read_tsv(
+  here("analysis/data/raw_data/ExpM_plate2_photometric2.csv"),
+  col_names = F)
+# starch counts in samples
+raw_counts <- readr::read_csv(here("analysis/data/raw_data/starch_counts.csv"))
+# starch counts in solutions
+sol_counts <- readr::read_csv(here("analysis/data/raw_data/solution_counts.csv"))
 # load R scripts
@@
 source(here("analysis/scripts/amylase-assay.R"))
+readr::write_csv(sol_corr, here("analysis/data/derived_data/sol_corr.csv"))
+readr::write_csv(corr_counts, here("analysis/data/derived_data/corr_counts.csv"))
+readr::write_csv(corr_comb, here("analysis/data/derived_data/corr_comb.csv"))
 ## Compile refs.bib using rbbt package ##

### 2022-06-03 analysis/scripts/amylase-assay.R

@@ -3,19 +3,19 @@ library(ggplot2)
 
 # Upload data -------------------------------------------------------------
 
-plt1_ph1 <- readr::read_tsv(
-  "analysis/data/raw_data/ExpM_plate1_photometric1.csv",
-  col_names = F)
-plt1_ph2 <- readr::read_tsv(
-  "analysis/data/raw_data/ExpM_plate1_photometric2.csv",
-  col_names = F)
-
-plt2_ph1 <- readr::read_tsv(
-  "analysis/data/raw_data/ExpM_plate2_photometric1.csv",
-  col_names = F)
-plt2_ph2 <- readr::read_tsv(
-  "analysis/data/raw_data/ExpM_plate2_photometric2.csv",
-  col_names = F)
+# plt1_ph1 <- readr::read_tsv(
+#   "analysis/data/raw_data/ExpM_plate1_photometric1.csv",
+#   col_names = F)
+# plt1_ph2 <- readr::read_tsv(
+#   "analysis/data/raw_data/ExpM_plate1_photometric2.csv",
+#   col_names = F)
+#
+# plt2_ph1 <- readr::read_tsv(
+#   "analysis/data/raw_data/ExpM_plate2_photometric1.csv",
+#   col_names = F)
+# plt2_ph2 <- readr::read_tsv(
+#   "analysis/data/raw_data/ExpM_plate2_photometric2.csv",
+#   col_names = F)
 
 
 # Functions ---------------------------------------------------------------

### 2022-06-03 analysis/scripts/analysis.R

@@ -5,14 +5,15 @@
 library(here)
 library(dplyr)
 library(broom)
+library(tidyr)
 
 # Upload data -------------------------------------------------------------
 
 # starch counts in samples
-raw_counts <- readr::read_csv("../data/raw_data/starch_counts.csv")
+#raw_counts <- readr::read_csv("../data/raw_data/starch_counts.csv")
 
 # starch counts in solutions
-sol_counts <- readr::read_csv("../data/raw_data/solution_counts.csv")
+#sol_counts <- readr::read_csv("../data/raw_data/solution_counts.csv")
 
 # Functions ---------------------------------------------------------------
 
@@ -47,7 +48,7 @@ raw_counts %>%
 
 sol_corr <- sol_counts %>%
   mutate(across(c(s, m, l, total), extrap_count, portion_slide, vol_slide, days = 16))
-readr::write_csv(sol_corr, "analysis/data/derived_data/sol_corr.csv")
+#readr::write_csv(sol_corr, "analysis/data/derived_data/sol_corr.csv")
 
 # combining potato and wheat counts from mixed treatment solution
 
@@ -83,7 +84,7 @@ corr_counts <- raw_counts %>%
   mutate(across(c(s, m, l, total), correct_count, portion_slide, vol)) %>%
   select(!c(vol, portion_slide)) %>%
   filter(sample != "st2C6")
-readr::write_csv(corr_counts, "analysis/data/derived_data/corr_counts.csv")
+#readr::write_csv(corr_counts, "analysis/data/derived_data/corr_counts.csv")
 
 corr_counts_long <- corr_counts %>%
   #filter(treatment != "control") %>%
@@ -101,7 +102,7 @@ corr_comb <- corr_counts %>%
 #   pivot_wider(names_sep = "", values_from = c(s, m, l, total),
 #               values_fill = NA,
 #               values_fn = function(x) sum(x, na.rm = T))
-readr::write_csv(corr_comb, "analysis/data/derived_data/corr_comb.csv")
+#readr::write_csv(corr_comb, "analysis/data/derived_data/corr_comb.csv")
 
 corr_comb_long <- corr_counts_long %>%
   filter(treatment != "control") %>%

### 2022-06-03 analysis/paper/discussion.Rmd
incorporate reviewer comments and resubmit
remove hyphen from citekey

@@
 morphology 
-[@jepsenCalculusRemoval2011; @simon-soroOralGeography2013; @proctorSpatialGradient2018; @fagernasMicrobialBiogeography2021; @haffajeeBiofilmPosition2009], 
-as well as evolutionary differences [@yatesOralMicrobiome2021].
+[@jepsenCalculusRemoval2011; @simonsoroOralGeography2013; @proctorSpatialGradient2018; @fagernasMicrobialBiogeography2021; @haffajeeBiofilmPosition2009], 
+as well as evolutionary differences [@yatesOralMicrobiome2021]. The set of
+limitations for our model differ from *in vivo* methods and, as such, we expect
+our model to complement the results and interpretations of existing and new
+*in vivo* studies.
 It can also facilitate training of students and researchers on methods of 

### 2022-06-08 analysis/paper/paper.Rmd
update references

@@
 try(
-  rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
-  #rbbt::bbt_update_bib(here("analysis/paper/paper.Rmd"), here("analysis/paper/refs.bib"))
+  #rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
+  rbbt::bbt_update_bib(here("analysis/paper/paper.Rmd"), here("analysis/paper/refs.bib"))
 )

### 2022-09-26 analysis/paper/paper.Rmd
binder fix - attempt 1

@@
 readr::write_csv(corr_comb, here("analysis/data/derived_data/corr_comb.csv"))
+
 ## Compile refs.bib using rbbt package ##
-cite_keys <- rbbt::bbt_detect_citations(list.files(pattern = ".Rmd"))
+#cite_keys <- rbbt::bbt_detect_citations(list.files(pattern = ".Rmd"))
 #cite_keys <- unlist(cite_keys, use.names = FALSE)
@@
   #rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
-  rbbt::bbt_update_bib(here("analysis/paper/paper.Rmd"), here("analysis/paper/refs.bib"))
+  rbbt::bbt_update_bib(
+    here("analysis/paper/paper.Rmd"),
+    here("analysis/paper/refs.bib"))
 )
 # ## end refs.bib ##
-# 
+
 # ## Word count ##

### 2022-09-26 analysis/paper/paper.Rmd
binder fix - attempt 2

@@
 runtitle: "Biases Associated with Starch Incorporation"
-documentclass: frontiers/frontiersSCNS  # or frontiersHLTH, or frontiersFPHY
+documentclass: ../templates/frontiersSCNS  # or frontiersHLTH, or frontiersFPHY
 author:
@@
 bibliography: refs.bib
-csl: frontiers/frontiers.csl
+csl: frontiers.csl
 params:

### 2022-09-27 README.Rmd
binder fix - attempt 1
update runtime

--- 2022-05-17 README.Rmd
@@
 
-<!--[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bbartholdy/byoc-starch/main?urlpath=rstudio)--> 
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/bbartholdy/byoc-starch/main?urlpath=rstudio) 
 [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.5604670.svg)](https://doi.org/10.5281/zenodo.5604670)
-
+![Current version](https://img.shields.io/github/v/release/bbartholdy/byoc-starch?label=version)
 <!-- DOI badge: [![DOI](.svg)](repo link) -->
@@
 You can download the compendium as a zip from from this URL:
-[master.zip](/archive/master.zip). After unzipping:
+[main.zip](/archive/refs/heads/main.zip). After unzipping:
 - open the `.Rproj` file in RStudio

### 2023-02-09 analysis/paper/paper.Rmd
update bibliography

@@
 #cite_keys <- unlist(cite_keys, use.names = FALSE)
+rmd_files <- list.files(path = here("analysis/paper/"), pattern = "*.Rmd")
+
 try(
@@
   rbbt::bbt_update_bib(
-    here("analysis/paper/paper.Rmd"),
+    rmd_files,
     here("analysis/paper/refs.bib"))

### 2023-02-09 analysis/paper/paper.tex
update bibliography

@@
 
-\documentclass[utf8]{frontiers/frontiersSCNS}
+\documentclass[utf8]{../templates/frontiersSCNS}
 
@@
 
-Statistical analysis was conducted in R version 4.2.0 (2022-04-22) (R Core Team, 2020) and
+Statistical analysis was conducted in R version 4.2.2 Patched (2022-11-10 r83330) (R Core Team, 2020) and
 the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
@@
 W., Haak, W., Bradshaw, C.
-J., Townsend, G., Soltysiak, A., Alt, K.
+J., Townsend, G., Sołtysiak, A., Alt, K.
 W., et al.
@@
 Sequencing ancient calcified dental plaque shows changes in oral microbiota with dietary shifts of the {Neolithic} and {Industrial} revolutions.
-\emph{Nature Genetics} 45, 450--5, 455e1.
+\emph{Nat Genet} 45, 450--5, 455e1.
 doi:\href{https://doi.org/10.1038/ng.2536}{10.1038/ng.2536}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-armitageExtractionIdentification1975}{}}%
-Armitage, P. L. (1975). The {Extraction} and {Identification} of {Opal Phytoliths} from the {Teeth} of {Ungulates}. \emph{Journal of Archaeological Science} 2, 187--197.
+Armitage, P. L. (1975). The {Extraction} and {Identification} of {Opal Phytoliths} from the {Teeth} of {Ungulates}. \emph{J Archaeol Sci} 2, 187--197.
 
 \leavevmode\vadjust pre{\hypertarget{ref-bernfeldAmylase1955}{}}%
-Bernfeld, P.
-(1955).
-{``Amylases, {\(\alpha\)} and {\(\beta\)},''} in \emph{Methods in {Enzymology}} ({Academic Press}), 149--158.
-doi:\href{https://doi.org/10.1016/0076-6879(55)01021-5}{10.1016/0076-6879(55)01021-5}.
+Bernfeld, P. (1955). {``Amylases, α and β,''} in \emph{Methods in {Enzymology}} ({Academic Press}), 149--158. doi:\href{https://doi.org/10.1016/0076-6879(55)01021-5}{10.1016/0076-6879(55)01021-5}.
 
@@
 Machado ({Spain}: {CSIC Press}).
+Available at: \url{https://books.google.com?id=j66CDVfVhwEC}.
 
@@
 The {Relationship Between Plaque pH}, {Plaque Acid Anion Profiles}, and {Oral Carbohydrate Retention After Ingestion} of {Several} '{Reference Foods}' by {Human Subjects}.
-\emph{Journal of Dental Research} 67, 861--865.
+\emph{J Dent Res} 67, 861--865.
 doi:\href{https://doi.org/10.1177/00220345880670051301}{10.1177/00220345880670051301}.
@@
 Dental calculus as a source of ancient alkaloids: {Detection} of nicotine by {LC-MS} in calculus samples from the {Americas}.
-\emph{Journal of Archaeological Science: Reports} 18, 509--515.
+\emph{J Archaeol Sci Rep} 18, 509--515.
 doi:\href{https://doi.org/10.1016/j.jasrep.2018.02.004}{10.1016/j.jasrep.2018.02.004}.
@@
 The evolution and changing ecology of the {African} hominid oral microbiome.
-\emph{Proceedings of the National Academy of Sciences} 118.
+\emph{PNAS} 118.
 doi:\href{https://doi.org/10.1073/pnas.2021655118}{10.1073/pnas.2021655118}.
@@
 Pleistocene dental calculus: {Recovering} information on {Paleolithic} food items, medicines, paleoenvironment and microbes.
-\emph{Evolutionary Anthropology: Issues, News, and Reviews} 27, 234--246.
+\emph{Evol Anthropol} 27, 234--246.
 doi:\href{https://doi.org/10.1002/evan.21718}{10.1002/evan.21718}.
@@
 Proteomic evidence of dietary sources in ancient dental calculus.
-\emph{Proceedings.
-Biological Sciences} 285, 20180977.
+\emph{Proc Biol Sci} 285, 20180977.
 doi:\href{https://doi.org/10.1098/rspb.2018.0977}{10.1098/rspb.2018.0977}.
@@
 Anthology of {Starch Granule Morphology} by {Scanning Electron Microscopy}.
-\emph{Starch - Stärke} 46, 121--129.
+\emph{Starch/Stärke} 46, 121--129.
 doi:\href{https://doi.org/10.1002/star.19940460402}{10.1002/star.19940460402}.
@@
 Lack of {Correlation Between Food Retention} on the {Human Dentition} and {Consumer Perception} of {Food Stickiness}.
-\emph{Journal of Dental Research} 70, 1314--1319.
+\emph{J Dent Res} 70, 1314--1319.
 doi:\href{https://doi.org/10.1177/00220345910700100101}{10.1177/00220345910700100101}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kashketFoodParticles1996}{}}%
-Kashket, S., Zhang, J., and Houte, J.
-V.
-(1996).
-Accumulation of {Fermentable Sugars} and {Metabolic Acids} in {Food Particles} that {Become Entrapped} on the {Dentition}.
-\emph{Journal of Dental Research}, 8.
+Kashket, S., Zhang, J., and Houte, J. V. (1996). Accumulation of {Fermentable Sugars} and {Metabolic Acids} in {Food Particles} that {Become Entrapped} on the {Dentition}. \emph{J Dent Res}, 8.
 
@@
 Plant microremains in dental calculus as a record of plant consumption: {A} test with {Twe} forager-horticulturalists.
-\emph{Journal of Archaeological Science: Reports} 2, 449--457.
+\emph{J Archaeol Sci Rep} 2, 449--457.
 doi:\href{https://doi.org/10.1016/j.jasrep.2015.03.009}{10.1016/j.jasrep.2015.03.009}.
@@
 Effect of {Frequent Consumption} of {Starchy Food Items} on {Enamel} and {Dentin Demineralization} and on {Plaque pH} in situ.
-\emph{Journal of Dental Research} 73, 652--660.
+\emph{J Dent Res} 73, 652--660.
 doi:\href{https://doi.org/10.1177/00220345940730031101}{10.1177/00220345940730031101}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Rrrtools}{}}%
-Marwick, B. (2019). \emph{Rrtools: {Creates} a reproducible research compendium}.
+Marwick, B. (2019). \emph{Rrtools: {Creates} a reproducible research compendium}. Available at: \url{https://github.com/benmarwick/rrtools}.
 
@@
 Extraction of {Opal Phytoliths} from {Herbivore Dental Calculus}.
-\emph{Journal of Archaeological Science} 21, 469--473.
+\emph{J Archaeol Sci} 21, 469--473.
 doi:\href{https://doi.org/10.1006/jasc.1994.1046}{10.1006/jasc.1994.1046}.
@@
 Combined methodologies for gaining much information from ancient dental calculus: Testing experimental strategies for simultaneously analysing {DNA} and food residues.
-\emph{Archaeological and Anthropological Sciences} 12, 10.
+\emph{Archaeol Anthropol Sci} 12, 10.
 doi:\href{https://doi.org/10.1007/s12520-019-00983-5}{10.1007/s12520-019-00983-5}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Rhere}{}}%
-Müller, K. (2020). \emph{Here: {A} simpler way to find your files}.
+Müller, K. (2020). \emph{Here: {A} simpler way to find your files}. Available at: \url{https://CRAN.R-project.org/package=here}.
 
@@
 (2013).
-Taking the {Starch} out of {Oral Biofilm Formation}: {Molecular Basis} and {Functional Significance} of {Salivary} {\(\alpha\)}-{Amylase Binding} to {Oral Streptococci}.
-\emph{Applied and Environmental Microbiology} 79, 416--423.
+Taking the {Starch} out of {Oral Biofilm Formation}: {Molecular Basis} and {Functional Significance} of {Salivary} α-{Amylase Binding} to {Oral Streptococci}.
+\emph{Appl.
+Environ.
+Microbiol.} 79, 416--423.
 doi:\href{https://doi.org/10.1128/AEM.02581-12}{10.1128/AEM.02581-12}.
@@
 The {Concomitant Deposition} of {Strontium} and {Fluoride} in {Dental Plaque}.
-\emph{Journal of Dental Research} 66, 1518--1522.
+\emph{J Dent Res} 66, 1518--1522.
 doi:\href{https://doi.org/10.1177/00220345870660100101}{10.1177/00220345870660100101}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Rpatchwork}{}}%
-Pedersen, T. L. (2020). \emph{Patchwork: {The} composer of plots}.
+Pedersen, T. L. (2020). \emph{Patchwork: {The} composer of plots}. Available at: \url{https://CRAN.R-project.org/package=patchwork}.
 
@@
 Dental calculus evidence of {Tai Forest Chimpanzee} plant consumption and life history transitions.
-\emph{Scientific Reports} 5, 15161.
+\emph{Sci Rep} 5, 15161.
 doi:\href{https://doi.org/10.1038/srep15161}{10.1038/srep15161}.
@@
 The representativeness of the dental calculus dietary record: Insights from {Taï} chimpanzee faecal phytoliths.
-\emph{Archaeological and Anthropological Sciences} 13, 104.
+\emph{Archaeol Anthropol Sci} 13, 104.
 doi:\href{https://doi.org/10.1007/s12520-021-01342-z}{10.1007/s12520-021-01342-z}.
@@
 A spatial gradient of bacterial diversity in the human oral cavity shaped by salivary flow.
-\emph{Nature Communications} 9, 681.
+\emph{Nat Commun} 9, 681.
 doi:\href{https://doi.org/10.1038/s41467-018-02900-1}{10.1038/s41467-018-02900-1}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Rbase}{}}%
-R Core Team (2020). \emph{R: {A} language and environment for statistical computing}. {Vienna, Austria}: {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing}.
+R Core Team (2020).
+\emph{R: {A} language and environment for statistical computing}.
+{Vienna, Austria}: {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing} Available at: \url{https://www.R-project.org/}.
 
@@
 Beyond food: {The} multiple pathways for inclusion of materials into ancient dental calculus.
-\emph{American Journal of Physical Anthropology} 162, 71--83.
+\emph{Am J Phys Anthropol} 162, 71--83.
 doi:\href{https://doi.org/10.1002/ajpa.23147}{10.1002/ajpa.23147}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Rbroom}{}}%
-Robinson, D., Hayes, A., and Couch, S. (2021). \emph{Broom: {Convert} statistical objects into tidy tibbles}.
+Robinson, D., Hayes, A., and Couch, S. (2021). \emph{Broom: {Convert} statistical objects into tidy tibbles}. Available at: \url{https://CRAN.R-project.org/package=broom}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-scannapiecoSalivaryAmylase1993}{}}%
-Scannapieco, F. A., Torres, G., and Levine, M. J. (1993). Salivary {\(\alpha\)}-amylase: Role in dental plaque and caries formation. \emph{Critical Reviews in Oral Biology \& Medicine} 4, 301--307.
+Scannapieco, F. A., Torres, G., and Levine, M. J. (1993). Salivary α-amylase: Role in dental plaque and caries formation. \emph{Critical Reviews in Oral Biology \& Medicine} 4, 301--307.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-simon-soroOralGeography2013}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-simonsoroOralGeography2013}{}}%
 Simón-Soro, A., Tomás, I., Cabrera-Rubio, R., Catalan, M.
@@
 Microbial geography of the oral cavity.
-\emph{Journal of Dental Research} 92, 616--621.
+\emph{J Dent Res} 92, 616--621.
 doi:\href{https://doi.org/10.1177/0022034513488119}{10.1177/0022034513488119}.
@@
 (1991).
-A {Multi-station Dental Plaque Microcosm} ({Artificial Mouth}) for the {Study} of {Plaque Growth}, {Metabolism}, {pH}, and {Mineralization}: \emph{Journal of Dental Research}.
+A {Multi-station Dental Plaque Microcosm} ({Artificial Mouth}) for the {Study} of {Plaque Growth}, {Metabolism}, {pH}, and {Mineralization}: \emph{J Dent Res}.
 doi:\href{https://doi.org/10.1177/00220345910700110301}{10.1177/00220345910700110301}.
@@
 A preliminary investigation into the ultrastructure of dental calculus and associated bacteria.
-\emph{Journal of Clinical Periodontology} 31, 364--369.
+\emph{J Clin Periodontol} 31, 364--369.
 doi:\href{https://doi.org/10.1111/j.1600-051X.2004.00484.x}{10.1111/j.1600-051X.2004.00484.x}.
@@
 Direct evidence of milk consumption from ancient human dental calculus.
-\emph{Scientific Reports} 4, 7104.
+\emph{Sci Rep} 4, 7104.
 doi:\href{https://doi.org/10.1038/srep07104}{10.1038/srep07104}.
@@
 Pathogens and host immunity in the ancient human oral cavity.
-\emph{Nature Genetics} 46, 336--44.
+\emph{Nat Genet} 46, 336--44.
 doi:\href{https://doi.org/10.1038/ng.2906}{10.1038/ng.2906}.
@@
 Evaluating microfossil content of dental calculus from {Brazilian} sambaquis.
-\emph{Journal of Archaeological Science} 37, 1326--1338.
+\emph{J Archaeol Sci} 37, 1326--1338.
 doi:\href{https://doi.org/10.1016/j.jas.2009.12.037}{10.1016/j.jas.2009.12.037}.

### 2023-02-09 R/amylase.R
implement more formal R functions

@@ -1,0 +1,89 @@
+#' Standard curve calculation
+#' @param std_data Data frame with the photometric readings of the standard curve.
+#' @param cols Vector containing the column numbers containing the standard curve photometric readings. If NULL, all columns assumed to be standard curve measurements.
+#' @param ml_maltose vector of length 7 containing the volume (ml) of 0.2% maltose solution added to the assay.
+#' @importFrom magrittr %>%
+#' @importFrom dplyr slice n
+stdcurve <- function(std_data, cols = NULL, ml_maltose = c(0.00375, 0.015, 0.030, 0.045, 0.060, 0.075, 0.150)){
+  if(!is.null(cols)){
+    std_data <- std_data[,cols]
+  }
+  # delta 540 calculation
+  mg_maltose <- 2 * ml_maltose # DOUBLE CHECK THIS
+  umol_maltose <- (mg_maltose * 1000) / 342.3 # calculate maltose in umoles
+  replicates <- dim(std_data)[2]
+  pht_blank <- std_data[8,] %>%
+    slice(rep(1:n(), each = 7))
+  pht_meas <- std_data[1:7,] - pht_blank
+
+  pht_meas <- c(pht_meas[, seq(1:replicates)]) %>%
+    unlist()
+  # Linear regression
+  stdCurve_data <- data.frame("d540" = pht_meas, "umol_maltose" = rep(umol_maltose, replicates))
+  std_curve <- lm(d540 ~ umol_maltose, stdCurve_data)
+  # Plot of standard curve
+  p_stdcurve <- stdCurve_data %>%
+    ggplot(aes(x = umol_maltose, y = d540)) +
+    geom_point() +
+    geom_smooth(formula = "y ~ x", method = "lm", se = F) +
+    theme_bw() +
+    labs(x = "umoles Maltose", y = "Absorbance (d540)")
+  out <- list("std_curve" = std_curve, "plot" = p_stdcurve, "blank" = pht_blank[1,], "replicates" = replicates, "model" = stdCurve_data)
+  return(out)
+}
+
+#' Amylase activity in sample calculation
+#' @param samp_data data frame with the photometric readings of the samples.
+#' @param cols vector containing the column numbers of the sample photometric readings. If NULL, all columns assumed to be standard curve measurements.
[…]

### 2023-02-09 R/counts.R

@@ -1,0 +1,26 @@
+#' Correct starch counts
+#'
+#' Calculates the number of starches on a whole microscope slide based on the
+#' proportion of slide that was actually counted.
+#'
+#' @param raw integer. absolute number of starches counted.
+#' @param prop numeric. Proportion of slide that was counted.
+#' @param vol numeric. Volume of EDTA on the mounted slide.
+#' @export
+
+correct_count <- function(raw, prop, vol){
+  corrected <- raw * (1 / prop) * (vol / 20)
+  corrected
+}
+
+#' Extrapolates the counts from a slide to the total count over the course of the
+#' experiment.
+#' @param @inheritParams correct_count
+#' @param vol numeric. Volume of starch added to the sample wells.
+#' @param days integer. Number of days starch was included in experiment treatments.
+#' @export
+
+extrap_count <- function(raw, prop, vol, days){
+  extrap <- raw * (1 / prop) * 100 * days
+  extrap
+}

### 2023-02-09 analysis/scripts/amylase-assay.R

@@ -20,107 +20,107 @@ library(ggplot2)
 
 # Functions ---------------------------------------------------------------
 
-#' Standard curve calculation
-#' @param std_data Data frame with the photometric readings of the standard curve.
-#' @param cols Vector containing the column numbers containing the standard curve photometric readings. If NULL, all columns assumed to be standard curve measurements.
-#' @param ml_maltose vector of length 7 containing the volume (ml) of 0.2% maltose solution added to the assay.
-#' @importFrom magrittr %>%
-#' @importFrom dplyr slice n
-stdcurve <- function(std_data, cols = NULL, ml_maltose = c(0.00375, 0.015, 0.030, 0.045, 0.060, 0.075, 0.150)){
-  if(!is.null(cols)){
-    std_data <- std_data[,cols]
-  }
-  # delta 540 calculation
-  mg_maltose <- 2 * ml_maltose # DOUBLE CHECK THIS
-  umol_maltose <- (mg_maltose * 1000) / 342.3 # calculate maltose in umoles
-  replicates <- dim(std_data)[2]
-  pht_blank <- std_data[8,] %>%
-    slice(rep(1:n(), each = 7))
-  pht_meas <- std_data[1:7,] - pht_blank
-
-  pht_meas <- c(pht_meas[, seq(1:replicates)]) %>%
-    unlist()
-  # Linear regression
-  stdCurve_data <- data.frame("d540" = pht_meas, "umol_maltose" = rep(umol_maltose, replicates))
-  std_curve <- lm(d540 ~ umol_maltose, stdCurve_data)
-  # Plot of standard curve
-  p_stdcurve <- stdCurve_data %>%
-    ggplot(aes(x = umol_maltose, y = d540)) +
-    geom_point() +
-    geom_smooth(formula = "y ~ x", method = "lm", se = F) +
-    theme_bw() +
-    labs(x = "umoles Maltose", y = "Absorbance (d540)")
-  out <- list("std_curve" = std_curve, "plot" = p_stdcurve, "blank" = pht_blank[1,], "replicates" = replicates, "model" = stdCurve_data)
-  return(out)
-}
-
-#' Amylase activity in sample calculation
-#' @param samp_data data frame with the photometric readings of the samples.
[…]

### 2023-02-09 analysis/scripts/analysis.R

@@ -19,17 +19,17 @@ library(tidyr)
 
 # function to correct counts for volume and proportion of slide counted
 
-correct_count <- function(raw, prop, vol){
-  corrected <- raw * (1 / prop) * (vol / 20)
-  corrected
-}
-
-# function to extrapolate solution counts to total exposure during the experiment
-  # i.e., 1 ml per day for 16 days
-extrap_count <- function(raw, prop, vol, days){
-  extrap <- raw * (1 / prop) * 100 * days
-  extrap
-}
+# correct_count <- function(raw, prop, vol){
+#   corrected <- raw * (1 / prop) * (vol / 20)
+#   corrected
+# }
+#
+# # function to extrapolate solution counts to total exposure during the experiment
+#   # i.e., 1 ml per day for 16 days
+# extrap_count <- function(raw, prop, vol, days){
+#   extrap <- raw * (1 / prop) * 100 * days
+#   extrap
+# }
 
 # Outlier detection and removal -------------------------------------------
 

### 2023-02-09 analysis/paper/paper.Rmd
add call to load functions

@@
 library(patchwork)
+devtools::load_all()
 plt1_ph1 <- readr::read_tsv(

### 2023-02-09 analysis/scripts/analysis.R
cleanup

@@ -15,21 +15,6 @@ library(tidyr)
 # starch counts in solutions
 #sol_counts <- readr::read_csv("../data/raw_data/solution_counts.csv")
 
-# Functions ---------------------------------------------------------------
-
-# function to correct counts for volume and proportion of slide counted
-
-# correct_count <- function(raw, prop, vol){
-#   corrected <- raw * (1 / prop) * (vol / 20)
-#   corrected
-# }
-#
-# # function to extrapolate solution counts to total exposure during the experiment
-#   # i.e., 1 ml per day for 16 days
-# extrap_count <- function(raw, prop, vol, days){
-#   extrap <- raw * (1 / prop) * 100 * days
-#   extrap
-# }
 
 # Outlier detection and removal -------------------------------------------
 

### 2023-05-31 analysis/paper/paper.Rmd
remove functions to import data in setup chunk

--- 2023-02-09 analysis/paper/paper.Rmd
@@
 library(patchwork)
+library(tidyr)
+#library(byocstarch)
+devtools::load_all()
+
 plt1_ph1 <- readr::read_tsv(
@@
 # starch counts in samples
-raw_counts <- readr::read_csv(here("analysis/data/raw_data/starch_counts.csv"))
+#raw_counts <- readr::read_csv(here("analysis/data/raw_data/starch_counts.csv"))
 # starch counts in solutions
-sol_counts <- readr::read_csv(here("analysis/data/raw_data/solution_counts.csv"))
+#sol_counts <- readr::read_csv(here("analysis/data/raw_data/solution_counts.csv"))
 # load R scripts
+#source(system.file("R", "analysis.R", package = "byocstarch"))
+#source("../../analysis/scripts/analysis.R")
+#source("../../analysis/scripts/amylase-assay.R")
+#source("../../analysis/scripts/figures.R")
+#source("../../analysis/scripts/tables.R")
 source(here("analysis/scripts/analysis.R"))
@@
 source(here("analysis/scripts/amylase-assay.R"))
-readr::write_csv(sol_corr, here("analysis/data/derived_data/sol_corr.csv"))
-readr::write_csv(corr_counts, here("analysis/data/derived_data/corr_counts.csv"))
-readr::write_csv(corr_comb, here("analysis/data/derived_data/corr_comb.csv"))
+#readr::write_csv(sol_corr, here("analysis/data/derived_data/sol_corr.csv"))
+#readr::write_csv(corr_counts, here("analysis/data/derived_data/corr_counts.csv"))
+#readr::write_csv(corr_comb, here("analysis/data/derived_data/corr_comb.csv"))
 

### 2023-05-31 analysis/scripts/analysis.R

@@ -2,11 +2,11 @@
 
 # Dependencies ------------------------------------------------------------
 
-library(here)
-library(dplyr)
-library(broom)
-library(tidyr)
-devtools::load_all()
+#library(here)
+#library(dplyr)
+#library(broom)
+#library(tidyr)
+#devtools::load_all()
 
 # combining potato and wheat counts from mixed treatment solution
 

### 2023-05-31 analysis/paper/discussion.Rmd
fix: broken citekeys

@@
 archaeological remains 
-[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010], which could in part be due to the 
+[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossil2010], which could in part be due to the 
 lack of oral amylase activity in our model. 
@@
 standard deviations and low proportional incorporation).
-Leonard and colleagues [-@leonardDentalCalculus2015] found individual
+Leonard and colleagues [-@leonardPlantMicroremains2015] found individual
 calculus samples to be a poor predictor of diet in a population, as many of the 
@@
 granules generally only reach up to 35 $\mu$m 
-[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
+[@gismondiStarchGranules2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
 Granule morphology may also play a role. Large wheat granules 
@@
 Based on the comparatively low counts obtained by Leonard
-and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
+and colleagues [-@leonardPlantMicroremains2015, Supplement 2], processing and amylase
 may have a substantial effect on starch granule retention in the oral cavity.

### 2023-05-31 analysis/paper/intro.Rmd
fix: broken citekeys

@@
 [@adlerSequencingAncient2013; @warinnerPathogensHost2014],
-microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
+microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranules2009;
 @mickleburghNewInsights2012], 

### 2023-05-31 analysis/paper/matmet.Rmd
fix: broken citekeys

@@
 protocol from Sissons and colleagues [-@sissonsMultistationPlaque1991] and
-Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
+Shellis [-@shellisSyntheticSaliva1978]. In brief, a biofilm inoculated
 with whole saliva was grown on a substrate suspended in artificial saliva, and
@@
 5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationPlaque1991], 6.4 mmol/l NaHCO~3~ 
-[@shellisSyntheticSalivaCultural1978], 2.5 mg/l haemin. This is subsequently 
+[@shellisSyntheticSaliva1978], 2.5 mg/l haemin. This is subsequently 
 adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120&deg;C, 
@@
 and (0.04 g/l MgCl)
-[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationPlaque1991].
+[@pearceConcomitantDeposition1987; @sissonsMultistationPlaque1991].
 The substrata were submerged in 1 ml/well CPMU for six minutes, five times 

### 2023-06-01 analysis/paper/matmet.Rmd
fix broken citekey

@@
 concentration in the calculus [as per 
-@wesolowskiEvaluatingMicrofossilContent2010].
+@wesolowskiEvaluatingMicrofossil2010].

### 2024-01-04 analysis/paper/fig_tab.Rmd
update dependencies

@@
                       message = FALSE)
-library(tidyverse)
 library(broom)

### 2024-01-04 analysis/scripts/analysis.R

@@ -6,6 +6,7 @@ library(here)
 library(dplyr)
 library(broom)
 library(tidyr)
+library(ggplot2)
 
 # Upload data -------------------------------------------------------------
 

### 2024-01-04 analysis/scripts/amylase-assay.R
add functions to R

@@ -17,111 +17,6 @@ library(ggplot2)
 #   "analysis/data/raw_data/ExpM_plate2_photometric2.csv",
 #   col_names = F)
 
-
-# Functions ---------------------------------------------------------------
-
-#' #' Standard curve calculation
-#' #' @param std_data Data frame with the photometric readings of the standard curve.
-#' #' @param cols Vector containing the column numbers containing the standard curve photometric readings. If NULL, all columns assumed to be standard curve measurements.
-#' #' @param ml_maltose vector of length 7 containing the volume (ml) of 0.2% maltose solution added to the assay.
-#' #' @importFrom magrittr %>%
-#' #' @importFrom dplyr slice n
-#' stdcurve <- function(std_data, cols = NULL, ml_maltose = c(0.00375, 0.015, 0.030, 0.045, 0.060, 0.075, 0.150)){
-#'   if(!is.null(cols)){
-#'     std_data <- std_data[,cols]
-#'   }
-#'   # delta 540 calculation
-#'   mg_maltose <- 2 * ml_maltose # DOUBLE CHECK THIS
-#'   umol_maltose <- (mg_maltose * 1000) / 342.3 # calculate maltose in umoles
-#'   replicates <- dim(std_data)[2]
-#'   pht_blank <- std_data[8,] %>%
-#'     slice(rep(1:n(), each = 7))
-#'   pht_meas <- std_data[1:7,] - pht_blank
-#'
-#'   pht_meas <- c(pht_meas[, seq(1:replicates)]) %>%
-#'     unlist()
-#'   # Linear regression
-#'   stdCurve_data <- data.frame("d540" = pht_meas, "umol_maltose" = rep(umol_maltose, replicates))
-#'   std_curve <- lm(d540 ~ umol_maltose, stdCurve_data)
-#'   # Plot of standard curve
-#'   p_stdcurve <- stdCurve_data %>%
-#'     ggplot(aes(x = umol_maltose, y = d540)) +
-#'     geom_point() +
-#'     geom_smooth(formula = "y ~ x", method = "lm", se = F) +
-#'     theme_bw() +
-#'     labs(x = "umoles Maltose", y = "Absorbance (d540)")
-#'   out <- list("std_curve" = std_curve, "plot" = p_stdcurve, "blank" = pht_blank[1,], "replicates" = replicates, "model" = stdCurve_data)
[…]

### 2024-01-04 analysis/scripts/analysis.R

@@ -12,14 +12,14 @@
 
 sol_long <- sol_corr %>%
   group_by(solution, starch) %>%
-  summarise(across(c(s,m,l, total), mean, na.rm = T)) %>%
+  summarise(across(c(s,m,l, total), \(x) mean(x, na.rm = T))) %>%
   pivot_longer(cols = c(s,m,l, total), values_to = "count", names_to = "size") %>%
   rename(treatment = solution) # rename to be consistent with sample counts
 
 sol_comb <- sol_corr %>%
   filter(solution != "mix") %>%
   group_by(solution, starch) %>%
-  summarise(across(c("s","m","l","total"), mean)) %>%
+  summarise(across(c("s","m","l","total"), \(x) mean(x))) %>%
   ungroup() %>%
   add_row(sol_corr[5:6,c("solution", "starch", "s","m","l","total")], ) %>%
   rename(treatment = solution) %>% # I should have done better naming the raw data...

### 2024-01-04 analysis/paper/paper.Rmd

@@
 title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
-#author: |
-# Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
-# date: |
-#  $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
 #### Frontiers submission ####
@@
 library(dplyr)
-library(ggplot2)
+#library(ggplot2)
 library(broom)
@@
   col_names = F)
-# starch counts in samples
-#raw_counts <- readr::read_csv(here("analysis/data/raw_data/starch_counts.csv"))
-# starch counts in solutions
-#sol_counts <- readr::read_csv(here("analysis/data/raw_data/solution_counts.csv"))
-# load R scripts
-#source(system.file("R", "analysis.R", package = "byocstarch"))
-#source("../../analysis/scripts/analysis.R")
-#source("../../analysis/scripts/amylase-assay.R")
-#source("../../analysis/scripts/figures.R")
-#source("../../analysis/scripts/tables.R")
+
 source(here("analysis/scripts/analysis.R"))
@@
 source(here("analysis/scripts/amylase-assay.R"))
-#readr::write_csv(sol_corr, here("analysis/data/derived_data/sol_corr.csv"))
-#readr::write_csv(corr_counts, here("analysis/data/derived_data/corr_counts.csv"))
-#readr::write_csv(corr_comb, here("analysis/data/derived_data/corr_comb.csv"))
 
-## Compile refs.bib using rbbt package ##
-#cite_keys <- rbbt::bbt_detect_citations(list.files(pattern = ".Rmd"))
-#cite_keys <- unlist(cite_keys, use.names = FALSE)
 rmd_files <- list.files(path = here("analysis/paper/"), pattern = "*.Rmd")
@@
 
-# ## Word count ##
-# wd_count <- wordcountaddin::word_count()
 cor_r <- signif(count_cor$estimate, 3)

### 2024-01-04 analysis/paper/paper.tex
add raw data to package structure

@@
 
-%%% Version 3.4 Generated 2018/06/15 %%%
+%%% Version 3.4 Generated 2022/06/14 %%%
 %%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
 %%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%
+%%%  NB logo1.jpg is required in the path in order to correctly compile front page header %%%
 
@@
 % Below is from frontiers
-%
-\bibliographystyle{frontiersinSCNS}
-% Use doublespacing - comment out for single spacing
-%\usepackage{setspace}
-%\doublespacing
 
@@
 \makesavenoteenv{longtable}
-
-% Pandoc citation processing
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 dietary information in the form of plant microfossils
-(Henry and Piperno, 2008; Hardy et al., 2009),
-proteins (Hendy et al., 2018; Warinner et al., 2014a),
-and other organic residues (Buckley et al., 2014).
+\citep{henryCalculusSyria2008, hardyStarchGranules2009},
+proteins \citep{hendyProteomicCalculus2018, warinnerEvidenceMilk2014},
+and other organic residues \citep{buckleyDentalCalculus2014}.
 This dietary information can be preserved within the mineralised dental plaque
@@
 past populations
-(Henry and Piperno, 2008; Jovanović et al., 2021; Tao et al., 2020)
-and extinct species (Henry et al., 2014; Hardy et al., 2012).
+\citep{henryCalculusSyria2008, jovanovicNeolithicCalculus2021, taoWheatCalculus2020}
+and extinct species \citep{henryNeanderthalCalculus2014, hardyNeanderthalMedics2012}.
 
@@
 from the dental calculus of archaeological fauna to investigate diet was first
-noted by Armitage (1975),
-and later by Middleton and Rovner (1994),
-and Fox and colleagues (1996). Starches and phytoliths were
-extracted from human dental calculus by Cummings and Magennis (1997).\\
+noted by Armitage \citeyearpar{armitageExtractionIdentification1975},
+and later by Middleton and Rovner \citeyearpar{middletonOpalPhytoliths1994},
+and Fox and colleagues \citeyearpar{foxPhytolithCalculus1996}. Starches and phytoliths were
+extracted from human dental calculus by Cummings and Magennis \citeyearpar{cummingsMayanCalculus1997}.\\
 In more recent years, the study of dental calculus has increased exponentially,
@@
 characterisation (including pathogens) through the analysis of DNA and proteins
-(Adler et al., 2013; Warinner et al., 2014b),
-microbotanical remains (Henry and Piperno, 2008; Hardy et al., 2009; Mickleburgh and Pagán-Jiménez, 2012),
+\citep{adlerSequencingAncient2013, warinnerPathogensHost2014},
+microbotanical remains \citep{henryCalculusSyria2008, hardyStarchGranules2009, mickleburghNewInsights2012},
 other organic residues and proteins from dietary compounds
-(Buckley et al., 2014; Hendy et al., 2018),
-and nicotine use (Eerkens et al., 2018). Especially the extraction
+\citep{buckleyDentalCalculus2014, hendyProteomicCalculus2018},
+and nicotine use \citep{eerkensDentalCalculus2018}. Especially the extraction
 of starch granules has become a rich source of dietary
@@
 over a variety of geographical and temporal ranges
-(Piperno and Dillehay, 2008; Henry et al., 2014; Tao et al., 2020; Jovanović et al., 2021).
+\citep{pipernoStarchGrains2008, henryNeanderthalCalculus2014, taoWheatCalculus2020, jovanovicNeolithicCalculus2021}.
 
 Despite this, our knowledge of dental calculus and the incorporation pathways of
-the various markers is limited (Radini et al., 2017), as is our
+the various markers is limited \citep{radiniFoodPathways2017}, as is our
 knowledge of information-loss caused by these pathways. Additionally, the methods
@@
 Extraction methods were tested
-by Tromp and colleagues (2017), specifically regarding
+by Tromp and colleagues \citeyearpar{trompEDTACalculus2017}, specifically regarding
 decalcification using HCl or EDTA.
@@
 within individuals.
-Studies conducted on modern humans (Leonard et al., 2015)
-and non-human primates (Power et al., 2015; Power et al., 2021)
+Studies conducted on modern humans \citep{leonardPlantMicroremains2015}
+and non-human primates \citep{powerChimpCalculus2015, powerRepresentativenessDental2021}
 have explored how well microremains (phytoliths and starches)
@@
 oral biofilm with a multifactorial aetiology and variable formation rates both
-within and between individuals (Jepsen et al., 2011; Haffajee et al., 2009),
+within and between individuals \citep{jepsenCalculusRemoval2011, haffajeeBiofilmPosition2009},
 contributing to
@@
 Additionally, the concentration of oral \(\alpha\)-amylase differs both between and
-within individuals (Froehlich et al., 1987; Nater et al., 2005),
+within individuals \citep{froehlichEffectOral1987, naterHumanAmylase2005},
 causing different rates of hydrolysis of the starch granules present in the oral
 cavity. Add to this the effects of the many different methods
-of starch processing (Hardy et al., 2018), as well as post-depositional
+of starch processing \citep{hardyRecoveringInformation2018}, as well as post-depositional
 processes that are still being explored
-(García-Granero, 2020; Mercader et al., 2018),
+\citep{graneroStarchTaphonomy2020, mercaderExaggeratedExpectations2018},
 and it becomes clear that using
@@
 
-\hypertarget{materials-and-methods}{%
-\section{Materials and Methods}\label{materials-and-methods}}
-
-\hypertarget{biofilm-formation}{%
-\subsection{Biofilm formation}\label{biofilm-formation}}
+\section{Materials and Methods}\label{materials-and-methods}
+
+\subsubsection{Biofilm formation}\label{biofilm-formation}
 
 In this study we employ a multispecies oral biofilm model following a modified
-protocol from Sissons and colleagues (1991) and
-Shellis (1978). In brief, a biofilm inoculated
+protocol from Sissons and colleagues \citeyearpar{sissonsMultistationPlaque1991} and
+Shellis \citeyearpar{shellisSyntheticSaliva1978}. In brief, a biofilm inoculated
 with whole saliva was grown on a substrate suspended in artificial saliva, and
@@
 The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
-described by Sissons and colleagues (1991).
+described by Sissons and colleagues \citeyearpar{sissonsMultistationPlaque1991}.
 It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778),
@@
 5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl\textsubscript{2},
-5.2 mmol/l Na\textsubscript{2}HPO\textsubscript{4} (Sissons et al., 1991), 6.4 mmol/l NaHCO\textsubscript{3}
-(Shellis, 1978), 2.5 mg/l haemin. This is subsequently
+5.2 mmol/l Na\textsubscript{2}HPO\textsubscript{4} \citep{sissonsMultistationPlaque1991}, 6.4 mmol/l NaHCO\textsubscript{3}
+\citep{shellisSyntheticSaliva1978}, 2.5 mg/l haemin. This is subsequently
 adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120°C,
 1 bar overpressure), and supplemented with 5.8 \(\mu\)mol/l menadione, 5 mmol/l urea,
-and 1 mmol/l arginine (Sissons et al., 1991).
+and 1 mmol/l arginine \citep{sissonsMultistationPlaque1991}.
 
@@
 and (0.04 g/l MgCl)
-(Pearce and Sissons, 1987; Sissons et al., 1991).
+\citep{pearceConcomitantDeposition1987, sissonsMultistationPlaque1991}.
 The substrata were submerged in 1 ml/well CPMU for six minutes, five times
@@
 
-\hypertarget{amylase-activity-detection}{%
-\subsection{Amylase activity detection}\label{amylase-activity-detection}}
+\subsubsection{Amylase activity detection}\label{amylase-activity-detection}
 
@@
 The protocol is a modified version of an Enzymatic Assay of \(\alpha\)-Amylase
-(\url{https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase}) (Bernfeld, 1955), which measures the amount of
+(\url{https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase}) \citep{bernfeldAmylase1955}, which measures the amount of
 maltose released from starch by \(\alpha\)-amylase activity. Results are reported
@@
 
-\hypertarget{treatment-solutions}{%
-\subsection{Treatment solutions}\label{treatment-solutions}}
+\subsubsection{Treatment solutions}\label{treatment-solutions}
 
@@
 
-\hypertarget{extraction-method}{%
-\subsection{Extraction method}\label{extraction-method}}
+\subsubsection{Extraction method}\label{extraction-method}
 
@@
 calculus in 0.5 \(\tiny{M}\) ethylenediaminetetraacetic acid (EDTA)
-(Tromp et al., 2017; Le Moyne and Crowther, 2021; Modi et al., 2020),
+\citep{trompEDTACalculus2017, lemoyneCalculusPretreatments2021, modiCalculusMethodologies2020},
 and vortexing for 3 days until the sample was completely dissolved.
@@
 
-\hypertarget{statistical-analysis}{%
-\subsection{Statistical analysis}\label{statistical-analysis}}
-
-Statistical analysis was conducted in R version 4.2.2 Patched (2022-11-10 r83330) (R Core Team, 2020) and
-the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
-here (Müller, 2020), and patchwork (Pedersen, 2020).
+\subsubsection{Statistical analysis}\label{statistical-analysis}
+
+Statistical analysis was conducted in R version 4.3.2 (2023-10-31) \citep{Rbase} and
+the following packages: tidyverse \citep{tidyverse2019}, broom \citep{Rbroom},
+here \citep{Rhere}, and patchwork \citep{Rpatchwork}.
 
@@
 calculus (also z-score standardised) to account for differences in starch
-concentration in the calculus (as per Wesolowski et al., 2010).
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
+concentration in the calculus \citep[as per][]{wesolowskiEvaluatingMicrofossil2010}.
+
+\section{Results}\label{results}
 
@@
 
-\hypertarget{no-amylase-activity-detected-in-the-model}{%
-\subsection{No amylase activity detected in the model}\label{no-amylase-activity-detected-in-the-model}}
+\subsubsection{No amylase activity detected in the model}\label{no-amylase-activity-detected-in-the-model}
 
@@
 
-\hypertarget{treatment-type-had-minimal-effect-on-biofilm-growth}{%
-\subsection{Treatment type had minimal effect on biofilm growth}\label{treatment-type-had-minimal-effect-on-biofilm-growth}}
+\subsubsection{Treatment type had minimal effect on biofilm growth}\label{treatment-type-had-minimal-effect-on-biofilm-growth}
 
@@
 
-\hypertarget{starch-counts}{%
-\subsection{Starch counts}\label{starch-counts}}
+\subsubsection{Starch counts}\label{starch-counts}
 
@@
 
-\hypertarget{proportion-of-available-starches-incorporated-in-samples}{%
-\subsubsection{Proportion of available starches incorporated in samples}\label{proportion-of-available-starches-incorporated-in-samples}}
+\paragraph{Proportion of available starches incorporated in samples}\label{proportion-of-available-starches-incorporated-in-samples}
 
@@
 
-\hypertarget{size-ratios-differ-between-solutions-and-samples}{%
-\subsubsection{Size ratios differ between solutions and samples}\label{size-ratios-differ-between-solutions-and-samples}}
+\paragraph{Size ratios differ between solutions and samples}\label{size-ratios-differ-between-solutions-and-samples}
 
@@
 
-\hypertarget{biofilm-weight-correlated-positively-with-extracted-starch-counts}{%
-\subsubsection{Biofilm weight correlated positively with extracted starch counts}\label{biofilm-weight-correlated-positively-with-extracted-starch-counts}}
+\paragraph{Biofilm weight correlated positively with extracted starch counts}\label{biofilm-weight-correlated-positively-with-extracted-starch-counts}
 
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\section{Discussion}\label{discussion}
 
@@
 archaeological remains
-(Tromp and Dudgeon, 2015; Tromp et al., 2017; Wesolowski et al., 2010), which could in part be due to the
+\citep{trompDietaryNondietary2015, trompEDTACalculus2017, wesolowskiEvaluatingMicrofossil2010}, which could in part be due to the
 lack of oral amylase activity in our model.
@@
 standard deviations and low proportional incorporation).
-Leonard and colleagues (2015) found individual
+Leonard and colleagues \citeyearpar{leonardPlantMicroremains2015} found individual
 calculus samples to be a poor predictor of diet in a population, as many of the
 consumed plants were missing from some individual samples, but were present in others.\\
-Power and colleagues (2015)
+Power and colleagues \citeyearpar{powerChimpCalculus2015}
 presented similar findings in non-human primates, where phytoliths were more
@@
 The size bias is also consistent with the findings by Power and colleagues
-(2015),
+\citeyearpar{powerChimpCalculus2015},
 who found that plants producing starches 10--20 \(\mu\)m in
@@
 granules generally only reach up to 35 \(\mu\)m
-(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, 174--176).
+\citep[174-176]{gismondiStarchGranules2019, haslamDecompositionStarch2004, seidemannStarchAtlas1966}.
 Granule morphology may also play a role. Large wheat granules
@@
 compared to surface area
-(van de Velde et al., 2002; Jane et al., 1994; Reichert, 1913, 364--365; Seidemann, 1966, 174--176).
+\citetext{\citealp{vandeveldeStarchMorphology2002}; \citealp{janeAnthologyStarch1994}; \citealp[364-365]{reichertStarchBible1913b}; \citealp[174-176]{seidemannStarchAtlas1966}}.
 Another potentially important factor
@@
 a result that contradicts findings from archaeological contexts
-(Wesolowski et al., 2010; Dudgeon and Tromp, 2014).
+\citep{wesolowskiEvaluatingMicrofossil2010, dudgeonDietGeography2014}.
 When the concentration of starch granules
@@
 archaeological samples of considerable age
-(Henry et al., 2014; Buckley et al., 2014; Wu et al., 2021).
+\citep{henryNeanderthalCalculus2014, buckleyDentalCalculus2014, wuDietEarliest2021}.
 Studies have also shown that not all starch granules come from a dietary source.
@@
 as palm phytoliths adhering to the skin of sweet potatoes
-(Tromp and Dudgeon, 2015),
+\citep{trompDietaryNondietary2015},
 or accidental ingestion not related to food consumption
-(Radini et al., 2017; Radini et al., 2019).\\
+\citep{radiniFoodPathways2017, radiniMedievalWomen2019}.\\
 When starch granules enter the mouth, whether through ingestion of food or accidental
@@
 tooth surfaces (including plaque and dental calculus) and mucous membranes
-(Kashket et al., 1991; Dodds and Edgar, 1988).
+\citep{kashketFoodRetention1991, doddsCarbohydrateRetention1988}.
 Bacteria also have the ability to adhere to starch granules
-(Topping et al., 2003),
+\citep{toppingResistantStarch2003},
 which would allow starches to attach to bacterial communities within the biofilm.
@@
 susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
-by \(\alpha\)-amylase (Kashket et al., 1996).
+by \(\alpha\)-amylase \citep{kashketFoodParticles1996}.
 The susceptibility of granules to hydrolysis depends on the crystallinity and size
@@
 while dehydrated starches will have a reduced susceptibility
-(Lingstrom et al., 1994; Franco et al., 1992; Haslam, 2004; Björck et al., 1984; Henry et al., 2009).
+\citep{lingstromStarchyFood1994, francoStarchDegradation1992, haslamDecompositionStarch2004, bjorckStarchProcessing1984, henryCookingStarch2009}.
 Cracks on the surface of the dental calculus, as well as unmineralised islands
 and channels may also be able to contain starch granules
-(Power et al., 2014; Tan et al., 2004a; Charlier et al., 2010).
+\citep{powerSEMCalculus2014, tanCalculusUltrastructure2004, charlierSEMCalculus2010}.
 Starch granules that are trapped in these pockets are (at least to some extent)
@@
 larger granules, and can be stored in larger quantities.
-This was also suggested by Power and colleagues (2014), who
+This was also suggested by Power and colleagues \citeyearpar{powerSEMCalculus2014}, who
 observed clusters of starches within dental calculus, rather than an even
@@
 a portion of its hydrolytic activity
-(Scannapieco et al., 1993; Nikitkova et al., 2013; Tan et al., 2004b, 2004a).
+\citep{scannapiecoSalivaryAmylase1993, nikitkovaStarchBiofilms2013, tanBacterialViability2004, tanCalculusUltrastructure2004}.
 After the death of an individual, starches within dental calculus are susceptible
@@
 (pH, temperature, moisture content, microorganisms)
-(Henry et al., 2009; García-Granero, 2020; Haslam, 2004; Franco et al., 1992).
+\citep{henryCookingStarch2009, graneroStarchTaphonomy2020, haslamDecompositionStarch2004, francoStarchDegradation1992}.
 Future study should explore how burial affects the recovery of starch from the
@@
 susceptible to hydrolysis
-(Franco et al., 1992; Haslam, 2004). The absence may
+\citep{francoStarchDegradation1992, haslamDecompositionStarch2004}. The absence may
 also affect biofilm growth due to the lack of amylase-bacterium interactions
-(Nikitkova et al., 2013). Conversely, the model may benefit from the
+\citep{nikitkovaStarchBiofilms2013}. Conversely, the model may benefit from the
 absence of \(\alpha\)-amylase, because it can allow us to directly explore
@@
 added to the model in concentrations similar to those found in the oral cavity
-(Scannapieco et al., 1993).
+\citep{scannapiecoSalivaryAmylase1993}.
 We are able to show how absolute counts in the treatments cause a difference in
@@
 negligible morphological changes caused by exposure to EDTA
-(Tromp et al., 2017; Le Moyne and Crowther, 2021; Modi et al., 2020),
+\citep{trompEDTACalculus2017, lemoyneCalculusPretreatments2021, modiCalculusMethodologies2020},
 these studies have not considered changes to separate size
@@
 Finally, we only used native starches in the experimental procedure and the results
-will likely differ for processed starches (García-Granero, 2020).
+will likely differ for processed starches \citep{graneroStarchTaphonomy2020}.
 Based on the comparatively low counts obtained by Leonard
-and colleagues (2015, Supplement 2), processing and amylase
+and colleagues \citeyearpar[Supplement 2]{leonardPlantMicroremains2015}, processing and amylase
 may have a substantial effect on starch granule retention in the oral cavity.
@@
 baseline testing of biases associated with dietary research conducted on dental calculus
-(Le Moyne and Crowther, 2021).
+\citep{lemoyneCalculusPretreatments2021}.
 Our high-throughput experimental setup allows us a
@@
 morphology
-(Jepsen et al., 2011; Simón-Soro et al., 2013; Proctor et al., 2018; Fagernäs et al., 2021; Haffajee et al., 2009),
-as well as evolutionary differences (Fellows Yates et al., 2021). The set of
+\citep{jepsenCalculusRemoval2011, simonsoroOralGeography2013, proctorSpatialGradient2018, fagernasMicrobialBiogeography2021, haffajeeBiofilmPosition2009},
+as well as evolutionary differences \citep{yatesOralMicrobiome2021}. The set of
 limitations for our model differ from \emph{in vivo} methods and, as such, we expect
@@
 
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
+\section{Conclusions}\label{conclusions}
 
@@
 
-\hypertarget{open-science-statement}{%
-\section*{Open Science Statement}\label{open-science-statement}}
+\section*{Open Science Statement}\label{open-science-statement}
 \addcontentsline{toc}{section}{Open Science Statement}
@@
 (\url{https://osf.io/uc5qy/}) and Github (\url{https://github.com/bbartholdy/byoc-starch}),
-following the format provided by the rrtools package (Marwick, 2019).
+following the format provided by the rrtools package \citep{Rrrtools}.
 More detailed protocols are available on OSF (\url{https://osf.io/akevs/})
@@
 
-\hypertarget{author-contributions}{%
-\section*{Author Contributions}\label{author-contributions}}
+\section*{Author Contributions}\label{author-contributions}
 \addcontentsline{toc}{section}{Author Contributions}
@@
 
-\hypertarget{funding}{%
-\section*{Funding}\label{funding}}
+\section*{Funding}\label{funding}
 \addcontentsline{toc}{section}{Funding}
@@
 
-\hypertarget{conflict-of-interest}{%
-\section*{Conflict of Interest}\label{conflict-of-interest}}
+\section*{Conflict of Interest}\label{conflict-of-interest}
 \addcontentsline{toc}{section}{Conflict of Interest}
@@
 
-\hypertarget{acknowledgements}{%
-\section*{Acknowledgements}\label{acknowledgements}}
+\section*{Acknowledgements}\label{acknowledgements}
 \addcontentsline{toc}{section}{Acknowledgements}
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-adlerSequencingAncient2013}{}}%
-Adler, C.
-J., Dobney, K., Weyrich, L.
-S., Kaidonis, J., Walker, A.
-W., Haak, W., Bradshaw, C.
-J., Townsend, G., Sołtysiak, A., Alt, K.
-W., et al.
-(2013).
-Sequencing ancient calcified dental plaque shows changes in oral microbiota with dietary shifts of the {Neolithic} and {Industrial} revolutions.
-\emph{Nat Genet} 45, 450--5, 455e1.
-doi:\href{https://doi.org/10.1038/ng.2536}{10.1038/ng.2536}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-armitageExtractionIdentification1975}{}}%
-Armitage, P. L. (1975). The {Extraction} and {Identification} of {Opal Phytoliths} from the {Teeth} of {Ungulates}. \emph{J Archaeol Sci} 2, 187--197.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bernfeldAmylase1955}{}}%
-Bernfeld, P. (1955). {``Amylases, α and β,''} in \emph{Methods in {Enzymology}} ({Academic Press}), 149--158. doi:\href{https://doi.org/10.1016/0076-6879(55)01021-5}{10.1016/0076-6879(55)01021-5}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bjorckStarchProcessing1984}{}}%
-Björck, I., Asp, N.-G., Birkhed, D., Eliasson, A.-C., Sjöberg, L.-B., and Lundquist, I.
-(1984).
-Effects of processing on starch availability {In} vitro and {In} vivo.
-{II}.
-{Drum-drying} of wheat flour.
-\emph{Journal of Cereal Science} 2, 165--178.
-doi:\href{https://doi.org/10.1016/S0733-5210(84)80030-2}{10.1016/S0733-5210(84)80030-2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-buckleyDentalCalculus2014}{}}%
-Buckley, S., Usai, D., Jakob, T., Radini, A., and Hardy, K.
-(2014).
-Dental {Calculus Reveals Unique Insights} into {Food Items}, {Cooking} and {Plant Processing} in {Prehistoric Central Sudan}.
-\emph{PLOS ONE} 9, e100808.
-doi:\href{https://doi.org/10.1371/journal.pone.0100808}{10.1371/journal.pone.0100808}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-charlierSEMCalculus2010}{}}%
-Charlier, P., Huynh-Charlier, I., Munoz, O., Billard, M., Brun, L., and Grandmaison, G.
-L.
-de la (2010).
-The microscopic (optical and {SEM}) examination of dental calculus deposits ({DCD}).
-{Potential} interest in forensic anthropology of a bio-archaeological method.
-\emph{Legal Medicine} 12, 163--171.
-doi:\href{https://doi.org/10.1016/j.legalmed.2010.03.003}{10.1016/j.legalmed.2010.03.003}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cummingsMayanCalculus1997}{}}%
-Cummings, L.
-S., and Magennis, A.
-(1997).
-{``A phytolith and starch record of food and grit in {Mayan} human tooth tartar,''} in \emph{The {State-of-the-Art} of {Phytoliths} in {Soils} and {Plants}}, eds.
-A.
-Pinilla, J.
-Juan-Tresserras, and M.
-J.
-Machado ({Spain}: {CSIC Press}).
-Available at: \url{https://books.google.com?id=j66CDVfVhwEC}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-doddsCarbohydrateRetention1988}{}}%
-Dodds, M.
-W.
-J., and Edgar, W.
-M.
-(1988).
-The {Relationship Between Plaque pH}, {Plaque Acid Anion Profiles}, and {Oral Carbohydrate Retention After Ingestion} of {Several} '{Reference Foods}' by {Human Subjects}.
-\emph{J Dent Res} 67, 861--865.
-doi:\href{https://doi.org/10.1177/00220345880670051301}{10.1177/00220345880670051301}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-dudgeonDietGeography2014}{}}%
-Dudgeon, J.
-V., and Tromp, M.
-(2014).
-Diet, {Geography} and {Drinking Water} in {Polynesia}: {Microfossil Research} from {Archaeological Human Dental Calculus}, {Rapa Nui} ({Easter Island}).
-\emph{International Journal of Osteoarchaeology} 24, 634--648.
-doi:\href{https://doi.org/10.1002/oa.2249}{10.1002/oa.2249}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-eerkensDentalCalculus2018}{}}%
-Eerkens, J.
-W., Tushingham, S., Brownstein, K.
-J., Garibay, R., Perez, K., Murga, E., Kaijankoski, P., Rosenthal, J.
-S., and Gang, D.
-R.
-(2018).
-Dental calculus as a source of ancient alkaloids: {Detection} of nicotine by {LC-MS} in calculus samples from the {Americas}.
-\emph{J Archaeol Sci Rep} 18, 509--515.
-doi:\href{https://doi.org/10.1016/j.jasrep.2018.02.004}{10.1016/j.jasrep.2018.02.004}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-fagernasMicrobialBiogeography2021}{}}%
-Fagernäs, Z., Salazar-García, D.
-C., Avilés, A., Haber, M., Henry, A., Maurandi, J.
-L., Ozga, A., Velsko, I.
-M., and Warinner, C.
-(2021).
-Understanding the microbial biogeography of ancient human dentitions to guide study design and interpretation.
-\emph{bioRxiv}, 2021.08.16.456492.
-doi:\href{https://doi.org/10.1101/2021.08.16.456492}{10.1101/2021.08.16.456492}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-yatesOralMicrobiome2021}{}}%
-Fellows Yates, J.
-A., Velsko, I.
-M., Aron, F., Posth, C., Hofman, C.
-A., Austin, R.
-M., Parker, C.
-E., Mann, A.
-E., Nägele, K., Arthur, K.
-W., et al.
-(2021).
-The evolution and changing ecology of the {African} hominid oral microbiome.
-\emph{PNAS} 118.
-doi:\href{https://doi.org/10.1073/pnas.2021655118}{10.1073/pnas.2021655118}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-foxPhytolithCalculus1996}{}}%
-Fox, C.
-L., Juan, J., and Albert, R.
-M.
-(1996).
-Phytolith analysis on dental calculus, enamel surface, and burial soil: {Information} about diet and paleoenvironment.
-\emph{American Journal of Physical Anthropology} 101, 101--113.
-doi:\href{https://doi.org/10.1002/(SICI)1096-8644(199609)101:1\%3C101::AID-AJPA7\%3E3.0.CO;2-Y}{10.1002/(SICI)1096-8644(199609)101:1\textless101::AID-AJPA7\textgreater3.0.CO;2-Y}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-francoStarchDegradation1992}{}}%
-Franco, C.
-M.
-L., Preto, S.
-J.
-do R., and Ciacco, C.
-F.
-(1992).
-Factors that {Affect} the {Enzymatic Degradation} of {Natural Starch Granules} -{Effect} of the {Size} of the {Granules}.
-\emph{Starch - Stärke} 44, 422--426.
-doi:\href{https://doi.org/10.1002/star.19920441106}{10.1002/star.19920441106}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-froehlichEffectOral1987}{}}%
-Froehlich, D.
-A., Pangborn, R.
-M., and Whitaker, J.
-R.
-(1987).
-The effect of oral stimulation on human parotid salivary flow rate and alpha-amylase secretion.
-\emph{Physiology \& Behavior} 41, 209--217.
-doi:\href{https://doi.org/10.1016/0031-9384(87)90355-6}{10.1016/0031-9384(87)90355-6}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-graneroStarchTaphonomy2020}{}}%
-García-Granero, J.
-J.
-(2020).
-Starch taphonomy, equifinality and the importance of context: {Some} notes on the identification of food processing through starch grain analysis.
-\emph{Journal of Archaeological Science} 124, 105267.
-doi:\href{https://doi.org/10.1016/j.jas.2020.105267}{10.1016/j.jas.2020.105267}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gismondiStarchGranulesData2019}{}}%
-Gismondi, A., D'Agostino, A., Canuti, L., Di Marco, G., Basoli, F., and Canini, A.
-(2019).
-Starch granules: A data collection of 40 food species.
-\emph{Plant Biosystems - An International Journal Dealing with all Aspects of Plant Biology} 153, 273--279.
-doi:\href{https://doi.org/10.1080/11263504.2018.1473523}{10.1080/11263504.2018.1473523}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-haffajeeBiofilmPosition2009}{}}%
-Haffajee, A.
-D., Teles, R.
-P., Patel, M.
-R., Song, X., Yaskell, T., and Socransky, S.
-S.
-(2009).
-Factors affecting human supragingival biofilm composition.
-{II}.
-{Tooth} position.
-\emph{Journal of Periodontal Research} 44, 520--528.
-doi:\href{https://doi.org/10.1111/j.1600-0765.2008.01155.x}{10.1111/j.1600-0765.2008.01155.x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hardyStarchGranulesDental2009}{}}%
-Hardy, K., Blakeney, T., Copeland, L., Kirkham, J., Wrangham, R., and Collins, M.
-(2009).
-Starch granules, dental calculus and new perspectives on ancient diet.
-\emph{Journal of Archaeological Science} 36, 248--255.
-doi:\href{https://doi.org/10.1016/j.jas.2008.09.015}{10.1016/j.jas.2008.09.015}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hardyNeanderthalMedics2012}{}}%
-Hardy, K., Buckley, S., Collins, M.
-J., Estalrrich, A., Brothwell, D., Copeland, L., García-Tabernero, A., García-Vargas, S., de la Rasilla, M., Lalueza-Fox, C., et al.
-(2012).
-Neanderthal medics?
-{Evidence} for food, cooking, and medicinal plants entrapped in dental calculus.
-\emph{Naturwissenschaften} 99, 617--626.
-doi:\href{https://doi.org/10.1007/s00114-012-0942-0}{10.1007/s00114-012-0942-0}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hardyRecoveringInformation2018}{}}%
-Hardy, K., Buckley, S., and Copeland, L.
-(2018).
-Pleistocene dental calculus: {Recovering} information on {Paleolithic} food items, medicines, paleoenvironment and microbes.
-\emph{Evol Anthropol} 27, 234--246.
-doi:\href{https://doi.org/10.1002/evan.21718}{10.1002/evan.21718}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-haslamDecompositionStarch2004}{}}%
-Haslam, M.
-(2004).
-The decomposition of starch grains in soils: Implications for archaeological residue analyses.
-\emph{Journal of Archaeological Science} 31, 1715--1734.
-doi:\href{https://doi.org/10.1016/j.jas.2004.05.006}{10.1016/j.jas.2004.05.006}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hendyProteomicCalculus2018}{}}%
-Hendy, J., Warinner, C., Bouwman, A., Collins, M.
-J., Fiddyment, S., Fischer, R., Hagan, R., Hofman, C.
-A., Holst, M., Chaves, E., et al.
-(2018).
-Proteomic evidence of dietary sources in ancient dental calculus.
-\emph{Proc Biol Sci} 285, 20180977.
-doi:\href{https://doi.org/10.1098/rspb.2018.0977}{10.1098/rspb.2018.0977}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-henryNeanderthalCalculus2014}{}}%
-Henry, A.
-G., Brooks, A.
-S., and Piperno, D.
-R.
-(2014).
-Plant foods and the dietary ecology of {Neanderthals} and early modern humans.
-\emph{Journal of Human Evolution} 69, 44--54.
-doi:\href{https://doi.org/10.1016/j.jhevol.2013.12.014}{10.1016/j.jhevol.2013.12.014}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-henryCookingStarch2009}{}}%
-Henry, A.
-G., Hudson, H.
-F., and Piperno, D.
-R.
-(2009).
-Changes in starch grain morphologies from cooking.
-\emph{Journal of Archaeological Science} 36, 915--922.
-doi:\href{https://doi.org/10.1016/j.jas.2008.11.008}{10.1016/j.jas.2008.11.008}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-henryCalculusSyria2008}{}}%
-Henry, A.
-G., and Piperno, D.
-R.
-(2008).
-Using plant microfossils from dental calculus to recover human diet: A case study from {Tell} al-{Raqā}'i, {Syria}.
-\emph{Journal of Archaeological Science} 35, 1943--1950.
-doi:\href{https://doi.org/10.1016/j.jas.2007.12.005}{10.1016/j.jas.2007.12.005}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-janeAnthologyStarch1994}{}}%
-Jane, J.-L., Kasemsuwan, T., Leas, S., Zobel, H., and Robyt, J.
-F.
-(1994).
-Anthology of {Starch Granule Morphology} by {Scanning Electron Microscopy}.
-\emph{Starch/Stärke} 46, 121--129.
-doi:\href{https://doi.org/10.1002/star.19940460402}{10.1002/star.19940460402}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jepsenCalculusRemoval2011}{}}%
-Jepsen, S., Deschner, J., Braun, A., Schwarz, F., and Eberhard, J.
-(2011).
-Calculus removal and the prevention of its formation.
-\emph{Periodontology 2000} 55, 167--188.
-doi:\href{https://doi.org/10.1111/j.1600-0757.2010.00382.x}{10.1111/j.1600-0757.2010.00382.x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jovanovicNeolithicCalculus2021}{}}%
-Jovanović, J., Power, R.
-C., de Becdelièvre, C., Goude, G., and Stefanović, S.
-(2021).
-Microbotanical evidence for the spread of cereal use during the {Mesolithic-Neolithic} transition in the {Southeastern Europe} ({Danube Gorges}): {Data} from dental calculus analysis.
-\emph{Journal of Archaeological Science} 125, 105288.
-doi:\href{https://doi.org/10.1016/j.jas.2020.105288}{10.1016/j.jas.2020.105288}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kashketFoodRetention1991}{}}%
-Kashket, S., Van Houte, J., Lopez, L.
-R., and Stocks, S.
-(1991).
-Lack of {Correlation Between Food Retention} on the {Human Dentition} and {Consumer Perception} of {Food Stickiness}.
-\emph{J Dent Res} 70, 1314--1319.
-doi:\href{https://doi.org/10.1177/00220345910700100101}{10.1177/00220345910700100101}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kashketFoodParticles1996}{}}%
-Kashket, S., Zhang, J., and Houte, J. V. (1996). Accumulation of {Fermentable Sugars} and {Metabolic Acids} in {Food Particles} that {Become Entrapped} on the {Dentition}. \emph{J Dent Res}, 8.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lemoyneCalculusPretreatments2021}{}}%
-Le Moyne, C., and Crowther, A.
-(2021).
-Effects of chemical pre-treatments on modified starch granules: {Recommendations} for dental calculus decalcification for ancient starch research.
-\emph{Journal of Archaeological Science: Reports} 35, 102762.
-doi:\href{https://doi.org/10.1016/j.jasrep.2020.102762}{10.1016/j.jasrep.2020.102762}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-leonardDentalCalculus2015}{}}%
-Leonard, C., Vashro, L., O'Connell, J.
-F., and Henry, A.
-G.
-(2015).
-Plant microremains in dental calculus as a record of plant consumption: {A} test with {Twe} forager-horticulturalists.
-\emph{J Archaeol Sci Rep} 2, 449--457.
-doi:\href{https://doi.org/10.1016/j.jasrep.2015.03.009}{10.1016/j.jasrep.2015.03.009}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lingstromStarchyFood1994}{}}%
-Lingstrom, P., Birkhed, D., Ruben, J., and Arends, J.
-(1994).
-Effect of {Frequent Consumption} of {Starchy Food Items} on {Enamel} and {Dentin Demineralization} and on {Plaque pH} in situ.
-\emph{J Dent Res} 73, 652--660.
-doi:\href{https://doi.org/10.1177/00220345940730031101}{10.1177/00220345940730031101}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Rrrtools}{}}%
-Marwick, B. (2019). \emph{Rrtools: {Creates} a reproducible research compendium}. Available at: \url{https://github.com/benmarwick/rrtools}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mercaderExaggeratedExpectations2018}{}}%
-Mercader, J., Akeju, T., Brown, M., Bundala, M., Collins, M.
-J., Copeland, L., Crowther, A., Dunfield, P., Henry, A., Inwood, J., et al.
-(2018).
-Exaggerated expectations in ancient starch research and the need for new taphonomic and authenticity criteria.
-\emph{FACETS} 3, 777--798.
-doi:\href{https://doi.org/10.1139/facets-2017-0126}{10.1139/facets-2017-0126}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mickleburghNewInsights2012}{}}%
-Mickleburgh, H.
-L., and Pagán-Jiménez, J.
-R.
-(2012).
-New insights into the consumption of maize and other food plants in the pre-{Columbian Caribbean} from starch grains trapped in human dental calculus.
-\emph{Journal of Archaeological Science} 39, 2468--2478.
-doi:\href{https://doi.org/10.1016/j.jas.2012.02.020}{10.1016/j.jas.2012.02.020}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-middletonOpalPhytoliths1994}{}}%
-Middleton, W.
-D., and Rovner, I.
-(1994).
-Extraction of {Opal Phytoliths} from {Herbivore Dental Calculus}.
-\emph{J Archaeol Sci} 21, 469--473.
-doi:\href{https://doi.org/10.1006/jasc.1994.1046}{10.1006/jasc.1994.1046}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-modiCalculusMethodologies2020}{}}%
-Modi, A., Pisaneschi, L., Zaro, V., Vai, S., Vergata, C., Casalone, E., Caramelli, D., Moggi-Cecchi, J., Mariotti Lippi, M., and Lari, M.
-(2020).
-Combined methodologies for gaining much information from ancient dental calculus: Testing experimental strategies for simultaneously analysing {DNA} and food residues.
-\emph{Archaeol Anthropol Sci} 12, 10.
-doi:\href{https://doi.org/10.1007/s12520-019-00983-5}{10.1007/s12520-019-00983-5}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Rhere}{}}%
-Müller, K. (2020). \emph{Here: {A} simpler way to find your files}. Available at: \url{https://CRAN.R-project.org/package=here}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-naterHumanAmylase2005}{}}%
-Nater, U.
-M., Rohleder, N., Gaab, J., Berger, S., Jud, A., Kirschbaum, C., and Ehlert, U.
-(2005).
-Human salivary alpha-amylase reactivity in a psychosocial stress paradigm.
-\emph{International Journal of Psychophysiology} 55, 333--342.
-doi:\href{https://doi.org/10.1016/j.ijpsycho.2004.09.009}{10.1016/j.ijpsycho.2004.09.009}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nikitkovaStarchBiofilms2013}{}}%
-Nikitkova, A.
-E., Haase, E.
-M., and Scannapieco, F.
-A.
-(2013).
-Taking the {Starch} out of {Oral Biofilm Formation}: {Molecular Basis} and {Functional Significance} of {Salivary} α-{Amylase Binding} to {Oral Streptococci}.
-\emph{Appl.
-Environ.
-Microbiol.} 79, 416--423.
-doi:\href{https://doi.org/10.1128/AEM.02581-12}{10.1128/AEM.02581-12}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pearceConcomitantDepositionStrontium1987}{}}%
-Pearce, E.
-I.
-F., and Sissons, C.
-H.
-(1987).
-The {Concomitant Deposition} of {Strontium} and {Fluoride} in {Dental Plaque}.
-\emph{J Dent Res} 66, 1518--1522.
-doi:\href{https://doi.org/10.1177/00220345870660100101}{10.1177/00220345870660100101}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Rpatchwork}{}}%
-Pedersen, T. L. (2020). \emph{Patchwork: {The} composer of plots}. Available at: \url{https://CRAN.R-project.org/package=patchwork}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pipernoStarchGrains2008}{}}%
-Piperno, D.
-R., and Dillehay, T.
-D.
-(2008).
-Starch grains on human teeth reveal early broad crop diet in northern {Peru}.
-\emph{Proceedings of the National Academy of Sciences} 105, 19622--19627.
-doi:\href{https://doi.org/10.1073/pnas.0808752105}{10.1073/pnas.0808752105}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-powerChimpCalculus2015}{}}%
-Power, R.
-C., Salazar-Garcia, D.
-C., Wittig, R.
-M., Freiberg, M., and Henry, A.
-G.
-(2015).
-Dental calculus evidence of {Tai Forest Chimpanzee} plant consumption and life history transitions.
-\emph{Sci Rep} 5, 15161.
-doi:\href{https://doi.org/10.1038/srep15161}{10.1038/srep15161}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-powerSEMCalculus2014}{}}%
-Power, R.
-C., Salazar-García, D.
-C., Wittig, R.
-M., and Henry, A.
-G.
-(2014).
-Assessing use and suitability of scanning electron microscopy in the analysis of micro remains in dental calculus.
-\emph{Journal of Archaeological Science} 49, 160--169.
-doi:\href{https://doi.org/10.1016/j.jas.2014.04.016}{10.1016/j.jas.2014.04.016}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-powerRepresentativenessDental2021}{}}%
-Power, R.
-C., Wittig, R.
-M., Stone, J.
-R., Kupczik, K., and Schulz-Kornas, E.
-(2021).
-The representativeness of the dental calculus dietary record: Insights from {Taï} chimpanzee faecal phytoliths.
-\emph{Archaeol Anthropol Sci} 13, 104.
-doi:\href{https://doi.org/10.1007/s12520-021-01342-z}{10.1007/s12520-021-01342-z}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-proctorSpatialGradient2018}{}}%
-Proctor, D.
-M., Fukuyama, J.
-A., Loomer, P.
-M., Armitage, G.
-C., Lee, S.
-A., Davis, N.
-M., Ryder, M.
-I., Holmes, S.
-P., and Relman, D.
-A.
-(2018).
-A spatial gradient of bacterial diversity in the human oral cavity shaped by salivary flow.
-\emph{Nat Commun} 9, 681.
-doi:\href{https://doi.org/10.1038/s41467-018-02900-1}{10.1038/s41467-018-02900-1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Rbase}{}}%
-R Core Team (2020).
-\emph{R: {A} language and environment for statistical computing}.
-{Vienna, Austria}: {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing} Available at: \url{https://www.R-project.org/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-radiniFoodPathways2017}{}}%
-Radini, A., Nikita, E., Buckley, S., Copeland, L., and Hardy, K.
-(2017).
-Beyond food: {The} multiple pathways for inclusion of materials into ancient dental calculus.
-\emph{Am J Phys Anthropol} 162, 71--83.
-doi:\href{https://doi.org/10.1002/ajpa.23147}{10.1002/ajpa.23147}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-radiniMedievalWomenEarly2019}{}}%
-Radini, A., Tromp, M., Beach, A., Tong, E., Speller, C., McCormick, M., Dudgeon, J.
-V., Collins, M.
-J., Rühli, F., Kröger, R., et al.
-(2019).
-Medieval women's early involvement in manuscript production suggested by lapis lazuli identification in dental calculus.
-\emph{Science Advances} 5, eaau7126.
-doi:\href{https://doi.org/10.1126/sciadv.aau7126}{10.1126/sciadv.aau7126}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-reichertStarchBible1913b}{}}%
-Reichert, E.
-T.
-(1913).
-\emph{The differentiation and specificity of starches in relation to genera, species, etc: Stereochemistry applied to protoplasmic processes and products, and as a strictly scientific basis for the classification of plants and animals}.
-{Washington, D.C.}: {Carnegie institution of Washington}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Rbroom}{}}%
-Robinson, D., Hayes, A., and Couch, S. (2021). \emph{Broom: {Convert} statistical objects into tidy tibbles}. Available at: \url{https://CRAN.R-project.org/package=broom}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-scannapiecoSalivaryAmylase1993}{}}%
-Scannapieco, F. A., Torres, G., and Levine, M. J. (1993). Salivary α-amylase: Role in dental plaque and caries formation. \emph{Critical Reviews in Oral Biology \& Medicine} 4, 301--307.
-
-\leavevmode\vadjust pre{\hypertarget{ref-seidemannStarchAtlas1966}{}}%
-Seidemann, J.
-(1966).
-\emph{St\{\textbackslash"a\}rke-{Atlas}: {Grundlagen} der {St}\{\textbackslash"a\}rke-{Mikroskopie} und {Beschreibung} der wichtigsten {St}\{\textbackslash"a\}rkearten}.
-{Berlin}: {Parey}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-shellisSyntheticSalivaCultural1978}{}}%
-Shellis, R.
-P.
-(1978).
-A synthetic saliva for cultural studies of dental plaque.
-\emph{Archives of Oral Biology} 23, 485--489.
-doi:\href{https://doi.org/10.1016/0003-9969(78)90081-X}{10.1016/0003-9969(78)90081-X}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-simonsoroOralGeography2013}{}}%
-Simón-Soro, A., Tomás, I., Cabrera-Rubio, R., Catalan, M.
-D., Nyvad, B., and Mira, A.
-(2013).
-Microbial geography of the oral cavity.
-\emph{J Dent Res} 92, 616--621.
-doi:\href{https://doi.org/10.1177/0022034513488119}{10.1177/0022034513488119}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sissonsMultistationPlaque1991}{}}%
-Sissons, C.
-H., Cutress, T.
-W., Hoffman, M.
-P., and Wakefield, J.
-S.
-J.
-(1991).
-A {Multi-station Dental Plaque Microcosm} ({Artificial Mouth}) for the {Study} of {Plaque Growth}, {Metabolism}, {pH}, and {Mineralization}: \emph{J Dent Res}.
-doi:\href{https://doi.org/10.1177/00220345910700110301}{10.1177/00220345910700110301}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tanCalculusUltrastructure2004}{}}%
-Tan, B.
-T.
-K., Gillam, D.
-G., Mordan, N.
-J., and Galgut, P.
-N.
-(2004a).
-A preliminary investigation into the ultrastructure of dental calculus and associated bacteria.
-\emph{J Clin Periodontol} 31, 364--369.
-doi:\href{https://doi.org/10.1111/j.1600-051X.2004.00484.x}{10.1111/j.1600-051X.2004.00484.x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tanBacterialViability2004}{}}%
-Tan, B.
-T.
-K., Mordan, N.
-J., Embleton, J., Pratten, J., and Galgut, P.
-N.
-(2004b).
-Study of {Bacterial Viability} within {Human Supragingival Dental Calculus}.
-\emph{Journal of Periodontology} 75, 23--29.
-doi:\href{https://doi.org/10.1902/jop.2004.75.1.23}{10.1902/jop.2004.75.1.23}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-taoWheatCalculus2020}{}}%
-Tao, D., Zhang, G., Zhou, Y., and Zhao, H.
-(2020).
-Investigating wheat consumption based on multiple evidences: {Stable} isotope analysis on human bone and starch grain analysis on dental calculus of humans from the {Laodaojing} cemetery, {Central Plains}, {China}.
-\emph{International Journal of Osteoarchaeology} 30, 594--606.
-doi:\href{https://doi.org/10.1002/oa.2884}{10.1002/oa.2884}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-toppingResistantStarch2003}{}}%
-Topping, D.
-L., Fukushima, M., and Bird, A.
-R.
-(2003).
-Resistant starch as a prebiotic and synbiotic: State of the art.
-\emph{Proceedings of the Nutrition Society} 62, 171--176.
-doi:\href{https://doi.org/10.1079/PNS2002224}{10.1079/PNS2002224}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-trompEDTACalculus2017}{}}%
-Tromp, M., Buckley, H., Geber, J., and Matisoo-Smith, E.
-(2017).
-{EDTA} decalcification of dental calculus as an alternate means of microparticle extraction from archaeological samples.
-\emph{Journal of Archaeological Science: Reports} 14, 461--466.
-doi:\href{https://doi.org/10.1016/j.jasrep.2017.06.035}{10.1016/j.jasrep.2017.06.035}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-trompDietaryNondietary2015}{}}%
-Tromp, M., and Dudgeon, J.
-V.
-(2015).
-Differentiating dietary and non-dietary microfossils extracted from human dental calculus: The importance of sweet potato to ancient diet on {Rapa Nui}.
-\emph{Journal of Archaeological Science} 54, 54--63.
-doi:\href{https://doi.org/10.1016/j.jas.2014.11.024}{10.1016/j.jas.2014.11.024}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-vandeveldeStarchMorphology2002}{}}%
-van de Velde, F., van Riel, J., and Tromp, R.
-H.
-(2002).
-Visualisation of starch granule morphologies using confocal scanning laser microscopy ({CSLM}).
-\emph{Journal of the Science of Food and Agriculture} 82, 1528--1536.
-doi:\href{https://doi.org/10.1002/jsfa.1165}{10.1002/jsfa.1165}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-warinnerEvidenceMilk2014}{}}%
-Warinner, C., Hendy, J., Speller, C., Cappellini, E., Fischer, R., Trachsel, C., Arneborg, J., Lynnerup, N., Craig, O.
-E., Swallow, D.
-M., et al.
-(2014a).
-Direct evidence of milk consumption from ancient human dental calculus.
-\emph{Sci Rep} 4, 7104.
-doi:\href{https://doi.org/10.1038/srep07104}{10.1038/srep07104}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-warinnerPathogensHost2014}{}}%
-Warinner, C., Rodrigues, J.
-F., Vyas, R., Trachsel, C., Shved, N., Grossmann, J., Radini, A., Hancock, Y., Tito, R.
-Y., Fiddyment, S., et al.
-(2014b).
-Pathogens and host immunity in the ancient human oral cavity.
-\emph{Nat Genet} 46, 336--44.
-doi:\href{https://doi.org/10.1038/ng.2906}{10.1038/ng.2906}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wesolowskiEvaluatingMicrofossilContent2010}{}}%
-Wesolowski, V., Ferraz Mendonça de Souza, S.
-M., Reinhard, K.
-J., and Ceccantini, G.
-(2010).
-Evaluating microfossil content of dental calculus from {Brazilian} sambaquis.
-\emph{J Archaeol Sci} 37, 1326--1338.
-doi:\href{https://doi.org/10.1016/j.jas.2009.12.037}{10.1016/j.jas.2009.12.037}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tidyverse2019}{}}%
-Wickham, H., Averick, M., Bryan, J., Chang, W., McGowan, L.
-D., François, R., Grolemund, G., Hayes, A., Henry, L., Hester, J., et al.
-(2019).
-Welcome to the {tidyverse}.
-\emph{Journal of Open Source Software} 4, 1686.
-doi:\href{https://doi.org/10.21105/joss.01686}{10.21105/joss.01686}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wuDietEarliest2021}{}}%
-Wu, Y., Tao, D., Wu, X., and Liu, W. (2021). Diet of the earliest modern humans in {East Asia}. {In Review} doi:\href{https://doi.org/10.21203/rs.3.rs-442096/v1}{10.21203/rs.3.rs-442096/v1}.
-
-\end{CSLReferences}
-
 \clearpage
 
-\hypertarget{tables}{%
-\section*{Tables}\label{tables}}
+\section*{Tables}\label{tables}
 \addcontentsline{toc}{section}{Tables}
@@
 
-\hypertarget{figures}{%
-\section*{Figures}\label{figures}}
+\section*{Figures}\label{figures}
 \addcontentsline{toc}{section}{Figures}
@@
 
+\bibliographystyle{plain}
+\bibliography{refs.bib}
+
 \end{document}

### 2024-01-04 data-raw/starch-counts.R

@@ -54,7 +54,8 @@ corr_comb <- corr_counts %>%
 #               values_fn = function(x) sum(x, na.rm = T))
 #readr::write_csv(corr_comb, "analysis/data/derived_data/corr_comb.csv")
 
-
+usethis::use_data(raw_counts, overwrite = TRUE)
+usethis::use_data(sol_counts, overwrite = TRUE)
 usethis::use_data(corr_comb, overwrite = TRUE)
 usethis::use_data(corr_counts, overwrite = TRUE)
 usethis::use_data(sol_corr, overwrite = TRUE)

### 2024-01-04 analysis/paper/discussion.qmd
rename sections to .qmd

--- 2023-05-31 analysis/paper/discussion.Rmd
@@
 a result that contradicts findings from archaeological contexts
-[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeography2014].
+[@wesolowskiEvaluatingMicrofossil2010; @dudgeonDietGeography2014].
 When the concentration of starch granules
@@
 or accidental ingestion not related to food consumption 
-[@radiniFoodPathways2017; @radiniMedievalWomenEarly2019].  
+[@radiniFoodPathways2017; @radiniMedievalWomen2019].  
 When starch granules enter the mouth, whether through ingestion of food or accidental

### 2024-01-04 analysis/paper/intro.qmd
rename sections to .qmd

--- 2023-05-31 analysis/paper/intro.Rmd
@@
 dietary information in the form of plant microfossils 
-[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
+[@henryCalculusSyria2008; @hardyStarchGranules2009], 
 proteins [@hendyProteomicCalculus2018; @warinnerEvidenceMilk2014],
@@
 within individuals. 
-Studies conducted on modern humans [@leonardDentalCalculus2015] 
+Studies conducted on modern humans [@leonardPlantMicroremains2015] 
 and non-human primates [@powerChimpCalculus2015; @powerRepresentativenessDental2021] 

### 2024-01-04 analysis/paper/results.qmd
rename sections to .qmd

--- 2022-05-17 analysis/paper/results.Rmd
@@
 
-## No amylase activity detected in the model
+### No amylase activity detected in the model
 
@@
 
-## Treatment type had minimal effect on biofilm growth
+### Treatment type had minimal effect on biofilm growth
 
@@
 
-## Starch counts
+### Starch counts
 
@@
 
-### Proportion of available starches incorporated in samples
+#### Proportion of available starches incorporated in samples
 
@@
 
-### Size ratios differ between solutions and samples
+#### Size ratios differ between solutions and samples
 
@@
 
-### Biofilm weight correlated positively with extracted starch counts
+#### Biofilm weight correlated positively with extracted starch counts
 

### 2024-01-04 analysis/paper/discussion.qmd
fix: convert cross-refs to quarto and add figures and tables

@@
 also had the highest absolute mean count of starch granules extracted from the 
-dental calculus
-(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
+dental calculus (@tbl-solution-count and @tbl-sample-count).
 This suggests that the starches that are more frequently consumed will be present
@@
 granules (10--20 $\mu$m) were often over-represented
-(Table \@ref(tab:sample-prop-tbl), 
-and 
-Figure \@ref(fig:ratio-plots)).
+(@tbl-sample-prop, and @fig-ratio-plots).
 Large potato granules were most affected, potentially because of the greater
@@
 `r strength_cor` `r direct_cor` correlation between size of biofilm deposit and
-retained starch granules (Figure \@ref(fig:cor-plot)A), meaning larger calculus
-deposits contain a higher quantity of granules;
-a result that contradicts findings from archaeological contexts
+retained starch granules ([@fig-cor-plot]A), meaning larger calculus deposits
+contain a higher quantity of granules; a result that contradicts findings from
+archaeological contexts
 [@wesolowskiEvaluatingMicrofossil2010; @dudgeonDietGeography2014].
-When the concentration of starch granules
-per mg calculus is considered, the correlation is weaker, but still present
-(Figure \@ref(fig:cor-plot)B).
+When the concentration of starch granules per mg calculus is considered, the
+correlation is weaker, but still present ([@fig-cor-plot]B).
 While the larger deposits contain a higher absolute count, our findings also suggest
@@
 The size bias against large granules (>20 $\mu$m) from both wheat and potato 
-(Table \@ref(tab:sample-prop-tbl)) may give further credence to
+(@tbl-sample-prop) may give further credence to
 this incorporation pathway, as the smaller starch granules have an advantage over

### 2024-01-04 analysis/paper/matmet.qmd
rename sections to .qmd
fix: convert cross-refs to quarto and add figures and tables

--- 2023-06-01 analysis/paper/matmet.Rmd
@@
-## Biofilm formation
+### Biofilm formation
 
@@
 for 10 days until the end of the experiment on day 24 
-(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).
+(see @fig-protocol for an overview of the protocol).
+
+```{r}
+#| label: fig-protocol
+#| fig-cap: "Overview of experiment protocol including the plate setup."
+knitr::include_graphics(here("figures/protocol_overview.png"))
+```
 
@@
 
-## Amylase activity detection
+### Amylase activity detection
 
@@
 
-## Treatment solutions
+### Treatment solutions
 
@@
 
-## Extraction method
+### Extraction method
 
@@
 
-## Statistical analysis
+### Statistical analysis
 

### 2024-01-04 analysis/paper/results.qmd
fix: convert cross-refs to quarto and add figures and tables

@@
 All samples yielded sufficient biofilm growth and starch incorporation to be 
-included in the analysis (Figure \@ref(fig:microscope-fig)), resulting in a total of 48 biofilm samples (two plates of 24),
-45 of which were used for analysis (three samples were set aside for later
- analysis).
-Most control samples contained no starch granules, while some contained negligible
-quantities (see Supplementary Material).
+included in the analysis (@fig-microscope), resulting in a total of 48 biofilm
+samples (two plates of 24), 45 of which were used for analysis (three samples
+were set aside for later analysis). Most control samples contained no starch
+granules, while some contained negligible quantities (see Supplementary Material).
+
+::: {#fig-microscope layout-ncol=2}
+
+```{r}
+knitr::include_graphics(here("figures/starches_w_bar.jpg"))#,
+                          #here("figures/st2C3.2-mix.jpg"),
+                          #here("figures/st1B4-wheat.jpg"),
+                          #here("figures/2D2-potato.jpg")))
+```
+
+```{r}
+knitr::include_graphics(here("figures/st2C3.2-mix.jpg"))
+```
+
+```{r}
+knitr::include_graphics(here("figures/st1B4-wheat.jpg"))
+```
+
+```{r}
+knitr::include_graphics(here("figures/2D2-potato.jpg"))
+```
+
+Microscope images of biofilm samples that were exposed to the starch solutions.
+Starch granules can be seen within bacterial communities and isolated. Scale bar = 20 μm.
+
+:::
 
@@
 = `r treat_f.stat`, p = `r treat_p`.
-A summary of sample weights is available in Table \@ref(tab:anova-tbl).
+A summary of sample weights is available in @tbl-anova.
+
+```{r}
+#| label: tbl-anova
+#| tbl-cap: "Summary statistics for biofilm dry-weights (in mg) by treatment."
+knitr::kable(anova_tbl, 
+             col.names = c("Treatment", "Mean", "SD", "Min", "Max"))
+```
 
@@
 (`r signif(subset(samp_counts, treatment == "potato")$total, 3)`) 
-(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
+(@tbl-solution-count and @tbl-sample-count).
+
+```{r}
+#| label: tbl-solution-count
+#| tbl-cap: "Mean starch counts from solutions, including the proportional makeup of the different sizes of granules."
+cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
+options(knitr.kable.NA = "")
+knitr::kable(sol_tbl, col.names = cnames)
+```
+
+```{r sample-count-tbl}
+#| label: tbl-sample-count
+#| tbl-cap: "Mean starch counts extracted from samples with standard deviation (SD), including the proportion of granule sizes of the total count."
+cnames <- c("Treatment", "Starch", "Small (%)", "SD", "Medium (%)", "SD", 
+            "Large (%)", "SD", "Total (%)", "SD")
+knitr::kable(samp_counts_tbl, col.names = cnames)
+```
 
@@
 with potato granules being more readily incorporated than wheat in both the
-separated- and mixed-treatment samples
-(Table \@ref(tab:sample-prop-tbl)). 
+separated- and mixed-treatment samples (@tbl-sample-prop). 
 There is an inverse relationship between the absolute starch count in the solutions 
@@
 incorporation, and vice versa for the mixed treatment.
+
+```{r}
+#| label: tbl-sample-prop
+#| tbl-cap: "The mean percentage of starches from the solutions that were incorporated into the samples."
+cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
+knitr::kable(perc_tbl, col.names = cnames)
+```
 
@@
 increased in potato treatments 
-(Figure \@ref(fig:ratio-plots)).
+(@fig-ratio-plots).
 
 <!-- Proportion and size ratios are the most important points to be made -->
+```{r}
+#| label: fig-ratio-plots
+#| fig-cap: "Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."
+#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
+(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+```
 

### 2024-01-04 analysis/paper/results.qmd
fix: convert cross-refs to quarto and add figures and tables

@@
 p `r cor_p`
-(Figure \@ref(fig:cor-plot)A).
+([@fig-cor-plot]A).
+
+```{r}
+#| label: fig-cor-plot
+#| warning: false
+#| fig-cap: "Scatter plots of (A) sample weight in mg and standardised starch count by z-score for separated treatments, and (B) sample weight in mg and standardised count of starch grains per mg calculus."
+
+pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
+```
 
@@
 p `r cor2_p`
-(Figure \@ref(fig:cor-plot)B).
+([@fig-cor-plot]B).
